### 2020-05-11 ms/ms.tex
revise fig1-3 and their text

\documentclass[12pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage{longtable}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=Blue,citecolor=black,linkcolor=black]{hyperref}
\def\equationautorefname~#1\null{Eq.~(#1)\null}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{lipsum}
\usepackage{multicol}
\usepackage{titlesec}	
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{enumitem}
\usepackage{soul}
\usepackage{booktabs}
\usepackage{pgfplotstable} % http://mirrors.ctan.org/graphics/pgf/contrib/pgfplots/doc/pgfplotstable.pdf
\usepackage{subcaption}

%SetFonts
% newtxtext+newtxmath
\usepackage{newtxtext} %loads helv for ss, txtt for tt
\usepackage{amsmath}
\usepackage[bigdelims]{newtxmath}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
%SetFonts

% less space before sections 
% \@startsection {NAME}{LEVEL}{INDENT}{BEFORESKIP}{AFTERSKIP}{STYLE} 
%            optional * [ALTHEADING]{HEADING} 
\makeatletter
 \renewcommand\section{\@startsection {section}{1}{\z@}%
     {-2.5ex \@plus -1ex \@minus -.2ex}%
     {1.3ex \@plus.2ex}%
    {\Large\bfseries}}

% Species names
%% Meta-Command for defining new species macros
\usepackage{xspace}

\newcommand{\species}[3]{%
  \newcommand{#1}{\gdef#1{\textit{#3}\xspace}\textit{#2}\xspace}}

\species{\yeast}{Saccharomyces cerevisiae}{S.~cerevisiae}
\species{\calbicans}{Candida albicans}{C.~albicans}
\species{\cneoformans}{Cryptococcus neoformans}{C.~neoformans}

% line numbers
 \usepackage[displaymath, mathlines]{lineno}
 \renewcommand\linenumberfont{\normalfont\small\sffamily}
\linenumbers
% \modulolinenumbers[2]

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\newtheorem{result}{Result}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{I}

% Supplementary
\newcommand{\beginsupplement}{%
      	\setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
}

% NatBib
\usepackage[numbers,super,sort&compress]{natbib}
%\usepackage[round,colon,authoryear]{natbib}
\renewcommand{\bibsection}{}
\renewcommand{\bibfont}{\small}

% unbreakable dashes
\usepackage[shortcuts]{extdash}

% Title page
\title{Late and early effective start of non-pharmaceutical interventions inferred during COVID-19 outbreaks }

% Authors
\renewcommand\Affilfont{\small}

\author[a]{Ilia Kohanovski}
\author[b,c]{Uri Obolski}
\author[a,*]{Yoav Ram}

\affil[a]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya 4610101, Israel}
\affil[b]{School of Public Health, Tel Aviv University, Tel Aviv 6997801, Israel}{
\affil[c]{Porter School of the Environment and Earth Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
\affil[*]{Corresponding author: yoav@yoavram.com}

% Document
\begin{document}
\maketitle

% Abstract
\begin{abstract}
During February and March 2020, several countries implemented a variety of non-pharmaceutical interventions with variable schedules to control the COVID\=/-19 pandemic caused by the SARA\=/CoV\=/2 virus.
Overall, these interventions have successfully reduced the ...
\end{abstract}

\pagebreak
% Introduction
\section*{Introduction}

The COVID-19 pandemic has resulted in implementation of extreme non-pharmaceutical interventions (NPIs) in many affected countries. These interventions, from social distancing to lockdowns, are applied in a rapid and widespread fashion.
The NPIs are designed and assessed using epidemiological models, which follow the dynamics of the viral infection to forecast the effect of different mitigation and suppression strategies on the levels of infection, hospitalization, and fatality.
These epidemiological models usually assume that the effect of NPIs on disease transmission begins at the officially declared date (e.g.~\citet{Li2020,Gatto2020,Flaxman2020}).

Adoption of public health recommendations is often critical for effective response to infectious diseases, and has been studied in the context of HIV~\citep{Kaufman2014} and vaccination~\citep{Dunn2015,Wiyeh2018}, for example.
However, behavioral and social change does not occur immediately, but rather requires time to diffuse in the population through media, social networks, and social interactions. 
Moreover, compliance to NPIs may differ between different interventions and between people.
For example, in a survey of 2,108 adults in the UK during Mar 2020, \citet{Atchison2020} found that those over 70 years old were more likely to adopt social distancing than young adults (18-34 years old), and that those with lower income were less likely to be able to work from home and to self-isolate.
Similarly, compliance to NPIs may be impacted by personal experiences. \citet{Smith2020} have surveyed 6,149 UK adults in late April and found that people who believe they have already had COVID\=/19 are more likely to think they are immune, and less likely to comply with social distancing measures. 
Compliance may also depend on risk perception as perceived by the the number of domestic cases or even by reported cases in other regions and countries.
Interestingly, the perceived risk of COVID\=/19 infection has likely caused a reduction in the number of influenza-like illness cases in the US starting from mid-February~\citep{Zipfel2020}.

Here, we hypothesize that there is a significant difference between the official start of NPIs and their adoption by the public and therefore their effect on transmission dynamics.
We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to estimate the effective start date of NPIs from publicly available COVID-19 case data in several geographical regions.
We compare these estimates to the official dates and find both delayed and advanced effect of NPIs on COVID\=/19 transmission dynamics.
We conclude by demonstrating how differences between the official and effective start of NPIs can confuse assessments of the effectiveness of the NPIs in a simple epidemic control framework. % TODO explain why important

%\pagebreak
% Models and Methods
\section*{Models and Methods}

%%% Data %%%%
\paragraph*{Data.} 
We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from several different countries. These incidence data summarize the number of individuals $X_t$ tested positive for SARS\=/CoV\=/2 RNA (using RT-qPCR) at each day $t$.
Data for Wuhan, China retrieved from \citet{Pei2020}, data for 11 European countries retrieved from \citet{Flaxman2020}. 
Regions in which there were multiple sequences of days with zero confirmed cases (e.g. France), we cropped the data to begin with the last sequence so that our analysis focuses on the first sustained outbreak rather than isolated imported cases. 
For dates of official NPI dates see \autoref{table:table1}.

% Table 1 %%%
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../data/NPI_dates.csv}
\caption{
\textbf{Official start of non-pharmaceutical interventions.}
The date of the first intervention is for a ban of public events, or encouragement of social distancing, or for school closures.
In all countries except Sweden, the date of the last intervention is for a lockdown. In Sweden, where a lockdown was not ordered during the studied dates, the last date is for school closures. Dates for European countries from \citet{Flaxman2020}, date for Wuhan, China from \citet{Pei2020}.
}
\label{table:table1}
\end{table}

%%% SEIR model %%%%
\paragraph*{SEIR model.}
We model SARS\=/CoV\=/2 infection dynamics by following the number of susceptible $S$, exposed $E$, reported infected $I_r$, and unreported infected $I_u$ individuals in a population of size $N$.
This model distinguishes between reported and unreported infected individuals: the reported infected are those that have enough symptoms to eventually be tested and thus appear in daily case reports, to which we fit the model.

Susceptible ($S$) individuals become exposed due to contact with reported or unreported infected individuals ($I_r$ or $I_u$) at a rate $\beta_t$ or $\mu \beta_t$.
The parameter $0 < \mu < 1$ represents the decreased transmission rate from unreported infected individuals, who are often subclinical or even asymptomatic.
The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioural changes of both susceptible and infected individuals.
Exposed individuals, after an average incubation period of $Z$ days, become reported infected with probability $\alpha_t$ or unreported infected with probability $(1-\alpha_t)$.
The reporting rate $0 < \alpha_t < 1$ may also change over time due to changes in human behavior.
Infected individuals remain infectious for an average period of $D$ days, after which they either recover, or becomes ill enough to be quarantined.
They therefore no longer infect other individuals, and the model does not track their frequency.
The model is described by the following equations:

\begin{equation} \label{eq:model}
\begin{aligned}
\frac{dS}{dt} & = -\beta_t S \frac{I_p}{N} - \mu \beta_t S \frac{I_s}{N} \\
\frac{dE}{dt} & = \beta_t S \frac{I_p}{N} + \mu \beta_t S \frac{I_s}{N}  - \frac{E}{Z} \\
\frac{dI_r}{dt} & = \alpha_t \frac{E}{Z} - \frac{I_r}{D} \\
\frac{dI_u}{dt} & = (1-\alpha_t) \frac{E}{Z} - \frac{I_r}{D} .
\end{aligned}
\end{equation}
The initial numbers of exposed $E(0)$ and unreported infected $I_u(0)$ are considered model parameters, whereas the initial number of reported infected is assumed to be zero $I_r(0)=0$, and the number of susceptible is $S(0)=N-E(0)-I_u(0)$.
This model is inspired by \citet{Li2020} and \citet{Pei2020}, who used a similar model with multiple regions and constant transmission $\beta$ and reporting rate $\alpha$ to infer COVID\=/19 dynamics in China and the continental US, respectively.

%%% Likelihood %%%
\paragraph*{Likelihood function.}
The \emph{expected} cumulative number of reported infected individuals until day $t$ is 
\begin{equation} \label{eq:Yt}
Y_t=\int_{0}^{t}{\alpha_s \frac{E(s)}{Z} \; ds}, \quad Y_0 = 0.
\end{equation}
We assume that reported infected individuals are confirmed and therefore observed in the daily case report of day $t$ with probability $p_t$ (note that an individual can only be observed once, and that $p_t$ may change over time, but $t$ is a specific date rather than the time elapsed since the individual was infected).
Hence, we assume that the number of confirmed cases in day $t$ is binomially distributed,
$$
X_t \sim \mathit{Bin}\big(n_t, p_t\big),
$$
where $n_t$ is the \emph{realized} (rather than expected) number of reported infected individuals yet to appear in daily reports by day~$t$.
The cumulative number of confirmed cases until day $t$ is
$$
\tilde{X}_t=\sum_{i=1}^{t}X_i, \quad X_0=0.
$$
Given $\tilde{X}_{t-1}$, we assume $n_t$ is Poisson distributed,
$$
\big(n_t \mid \tilde{X}_{t-1}\big) \sim \mathit{Poi}\Big( Y_t - \tilde{X}_{t-1} \Big), 
\quad n_1 \sim \mathit{Poi}(Y_1).
$$ 
Therefore, $\big(X_t \mid \tilde{X}_{t-1} \big)$ is a binomial conditioned on a Poisson, which reduces to a Poisson with
\begin{equation} \label{eq:P(Xt|Xt-1)}
\big(X_t \mid \tilde{X}_{t-1} \big) \sim \mathit{Poi}\Big( \big( Y_t - \tilde{X}_{t-1} \big) \cdot p_t \Big), 
\quad X_1 \sim \mathit{Poi}(Y_1 \cdot p_1).
\end{equation}

For given vector $\theta$ of model parameters (\autoref{eq:theta}), we compute the expected cumulative number of reported infected individuals $\{Y_t\}_{t=1}^{T}$ for each day (\autoref{eq:Yt}). Then, since $\tilde{X}_{t-1}$ is a function of $X_1, \ldots, X_{t-1}$, we can use \autoref{eq:P(Xt|Xt-1)} to write the probability to observe the confirmed case data $\vec{X} = (X_1, \ldots, X_T)$ as 
\begin{equation} \label{eq:likelihood}
\mathbb{L}(\theta \mid \vec{X}) = P(\vec{X} \mid \theta) = P(X_1 \mid \theta) P(X_2 \mid \tilde{X_1}, \theta) \cdots P(X_T \mid \tilde{X}_{T-1}, \theta).
\end{equation}
This defines a \emph{likelihood function} $\mathbb{L}(\theta \mid \vec{X})$ for the parameter vector $\theta$ given the data $\vec{X}$.

%%% NPI %%%
\paragraph*{NPI model.}
To model non-pharmaceutical interventions (NPIs), we set the beginning of the NPIs to day $\tau$ and define
\begin{equation} \label{eq:NPI_model}
\beta_t = \begin{cases} 
  \beta, & t < \tau \\ % TODO make sure < and not \le, and the next line \ge and not >
  \beta \lambda, & t \ge \tau
\end{cases},
\quad
\alpha_t = \begin{cases} 
  \alpha_1, & t < \tau \\ % TODO make sure < and not \le, and the next line \ge and not >
  \alpha_2, & t \ge \tau
\end{cases},
\quad
p_t = \begin{cases} 
  1/9, & t < \tau \\ % TODO make sure < and not \le, and the next line \ge and not >
  1/6, & t \ge \tau
\end{cases},
\end{equation}
where $0 < \lambda < 1$.
The values for $p_t$ follow \citet{Li2020}, who estimated the average time between infection and reporting in Wuhan, China, at 9 days before the start of NPIs (Jan 23, 2020) and 6 days after start of NPIs.
The parameter $\tau$ is then added to the parameter vector $\theta$ (\autoref{eq:theta}).

%%% Model fitting %%%
\paragraph*{Parameter estimation.}
To estimate the parameters of our model from the data $\vec{X}$, we apply a Bayesian inference approach.
We start our model $\Delta t$ days before the outbreak (defined as consecutive days with increasing confirmed cases) in each country~\citep{Gatto2020}.
The model in \autoref{eq:model} is parameterized by the vector $\theta$, where
\begin{equation} \label{eq:theta}
\theta=\Big(Z, D, \mu, \{\beta_t\}, \{\alpha_t\}, \{p_t\}, E(0), I_u(0)\Big), \tau, \Delta t.
\end{equation}

The likelihood function is defined in \autoref{eq:likelihood}.
We define the following prior distributions on the model parameters $P(\theta)$: 
\begin{equation} \label{eq:priors}
\begin{aligned} % TODO verify before submitting
Z & \sim \mathit{Uniform}(2, 5) \\
D & \sim \mathit{Uniform}(2, 5) \\
\mu & \sim \mathit{Uniform}(0.2, 1) \\
\beta & \sim \mathit{Uniform}(0.8, 1.5) \\
\lambda & \sim \mathit{Uniform}(0, 1) \\
\alpha_1, \alpha_2 & \sim \mathit{Uniform}(0.02, 1)\\
E(0) & \sim \mathit{Uniform}(0, 3000) \\
I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
\tau &\sim \mathit{TruncatedNormal}(\tau^*, 5, 1, T-2),
\end{aligned}
\end{equation}
where $\mathit{TruncatedNormal}(\mu, \sigma, a, b)$ is a truncated normal distribution with mean $\mu$ and standard deviation $\sigma$ taking values between $a$ and $b$; $T$ is the number of days in the data $\vec{X}$; and $\tau^*$ is the official start of the NPI.
Most priors follow \citet{Li2020}, with the following exceptions.
$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates.
For the effective start of NPIs $\tau$ we have also tested an uninformative uniform prior $U(1,T-1)$. DIC (see definition below) was lower for the truncated normal prior in most countries, except \hl{Germany?}. More importantly, the uninformative prior could result in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom (this was due to MCMC chains being stuck in low posterior regions of the parameter space).
We therefore decided to use the more informative truncated normal prior.

The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is then estimated using an \textit{affine-invariant ensemble sampler for Markov chain Monte Carlo} (MCMC) implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
The maximum a posteriori

%%% Model selection %%%
\paragraph*{Model selection.}
We perform model selection using DIC (deviance information criterion)~\citep{Spiegelhalter2002},
\begin{equation} \label{eq:DIC}
\begin{aligned}
DIC(\theta, \vec{X}) &= 2\mathbb{E}[D(\theta)] - D(\mathbb{E}[\theta])   \\
&= 2 \log\mathcal{L}(\mathbb{E}[\theta] \mid \vec{X}) - 4\mathbb{E}[\log\mathcal{L}(\theta \mid \vec{X})],
\end{aligned}
\end{equation}
where $D(\theta)=-2\log\mathcal{L}(\theta \mid \vec{X})$ is the Bayesian deviance, and expectations $\mathbb{E}[\cdot]$ are taken over the posterior distribution $P(\theta \mid \vec{X})$.
We compare models by reporting their relative DIC; lower is better.

%%% Source code %%%
\paragraph*{Source code.} 
We use Python 3 (Anaconda) with the NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee packages.
All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.

%%% Fig 1 %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig1.pdf}
    \caption{
    \textbf{Official and effective start of non-pharmaceutical interventions.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPI is shown for different regions. The effective NPI dates in Italy and Wuhan are significantly delayed compared to the official dates, whereas in Denmark, France, Spain, and Germany, the effective date is earlier than the official date.
	$\hat{\tau}$ is the posterior median, see \hl{TABLE2}. $\tau^*$ is the last NPI date, see \autoref{table:table1}. Thin and bold lines show 95\% and 75\% credible intervals (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
    }
    \label{fig:fig1}
\end{figure}

%\pagebreak
% Results
\section*{Results}

Several studies have described the effects of non-pharmaceutical interventions in different geographical regions~\citep{Flaxman2020,Gatto2020,Li2020}. 
These studies have assumed that the parameters of the epidemiological model change at a specific date, as in \autoref{eq:NPI_model}, and set the change date $\tau$ to the official NPI date $\tau^*$ (\autoref{table:table1}).
They then fit the model once for time $t<\tau^*$ and once for time $t \ge \tau^*$.
For example, \citet{Li2020} estimate the dynamics in China before and after $\tau^*$ at Jan 23. Thereby, they effectively estimate $(\beta, \alpha_1)$ and $(\lambda, \alpha_2)$ separately.
Here we estimate the posterior distribution $P(\tau \mid \vec{X})$ of the \emph{effective} start date of the NPIs by jointly estimating $\tau, \beta, \lambda, \alpha_1, \alpha_2$ on the entire data per region (e.g. Italy, Austria), rather than splitting the data at $\tau^*$.
We then estimate the maximum a posteriori (MAP) estimate $\hat{\tau}=\mathit{argmax}_{\tau}{P(\tau \mid \vec{X})}$. \hl{using median or argmax?}

We find that a model that considers an NPI (\autoref{eq:NPI_model}) is a better fit to the data than a model without an NPI, i.e. with constant $\beta$ and $\alpha$ (\hl{$\Delta DIC > ?$ for all regions}.)
We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}): the \hl{95\%} credible interval on $\hat{\tau}$ does not include $\tau^*$, and the DIC of the model with free $\tau$ parameter is lower than that of a model with a fixed $\tau \equiv \tau^*$ (\hl{$\Delta DIC > ?$}.) The exception that proves the rule is \hl{Switzerland}, where the effective and official dates are the same. Another important exception is the \hl{United~Kingdom?} % TODO need more data? Maybe not such an exception?

In the following, we describe our findings on delayed and advanced start of NPI in detail.

%%% Delayed
\paragraph*{Delayed effective start of NPI.}
In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$ (\autoref{fig:fig1}).

In Italy, the first case officially confirmed on Feb~21, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
That is, the official date $\tau^*$ is either Mar~8 or~11.
However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.7 days 95\% CI ; \autoref{fig:fig2}).
Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate  the effective start of NPIs to be several days layer at around Mar~2 ($\pm$2.65 days 95\% CI \autoref{fig:fig2}).

%%% Fig 2 %%%
\begin{figure}[h]
    \centering
    % Wuhan
    % τ_hat = 19.33
	% +-2.65 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig2a.pdf}
    \end{subfigure}
  	~
	% Italy
	% τ_hat = 24.74
	% +-0.71 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig2b.pdf}
    \end{subfigure}
    \caption{
	\textbf{Delayed effect of non-pharmaceutical interventions in Italy and Wuhan, China.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:fig2}
\end{figure}

%%% Advanced
\paragraph*{Advanced effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:fig1}).
In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ at Mar~8 or~9 ($\pm$2.96 95\%CI), rather than Mar~14 (\autoref{fig:fig3}).

Similarly, in France the official lockdown started at Mar~17 ($\tau^*$), with initial NPIs at Mar 13~\citep{Flaxman2020}. 
However, we estimate the effective start of NPIs $\hat{\tau}$ at Mar~8 ($\pm$5.9 days 95\% CI). Although the credible interval is wide, spanning from Mar~2 to Mar~13, the official lockdown start at Mar~17 is later still (\autoref{fig:fig3}).

Interestingly, the effective start of NPIs $\hat{\tau}$ in both France and Spain is estimated at Mar~8, although the official dates are differ by three days. Moreover, the number of daily cases was similar until Mar~8 in both countries, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).

%%% Fig 3 %%%
\begin{figure}[h]
    \centering
    % France
    % τ_hat = 12.38
	% +-5.90 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig3a.pdf}
    \end{subfigure}
    ~
    % Spain
    % τ_hat = 13.95
	% +-2.96 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig3b.pdf}
    \end{subfigure}
    \caption{
    \textbf{Advanced effect of non-pharmaceutical interventions in France and Spain.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
	}
	\label{fig:fig3}
\end{figure}

%%% Exact
\paragraph*{The exception that proves the rule.}
We find one case in which the official and effective dates match: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
Indeed, our MAP estimate $\hat{\tau}$ is Mar~20, and the posterior distribution shows two density peaks: a smaller one between Mar~10 and Mar~14, and a taller one between Mar~17 and Mar~22. It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.

%%% Assessment
\paragraph*{Effect of delays and advances of real-time assessment.}

The success of non-pharmaceutical interventions is assessed by health officials using various metrics, such as the decline in the growth rate of daily cases. These assessments are made a specific number of days after the intervention began, to accommodate for the expected serial interval~\citep{Banholzer2020} (i.e. time between successive cases in a chain of transmission), which is estimated at about 4-7 days~\citep{Gatto2020}. % Table 1

However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the delayed or advanced population response to the NPI.
Such a case is illustrated in \autoref{fig:fig4} using data and parameters from Italy.
Here, a lockdown is officially ordered on Mar~10 ($\tau^*$, but its delayed effect on the transmission dynamics starts on Mar~15 ($\hat{\tau}$). If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases to follow the dashed red line. However, the number of cases will actually follow the black line, leading to a significant different ($\Delta$) between the projections and the realization.

%%% Fig 4 %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig4.pdf}
    \caption{
    \textbf{Delayed effective start of NPI leads to under-estimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). Model predictions, assuming a 50\% decrease in transmission rate after the NPI starts, are shown as colored lines with 95\% confidence intervals. 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model under-estimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are MAP estimates for Italy (\hl{TABLE}) but with $\lambda=0.5$ and $\alpha_1=\alpha_2$.
    } 
    \label{fig:fig4}
\end{figure}

\pagebreak
% Discussion
\section*{Discussion}

We have estimated the effective start date of NPIs in several geographical regions using an SEIR epidemiological model and an MCMC parameter estimation framework.
We find examples of both advanced and delayed response to NPIs (\autoref{fig:fig1}).

For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred 3-5 after the official date (\autoref{fig:fig2}). This could be explained by low compliance. In Italy, for example, a leak about the intent to lockdown Northern provinces results in people leaving those provinces~\citep{Gatto2020}. However, delayed effect of NPIs could also be due to the time required by both the government and the citizens to organize for a lockdown.

In contrast, in most investigated countries, such as Spain and France, transmission rates seem to have been reduced even before official lockdowns were implemented (\autoref{fig:fig3}).
This advanced response is possibly due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response increased risk perception due to domestic or international COVID\=/19-related reports.
This finding may also suggest that severe NPIs, such as lockdowns, were unnecessary, and that milder measures that were adopted by the population, possibly due to government recommendations, media coverage, and social networks, could have been sufficient for epidemic control.
\hl{check if this is true} Indeed, the evidence supports a change in transmission dynamics (i.e. a model with $\tau$) even for Sweden, in which a lockdown was not implemented, suggesting that lockdowns may not be necessary if other NPIs are adopted early enough during the outbreak~\citep{Banholzer2020} (Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.)

Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a 7 day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
However, the delays and advances we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).

We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point. It may be interesting to investigate if the evidence favors a model with \emph{two} change-points, rather than one. 
Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), a mix of NPIs and changes in weather, a mix of domestic and international effects on risk perception, or other similar factors.

As several countries (e.g. Austria, Israel) begin to relieve lockdowns and ease restrictions, we expect similar delays and advances to occur: in some countries people will begin to behave as if restrictions were eased even before the official date, and in some countries people will continue to self-restrict even after restrictions are officially removed.

\paragraph*{Conclusions.}
We have estimated the effective start date of NPIs and found that they often differ from the official dates.
Our results emphasize the complex interaction between personal, regional, and global determinants of behavioral. Thus, our results highlight the need to further study variability in compliance and behavior over both time and space. This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Fenichela2011,Arthur2020}.

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
%We thank XXX for discussions and comments.
This work was supported in part by the Israel Science Foundation 552/19 and 1399/17.
}

\pagebreak
\begin{multicols}{2}[\section*{References}]
\nolinenumbers
\bibliographystyle{agsm}
%\bibliography{/Users/yoavram/Documents/library}
\bibliography{ms}
\end{multicols}

\pagebreak
\section*{Supplementary Material}
\beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/

%%% Fig Fr-vs-Es %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=\textwidth]{../figures/Fig-fr-vs-es.pdf}
    \caption{
    \textbf{COVID\=/19 confirmed cases in France and Spain.}
    Number of cases proportional to population size (as of 2018). 
    Vertical line shows Mar~8, the effective start of NPIs $\hat{\tau}$ in both countries.
    } 
    \label{fig:fig-fr-vs-es}
\end{figure}

\end{document}

### 2020-05-11 ms/ms.tex
delayed-advanced -> late-early

@@
 We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to estimate the effective start date of NPIs from publicly available COVID-19 case data in several geographical regions.
-We compare these estimates to the official dates and find both delayed and advanced effect of NPIs on COVID\=/19 transmission dynamics.
+We compare these estimates to the official dates and find both late and early effects of NPIs on COVID\=/19 transmission dynamics.
 We conclude by demonstrating how differences between the official and effective start of NPIs can confuse assessments of the effectiveness of the NPIs in a simple epidemic control framework.
@@
 
-In the following, we describe our findings on delayed and advanced start of NPI in detail.
+In the following, we describe our findings on late and early effective start of NPI in detail.
 
-%%% Delayed
-\paragraph*{Delayed effective start of NPI.}
+%%% Late
+\paragraph*{Late effective start of NPIs.}
 In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$ (\autoref{fig:fig1}).
@@
     \caption{
-	\textbf{Delayed effect of non-pharmaceutical interventions in Italy and Wuhan, China.}
+	\textbf{Late effect of non-pharmaceutical interventions in Italy and Wuhan, China.}
     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
@@
 
-%%% Advanced
-\paragraph*{Advanced effective start of NPIs.}
+%%% Early
+\paragraph*{Early effective start of NPIs.}
 In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:fig1}).
@@
     \caption{
-    \textbf{Advanced effect of non-pharmaceutical interventions in France and Spain.}
+    \textbf{Early effect of non-pharmaceutical interventions in France and Spain.}
     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
@@
 %%% Assessment
-\paragraph*{Effect of delays and advances of real-time assessment.}
+\paragraph*{Effect of late and early effect of NPIs on real-time assessment.}
 
@@
 
-However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the delayed or advanced population response to the NPI.
+However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
 Such a case is illustrated in \autoref{fig:fig4} using data and parameters from Italy.
-Here, a lockdown is officially ordered on Mar~10 ($\tau^*$, but its delayed effect on the transmission dynamics starts on Mar~15 ($\hat{\tau}$).
+Here, a lockdown is officially ordered on Mar~10 ($\tau^*$, but its late effect on the transmission dynamics starts on Mar~15 ($\hat{\tau}$).
 If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases to follow the dashed red line.
@@
     \caption{
-    \textbf{Delayed effective start of NPI leads to under-estimation of daily confirmed cases.}
+    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
     Real number of daily cases in Italy in black (markers: data, line: time moving average).
@@
 We have estimated the effective start date of NPIs in several geographical regions using an SEIR epidemiological model and an MCMC parameter estimation framework.
-We find examples of both advanced and delayed response to NPIs (\autoref{fig:fig1}).
+We find examples of both late and early effect of NPIs (\autoref{fig:fig1}).
 
@@
 In Italy, for example, a leak about the intent to lockdown Northern provinces results in people leaving those provinces~\citep{Gatto2020}.
-However, delayed effect of NPIs could also be due to the time required by both the government and the citizens to organize for a lockdown.
+However, late effect of NPIs could also be due to the time required by both the government and the citizens to organize for a lockdown.
 
 In contrast, in most investigated countries, such as Spain and France, transmission rates seem to have been reduced even before official lockdowns were implemented (\autoref{fig:fig3}).
-This advanced response is possibly due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response increased risk perception due to domestic or international COVID\=/19-related reports.
+This early response is possibly due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response increased risk perception due to domestic or international COVID\=/19-related reports.
 This finding may also suggest that severe NPIs, such as lockdowns, were unnecessary, and that milder measures that were adopted by the population, possibly due to government recommendations, media coverage, and social networks, could have been sufficient for epidemic control.
@@
 Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a 7 day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
-However, the delays and advances we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).
+However, the late and early effects we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).
 

### 2020-05-11 ms/ms.tex
finished abstract

@@
 \begin{abstract}
-During February and March 2020, several countries implemented a variety of non-pharmaceutical interventions with variable schedules to control the COVID\=/-19 pandemic caused by the SARA\=/CoV\=/2 virus.
-Overall, these interventions have successfully reduced the ...
+
+During February and March 2020, several countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules to control the COVID\=/-19 pandemic caused by the SARA\=/CoV\=/2 virus.
+Overall, these interventions seem to have successfully reduced the spread of the pandemic.
+We hypothesize that the official and effective start date of such interventions can significantly differ, for example due to slow diffusion of guidelines in the population, or due to unpreparedness of the authorities and the public.
+We use an SEIR epidemiological model and an MCMC inference framework to estimate the effective start of NPIs in several countries, and compare this effective dates to the official dates.
+We report our finding of both late and early effects of NPIs, and discuss potential causes and consequences of our results.
+
 \end{abstract}

### 2020-05-11 src/Table2.py
create table csv and tex from csv output

@@ -1,0 +1,40 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+
+import os
+import sys
+
+sns.set_context('paper', font_scale=1.3)
+red, blue, green = sns.color_palette('Set1', 3)
+
+def int_to_dt(t):
+    return pd.to_datetime(start_date) + timedelta(days=t)
+
+if __name__ == '__main__':
+	job_id = sys.argv[1]
+	output_folder = r'/Users/yoavram/Library/Mobile Documents/com~apple~CloudDocs/EffectiveNPI-Data/output/{}/'.format(job_id)
+
+	NPI_dates = pd.read_csv('../data/NPI_dates.csv')
+
+	data = pd.read_csv(os.path.join(output_folder, 'tables', 'all-countries-{}.csv'.format(job_id)))
+	features = ['country', 'official_τ', 'τ median', 'τ CI (75%)', 'τ CI (95%)']
+	features += [x for x in data.columns if 'median' in x and 'τ' not in x]
+
+	table = data[features].copy()
+	table['country'] = [x.replace('_', ' ') for x in table['country']]
+	table.columns = [x.replace(' median', '') for x in table.columns]
+	table = table.rename(columns=
+	    {'country': 'Country', 'official_τ': r'$\tau^*$', 'τ' : r'$\tau$', 'τ CI (75%)': '75\% CI', 'τ CI (95%)': '95\% CI', 
+	     'Δt0': r'$\Delta t$', 'E0': r'$E(0)$', 'Iu0': r'$I_u(0)$', 'μ': r'$\mu$', 'β': r'$\beta$', 'λ': r'$\lambda$', 
+	     'α1': r'$\alpha_1$', 'α2': r'$\alpha_2$', 'Z': '$Z$', 'D': '$D$'
+	    })
+	table.columns = ["\ccell{"+x+"}" for x in table.columns]
+
+	csv_filename = '../figures/Table2.csv'
+	tex_filename = '../figures/Table2.tex'
+	table.to_csv(csv_filename, index=False)
+	with open(tex_filename, 'wt') as f:
+	    print(table.to_latex(escape=False, index=False), file=f)
+	print("Saved output to {} and {}".format(csv_filename, tex_filename))

### 2020-05-11 ms/ms.tex
table 2 read from tex

@@
 \usepackage{subcaption}
+\usepackage{dcolumn}
+\newcolumntype{d}[1]{D{.}{.}{#1}}
+\newcommand\ccell[1]{\multicolumn{1}{c}{#1}} % a wild guess...
+\usepackage{adjustbox}
 
@@
 %%% SEIR model %%%%
-\paragraph*{SEIR model.}
+\paragraph*{SEIR model.} \label{sec:model}
 We model SARS\=/CoV\=/2 infection dynamics by following the number of susceptible $S$, exposed $E$, reported infected $I_r$, and unreported infected $I_u$ individuals in a population of size $N$.
@@
 The parameter $0 < \mu < 1$ represents the decreased transmission rate from unreported infected individuals, who are often subclinical or even asymptomatic.
-The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioural changes of both susceptible and infected individuals.
+The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioral changes of both susceptible and infected individuals.
 Exposed individuals, after an average incubation period of $Z$ days, become reported infected with probability $\alpha_t$ or unreported infected with probability $(1-\alpha_t)$.
@@
 \begin{equation} \label{eq:theta}
-\theta=\Big(Z, D, \mu, \{\beta_t\}, \{\alpha_t\}, \{p_t\}, E(0), I_u(0)\Big), \tau, \Delta t.
+\theta=\Big(Z, D, \mu, \{\beta_t\}, \{\alpha_t\}, \{p_t\}, E(0), I_u(0), \tau, \Delta t \Big).
 \end{equation}
@@
 The effective NPI dates in Italy and Wuhan are significantly delayed compared to the official dates, whereas in Denmark, France, Spain, and Germany, the effective date is earlier than the official date.
-	$\hat{\tau}$ is the posterior median, see \hl{TABLE2}.
+	$\hat{\tau}$ is the posterior median, see \autoref{table:table2}.
 $\tau^*$ is the last NPI date, see \autoref{table:table1}.
@@
 In the following, we describe our findings on late and early effective start of NPI in detail.
+
+%% Table 2 %%%
+\begin{table}[h]
+\centering
+\scriptsize
+\input{../figures/Table2.tex}
+\caption{
+\textbf{Parameter estimates for different regions.}
+See \autoref{eq:model} for model parameters.
+All estimates are posterior medians.
+75\% and 95\% credible intervals given only for $\tau$, in days.
+$\tau^*$ is the official last NPI date, see \autoref{table:table1}.
+}
+\label{table:table2}
+\end{table}
 
@@
     \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
-    Here, model parameters are MAP estimates for Italy (\hl{TABLE}) but with $\lambda=0.5$ and $\alpha_1=\alpha_2$.
+    Here, model parameters are estimates for Italy (\autoref{table:table2}) but with $\lambda=0.5$ and $\alpha_1=\alpha_2$.
     } 

### 2020-05-11 ms/ms.tex
revised median, highlights for todos

@@
 $\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
-We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates.
+We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood, higher DIC (see below), and unreasonable parameter estimates.
 For the effective start of NPIs $\tau$ we have also tested an uninformative uniform prior $U(1,T-1)$.
-DIC (see definition below) was lower for the truncated normal prior in most countries, except \hl{Germany?}.
-More importantly, the uninformative prior could result in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom (this was due to MCMC chains being stuck in low posterior regions of the parameter space).
+DIC (see below) was lower for the truncated normal prior in \hl{most countries}.
+More importantly, the uninformative prior could result in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom.
+This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
 We therefore decided to use the more informative truncated normal prior.
@@
 Italy, Austria), rather than splitting the data at $\tau^*$.
-We then estimate the maximum a posteriori (MAP) estimate $\hat{\tau}=\mathit{argmax}_{\tau}{P(\tau \mid \vec{X})}$. \hl{using median or argmax?}
+We then estimate the posterior probability $P(\tau \mid \vec{X})$ by marginalizing the joint posterior, and estimate $\hat{\tau}$ as the posterior median.
 
@@
 with constant $\beta$ and $\alpha$ (\hl{$\Delta DIC > ?$ for all regions}.)
-We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}): the \hl{95\%} credible interval on $\hat{\tau}$ does not include $\tau^*$, and the DIC of the model with free $\tau$ parameter is lower than that of a model with a fixed $\tau \equiv \tau^*$ (\hl{$\Delta DIC > ?$}.) The exception that proves the rule is \hl{Switzerland}, where the effective and official dates are the same.
-Another important exception is the \hl{United~Kingdom?} % TODO need more data?
-Maybe not such an exception?
+We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in some regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}): the credible interval on $\hat{\tau}$ does not include $\tau^*$, and the DIC of the model with free $\tau$ parameter is lower than that of a model with a fixed $\tau \equiv \tau^*$ (\hl{$\Delta DIC > ?$}.)
 
@@
 
-In Italy, the first case officially confirmed on Feb~21, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
+\hl{In Italy}, the first case officially confirmed on Feb~21, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
 That is, the official date $\tau^*$ is either Mar~8 or~11.
 However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.7 days 95\% CI ; \autoref{fig:fig2}).
-Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate  the effective start of NPIs to be several days layer at around Mar~2 ($\pm$2.65 days 95\% CI \autoref{fig:fig2}).
+Similarly, \hl{in Wuhan, China}, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate  the effective start of NPIs to be several days layer at around Mar~2 ($\pm$2.65 days 95\% CI \autoref{fig:fig2}).
 
@@
     % Wuhan 
-    % τ_hat = 19.33
-	% +-2.65 days for 0.95 quantile
     \begin{subfigure}{0.45\textwidth}
@@
 	% Italy
-	% τ_hat = 24.74
-	% +-0.71 days for 0.95 quantile
     \begin{subfigure}{0.45\textwidth}
@@
 A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ at Mar~8 or~9 ($\pm$2.96 95\%CI), rather than Mar~14 (\autoref{fig:fig3}).
+Nevertheless, we estimate the \hl{effective start of NPI} $\hat{\tau}$ at Mar~8 or~9 ($\pm$2.96 95\%CI), rather than Mar~14 (\autoref{fig:fig3}).
 
-Similarly, in France the official lockdown started at Mar~17 ($\tau^*$), with initial NPIs at Mar 13~\citep{Flaxman2020}. 
+Similarly, in \hl{France} the official lockdown started at Mar~17 ($\tau^*$), with initial NPIs at Mar 13~\citep{Flaxman2020}. 
 However, we estimate the effective start of NPIs $\hat{\tau}$ at Mar~8 ($\pm$5.9 days 95\% CI).
@@
 Moreover, the number of daily cases was similar until Mar~8 in both countries, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
+This may suggest that correlation exist between effective start in NPIs due to global or international events.
 
@@
 We find one case in which the official and effective dates match: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
-Indeed, our MAP estimate $\hat{\tau}$ is Mar~20, and the posterior distribution shows two density peaks: a smaller one between Mar~10 and Mar~14, and a taller one between Mar~17 and Mar~22.
+Indeed, we estimate that $\hat{\tau}$ is \hl{Mar~20}, and the posterior distribution shows two density peaks: a smaller one between Mar~10 and Mar~14, and a taller one between Mar~17 and Mar~22.
 It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.

### 2020-05-11 src/plot_utils.py
more fixes and additions in plots and csv

@@ -52,7 +52,7 @@ def write_csv_header(file_name):
 
     with open(file_name, mode='w') as file: # use pd to write csv files?
         writer = csv.writer(file, delimiter=',', quotechar='"', quoting=csv.QUOTE_MINIMAL)
-        writer.writerow(['country','DIC','MAP loglik','N','p_steps','p_τ_model','p_Td1','p_Td2','official_τ','τ mean','τ median','τ MAP','τ mean from 1 Jan','τ median from 1 Jan','τ MAP from 1 Jan','τ CI (75%)','τ CI (95%)', *params_headers])
+        writer.writerow(['country','DIC','loglik(MAP)','loglik(mean)','loglik(median)','N','p_steps','p_τ_model','p_Td1','p_Td2','official_τ','τ mean','τ median','τ MAP','official τ from 1 Jan','τ mean from 1 Jan','τ median from 1 Jan','τ MAP from 1 Jan','τ CI median (75%)','τ CI median (95%)','τ CI mean (75%)','τ CI mean (95%)', *params_headers])
 
 def write_csv_data(file_name):
     #params means and medians
@@ -67,6 +67,7 @@ def write_csv_data(file_name):
     τ_median =  format(τ_to_string(np.median(τ_posterior)))
     τ_MAP =  format(τ_to_string(MAPs[var_names.index('τ')]))
 
+    τ_official_from1Jar = (pd.to_datetime(start_date) - pd.Timestamp('2020-01-01')).days + official_τ
     τ_mean_from1Jar = (pd.to_datetime(start_date) - pd.Timestamp('2020-01-01')).days + τ_posterior.mean()
     τ_median_from1Jar =  (pd.to_datetime(start_date) - pd.Timestamp('2020-01-01')).days + np.median(τ_posterior)
     τ_MAP_from1Jar =  (pd.to_datetime(start_date) - pd.Timestamp('2020-01-01')).days + MAPs[var_names.index('τ')]
@@ -76,12 +77,12 @@ def write_csv_data(file_name):
     τ_MAP_from1Jar = round(τ_MAP_from1Jar,2)
     with open(file_name, mode='a') as file:
         writer = csv.writer(file, delimiter=',', quotechar='"', quoting=csv.QUOTE_MINIMAL)
[…]

### 2020-05-11 src/inference.py
loading official dates from NPI_dates.csv; normal prior around the middle of NPI

@@ -27,21 +27,14 @@ Td1 = 9
 Td2 = 6
 seed_max = 3000
 
-# TODO read from ../data/NPI_dates.csv, which is also fed to ms.tex
-official_τ_dates = {
-    'Austria' : datetime(2020, 3, 16),
-    'Belgium' : datetime(2020, 3, 18),
-    'Denmark' : datetime(2020, 3, 18),
-    'France' : datetime(2020, 3, 17),
-    'Germany' : datetime(2020, 3, 22),
-    'Italy' : datetime(2020, 3, 11), 
-    'Norway' : datetime(2020, 3, 24),
-    'Spain': datetime(2020, 3, 14),
-    'Sweden': datetime(2020, 3, 18), # only school closure
-    'Switzerland': datetime(2020, 3, 20),
-    'United_Kingdom': datetime(2020, 3, 24),
-    'Wuhan' : datetime(2020, 1, 23)
-}
+
+def get_first_NPI_date(country_name):
+    df = pd.read_csv('../data/NPI_dates.csv',parse_dates=['First','Last'])
+    return df[df['Country']==country_name]['First'].iloc[0].to_pydatetime()
+
+def get_last_NPI_date(country_name):
+    df = pd.read_csv('../data/NPI_dates.csv',parse_dates=['First','Last'])
+    return df[df['Country']==country_name]['Last'].iloc[0].to_pydatetime()
 
 var_names = ['Z', 'D', 'μ', 'β', 'α1', 'λ', 'α2', 'E0', 'Iu0','Δt0','τ']
 
@@ -68,17 +61,24 @@ def find_start_day(cases_and_dates):
     return cases_and_dates.iloc[ind-zeros]['date']
 
 
+def τ_to_string(τ, start_date):
+    return (pd.to_datetime(start_date) + timedelta(days=τ)).strftime('%b %d')
+
 def get_τ_prior(start_date, ndays, country_name, τ_model):
     if τ_model==TauModel.uniform_prior:
         return randint(1,ndays) #[including,not-including]
 
     #normal_prior
-    official_τ_date = official_τ_dates[country_name]
-    official_τ = (official_τ_date-pd.to_datetime(start_date)).days
+    last_τ = (get_last_NPI_date(country_name) - pd.to_datetime(start_date)).days
+    first_τ = (get_first_NPI_date(country_name) - pd.to_datetime(start_date)).days
+
[…]

### 2020-05-11 src/plot_utils.py

@@ -18,7 +18,7 @@ from rakott.mpl import fig_panel_labels
 import warnings
 warnings.filterwarnings('ignore')
 
-from inference import ode, simulate, simulate_one, official_τ_dates, TauModel, log_likelihood, log_prior, get_τ_prior
+from inference import ode, simulate, simulate_one, TauModel, log_likelihood, log_prior, get_τ_prior, get_first_NPI_date, get_last_NPI_date
 
 def load_data(file_name, country_name, burn_fraction=0.6, lim_steps=None):
     # it's the only global point. we initialize all the params here once and don't update it later (only when load_data again for different file_name)
@@ -41,7 +41,7 @@ def load_data(file_name, country_name, burn_fraction=0.6, lim_steps=None):
         sample = chain[:, int(lim_steps * burn_fraction):lim_steps, :].reshape(-1, ndim)
         lnprobability = data['lnprobability'][:, int(lim_steps * burn_fraction):lim_steps]
         logliks = data['logliks'].reshape(nwalkers,nsteps)[:,int(lim_steps * burn_fraction):lim_steps].reshape(-1)
-    official_τ_date = official_τ_dates[country_name]
+    official_τ_date = get_last_NPI_date(country_name)
     official_τ = (official_τ_date - pd.to_datetime(start_date)).days
 
 def write_csv_header(file_name):

### 2020-05-12 src/Table2.py
table header changed

@@ -19,8 +19,8 @@ if __name__ == '__main__':
 	NPI_dates = pd.read_csv('../data/NPI_dates.csv')
 
 	data = pd.read_csv(os.path.join(output_folder, 'tables', 'all-countries-{}.csv'.format(job_id)))
-	features = ['country', 'official_τ', 'τ median', 'τ CI (75%)', 'τ CI (95%)']
-	features += [x for x in data.columns if 'median' in x and 'τ' not in x]
+	features = ['country', 'official_τ', 'τ median', 'τ CI median (75%)', 'τ CI median (95%)']
+	features += [x for x in data.columns if 'median' in x and 'τ' not in x and 'loglik' not in x]
 
 	table = data[features].copy()
 	table['country'] = [x.replace('_', ' ') for x in table['country']]

### 2020-05-12 ms/ms.tex
revised title by UO suggestions
udpate table2 and fig1, revised prior desciption for the new normal we expect to use

@@
 % Title page
-\title{Late and early effective start of non-pharmaceutical interventions inferred during COVID-19 outbreaks }
+\title{Inferring the effective start dates of non-pharmaceutical interventions during COVID-19 outbreaks}
 
@@
 \begin{equation} \label{eq:likelihood}
-\mathbb{L}(\theta \mid \vec{X}) = P(\vec{X} \mid \theta) = P(X_1 \mid \theta) P(X_2 \mid \tilde{X_1}, \theta) \cdots P(X_T \mid \tilde{X}_{T-1}, \theta).
+\mathbb{L}(\theta \mid \vec{X}) = 
+P(\vec{X} \mid \theta) = 
+P(X_1 \mid \theta) \cdot P(X_2 \mid \tilde{X_1}, \theta) \cdots P(X_T \mid \tilde{X}_{T-1}, \theta).
 \end{equation}
@@
 The likelihood function is defined in \autoref{eq:likelihood}.
-We define the following prior distributions on the model parameters $P(\theta)$: 
+The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using an \textit{affine-invariant ensemble sampler for Markov chain Monte Carlo} (MCMC) implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
+
+We defined the following prior distributions on the model parameters $P(\theta)$: 
 \begin{equation} \label{eq:priors}
@@
 I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
-\tau &\sim \mathit{TruncatedNormal}(\tau^*, 5, 1, T-2),
+%\tau &\sim \mathit{TruncatedNormal}(\tau^*, 5, 1, T-2),
+\tau &\sim \mathit{TruncatedNormal}\Big(\frac{\tau^*+\tau^0}{2}, \frac{\tau^*-\tau^0}{2}, 1, T-2\Big),
 \end{aligned}
 \end{equation}
-where $\mathit{TruncatedNormal}(\mu, \sigma, a, b)$ is a truncated normal distribution with mean $\mu$ and standard deviation $\sigma$ taking values between $a$ and $b$; $T$ is the number of days in the data $\vec{X}$; and $\tau^*$ is the official start of the NPI.
-Most priors follow \citet{Li2020}, with the following exceptions.
-$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
-We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood, higher DIC (see below), and unreasonable parameter estimates.
-For the effective start of NPIs $\tau$ we have also tested an uninformative uniform prior $U(1,T-1)$.
+where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:table1}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
+We have also tested an uninformative uniform prior $U(1,T-2)$.
 DIC (see below) was lower for the truncated normal prior in \hl{most countries}.
@@
 We therefore decided to use the more informative truncated normal prior.
-
-The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is then estimated using an \textit{affine-invariant ensemble sampler for Markov chain Monte Carlo} (MCMC) implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
-The maximum a posteriori
+Other priors follow \citet{Li2020}, with the following exceptions.
+$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
+We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood, higher DIC (see below), and unreasonable parameter estimates, and therefore chose $U(1,5)$ as the prior.
 

### 2020-05-12 src/inference.py
fixed delta_t0

@@ -153,7 +153,12 @@ def log_likelihood(θ, X, N):
     τ = int(τ) # for explanation see https://github.com/dfm/emcee/issues/150
     Δt0 = int(Δt0)
 
+    total_zeros = params_bounds['Δt0'][1]
+    unrellevant_zeros = total_zeros - Δt0
+    τ = τ - unrellevant_zeros
+    X = X[unrellevant_zeros:]
     ndays = len(X)
+
     S, E, Ir, Iu, R, Y = simulate(*θ, ndays, N)
     p1 = 1/Td1
     p2 = 1/Td2
@@ -178,7 +183,7 @@ def log_posterior(θ, X, τ_prior, N):
     return logpost
 
 
-def guess_one():
+def guess_one(ndays):
     while True:
         res = prior(ndays,τ_prior)
         if np.isfinite(log_likelihood(res,X,N)):
@@ -221,8 +226,7 @@ if __name__ == '__main__':
     cases_and_dates = df.iloc[::-1][['cases','date']]
     start_date = find_start_day(cases_and_dates)
     X = np.array(cases_and_dates[cases_and_dates['date'] >= start_date]['cases'])
-    ndays = len(X)
-    τ_prior = get_τ_prior(start_date, ndays, country_name,τ_model)
+    τ_prior = get_τ_prior(start_date, len(X), country_name,τ_model)
 
     ndim = len(var_names)
     nwalkers = 50
@@ -232,7 +236,7 @@ if __name__ == '__main__':
     if args.steps:
         nsteps = args.steps
 
-    guesses = np.array([guess_one() for _ in range(nwalkers)])
+    guesses = np.array([guess_one(len(X)) for _ in range(nwalkers)])
     if cores and cores!=1:
         with Pool(cores) as pool:
             sampler = emcee.EnsembleSampler(nwalkers, ndim, log_posterior, args=[X, τ_prior, N], pool=pool)

### 2020-05-12 src/inference.py
fix

@@ -66,7 +66,7 @@ def τ_to_string(τ, start_date):
 
 def get_τ_prior(start_date, ndays, country_name, τ_model):
     if τ_model==TauModel.uniform_prior:
-        return randint(1,ndays) #[including,not-including]
+        return randint(params_bounds['Δt0'][1], ndays) #[including,not-including]
 
     #normal_prior
     last_τ = (get_last_NPI_date(country_name) - pd.to_datetime(start_date)).days
@@ -159,7 +159,7 @@ def log_likelihood(θ, X, N):
     X = X[unrellevant_zeros:]
     ndays = len(X)
 
-    S, E, Ir, Iu, R, Y = simulate(*θ, ndays, N)
+    S, E, Ir, Iu, R, Y = simulate(Z, D, μ, β, α1, λ, α2, E0, Iu0, Δt0, τ, ndays, N)
     p1 = 1/Td1
     p2 = 1/Td2
     Xsum = X.cumsum() 

### 2020-05-12 src/inference.py
f

@@ -29,10 +29,12 @@ seed_max = 3000
 
 
 def get_first_NPI_date(country_name):
+    country_name = 'United Kingdom' if country_name == 'United_Kingdom' else country_name
     df = pd.read_csv('../data/NPI_dates.csv',parse_dates=['First','Last'])
     return df[df['Country']==country_name]['First'].iloc[0].to_pydatetime()
 
 def get_last_NPI_date(country_name):
+    country_name = 'United Kingdom' if country_name == 'United_Kingdom' else country_name
     df = pd.read_csv('../data/NPI_dates.csv',parse_dates=['First','Last'])
     return df[df['Country']==country_name]['Last'].iloc[0].to_pydatetime()
 
@@ -74,11 +76,11 @@ def get_τ_prior(start_date, ndays, country_name, τ_model):
 
     lower = params_bounds['Δt0'][1]
     upper = ndays - 2
-    μ = (last_τ + first_τ) / 2
-    σ = (last_τ - first_τ) / 2
-    σ = σ if σ!=0 else 5 # when last_tau == first_tau
-    print(τ_to_string(μ,start_date), τ_to_string(last_τ,start_date), τ_to_string(first_τ,start_date))
-    print(μ, σ)
+    # μ = (last_τ + first_τ) / 2
+    # σ = (last_τ - first_τ) / 2
+    # σ = 5 if σ<5 else σ 
+    μ = last_τ
+    σ = 5
     return truncnorm(
         (lower - μ) / σ, (upper - μ) / σ, loc=μ, scale=σ)
 

### 2020-05-12 src/plot_utils.py
fixed plot_incidences

@@ -18,7 +18,7 @@ from rakott.mpl import fig_panel_labels
 import warnings
 warnings.filterwarnings('ignore')
 
-from inference import ode, simulate, simulate_one, TauModel, log_likelihood, log_prior, get_τ_prior, get_first_NPI_date, get_last_NPI_date
+from inference import ode, simulate, simulate_one, TauModel, log_likelihood, log_prior, get_τ_prior, get_first_NPI_date, get_last_NPI_date, params_bounds
 
 def load_data(file_name, country_name, burn_fraction=0.6, lim_steps=None):
     # it's the only global point. we initialize all the params here once and don't update it later (only when load_data again for different file_name)
@@ -414,15 +414,24 @@ def print_τ_dist():
     return sorted(s,key=lambda t: pd.Timestamp('2020 '+t[0]))
 
 def plot_incidences(ax=None, color=blue):
+    global daily_cases, theta
     if ax is None: fig, ax = plt.subplots()
 
     np.random.seed(10)
-    num_simulations = 300
+    num_simulations = 128
     daily_cases = []
-    for _ in range(num_simulations):
+    for i in range(num_simulations):
         idx = np.random.choice(sample.shape[0])
-        y = generate(*sample[idx, :],ndays,N)
-        daily_cases.append(y)
+        
+        Z, D, μ, β, α1, λ, α2, E0, Iu0, Δt0, τ = sample[idx, :]
+        Δt0 = int(Δt0)
+        τ = int(τ)
+        total_zeros = params_bounds['Δt0'][1]
+        unrellevant_zeros = total_zeros - Δt0
+        τ = τ - unrellevant_zeros
+        y = generate(Z, D, μ, β, α1, λ, α2, E0, Iu0, Δt0, τ, ndays-unrellevant_zeros ,N)       
+        
+        daily_cases.append(np.array([0]*unrellevant_zeros + list(y)))
     daily_cases = np.array(daily_cases)
 
     t = np.arange(ndays)

### 2020-05-17 ms/ms.tex
ms typo
update likelihood; added citation

@@
 \begin{aligned}
-\frac{dS}{dt} & = -\beta_t S \frac{I_p}{N} - \mu \beta_t S \frac{I_s}{N} \\
-\frac{dE}{dt} & = \beta_t S \frac{I_p}{N} + \mu \beta_t S \frac{I_s}{N}  - \frac{E}{Z} \\
+\frac{dS}{dt} & = -\beta_t S \frac{I_r}{N} - \mu \beta_t S \frac{I_u}{N} \\
+\frac{dE}{dt} & = \beta_t S \frac{I_r}{N} + \mu \beta_t S \frac{I_u}{N}  - \frac{E}{Z} \\
 \frac{dI_r}{dt} & = \alpha_t \frac{E}{Z} - \frac{I_r}{D} \\
@@
 \paragraph*{Likelihood function.}
-The \emph{expected} cumulative number of reported infected individuals until day $t$ is 
+% TODO update
+For a given vector $\theta$ of model parameters the \emph{expected} cumulative number of reported infected individuals ($I_r$) until day $t$ is, following~\autoref{eq:model},
 \begin{equation} \label{eq:Yt}
-Y_t=\int_{0}^{t}{\alpha_s \frac{E(s)}{Z} \; ds}, \quad Y_0 = 0.
+Y_t(\theta) = \int_{0}^{t}{\alpha_s \frac{E(s)}{Z} \; ds}, \quad Y_0 = 0.
 \end{equation}
 We assume that reported infected individuals are confirmed and therefore observed in the daily case report of day $t$ with probability $p_t$ (note that an individual can only be observed once, and that $p_t$ may change over time, but $t$ is a specific date rather than the time elapsed since the individual was infected).
-Hence, we assume that the number of confirmed cases in day $t$ is binomially distributed,
-$$
-X_t \sim \mathit{Bin}\big(n_t, p_t\big),
-$$
-where $n_t$ is the \emph{realized} (rather than expected) number of reported infected individuals yet to appear in daily reports by day~$t$.
-The cumulative number of confirmed cases until day $t$ is
-$$
-\tilde{X}_t=\sum_{i=1}^{t}X_i, \quad X_0=0.
-$$
-Given $\tilde{X}_{t-1}$, we assume $n_t$ is Poisson distributed,
-$$
-\big(n_t \mid \tilde{X}_{t-1}\big) \sim \mathit{Poi}\Big( Y_t - \tilde{X}_{t-1} \Big), 
-\quad n_1 \sim \mathit{Poi}(Y_1).
-$$ 
-Therefore, $\big(X_t \mid \tilde{X}_{t-1} \big)$ is a binomial conditioned on a Poisson, which reduces to a Poisson with
-\begin{equation} \label{eq:P(Xt|Xt-1)}
-\big(X_t \mid \tilde{X}_{t-1} \big) \sim \mathit{Poi}\Big( \big( Y_t - \tilde{X}_{t-1} \big) \cdot p_t \Big), 
-\quad X_1 \sim \mathit{Poi}(Y_1 \cdot p_1).
+We denote by $X_t$ the number of confirmed cases in day $t$, and by $\tilde{X}_t$ the cumulative number of confirmed cases until day $t$,
+\begin{equation} \label{eq:Xsumt}
+\tilde{X}_t=\sum_{i=1}^{t}X_i.
 \end{equation}
+Therefore, at day $t$ the number of reported infected yet-to-be confirmed individuals is
+$(Y_t(\theta) - \tilde{X}_{t-1})$.
+We therefore assume that $X_t$ conditioned on $\tilde{X}_{t-1}$ is Poisson distributed,
+\begin{equation} \label{eq:Xt} \begin{aligned}
+\Big(X_1 \mid \theta \Big) & \sim \mathit{Poi}\big( Y_1(\theta) \cdot p_1 \big), \\
+\Big(X_t \mid \tilde{X}_{t-1}, \theta \Big) & \sim 
+\mathit{Poi}\Big( \big(Y_t(\theta) - \tilde{X}_{t-1}\big) \cdot p_t \Big), \quad t>1.
+\end{aligned}\end{equation}
 
-For given vector $\theta$ of model parameters (\autoref{eq:theta}), we compute the expected cumulative number of reported infected individuals $\{Y_t\}_{t=1}^{T}$ for each day (\autoref{eq:Yt}).
-Then, since $\tilde{X}_{t-1}$ is a function of $X_1, \ldots, X_{t-1}$, we can use \autoref{eq:P(Xt|Xt-1)} to write the probability to observe the confirmed case data $\vec{X} = (X_1, \ldots, X_T)$ as
+Hence, the \emph{likelihood function} $\mathbb{L}(\theta \mid \vec{X})$ for the parameter vector $\theta$ given the confirmed case data $\vec{X} = (X_1, \ldots, X_T)$ is defined by the probability to observe  $\vec{X}$ given $\theta$,
 \begin{equation} \label{eq:likelihood}
@@
 \end{equation}
-This defines a \emph{likelihood function} $\mathbb{L}(\theta \mid \vec{X})$ for the parameter vector $\theta$ given the data $\vec{X}$.
 
@@
 where $0 < \lambda < 1$.
-The values for $p_t$ follow \citet{Li2020}, who estimated the average time between infection and reporting in Wuhan, China, at 9 days before the start of NPIs (Jan 23, 2020) and 6 days after start of NPIs.
-The parameter $\tau$ is then added to the parameter vector $\theta$ (\autoref{eq:theta}).
+The values for $p_t$ follow \citet{Li2020}, who estimated the average time between infection and reporting in Wuhan, China, at 9 days before the start of NPIs and 6 days after start of NPIs.
 
@@
 \paragraph*{Parameter estimation.}
-To estimate the parameters of our model from the data $\vec{X}$, we apply a Bayesian inference approach.
-We start our model $\Delta t$ days before the outbreak (defined as consecutive days with increasing confirmed cases) in each country~\citep{Gatto2020}.
+To estimate the model parameters from the daily case data $\vec{X}$, we apply a Bayesian inference approach.
+We start our model $\Delta t$ days~\citep{Gatto2020} before the outbreak (defined as consecutive days with increasing confirmed cases) in each country.
 The model in \autoref{eq:model} is parameterized by the vector $\theta$, where
@@
 The likelihood function is defined in \autoref{eq:likelihood}.
-The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using an \textit{affine-invariant ensemble sampler for Markov chain Monte Carlo} (MCMC) implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
+The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using an \textit{affine-invariant ensemble sampler for Markov chain Monte Carlo} (MCMC)~\citep{Goodman2010} implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
 
@@
 \paragraph*{Source code.} 
-We use Python 3 (Anaconda) with the NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee packages.
+We use Python 3 with the NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee packages.
 All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
@@
 Thus, our results highlight the need to further study variability in compliance and behavior over both time and space.
-This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Fenichela2011,Arthur2020}.
+This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Fenichela2011,Walters2013,Arthur2020}.
 

### 2020-05-19 ms/ms.tex
added ppc figures

@@
 \usepackage{subcaption}
+\captionsetup[subfigure]{labelformat=empty}
 \usepackage{dcolumn}
@@
 
+%%% Fig PPC %%%
+\begin{figure}[h]
+    \centering
+	\subcaption{Austria}
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Austria_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Austria_ppc.pdf}
+    \end{subfigure}
+    %%%
+	\subcaption{Belgium}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Belgium_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Belgium_ppc.pdf}
+    \end{subfigure}
+    %%%
+	\subcaption{Denmark}    
+     \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Denmark_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Denmark_ppc.pdf}
+    \end{subfigure}
+\end{figure}
+    %%%
+\begin{figure}
+    \ContinuedFloat 
+	\subcaption{France}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/France_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/France_ppc.pdf}
+    \end{subfigure}
+    %%%  
+	\subcaption{Germany}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Germany_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Germany_ppc.pdf}
+    \end{subfigure}
+    %%% 
+	\subcaption{Italy}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Italy_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Italy_ppc.pdf}
+    \end{subfigure}
+    %%% 
+	\subcaption{Norway}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Norway_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Norway_ppc.pdf}
+    \end{subfigure}
+\end{figure}
+    %%%
+\begin{figure}
+    \ContinuedFloat   
+	\subcaption{Spain}     
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Spain_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Spain_ppc.pdf}
+    \end{subfigure}
+    %%%  
+	\subcaption{Sweden}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Sweden_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Sweden_ppc.pdf}
+    \end{subfigure}
+    %%%  
+	\subcaption{Switzerland}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Switzerland_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Switzerland_ppc.pdf}
+    \end{subfigure}
+    %%%  
+	\subcaption{United Kingdom}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/United_Kingdom_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/United_Kingdom_ppc.pdf}
+    \end{subfigure}
+    %%%
+    \caption{
+     \textbf{Figure S1. Posterior predictive plots.}
+	}
+	\label{fig:ppc}
+\end{figure}
+
+\pagebreak
+
 %%% Fig Fr-vs-Es %%%

### 2020-05-19 ms/ms.tex
better captions for ppc fig

@@
     \caption{
-     \textbf{Figure S1. Posterior predictive plots.}
-	}
+     \textbf{Figure S1. Posterior prediction check plots}
+     Markers represent data ($\vec{X}$).
+Black line represent a smoothing of the data points using a Savitzky-Golay filter.
+Color lines represent posterior predictions from a model with fixed $\tau$, in red, and free $\tau$, in blue.
+These predictions are made by drawing 1,000 samples from the parameter posterior distribution and then generating a daily case count using the SEIR model in \autoref{eq:model}.
+Note the differences in the y-axis scale.}
 	\label{fig:ppc}

### 2020-05-19 ms/ms.tex
added fig for NPI dates; PPC plots; Swiss posterior fig; landscape table 2; revised text according to May 14-15 inference results

@@
 \usepackage{adjustbox}
+\usepackage{lscape}
 
@@
 Dates for European countries from \citet{Flaxman2020}, date for Wuhan, China from \citet{Pei2020}.
+See \autoref{fig:NPI_dates} for a visual presentation.
 }
@@
 I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
-%\tau &\sim \mathit{TruncatedNormal}(\tau^*, 5, 1, T-2),
+\Delta t & \sim \mathit{Uniform}(1, 5) \\
 \tau &\sim \mathit{TruncatedNormal}\Big(\frac{\tau^*+\tau^0}{2}, \frac{\tau^*-\tau^0}{2}, 1, T-2\Big),
@@
 where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:table1}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
+% TODO something formal for the next line?
 We have also tested an uninformative uniform prior $U(1,T-2)$.
-DIC (see below) was lower for the truncated normal prior in \hl{most countries}.
-More importantly, the uninformative prior could result in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom.
+The uninformative prior could result in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom.
 This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
@@
 All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
+Files containing samples from the posterior distributions will be deposited on \hl{FigShare}.
 
@@
 
-We find that a model that considers an NPI (\autoref{eq:NPI_model}) is a better fit to the data than a model without an NPI, i.e.
-with constant $\beta$ and $\alpha$ (\hl{$\Delta DIC > ?$ for all regions}.)
-We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in some regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}): the credible interval on $\hat{\tau}$ does not include $\tau^*$, and the DIC of the model with free $\tau$ parameter is lower than that of a model with a fixed $\tau \equiv \tau^*$ (\hl{$\Delta DIC > ?$}.)
+We find that a model that considers an NPI (\autoref{eq:NPI_model}) is a better fit to the data than a model without an NPI, i.e. with constant $\beta$ and $\alpha$ (\hl{???}.)
+We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}).
+Indeed, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:fig1}).
+Moreover, we compared the posterior predictive plots of a model with a free $\tau$ with those of a model with $\tau$ fixed at $\tau^*$.
+The model with free $\tau$ clearly produces better and less variable predictions (\autoref{fig:ppc}).
 
@@
 %% Table 2 %%%
+\begin{landscape}
 \begin{table}[h]
 \centering
-\scriptsize
 \input{../figures/Table2.tex}
@@
 \end{table}
+\end{landscape}
 
@@
 
-\hl{In Italy}, the first case officially confirmed on Feb~21, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
-That is, the official date $\tau^*$ is either Mar~8 or~11.
-However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.7 days 95\% CI ; \autoref{fig:fig2}).
-Similarly, \hl{in Wuhan, China}, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate  the effective start of NPIs to be several days layer at around Mar~2 ($\pm$2.65 days 95\% CI \autoref{fig:fig2}).
+In Italy, the first case was officially confirmed on Feb~21.
+School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
+That is, the first and last official dates are Mar~8 and Mar~11.
+However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.47 days 95\% CI ; \autoref{fig:fig2}).
+Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be several days layer at Feb~2 ($\pm$2.85 days 95\% CI \autoref{fig:fig2}).
 
@@
     \centering
-    % Wuhan 
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/Fig2a.pdf}
+        \includegraphics[width=\textwidth]{../figures/posterior/Wuhan_τ_posterior.pdf}
     \end{subfigure}
   	~
-	% Italy
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/Fig2b.pdf}
+        \includegraphics[width=\textwidth]{../figures/posterior/Italy_τ_posterior.pdf}
     \end{subfigure}
@@
 A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the \hl{effective start of NPI} $\hat{\tau}$ at Mar~8 or~9 ($\pm$2.96 95\%CI), rather than Mar~14 (\autoref{fig:fig3}).
+Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8 or~9 ($\pm$1.08 95\%CI), rather than Mar~14 (\autoref{fig:fig3}).
 
-Similarly, in \hl{France} the official lockdown started at Mar~17 ($\tau^*$), with initial NPIs at Mar 13~\citep{Flaxman2020}. 
-However, we estimate the effective start of NPIs $\hat{\tau}$ at Mar~8 ($\pm$5.9 days 95\% CI).
-Although the credible interval is wide, spanning from Mar~2 to Mar~13, the official lockdown start at Mar~17 is later still (\autoref{fig:fig3}).
+Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ on Mar~8 or Mar~9 ($\pm$6.27 days 95\% CI, \autoref{fig:fig3}).
+Although the credible interval is wider compared to Spain, spanning from Mar~2 to Mar~15, the official lockdown start at Mar~17 is later still, and even the earliest NPI, banning of public events, only started on Mar~13~\citep{Flaxman2020}.
 
-Interestingly, the effective start of NPIs $\hat{\tau}$ in both France and Spain is estimated at Mar~8, although the official dates are differ by three days.
-Moreover, the number of daily cases was similar until Mar~8 in both countries, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
-This may suggest that correlation exist between effective start in NPIs due to global or international events.
+Interestingly, the effective start of NPIs $\hat{\tau}$ in both France and Spain is estimated at Mar~8, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
+The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
+This may suggest that correlation exist between effective start of NPIs due to global or international events.
 
@@
     \centering
-    % France
-    % τ_hat = 12.38
-	% +-5.90 days for 0.95 quantile
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/Fig3a.pdf}
+        \includegraphics[width=\textwidth]{../figures/posterior/France_τ_posterior.pdf}
     \end{subfigure}
     ~
-    % Spain
-    % τ_hat = 13.95
-	% +-2.96 days for 0.95 quantile
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/Fig3b.pdf}
+        \includegraphics[width=\textwidth]{../figures/posterior/Spain_τ_posterior.pdf}
     \end{subfigure}
@@
 We find one case in which the official and effective dates match: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
-Indeed, we estimate that $\hat{\tau}$ is \hl{Mar~20}, and the posterior distribution shows two density peaks: a smaller one between Mar~10 and Mar~14, and a taller one between Mar~17 and Mar~22.
+Indeed, the posterior median $\hat{\tau}$ is Mar~20 ($\pm$8.46 days 95\% CI), and the posterior distribution shows two density peaks: a smaller one between Mar~10 and Mar~14, and a bigger one between Mar~17 and Mar~22~(\autoref{fig:Switzerland}).
 It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
@@
 
-For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred 3-5 after the official date (\autoref{fig:fig2}).
-This could be explained by low compliance.
-In Italy, for example, a leak about the intent to lockdown Northern provinces results in people leaving those provinces~\citep{Gatto2020}.
-However, late effect of NPIs could also be due to the time required by both the government and the citizens to organize for a lockdown.
+For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred more than five days after the official date (\autoref{fig:fig2}).
+This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
+Late effect of NPIs might also be due to the time required by both the government and the citizens to organize for a lockdown, and for the new guidelines to diffuse in the population.
 
-In contrast, in most investigated countries, such as Spain and France, transmission rates seem to have been reduced even before official lockdowns were implemented (\autoref{fig:fig3}).
-This early response is possibly due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response increased risk perception due to domestic or international COVID\=/19-related reports.
-This finding may also suggest that severe NPIs, such as lockdowns, were unnecessary, and that milder measures that were adopted by the population, possibly due to government recommendations, media coverage, and social networks, could have been sufficient for epidemic control.
+In contrast, in most investigated countries (e.g., Spain and France), we infer reduced transmission rates even before official lockdowns were implemented (\autoref{fig:fig3}).
+This early response might be due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response to increased risk perception due to domestic or international COVID\=/19-related reports.
+This finding may also suggest that severe NPIs, such as lockdowns, were unnecessary, and that less extreme measures adopted by the population could have been sufficient for epidemic control.
+These less extreme measures may have been implemented due to government recommendations, media coverage, and social networks, rather than official NPIs.
 \hl{check if this is true} Indeed, the evidence supports a change in transmission dynamics (i.e.
@@
 
-Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a 7 day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
+\hl{Attempts to asses the effect} of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a 7 day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
 However, the late and early effects we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).
 
-We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point.
+We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ \hl{over a model without such a change-point}.
 It may be interesting to investigate if the evidence favors a model with \emph{two} change-points, rather than one.
 Two such change-points could reflect escalating NPIs (e.g.
-school closures followed by lockdowns), a mix of NPIs and changes in weather, a mix of domestic and international effects on risk perception, or other similar factors.
+school closures followed by lockdowns), or a mix of NPIs and other events, such as weather, or domestic and international events that affect risk perception.
 
@@
 We have estimated the effective start date of NPIs and found that they often differ from the official dates.
-Our results emphasize the complex interaction between personal, regional, and global determinants of behavioral.
-Thus, our results highlight the need to further study variability in compliance and behavior over both time and space.
+Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to infectious disease.
+Therefore, we emphasize the need to further study variability in compliance and behavior over both time and space.
 This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Fenichela2011,Walters2013,Arthur2020}.
@@
 \section*{Acknowledgements}
-%We thank XXX for discussions and comments.
+We thank Lilach Hadany and Oren Kolodny for discussions and comments.
 This work was supported in part by the Israel Science Foundation 552/19 and 1399/17.
@@
 These predictions are made by drawing 1,000 samples from the parameter posterior distribution and then generating a daily case count using the SEIR model in \autoref{eq:model}.
-Note the differences in the y-axis scale.}
+Note the differences in the y-axis scale.
+    }
 	\label{fig:ppc}
@@
 \pagebreak
+
+%%% Fig NPI dates %%%
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=0.5\textwidth]{../figures/Fig-NPI_dates.pdf}
+    \caption{
+    \textbf{Official start of non-pharmaceutical interventions.}
+	See \autoref{table:table1} for more details. Wuhan, China is not shown.
+    } 
+    \label{fig:NPI_dates}
+\end{figure}
 
@@
 
+%%% Fig Switzerland %%%
+\begin{figure}[h]
+    \centering
+        \includegraphics[width=0.75\textwidth]{../figures/posterior/Switzerland_τ_posterior.pdf}
+    \caption{
+	\textbf{Late effect of non-pharmaceutical interventions in Switzerland.}
+    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
+Red line shows the official last NPI date $\tau^*$.
+Black line shows the estimated $\hat{\tau}$.
+Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
+    }
+    \label{fig:Switzerland}
+\end{figure}
+
 \end{document}

### 2020-05-19 src/Table2.py
wuhan china, sort by name

@@ -32,6 +32,8 @@ if __name__ == '__main__':
 	     'Δt0': r'$\Delta t$', 'E0': r'$E(0)$', 'Iu0': r'$I_u(0)$', 'μ': r'$\mu$', 'β': r'$\beta$', 'λ': r'$\lambda$', 
 	     'α1': r'$\alpha_1$', 'α2': r'$\alpha_2$', 'Z': '$Z$', 'D': '$D$'
 	    })
+	table = table.sort_values('Country')
+	table.loc[table['Country'] == 'Wuhan', 'Country'] = 'Wuhan, China'
 	table.columns = ["\ccell{"+x+"}" for x in table.columns]
 
 	csv_filename = '../figures/Table2.csv'

### 2020-05-19 ms/ms.tex
fix fig 4

@@
 However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
-Such a case is illustrated in \autoref{fig:fig4} using data and parameters from Italy.
-Here, a lockdown is officially ordered on Mar~10 ($\tau^*$, but its late effect on the transmission dynamics starts on Mar~15 ($\hat{\tau}$).
-If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases to follow the dashed red line.
-However, the number of cases will actually follow the black line, leading to a significant different ($\Delta$) between the projections and the realization.
+This is illustrated in \autoref{fig:fig4} using data and parameters from Italy.
+Here, a lockdown is officially ordered on Mar~10 ($\tau^*$), but its late effect on the transmission dynamics starts on Mar~16 ($\hat{\tau}$).
+If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases be within the red lines (posterior predictions assuming $\tau=\tau^*$).
+This leads to a significant underestimation, which might be interpreted by officials as ineffectiveness of NPIs, leading to further escalations.
+However, the number of cases will actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data.
 
@@
     Real number of daily cases in Italy in black (markers: data, line: time moving average). 
-Model predictions, assuming a 50\% decrease in transmission rate after the NPI starts, are shown as colored lines with 95\% confidence intervals.
+    Model posterior predictions are shown as colored lines (1,000 draws from the posterior distribution). 
     Shaded box illustrates a serial interval of seven days.
@@
     \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
-    Here, model parameters are estimates for Italy (\autoref{table:table2}) but with $\lambda=0.5$ and $\alpha_1=\alpha_2$.
+    Here, model parameters are estimates for Italy (\autoref{table:table2}).
     } 
@@
 
-\hl{Attempts to asses the effect} of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a 7 day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
+Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a seven days delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
 However, the late and early effects we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).

### 2020-05-19 src/plot_utils.py
ppc refactoring

@@ -22,10 +22,12 @@ from model.normal_prior_model import NormalPriorModel
 from model.uniform_prior_model import UniformPriorModel
 from inference import get_first_NPI_date, get_last_NPI_date, params_bounds, params_bounds, get_model_class
 
-def load_data(file_name, country_name, burn_fraction=0.6, lim_steps=None):
+country_name = None
+def load_data(file_name, _country_name, burn_fraction=0.6, lim_steps=None):
     # it's the only global point. we initialize all the params here once and don't update it later (only when load_data again for different file_name)
     # TODO PLEASE DONT USE global anywhere in your code
-    global first_NPI, last_NPI, incidences, start_date, var_names, nsteps, ndim, N, Td1, Td2, ndays, sample, lnprobability, logliks, model_type, model
+    global first_NPI, last_NPI, incidences, start_date, var_names, nsteps, ndim, N, Td1, Td2, ndays, sample, lnprobability, logliks, model_type, model, country_name
+    country_name = _country_name
     data = np.load(file_name)
     incidences = data['incidences']
     start_date = data['start_date']

### 2020-05-19 src/ppc.py

@@ -22,7 +22,7 @@ from click_spinner import spinner
 from inference import get_last_NPI_date
 from inference import get_first_NPI_date
 from inference import params_bounds
-from inference import find_start_day
+from inference import get_model_class
 from model.normal_prior_model import NormalPriorModel
 from model.fixed_tau_model import FixedTauModel
 
@@ -42,34 +42,22 @@ def date_to_date(x):
 def τ_to_string(τ, start_date):
     return (pd.to_datetime(start_date) + timedelta(days=τ)).strftime('%b %d')
 
-
-def load_data(country):
-	url = 'https://github.com/ImperialCollegeLondon/covid19model/raw/v1.0/data/COVID-19-up-to-date.csv'
-	fname = '../data/COVID-19-up-to-date.csv'
-	if not os.path.exists(fname):
-	    urllib.request.urlretrieve(url, fname)
-	data = pd.read_csv(fname, encoding='iso-8859-1')
-	data['date'] = pd.to_datetime(data['dateRep'], format='%d/%m/%Y')
-	data = data[data['countriesAndTerritories'] == country]
-	N = data.iloc[0]['popData2018']
-	cases_and_dates = data.iloc[::-1][['cases','date']]
-	start_date = find_start_day(cases_and_dates)
-	X = np.array(cases_and_dates[cases_and_dates['date'] >= start_date]['cases'])	
-	return X, start_date, N
-
-def load_chain(job_id):
+def load_chain(job_id, burn_fraction=0.6):
 	with spinner():
 		fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
 		inference_data = np.load(fname)
 		chain = inference_data['chain']
 		var_names = inference_data['var_names']
+		nsteps, ndim, N, Td1, Td2, model_type = inference_data['params']
+		X = inference_data['incidences']
+		start_date = inference_data['start_date']
 		# print("Loaded {} with parameters:".format(fname))
 		# print(var_names)
 		nchains, nsteps, ndim = chain.shape
-		chain = chain[:, nsteps//2:, :]
+		nburn = int(nsteps * burn_fraction)
+		chain = chain[:, nburn:, :]
 		chain = chain.reshape((-1, ndim))
-		return chain, var_names
-
+		return chain, Td1, Td2, model_type, X, start_date, N
 
[…]

### 2020-05-20 src/ppc.py
fname for load_chain

@@ -42,9 +42,10 @@ def date_to_date(x):
 def τ_to_string(τ, start_date):
     return (pd.to_datetime(start_date) + timedelta(days=τ)).strftime('%b %d')
 
-def load_chain(job_id, burn_fraction=0.6):
+def load_chain(job_id=None, fname=None, burn_fraction=0.6):
 	with spinner():
-		fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
+		if fname is None:
+			fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
 		inference_data = np.load(fname)
 		chain = inference_data['chain']
 		var_names = inference_data['var_names']
@@ -76,7 +77,8 @@ if __name__ == '__main__':
 			color = colors[color]
 	else:
 		color = blue
-	chain, Td1, Td2, model_type, X, start_date, N = load_chain(job_id)
+	chain_fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
+	chain, Td1, Td2, model_type, X, start_date, N = load_chain(fname=chain_fname)
 	ndays = len(X)
 	X_mean = scipy.signal.savgol_filter(X, 3, 1)
 	

### 2020-05-20 src/inference.py
check folder exists before removing it

@@ -169,5 +169,7 @@ if __name__ == '__main__':
     if not os.path.exists(os.path.join(output_folder,'src')):
         os.mkdir(os.path.join(output_folder,'src'))
     copyfile(sys.argv[0], os.path.join(output_folder,'src', sys.argv[0])) 
-    rmtree(os.path.join(output_folder,'src', 'model'))
-    copytree('model', os.path.join(output_folder,'src', 'model'))
+    model_path = os.path.join(output_folder,'src', 'model')
+    if os.path.exists(model_path): 
+        rmtree(model_path)
+    copytree('model', model_path)

### 2020-05-20 src/ppc.py
predict one week after last inference date; compute P(max pred > max data); plot one week ahead

@@ -6,6 +6,7 @@ from datetime import datetime, timedelta
 import urllib
 
 import matplotlib as mpl
+mpl.use("Agg")
 import matplotlib.pyplot as plt
 import numpy as np
 import scipy.stats
@@ -23,6 +24,7 @@ from inference import get_last_NPI_date
 from inference import get_first_NPI_date
 from inference import params_bounds
 from inference import get_model_class
+from inference import find_start_day
 from model.normal_prior_model import NormalPriorModel
 from model.fixed_tau_model import FixedTauModel
 
@@ -66,9 +68,39 @@ def posterior_prediction(chain, model, nreps):
 		model.generate_daily_cases(θi) for θi in θ
 	])
 
+
+def load_data(country_name, up_to_date=None):
+	if country_name=='Wuhan':
+	    df = pd.read_csv('../data/Incidence.csv')
+	    df['date'] = pd.to_datetime(df['Date'], dayfirst=True)
+	    df['cases'] = df[country_name]
+	    df = df[::-1] # TODO why?
+	    N = pd.read_csv('../data/pop.csv', index_col='City').loc[country_name].values[0]
+	else:
+	    url = 'https://github.com/ImperialCollegeLondon/covid19model/raw/master/data/COVID-19-up-to-date.csv'
+	    fname = '../data/COVID-19-up-to-date_master.csv'
+	    if not os.path.exists(fname):
+	        urllib.request.urlretrieve(url, fname)
+	    df = pd.read_csv(fname, encoding='iso-8859-1')
+	    df['date'] = pd.to_datetime(df['dateRep'], format='%d/%m/%Y')
+	    df = df[df['countriesAndTerritories'] == country_name]
+	    N = df.iloc[0]['popData2018']
+
+	cases_and_dates = df.iloc[::-1][['cases','date']]
+	if up_to_date:
+	    cases_and_dates = cases_and_dates[cases_and_dates['date']<=up_to_date]
+	start_date = find_start_day(cases_and_dates)
+	X = cases_and_dates.loc[cases_and_dates['date'] >= start_date, 'cases'].values
+	T = cases_and_dates.loc[cases_and_dates['date'] >= start_date, 'date']
+	return X, T
+
+
 if __name__ == '__main__':
 	nreps = 1000
[…]

### 2020-05-20 src/ppc.sh
run all

@@ -1,0 +1,9 @@
+for country in Austria Belgium Denmark France Germany Italy Norway Spain Sweden Switzerland United_Kingdom
+do
+	python ppc.py 2020-05-14-n1-normal-1M $country blue >> free_tau_loss.txt
+	python ppc.py 2020-05-14-n1-notau-1M $country green >> no_tau_loss.txt
+	python ppc.py 2020-05-15-n1-fixed-tau-1M $country red >> fixed_tau_loss.txt
+done
+
+
+

### 2020-05-20 ms/ms.tex
added WAIC

@@
 % \modulolinenumbers[2]
+
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
 
@@
 In contrast, in most investigated countries (e.g., Spain and France), we infer reduced transmission rates even before official lockdowns were implemented (\autoref{fig:fig3}).
-This early response might be due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response to increased risk perception due to domestic or international COVID\=/19-related reports.
+An early effective date might be due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response to increased risk perception due to domestic or international COVID\=/19-related reports.
 This finding may also suggest that severe NPIs, such as lockdowns, were unnecessary, and that less extreme measures adopted by the population could have been sufficient for epidemic control.
 These less extreme measures may have been implemented due to government recommendations, media coverage, and social networks, rather than official NPIs.
-\hl{check if this is true} Indeed, the evidence supports a change in transmission dynamics (i.e.
-a model with $\tau$) even for Sweden, in which a lockdown was not implemented, suggesting that lockdowns may not be necessary if other NPIs are adopted early enough during the outbreak~\citep{Banholzer2020} (Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.)
+Indeed, the evidence supports a change in transmission dynamics (i.e.
+a model with free $\tau$) even for Sweden~(\autoref{fig:ppc}), in which a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}, suggesting that lockdowns may not be necessary if other NPIs are adopted early enough during the outbreak~\citep{Banholzer2020}.
 
@@
 
+%%% Fig NPI dates %%%
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=0.5\textwidth]{../figures/Fig-NPI_dates.pdf}
+    \caption{
+    \textbf{Official start of non-pharmaceutical interventions.}
+	See \autoref{table:table1} for more details. Wuhan, China is not shown.
+    } 
+    \label{fig:NPI_dates}
+\end{figure}
+
 %%% Fig PPC %%%
@@
     \end{subfigure}
+    \label{fig:ppc}
     ~
@@
     \end{subfigure}
-\end{figure}
 	%%%
-\begin{figure}
-    \ContinuedFloat 
 	\subcaption{France}    
@@
     \end{subfigure}
+\end{figure}
     %%%
+\begin{figure}
+    \ContinuedFloat 
 	\subcaption{Germany}    
@@
     \end{subfigure}
-\end{figure}
     %%%  
-\begin{figure}
-    \ContinuedFloat   
 	\subcaption{Spain}     
@@
     \end{subfigure}
+\end{figure}
     %%%
+\begin{figure}
+    \ContinuedFloat 
 	\subcaption{Sweden}    
@@
     %%%  
+	\subcaption{Wuhan, China}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Wuhan_ppc.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Wuhan_ppc.pdf}
+    \end{subfigure}
+    %%%
     \caption{
-    \textbf{Figure S1. Posterior prediction check plots}
+    \textbf{Figure S2. Posterior prediction check plots}
     Markers represent data ($\vec{X}$).
@@
     }	
-	\label{fig:ppc}
 \end{figure}
@@
 \pagebreak
-
-%%% Fig NPI dates %%%
-\begin{figure}[h]
-    \centering
-	\includegraphics[width=0.5\textwidth]{../figures/Fig-NPI_dates.pdf}
-    \caption{
-    \textbf{Official start of non-pharmaceutical interventions.}
-	See \autoref{table:table1} for more details. Wuhan, China is not shown.
-    } 
-    \label{fig:NPI_dates}
-\end{figure}
 
@@
 
-%%% Fig Switzerland %%%
-\begin{figure}[h]
+% Table WAIC %%%
+\begin{table}[h]
 \centering
-        \includegraphics[width=0.75\textwidth]{../figures/posterior/Switzerland_τ_posterior.pdf}
+\pgfplotstabletypeset[
+    col sep=comma,
+    string type,
+    every head row/.style={before row=\hline,after row=\hline},
+    every last row/.style={after row=\hline},
+    ]{../figures/Table-WAIC.csv}
 \caption{
-	\textbf{Late effect of non-pharmaceutical interventions in Switzerland.}
-    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
-Red line shows the official last NPI date $\tau^*$.
-Black line shows the estimated $\hat{\tau}$.
-Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
+\textbf{WAIC values for the different models.}
+	WAIC (widely applicable information criterion)~\citep{gelman2013bayesian} values for models  with: \emph{No}: no $\tau$ at all; \emph{Fixed}: $\tau$ fixed at the official last NPI date $\tau^*$; and \emph{Free}: free $\tau$.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
 }
-    \label{fig:Switzerland}
-\end{figure}
+\label{table:table1}
+\end{table}
 

### 2020-05-20 ms/ms.tex
revised model comparison par

@@
 
-We find that a model that considers an NPI (\autoref{eq:NPI_model}) is a better fit to the data than a model without an NPI, i.e. with constant $\beta$ and $\alpha$ (\hl{???}.)
+We compared the \hl{posterior predictive plots} of a model with a free $\tau$ with those of a model with $\tau$ fixed at $\tau^*$ and without $\tau$.
+The model with free $\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}).
+\hl{A similar comparison to no $\tau$.}
+However, when we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free parameter for the start of the NPI was better than the other models only in half of the regions.
+
 We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}).
 Indeed, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:fig1}).
-Moreover, we compared the posterior predictive plots of a model with a free $\tau$ with those of a model with $\tau$ fixed at $\tau^*$.
-The model with free $\tau$ clearly produces better and less variable predictions (\autoref{fig:ppc}).
-
 In the following, we describe our findings on late and early effective start of NPI in detail.
@@
 \bibliographystyle{agsm}
-%\bibliography{/Users/yoavram/Documents/library}
-\bibliography{ms}
+\bibliography{/Users/yoavram/Documents/library}
+%\bibliography{ms}
 \end{multicols}

### 2020-05-20 src/Table-estimated-params.py
rename

@@ -36,8 +36,8 @@ if __name__ == '__main__':
 	table.loc[table['Country'] == 'Wuhan', 'Country'] = 'Wuhan, China'
 	table.columns = ["\ccell{"+x+"}" for x in table.columns]
 
-	csv_filename = '../figures/Table2.csv'
-	tex_filename = '../figures/Table2.tex'
+	csv_filename = '../figures/Table-estimated-params.csv'
+	tex_filename = '../figures/Table-estimated-params.tex'
 	table.to_csv(csv_filename, index=False)
 	with open(tex_filename, 'wt') as f:
 	    print(table.to_latex(escape=False, index=False), file=f)

### 2020-05-20 src/ppc.py
change load_chain

@@ -54,13 +54,16 @@ def load_chain(job_id=None, fname=None, burn_fraction=0.6):
 		nsteps, ndim, N, Td1, Td2, model_type = inference_data['params']
 		X = inference_data['incidences']
 		start_date = inference_data['start_date']
+		logliks = inference_data['logliks']
 		# print("Loaded {} with parameters:".format(fname))
 		# print(var_names)
 		nchains, nsteps, ndim = chain.shape
 		nburn = int(nsteps * burn_fraction)
 		chain = chain[:, nburn:, :]
 		chain = chain.reshape((-1, ndim))
-		return chain, Td1, Td2, model_type, X, start_date, N
+		logliks = logliks.reshape((nchains, nsteps))
+		logliks = logliks[:, nburn:].ravel()
+		return chain, logliks, Td1, Td2, model_type, X, start_date, N
 
 def posterior_prediction(chain, model, nreps):
 	θ = chain[np.random.choice(chain.shape[0], nreps)]
@@ -115,7 +118,7 @@ if __name__ == '__main__':
 	ndays = len(X)
 	
 	chain_fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
-	chain, Td1, Td2, model_type, _, start_date, N = load_chain(fname=chain_fname)
+	chain, _, Td1, Td2, model_type, _, start_date, N = load_chain(fname=chain_fname)
 	X_mean = scipy.signal.savgol_filter(X, 3, 1)
 	
 	model_class = get_model_class(model_type)

### 2020-05-20 ms/ms.tex
refreshed ms.bib
revise revise revise

@@
 We hypothesize that the official and effective start date of such interventions can significantly differ, for example due to slow diffusion of guidelines in the population, or due to unpreparedness of the authorities and the public.
-We use an SEIR epidemiological model and an MCMC inference framework to estimate the effective start of NPIs in several countries, and compare this effective dates to the official dates.
+We use an SEIR epidemiological model and an MCMC inference framework to estimate the effective start of NPIs in several countries, and compare these effective dates to the official dates.
 We report our finding of both late and early effects of NPIs, and discuss potential causes and consequences of our results.
@@
 France), we cropped the data to begin with the last sequence so that our analysis focuses on the first sustained outbreak rather than isolated imported cases.
-For dates of official NPI dates see \autoref{table:table1}.
+For dates of official NPI dates see \autoref{table:NPI_dates}.
 
@@
 }
-\label{table:table1}
+\label{table:NPI_dates}
 \end{table}
+
+%%% Fig NPI dates %%%
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=0.5\textwidth]{../figures/Fig-NPI_dates.pdf}
+    \caption{
+    \textbf{Official start of non-pharmaceutical interventions.}
+	See \autoref{table:NPI_dates} for more details. Wuhan, China is not shown.
+    } 
+    \label{fig:NPI_dates}
+\end{figure}
 
@@
 \end{equation}
-where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:table1}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
+where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
 % TODO something formal for the next line?
@@
 %%% Model selection %%%
-\paragraph*{Model selection.}
-We perform model selection using DIC (deviance information criterion)~\citep{Spiegelhalter2002},
-\begin{equation} \label{eq:DIC}
+\paragraph*{Model comparison.}
+We perform model selection using two methods.
+First, we compute WAIC (widely applicable information criterion)~\citep{gelman2013bayesian},
+\begin{equation} \label{eq:WAIC}
 \begin{aligned}
-DIC(\theta, \vec{X}) &= 2\mathbb{E}[D(\theta)] - D(\mathbb{E}[\theta])   \\
-&= 2 \log\mathcal{L}(\mathbb{E}[\theta] \mid \vec{X}) - 4\mathbb{E}[\log\mathcal{L}(\theta \mid \vec{X})],
+WAIC(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
 \end{aligned}
 \end{equation}
-where $D(\theta)=-2\log\mathcal{L}(\theta \mid \vec{X})$ is the Bayesian deviance, and expectations $\mathbb{E}[\cdot]$ are taken over the posterior distribution $P(\theta \mid \vec{X})$.
-We compare models by reporting their relative DIC; lower is better.
+where $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance operators taken over the posterior distribution $P(\theta \mid \vec{X})$.
+We compare models by reporting their relative WAIC; lower is better~(\autoref{table:WAIC}).
+A minority of MCMC chains that fail to fully converge can lead to overestimation of the variance (the second term in \autoref{eq:WAIC}).
+Therefore, we exclude from the WAIC computation chains with mean log-likelihood that is three standard deviations or more from the overall mean.
+
+We also apply posterior predictive plots, in which we sample 1,000 parameter vectors from the posterior distributions $P(\theta \mid \vec{X})$ and use them to simulate the SEIR model (\autoref{eq:model}).
+We plot these simulated dynamics together with the date~(\autoref{fig:ppc}).
+Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the compactness of the predictions) are good ways to visually compare models.
 
@@
 All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
-Files containing samples from the posterior distributions will be deposited on \hl{FigShare}.
+Files containing samples from the posterior distributions will be deposited on FigShare.
 
@@
 The effective NPI dates in Italy and Wuhan are significantly delayed compared to the official dates, whereas in Denmark, France, Spain, and Germany, the effective date is earlier than the official date.
-	$\hat{\tau}$ is the posterior median, see \autoref{table:table2}.
-$\tau^*$ is the last NPI date, see \autoref{table:table1}.
+	$\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}.
+$\tau^*$ is the last NPI date, see \autoref{table:NPI_dates}.
 Thin and bold lines show 95\% and 75\% credible intervals (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
@@
 Several studies have described the effects of non-pharmaceutical interventions in different geographical regions~\citep{Flaxman2020,Gatto2020,Li2020}. 
-These studies have assumed that the parameters of the epidemiological model change at a specific date, as in \autoref{eq:NPI_model}, and set the change date $\tau$ to the official NPI date $\tau^*$ (\autoref{table:table1}).
+These studies have assumed that the parameters of the epidemiological model change at a specific date, as in \autoref{eq:NPI_model}, and set the change date $\tau$ to the official NPI date $\tau^*$ (\autoref{table:NPI_dates}).
 They then fit the model once for time $t<\tau^*$ and once for time $t \ge \tau^*$.
@@
 
-We compared the \hl{posterior predictive plots} of a model with a free $\tau$ with those of a model with $\tau$ fixed at $\tau^*$ and without $\tau$.
+We compared the posterior predictive plots of a model with a free $\tau$ with those of a model with $\tau$ fixed at $\tau^*$ and without $\tau$.
 The model with free $\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}).
-\hl{A similar comparison to no $\tau$.}
-However, when we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free parameter for the start of the NPI was better than the other models only in half of the regions.
+When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free parameter for the start of the NPI was better than the other models in 8 out of 11 of the regions; the exceptions are Austria, Belgium, Norway, and United Kingdom.
 
-We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}).
-Indeed, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:fig1}).
+We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}),; that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:fig1}).
+The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well another country (see below).
+Norway also has a relatively wide credible interval, which could be expected as it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
 In the following, we describe our findings on late and early effective start of NPI in detail.
 
-%% Table 2 %%%
-\begin{landscape}
+% Table WAIC %%%
 \begin{table}[h]
 \centering
-\input{../figures/Table2.tex}
+\pgfplotstabletypeset[
+    col sep=comma,
+    string type,
+    every head row/.style={before row=\hline,after row=\hline},
+    every last row/.style={after row=\hline},
+    ]{../figures/Table-WAIC.csv}
 \caption{
-\textbf{Parameter estimates for different regions.}
-See \autoref{eq:model} for model parameters.
-All estimates are posterior medians.
-75\% and 95\% credible intervals given only for $\tau$, in days.
-$\tau^*$ is the official last NPI date, see \autoref{table:table1}.
+\textbf{WAIC values for the different models.}
+	WAIC (widely applicable information criterion)~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
 }
-\label{table:table2}
+\label{table:WAIC}
 \end{table}
-\end{landscape}
 
@@
 That is, the first and last official dates are Mar~8 and Mar~11.
-However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.47 days 95\% CI ; \autoref{fig:fig2}).
-Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be several days layer at Feb~2 ($\pm$2.85 days 95\% CI \autoref{fig:fig2}).
+However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.47 days 95\% CI ; \autoref{fig:late}).
+Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be several days layer at Feb~2 ($\pm$2.85 days 95\% CI \autoref{fig:late}).
 
-%%% Fig 2 %%%
+%%% Fig Late %%%
 \begin{figure}[h]
@@
     }
-    \label{fig:fig2}
+    \label{fig:late}
 \end{figure}
@@
 A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8 or~9 ($\pm$1.08 95\%CI), rather than Mar~14 (\autoref{fig:fig3}).
+Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8 or~9 ($\pm$1.08 95\%CI), rather than Mar~14 (\autoref{fig:early}).
 
-Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ on Mar~8 or Mar~9 ($\pm$6.27 days 95\% CI, \autoref{fig:fig3}).
+Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ on Mar~8 or Mar~9 ($\pm$6.27 days 95\% CI, \autoref{fig:early}).
 Although the credible interval is wider compared to Spain, spanning from Mar~2 to Mar~15, the official lockdown start at Mar~17 is later still, and even the earliest NPI, banning of public events, only started on Mar~13~\citep{Flaxman2020}.
@@
 	}
-	\label{fig:fig3}
+	\label{fig:early}
 \end{figure}
@@
 %%% Exact
-\paragraph*{The exception that proves the rule.}
+\paragraph*{Like a Swiss watch.}
 We find one case in which the official and effective dates match: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
@@
 It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
+
+%% Table 2 %%%
+\begin{landscape}
+\begin{table}[h]
+\centering
+\input{../figures/Table-estimated-params.tex}
+\caption{
+\textbf{Parameter estimates for different regions.}
+See \autoref{eq:model} for model parameters.
+All estimates are posterior medians.
+75\% and 95\% credible intervals given only for $\tau$, in days.
+$\tau^*$ is the official last NPI date, see \autoref{table:NPI_dates}.
+}
+\label{table:estimated-params}
+\end{table}
+\end{landscape}
 
@@
     \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
-    Here, model parameters are estimates for Italy (\autoref{table:table2}).
+    Here, model parameters are estimates for Italy (\autoref{table:estimated-params}).
     } 
@@
 
-For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred more than five days after the official date (\autoref{fig:fig2}).
+For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred more than five days after the official date (\autoref{fig:late}).
 This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
@@
 
-In contrast, in most investigated countries (e.g., Spain and France), we infer reduced transmission rates even before official lockdowns were implemented (\autoref{fig:fig3}).
+In contrast, in most investigated countries (e.g., Spain and France), we infer reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}).
 An early effective date might be due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response to increased risk perception due to domestic or international COVID\=/19-related reports.
@@
 
-Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a seven days delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
+Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
 However, the late and early effects we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).
 
-We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ \hl{over a model without such a change-point}.
-It may be interesting to investigate if the evidence favors a model with \emph{two} change-points, rather than one.
+We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 11 regions (\autoref{table:WAIC}).
+It will be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
 Two such change-points could reflect escalating NPIs (e.g.
-school closures followed by lockdowns), or a mix of NPIs and other events, such as weather, or domestic and international events that affect risk perception.
+school closures followed by lockdowns), an NPI followed by a relaxation of the intervention, or a mix of NPIs and other events, such as weather, or domestic and international events that affect risk perception.
 
@@
 
-\pagebreak
 \begin{multicols}{2}[\section*{References}]
@@
 \bibliographystyle{agsm}
-\bibliography{/Users/yoavram/Documents/library}
-%\bibliography{ms}
+%\bibliography{/Users/yoavram/Documents/library}
+\bibliography{ms}
 \end{multicols}
@@
 \beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
-
-%%% Fig NPI dates %%%
-\begin{figure}[h]
-    \centering
-	\includegraphics[width=0.5\textwidth]{../figures/Fig-NPI_dates.pdf}
-    \caption{
-    \textbf{Official start of non-pharmaceutical interventions.}
-	See \autoref{table:table1} for more details. Wuhan, China is not shown.
-    } 
-    \label{fig:NPI_dates}
-\end{figure}
 
@@
 
-% Table WAIC %%%
-\begin{table}[h]
+%%% Fig Switzerland %%%
+ \begin{figure}[h]
      \centering
-\pgfplotstabletypeset[
-    col sep=comma,
-    string type,
-    every head row/.style={before row=\hline,after row=\hline},
-    every last row/.style={after row=\hline},
-    ]{../figures/Table-WAIC.csv}
+         \includegraphics[width=0.75\textwidth]{../figures/posterior/Switzerland_τ_posterior.pdf}
      \caption{
-\textbf{WAIC values for the different models.}
-	WAIC (widely applicable information criterion)~\citep{gelman2013bayesian} values for models  with: \emph{No}: no $\tau$ at all; \emph{Fixed}: $\tau$ fixed at the official last NPI date $\tau^*$; and \emph{Free}: free $\tau$.
-WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
+ 	\textbf{Effective date of non-pharmaceutical interventions in Switzerland matches the official date}
+     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
+Red line shows the official last NPI date $\tau^*$.
+Black line shows the estimated $\hat{\tau}$.
+Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
      }
-\label{table:table1}
-\end{table}
+     \label{fig:Switzerland}
+ \end{figure}
 

### 2020-05-20 ms/ms.tex
typos and formatting

@@
 \begin{aligned}
-WAIC(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
+\mathit{WAIC}(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
 \end{aligned}
@@
 $\tau^*$ is the last NPI date, see \autoref{table:NPI_dates}.
-Thin and bold lines show 95\% and 75\% credible intervals (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
+Thin and bold lines show 95\% and 75\% credible intervals, respectively (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
     }
@@
 The model with free $\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}).
-When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free parameter for the start of the NPI was better than the other models in 8 out of 11 of the regions; the exceptions are Austria, Belgium, Norway, and United Kingdom.
+When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free parameter for the start of the NPI was better than the other models in 8 out of 12 of the regions; the exceptions are Austria, Belgium, Norway, and United Kingdom.
 
@@
 However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.47 days 95\% CI ; \autoref{fig:late}).
+
 Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be several days layer at Feb~2 ($\pm$2.85 days 95\% CI \autoref{fig:late}).
@@
 All estimates are posterior medians.
-75\% and 95\% credible intervals given only for $\tau$, in days.
+75\% and 95\% credible intervals given for $\tau$, in days.
 $\tau^*$ is the official last NPI date, see \autoref{table:NPI_dates}.
@@
 
-We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 11 regions (\autoref{table:WAIC}).
+We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (\autoref{table:WAIC}).
 It will be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 

### 2020-05-20 ms/ms.tex
revised biblio

@@
 % NatBib
-\usepackage[numbers,super,sort&compress]{natbib}
+\usepackage[super]{natbib}
+%\usepackage[numbers,super,sort&compress]{natbib}
 %\usepackage[round,colon,authoryear]{natbib}
 \renewcommand{\bibsection}{}
-\renewcommand{\bibfont}{\small}
+%\renewcommand{\bibfont}{\small}
 
@@
 
-\author[a]{Ilia Kohanovski}
-\author[b,c]{Uri Obolski}
-\author[a,*]{Yoav Ram}
+\author[1]{Ilia Kohanovski}
+\author[2,3]{Uri Obolski}
+\author[1,a]{Yoav Ram}
 
-\affil[a]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya 4610101, Israel}
-\affil[b]{School of Public Health, Tel Aviv University, Tel Aviv 6997801, Israel}{
-\affil[c]{Porter School of the Environment and Earth Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
-\affil[*]{Corresponding author: yoav@yoavram.com}
+\affil[1]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya 4610101, Israel}
+\affil[2]{School of Public Health, Tel Aviv University, Tel Aviv 6997801, Israel}{
+\affil[3]{Porter School of the Environment and Earth Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
+\affil[a]{Corresponding author: yoav@yoavram.com}
 
@@
 
-\begin{multicols}{2}[\section*{References}]
+\section*{References}
 \nolinenumbers
-\bibliographystyle{agsm}
+\bibliographystyle{Science}
 %\bibliography{/Users/yoavram/Documents/library}
 \bibliography{ms}
-\end{multicols}
 
@@
 \beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
+
+%%% Fig Fr-vs-Es %%%
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=\textwidth]{../figures/Fig-fr-vs-es.pdf}
+    \caption{
+    \textbf{COVID\=/19 confirmed cases in France and Spain.}
+    Number of cases proportional to population size (as of 2018). 
+    Vertical line shows Mar~8, the effective start of NPIs $\hat{\tau}$ in both countries.
+    } 
+    \label{fig:fig-fr-vs-es}
+\end{figure}
+
+%%% Fig Switzerland %%%
+ \begin{figure}[h]
+     \centering
+         \includegraphics[width=0.75\textwidth]{../figures/posterior/Switzerland_τ_posterior.pdf}
+     \caption{
+ 	\textbf{Effective date of non-pharmaceutical interventions in Switzerland matches the official date}
+     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
+Red line shows the official last NPI date $\tau^*$.
+Black line shows the estimated $\hat{\tau}$.
+Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
+     }
+     \label{fig:Switzerland}
+ \end{figure}
 
@@
     \caption{
-    \textbf{Figure S2. Posterior prediction check plots}
+    \textbf{Figure S3. Posterior prediction check plots}
     Markers represent data ($\vec{X}$).
@@
 
-\pagebreak
-
-%%% Fig Fr-vs-Es %%%
-\begin{figure}[h]
-    \centering
-	\includegraphics[width=\textwidth]{../figures/Fig-fr-vs-es.pdf}
-    \caption{
-    \textbf{COVID\=/19 confirmed cases in France and Spain.}
-    Number of cases proportional to population size (as of 2018). 
-    Vertical line shows Mar~8, the effective start of NPIs $\hat{\tau}$ in both countries.
-    } 
-    \label{fig:fig-fr-vs-es}
-\end{figure}
-
-%%% Fig Switzerland %%%
- \begin{figure}[h]
-     \centering
-         \includegraphics[width=0.75\textwidth]{../figures/posterior/Switzerland_τ_posterior.pdf}
-     \caption{
- 	\textbf{Effective date of non-pharmaceutical interventions in Switzerland matches the official date}
-     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
-Red line shows the official last NPI date $\tau^*$.
-Black line shows the estimated $\hat{\tau}$.
-Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
-     }
-     \label{fig:Switzerland}
- \end{figure}
-
 \end{document}

### 2020-05-20 ms/ms.tex
typos in ms.tex: 12 regions; added digits to csv reports

--- 2020-05-20 ms/ms.tex
@@
 \begin{aligned}
-\mathit{WAIC}(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
+WAIC(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
 \end{aligned}
@@
 $\tau^*$ is the last NPI date, see \autoref{table:NPI_dates}.
-Thin and bold lines show 95\% and 75\% credible intervals, respectively (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
+Thin and bold lines show 95\% and 75\% credible intervals (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
     }
@@
 
-We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}),; that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:fig1}).
+We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:fig1}).
 The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well another country (see below).
@@
 However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.47 days 95\% CI ; \autoref{fig:late}).
-
 Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be several days layer at Feb~2 ($\pm$2.85 days 95\% CI \autoref{fig:late}).
@@
 All estimates are posterior medians.
-75\% and 95\% credible intervals given for $\tau$, in days.
+75\% and 95\% credible intervals given only for $\tau$, in days.
 $\tau^*$ is the official last NPI date, see \autoref{table:NPI_dates}.

### 2020-05-20 src/plot_utils.py

@@ -71,9 +71,9 @@ def write_csv_header(file_name):
 
 def write_csv_data(file_name):
     #params means and medians
-    means = [round(sample[:,i].mean(),2) for i in range(len(var_names))]
-    medians = [round(np.median(sample[:,i]),2) for i in range(len(var_names))]
-    MAPs = [round(m,2) for m in get_MAP()]
+    means = [round(sample[:,i].mean(),4) for i in range(len(var_names))]
+    medians = [round(np.median(sample[:,i]),4) for i in range(len(var_names))]
+    MAPs = [round(m,4) for m in get_MAP()]
     params_values = [e for l in zip(means,medians,MAPs) for e in l]
 
     #tau
@@ -102,15 +102,15 @@ def write_csv_data(file_name):
     τ_median_from1Jar =  (pd.to_datetime(start_date) - pd.Timestamp('2020-01-01')).days + np.median(τ_posterior)
     τ_MAP_from1Jar =  (pd.to_datetime(start_date) - pd.Timestamp('2020-01-01')).days + τ_MAP_i
 
-    τ_mean_from1Jar = round(τ_mean_from1Jar,2)
-    τ_median_from1Jar = round(τ_median_from1Jar,2)
-    τ_MAP_from1Jar = round(τ_MAP_from1Jar,2)
+    τ_mean_from1Jar = round(τ_mean_from1Jar,4)
+    τ_median_from1Jar = round(τ_median_from1Jar,4)
+    τ_MAP_from1Jar = round(τ_MAP_from1Jar,4)
 
     with open(file_name, mode='a') as file:
         writer = csv.writer(file, delimiter=',', quotechar='"', quoting=csv.QUOTE_MINIMAL)
-        writer.writerow([country_name, '{:.2f}'.format(calc_DIC(calc_loglik_median)), '{:.2f}'.format(calc_DIC(calc_loglik_mean)), '{:.2f}'.format(calc_DIC(calc_loglikMAP)), '{:.2f}'.format(calc_loglikMAP()),'{:.2f}'.format(calc_loglik_mean()),'{:.2f}'.format(calc_loglik_median()), N, nsteps, model_type, Td1, Td2,
[…]

### 2020-05-24 src/Fig_tau_posterior.py
rename

@@ -1,6 +1,7 @@
 #!/usr/bin/env python
 # coding: utf-8
 import matplotlib as mpl
+mpl.use("Agg")s
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd

### 2020-05-24 src/Fig_NPI_dates.py
convert from ipynb

@@ -1,0 +1,52 @@
+#!/usr/bin/env python
+# coding: utf-8
+import os
+from datetime import datetime, timedelta
+
+import matplotlib as mpl
+mpl.use("Agg")
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+
+from rakott.mpl import savefig_bbox
+
+def int_to_dt(t):
+    return pd.to_datetime(start_date) + timedelta(days=t)
+def dt_to_days(t):
+    return (t - datetime(2020, 1, 1)).days
+def str_to_dt(t):
+    return datetime.strptime(t, '%b %d %Y')
+
+sns.set_context('paper', font_scale=1.3)
+red, blue, green = sns.color_palette('Set1', 3)
+colors = sns.color_palette('Paired')
+
+date_range = [datetime(2020, 3, 1) + timedelta(weeks=i) for i in range(5)]
+date_formatter = mpl.dates.DateFormatter('%b %d')
+
+def get_official_dates():
+	official = pd.read_csv('../data/NPI_dates.csv').sort_values('Country', ascending=False)
+	official['First'] = [str_to_dt(t) for t in official['First']]
+	official['Last'] = [str_to_dt(t) for t in official['Last']]
+	official['First (days)'] = [dt_to_days(t) for t in official['First']]
+	official['Last (days)'] = [dt_to_days(t) for t in official['Last']]
+	return official
+
+if __name__ == '__main__':
+
+	official = get_official_dates()
+
+	fig, ax = plt.subplots(figsize=(4, 6))
+	data_ = official[official['First'] >= datetime(2020, 3, 1)]
+	ax.plot('First', 'Country', 'ok', data=data_)
+	ax.plot('Last', 'Country', 'ok', data=data_)
+	ax.hlines('Country', 'First', 'Last', data=data_)
+	ax.set_xticks(date_range)
+	ax.xaxis.set_major_formatter(date_formatter)
+	ax.grid(True)
+	txt = ax.get_yticklabels()
+	sns.despine()
+	fig.savefig('../figures/Fig-NPI_dates.pdf', dpi=100, **savefig_bbox(*txt))
+	print("Saved NPI dates to ../figures/Fig-NPI_dates.pdf")

### 2020-05-24 src/Fig_tau_summary.py

@@ -1,0 +1,122 @@
+#!/usr/bin/env python
+# coding: utf-8
+import os
+from datetime import datetime, timedelta
+import sys
+
+import matplotlib as mpl
+mpl.use("Agg")
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+
+from rakott.mpl import savefig_bbox
+
+from Fig_NPI_dates import get_official_dates
+
+def int_to_dt(t):
+    return pd.to_datetime(start_date) + timedelta(days=t)
+def dt_to_days(t):
+    return (t - datetime(2020, 1, 1)).days
+def str_to_dt(t):
+    return datetime.strptime(t, '%b %d %Y')
+def date_to_int(x):
+    dt = datetime.strptime(x + ' 2020', '%b %d %Y')
+    td = dt - datetime(2020, 1, 1)
+    return td.days
+def date_to_date(x):
+    dt = datetime.strptime(x + ' 2020', '%b %d %Y')
+    return dt
+
+sns.set_context('paper', font_scale=1.3)
+red, blue, green = sns.color_palette('Set1', 3)
+colors = sns.color_palette('Paired')
+
+date_range = [datetime(2020, 3, 1) + timedelta(weeks=i) for i in range(5)]
+date_formatter = mpl.dates.DateFormatter('%b %d')
+
+
+if __name__ == '__main__':
+	job_id = sys.argv[1]
+	verbose = len(sys.argv) > 2 and sys.argv[2] == '-v'
+
+	output_folder = r'/Users/yoavram/Library/Mobile Documents/com~apple~CloudDocs/EffectiveNPI-Data/output/{}'.format(job_id)
+	
+	official = get_official_dates()
+	table_path = os.path.join(output_folder, 'tables', 'all-countries-{}.csv'.format(job_id))
+	df = pd.read_csv(table_path)
+	df = df.rename(columns={'official_τ': 'τ official'})
+	df['country'] = df['country'].str.replace("_", " ")
+
+	countries = df['country'].unique()
+	country_color = {country: colors[i] for i, country in enumerate(countries)}
+	df['color'] = [country_color[country] for country in df['country']]
+
+	df['τ mean days'] = [date_to_int(x) for x in df['τ mean']]
[…]

### 2020-05-24 src/Table-WAIC.py

@@ -1,0 +1,101 @@
+#!/usr/bin/env python
+# coding: utf-8
+import matplotlib as mpl
+mpl.use("Agg")
+import matplotlib.pyplot as plt
+import numpy as np
+import scipy.stats
+from scipy.special import logsumexp
+import pandas as pd
+import seaborn as sns
+from tqdm import tqdm
+
+import os
+from datetime import datetime, timedelta
+
+from rakott.mpl import fig_panel_labels, fig_xlabel, fig_ylabel, savefig_bbox
+
+from inference import find_start_day
+from Fig_ppc import load_data
+
+sns.set_context('paper', font_scale=1.3)
+red, blue, green = sns.color_palette('Set1', 3)
+
+
+def load_chain(job_id, country, burn_fraction=0.6):
+    fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
+    inference_data = np.load(fname)
+    nsteps, ndim, N, Td1, Td2, model_type = inference_data['params']
+    logliks = inference_data['logliks']
+    nchains = logliks.size // nsteps
+    logliks = logliks.reshape(nchains, nsteps)
+    nburn = int(nsteps*burn_fraction)
+    logliks = logliks[:, nburn:]
+    return logliks
+
+
+def inliers(logliks, PLOT=False):
+    chain_mean_loglik = logliks.mean(axis=1)
+    std_mean_loglikg = chain_mean_loglik.std(ddof=1)
+    mean_mean_loglikg = chain_mean_loglik.mean()
+    idx = abs(chain_mean_loglik - mean_mean_loglikg) < 3*std_mean_loglikg
+    if PLOT:
+        if idx.any():
+            plt.plot(logliks[idx, ::1000].T, '.k', label='inliers')
+        if (~idx).any():
+            plt.plot(logliks[~idx, ::1000].T, '.r', label='outliers')
+        plt.ylabel('Log-likelihood')
+        plt.legend()
+    return idx
+
+
+def WAIC(logliks):
+    logliks = logliks[inliers(logliks)]
+    S = logliks.size
+    llpd = -np.log(S) + logsumexp(logliks)
+    p1 = 2*(-np.log(S) + logsumexp(logliks) - logliks.mean())
+    p2 = np.var(logliks, ddof=1)
+    return -2*(llpd + -p1), -2*(llpd + -p2)
+
+if __name__ == '__main__':    
[…]

### 2020-05-24 ms/ms.tex
fig names and headings

@@
 
-% Table 1 %%%
+% Table - NPI dates %%%
 \begin{table}[h]
@@
 
-%%% Fig NPI dates %%%
+%%% Fig - NPI dates %%%
 \begin{figure}[h]
@@
 
-%%% Fig 1 %%%
+%%% Fig - tau summary %%%
 \begin{figure}[h]
     \centering
-	\includegraphics[width=0.6\textwidth]{../figures/Fig1.pdf}
+	\includegraphics[width=0.6\textwidth]{../figures/Fig-tau-summary.pdf}
     \caption{
-    \textbf{Official and effective start of non-pharmaceutical interventions.}
+    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
     	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPI is shown for different regions.
@@
     }
-    \label{fig:fig1}
+    \label{fig:tau-summary}
 \end{figure}
@@
 
-We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:fig1}).
+We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
 The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well another country (see below).
@@
 
-% Table WAIC %%%
+% Table - WAIC %%%
 \begin{table}[h]
@@
 \paragraph*{Late effective start of NPIs.}
-In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$ (\autoref{fig:fig1}).
+In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$ (\autoref{fig:tau-summary}).
 
@@
 
-%%% Fig Late %%%
+%%% Fig - Late %%%
 \begin{figure}[h]
@@
 \paragraph*{Early effective start of NPIs.}
-In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:fig1}).
+In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
 In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
@@
 
-%%% Fig 3 %%%
+%%% Fig - Early %%%
 \begin{figure}[h]
@@
 
-%% Table 2 %%%
+%% Table - estimated parameters %%%
 \begin{landscape}
@@
 
-%%% Fig 4 %%%
+%%% Fig - Prediction %%%
 \begin{figure}[h]
     \centering
-	\includegraphics[width=0.5\textwidth]{../figures/Fig4.pdf}
+	\includegraphics[width=0.5\textwidth]{../figures/Fig-prediction.pdf}
     \caption{
@@
     Real number of daily cases in Italy in black (markers: data, line: time moving average). 
-    Model posterior predictions are shown as colored lines (1,000 draws from the posterior distribution). 
+    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
     Shaded box illustrates a serial interval of seven days.
@@
     } 
-    \label{fig:fig4}
+    \label{fig:fig-prediction}
 \end{figure}
@@
 We have estimated the effective start date of NPIs in several geographical regions using an SEIR epidemiological model and an MCMC parameter estimation framework.
-We find examples of both late and early effect of NPIs (\autoref{fig:fig1}).
+We find examples of both late and early effect of NPIs (\autoref{fig:tau-summary}).
 

### 2020-05-24 ms/ms.tex
covid macro

@@
 
+% COVID-19
+\newcommand{\covid}{COVID\=/19 }
+\newcommand{\sars}{SARS\=/CoV\=/2 }
+
 % line numbers
@@
 % Title page
-\title{Inferring the effective start dates of non-pharmaceutical interventions during COVID-19 outbreaks}
+\title{Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks}
 
@@
 
-During February and March 2020, several countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules to control the COVID\=/-19 pandemic caused by the SARA\=/CoV\=/2 virus.
+During February and March 2020, several countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules to control the \covid pandemic caused by the \sars virus.
 Overall, these interventions seem to have successfully reduced the spread of the pandemic.
@@
 Similarly, compliance to NPIs may be impacted by personal experiences.
-\citet{Smith2020} have surveyed 6,149 UK adults in late April and found that people who believe they have already had COVID\=/19 are more likely to think they are immune, and less likely to comply with social distancing measures.
+\citet{Smith2020} have surveyed 6,149 UK adults in late April and found that people who believe they have already had \covid are more likely to think they are immune, and less likely to comply with social distancing measures.
 Compliance may also depend on risk perception as perceived by the the number of domestic cases or even by reported cases in other regions and countries.
-Interestingly, the perceived risk of COVID\=/19 infection has likely caused a reduction in the number of influenza-like illness cases in the US starting from mid-February~\citep{Zipfel2020}.
+Interestingly, the perceived risk of \covid infection has likely caused a reduction in the number of influenza-like illness cases in the US starting from mid-February~\citep{Zipfel2020}.
 
@@
 We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to estimate the effective start date of NPIs from publicly available COVID-19 case data in several geographical regions.
-We compare these estimates to the official dates and find both late and early effects of NPIs on COVID\=/19 transmission dynamics.
+We compare these estimates to the official dates and find both late and early effects of NPIs on \covid transmission dynamics.
 We conclude by demonstrating how differences between the official and effective start of NPIs can confuse assessments of the effectiveness of the NPIs in a simple epidemic control framework.
@@
 The initial numbers of exposed $E(0)$ and unreported infected $I_u(0)$ are considered model parameters, whereas the initial number of reported infected is assumed to be zero $I_r(0)=0$, and the number of susceptible is $S(0)=N-E(0)-I_u(0)$.
-This model is inspired by \citet{Li2020} and \citet{Pei2020}, who used a similar model with multiple regions and constant transmission $\beta$ and reporting rate $\alpha$ to infer COVID\=/19 dynamics in China and the continental US, respectively.
+This model is inspired by \citet{Li2020} and \citet{Pei2020}, who used a similar model with multiple regions and constant transmission $\beta$ and reporting rate $\alpha$ to infer \covid dynamics in China and the continental US, respectively.
 
@@
 In contrast, in most investigated countries (e.g., Spain and France), we infer reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}).
-An early effective date might be due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response to increased risk perception due to domestic or international COVID\=/19-related reports.
+An early effective date might be due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response to increased risk perception due to domestic or international \covid-related reports.
 This finding may also suggest that severe NPIs, such as lockdowns, were unnecessary, and that less extreme measures adopted by the population could have been sufficient for epidemic control.
@@
 
-Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
+Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of \covid~\citep{Gatto2020}.
 However, the late and early effects we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).
@@
     \caption{
-    \textbf{COVID\=/19 confirmed cases in France and Spain.}
+    \textbf{\covid confirmed cases in France and Spain.}
     Number of cases proportional to population size (as of 2018). 

### 2020-05-24 src/Table-WAIC.py
<2 or >2

@@ -88,14 +88,19 @@ if __name__ == '__main__':
     df.loc[df['Country']=='Wuhan', 'Country'] = 'Wuhan China'
     
     df = pd.pivot(df, index='Country', columns='Model')
-    
+    print(df)
+
     df = df.drop(columns='WAIC1')
     df = df.droplevel(0, axis=1)
 
-    idx = df['Free']==df.min(axis=1)
-    df.loc[idx, 'Free'] = ['\\textbf{'+'{:.2f}'.format(x)+'}' for x in df.loc[idx, 'Free']] 
-    df.loc[~idx, 'Free'] = ['{:.2f}'.format(x) for x in df.loc[~idx, 'Free']] 
+    for country, row in df.iterrows():
+        if row['Free'] + 2 < min(row['Fixed'], row['No']):
+            df.loc[country, 'Free'] = '\\textbf{'+'{:.2f}'.format(row['Free'])+'}**'
+        elif row['Free'] < min(row['Fixed'], row['No']):
+            df.loc[country, 'Free'] = '\\textbf{'+'{:.2f}'.format(row['Free'])+'}*'
+        else:
+            df.loc[country, 'Free'] ='{:.2f}'.format(row['Free'])
     
-    df.to_csv('../figures/Table-WAIC.csv', index='Country', float_format="%.2f")
+    df.to_csv('../figures/Table-WAIC_tmp.csv', index='Country', float_format="%.2f")
     print("Saved to ../figures/Table-WAIC.csv")
     

### 2020-05-24 ms/ms.tex
UO comments

@@
 
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+
 % less space before sections 
@@
 
-\renewcommand{\thefootnote}{\fnsymbol{footnote}}
-
 % Yoav & Lee commands
@@
 % NatBib
-\usepackage[super]{natbib}
+\usepackage[super,comma]{natbib}
 %\usepackage[numbers,super,sort&compress]{natbib}
@@
 Overall, these interventions seem to have successfully reduced the spread of the pandemic.
-We hypothesize that the official and effective start date of such interventions can significantly differ, for example due to slow diffusion of guidelines in the population, or due to unpreparedness of the authorities and the public.
-We use an SEIR epidemiological model and an MCMC inference framework to estimate the effective start of NPIs in several countries, and compare these effective dates to the official dates.
-We report our finding of both late and early effects of NPIs, and discuss potential causes and consequences of our results.
+We hypothesise that the official and effective start date of such interventions can significantly differ, for example due to slow adoption by the population, or due to unpreparedness of the authorities and the public.
+We fit an SEIR model to case data from~12 countries to infer the effective start dates of interventions and contrast them with the official dates.
+We find both late and early effects of interventions.
+For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective date on Mar~16 ($\pm$0.47 days 95\% CI).
+In contrast, Spain announced a lockdown on Mar~14, but we infer an effective start date on Mar~8 or~9 ($\pm$ 1.08 days 95\% CI).
+We discuss potential causes and consequences of our results.
 
@@
 
-The COVID-19 pandemic has resulted in implementation of extreme non-pharmaceutical interventions (NPIs) in many affected countries.
+The \covid pandemic has resulted in implementation of extreme non-pharmaceutical interventions (NPIs) in many affected countries.
 These interventions, from social distancing to lockdowns, are applied in a rapid and widespread fashion.
-The NPIs are designed and assessed using epidemiological models, which follow the dynamics of the viral infection to forecast the effect of different mitigation and suppression strategies on the levels of infection, hospitalization, and fatality.
-These epidemiological models usually assume that the effect of NPIs on disease transmission begins at the officially declared date (e.g.~\citet{Li2020,Gatto2020,Flaxman2020}).
+NPIs are designed and assessed using epidemiological models, which follow the dynamics of the viral infection to forecast the effect of different mitigation and suppression strategies on the levels of infection, hospitalization, and fatality.
+These epidemiological models usually assume that the effect of NPIs on disease transmission begins at the officially declared date~\citep{Li2020,Gatto2020,Flaxman2020}.
 
 Adoption of public health recommendations is often critical for effective response to infectious diseases, and has been studied in the context of HIV~\citep{Kaufman2014} and vaccination~\citep{Dunn2015,Wiyeh2018}, for example.
-However, behavioral and social change does not occur immediately, but rather requires time to diffuse in the population through media, social networks, and social interactions. 
-Moreover, compliance to NPIs may differ between different interventions and between people.
+However, behavioural and social change does not occur immediately, but rather requires time to diffuse in the population through media, social networks, and social interactions. 
+Moreover, compliance to NPIs may differ between different interventions and between people with different backgrounds.
 For example, in a survey of 2,108 adults in the UK during Mar 2020, \citet{Atchison2020} found that those over 70 years old were more likely to adopt social distancing than young adults (18-34 years old), and that those with lower income were less likely to be able to work from home and to self-isolate.
@@
 Here, we hypothesize that there is a significant difference between the official start of NPIs and their adoption by the public and therefore their effect on transmission dynamics.
-We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to estimate the effective start date of NPIs from publicly available COVID-19 case data in several geographical regions.
-We compare these estimates to the official dates and find both late and early effects of NPIs on \covid transmission dynamics.
+We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to infer the effective start date of NPIs from publicly available \covid case data in several geographical regions.
+We compare these estimates to the official dates, and find that they include both late and early effects of NPIs on \covid transmission dynamics.
 We conclude by demonstrating how differences between the official and effective start of NPIs can confuse assessments of the effectiveness of the NPIs in a simple epidemic control framework.
-% TODO explain why important
 
@@
 \paragraph*{Data.} 
-We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from several different countries.
-These incidence data summarize the number of individuals $X_t$ tested positive for SARS\=/CoV\=/2 RNA (using RT-qPCR) at each day $t$.
+We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions.
+These incidence data summarise the number of individuals $X_t$ tested positive for \sars RNA (using RT-qPCR) at each day $t$.
 Data for Wuhan, China retrieved from \citet{Pei2020}, data for 11 European countries retrieved from \citet{Flaxman2020}. 
@@
 Susceptible ($S$) individuals become exposed due to contact with reported or unreported infected individuals ($I_r$ or $I_u$) at a rate $\beta_t$ or $\mu \beta_t$.
-The parameter $0 < \mu < 1$ represents the decreased transmission rate from unreported infected individuals, who are often subclinical or even asymptomatic.
-The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioral changes of both susceptible and infected individuals.
+The parameter $0 < \mu < 1$ represents the decreased transmission rate from unreported infected individuals, who are often subclinical or even asymptomatic~\citep{Thompson2020,Ferretti2020}.
+The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioural changes of both susceptible and infected individuals.
 Exposed individuals, after an average incubation period of $Z$ days, become reported infected with probability $\alpha_t$ or unreported infected with probability $(1-\alpha_t)$.
@@
 \frac{dI_r}{dt} & = \alpha_t \frac{E}{Z} - \frac{I_r}{D} \\
-\frac{dI_u}{dt} & = (1-\alpha_t) \frac{E}{Z} - \frac{I_r}{D} .
+\frac{dI_u}{dt} & = (1-\alpha_t) \frac{E}{Z} - \frac{I_r}{D} \\
+\frac{dR}{dt} & = \frac{I_r}{D} + \frac{I_r}{D} .
 \end{aligned}
 \end{equation}
-The initial numbers of exposed $E(0)$ and unreported infected $I_u(0)$ are considered model parameters, whereas the initial number of reported infected is assumed to be zero $I_r(0)=0$, and the number of susceptible is $S(0)=N-E(0)-I_u(0)$.
+The initial numbers of exposed $E(0)$ and unreported infected $I_u(0)$ are free model parameters (i.e.
+estimated during model fitting), whereas the initial number of reported infected is assumed to be zero $I_r(0)=0$, and the number of susceptible is $S(0)=N-E(0)-I_u(0)$.
 This model is inspired by \citet{Li2020} and \citet{Pei2020}, who used a similar model with multiple regions and constant transmission $\beta$ and reporting rate $\alpha$ to infer \covid dynamics in China and the continental US, respectively.
@@
 We assume that reported infected individuals are confirmed and therefore observed in the daily case report of day $t$ with probability $p_t$ (note that an individual can only be observed once, and that $p_t$ may change over time, but $t$ is a specific date rather than the time elapsed since the individual was infected).
-We denote by $X_t$ the number of confirmed cases in day $t$, and by $\tilde{X}_t$ the cumulative number of confirmed cases until day $t$,
+We denote by $X_t$ the \emph{observed} number of confirmed cases in day $t$, and by $\tilde{X}_t$ the cumulative number of confirmed cases until end of day $t$,
 \begin{equation} \label{eq:Xsumt}
@@
 We compare models by reporting their relative WAIC; lower is better~(\autoref{table:WAIC}).
-A minority of MCMC chains that fail to fully converge can lead to overestimation of the variance (the second term in \autoref{eq:WAIC}).
-Therefore, we exclude from the WAIC computation chains with mean log-likelihood that is three standard deviations or more from the overall mean.
+A minority (<5\%) of MCMC chains that fail to fully converge can lead to overestimation of the variance (the second term in \autoref{eq:WAIC}).
+Therefore, we exclude from the WAIC computation chains with mean log-likelihood that is three standard deviations or more from the overall mean. %TODO
 
-We also apply posterior predictive plots, in which we sample 1,000 parameter vectors from the posterior distributions $P(\theta \mid \vec{X})$ and use them to simulate the SEIR model (\autoref{eq:model}).
-We plot these simulated dynamics together with the date~(\autoref{fig:ppc}).
+We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}), and plot the simulated dynamics~(\autoref{fig:ppc}).
 Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the compactness of the predictions) are good ways to visually compare models.
@@
 All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
-Files containing samples from the posterior distributions will be deposited on FigShare.
+Samples from the posterior distributions will be deposited on FigShare.
+
+%\pagebreak
+% Results
+\section*{Results}
 
 %%% Fig - tau summary %%%
-\begin{figure}[h]
+\begin{figure}[b!]
     \centering
@@
 $\tau^*$ is the last NPI date, see \autoref{table:NPI_dates}.
-Thin and bold lines show 95\% and 75\% credible intervals, respectively (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
+Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e.
+area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
     }
@@
 \end{figure}
-
-%\pagebreak
-% Results
-\section*{Results}
 
@@
 The model with free $\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}).
-When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free parameter for the start of the NPI was better than the other models in 8 out of 12 of the regions; the exceptions are Austria, Belgium, Norway, and United Kingdom.
+When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free parameter for the start of the NPI was preferred over the other models in 8 out of 12 of the regions (although only narrowly for 5 of the 8); the exceptions are Austria, Belgium, Norway, and United Kingdom.
 
@@
 In the following, we describe our findings on late and early effective start of NPI in detail.
-
-% Table - WAIC %%%
-\begin{table}[h]
-\centering
-\pgfplotstabletypeset[
-    col sep=comma,
-    string type,
-    every head row/.style={before row=\hline,after row=\hline},
-    every last row/.style={after row=\hline},
-    ]{../figures/Table-WAIC.csv}
-\caption{
-\textbf{WAIC values for the different models.}
-	WAIC (widely applicable information criterion)~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
-WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
-}
-\label{table:WAIC}
-\end{table}
 
@@
 
-Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be several days layer at Feb~2 ($\pm$2.85 days 95\% CI \autoref{fig:late}).
+Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later at Feb~2 ($\pm$2.85 days 95\% CI \autoref{fig:late}).
 
@@
 A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8 or~9 ($\pm$1.08 95\%CI), rather than Mar~14 (\autoref{fig:early}).
+Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8 or~9 ($\pm$1.08 days 95\%CI), rather than Mar~14 (\autoref{fig:early}).
 
@@
 %%% Assessment
-\paragraph*{Effect of late and early effect of NPIs on real-time assessment.}
+\paragraph*{Consequences of late and early effect of NPIs on real-time assessment.}
 
@@
 However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
-This is illustrated in \autoref{fig:fig4} using data and parameters from Italy.
+This is illustrated in \autoref{fig:prediction} using data and parameters from Italy.
 Here, a lockdown is officially ordered on Mar~10 ($\tau^*$), but its late effect on the transmission dynamics starts on Mar~16 ($\hat{\tau}$).
@@
     } 
-    \label{fig:fig-prediction}
+    \label{fig:prediction}
 \end{figure}
@@
 This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
-Late effect of NPIs might also be due to the time required by both the government and the citizens to organize for a lockdown, and for the new guidelines to diffuse in the population.
+Late effect of NPIs might also be due to the time required by both the government and the citizens to organise for a lockdown, and for the new guidelines to diffuse in the population.
 
 In contrast, in most investigated countries (e.g., Spain and France), we infer reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}).
-An early effective date might be due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response to increased risk perception due to domestic or international \covid-related reports.
-This finding may also suggest that severe NPIs, such as lockdowns, were unnecessary, and that less extreme measures adopted by the population could have been sufficient for epidemic control.
-These less extreme measures may have been implemented due to government recommendations, media coverage, and social networks, rather than official NPIs.
+An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
+These behavioural adaptations may have a stronger influence than previously thought.
+In addition, diffusion of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may have been influenced by increased risk perception due to domestic or international COVID\=/19-related reports.
 Indeed, the evidence supports a change in transmission dynamics (i.e.
-a model with free $\tau$) even for Sweden~(\autoref{fig:ppc}), in which a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}, suggesting that lockdowns may not be necessary if other NPIs are adopted early enough during the outbreak~\citep{Banholzer2020}.
+a model with free $\tau$) even for Sweden~(\autoref{fig:ppc}), in which a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
 
 Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of \covid~\citep{Gatto2020}.
-However, the late and early effects we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).
+However, the late and early effects we have estimated may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).
 
 We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (\autoref{table:WAIC}).
-It will be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
-Two such change-points could reflect escalating NPIs (e.g.
-school closures followed by lockdowns), an NPI followed by a relaxation of the intervention, or a mix of NPIs and other events, such as weather, or domestic and international events that affect risk perception.
+It could be interesting to check if the evidence favours a model with \emph{two} change-points, rather than one. 
+Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
+However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as weather, or domestic and international events that affect risk perception.
 
 As several countries (e.g.
-Austria, Israel) begin to relieve lockdowns and ease restrictions, we expect similar delays and advances to occur: in some countries people will begin to behave as if restrictions were eased even before the official date, and in some countries people will continue to self-restrict even after restrictions are officially removed.
+Austria) begin to relieve lockdowns and ease restrictions, we expect similar delays and advances to occur: in some countries people will begin to behave as if restrictions were eased even before the official date, and in some countries people will continue to self-restrict even after restrictions are officially removed.
 
@@
 \nolinenumbers
-\bibliographystyle{Science}
-%\bibliography{/Users/yoavram/Documents/library}
-\bibliography{ms}
+\bibliographystyle{agsm}
+\bibliography{/Users/yoavram/Documents/library}
+%\bibliography{ms}
 
@@
 \beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
+
+% Table - WAIC %%%
+\begin{table}[h]
+\centering
+\pgfplotstabletypeset[
+    col sep=comma,
+    string type,
+    every head row/.style={before row=\hline,after row=\hline},
+    every last row/.style={after row=\hline},
+    ]{../figures/Table-WAIC_tmp.csv}
+\caption{
+\textbf{WAIC values for the different models.}
+	WAIC (widely applicable information criterion)~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
+Bold emphasise cases in which the \emph{Free} model had the lowest WAIC.
+$*$ and $**$ mark if the difference was smaller or greater than 2, which is a popular significance level for model comparison~\citep{Kass1995}.
+}
+\label{table:WAIC}
+\end{table}
 

### 2020-05-24 src/Fig_tau_posterior.py
red shade for NPIs; HPI code added but commented out for now; quiet mode

@@ -1,15 +1,17 @@
 #!/usr/bin/env python
 # coding: utf-8
+import os
+from datetime import datetime, timedelta
+import sys
+
 import matplotlib as mpl
-mpl.use("Agg")s
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
 import seaborn as sns
-import os
-from datetime import datetime, timedelta
+from arviz import hpd
+
 from rakott.mpl import savefig_bbox
-import sys
 
 sns.set_context('paper', font_scale=1.3)
 red, blue, green = sns.color_palette('Set1', 3)
@@ -21,6 +23,7 @@ if __name__ == '__main__':
 	job_id = sys.argv[1]
 	output_folder = r'/Users/yoavram/Library/Mobile Documents/com~apple~CloudDocs/EffectiveNPI-Data/output/{}/'.format(job_id)
 	country = sys.argv[2]
+	quiet = len(sys.argv) > 3 and sys.argv[3] == '-q'	
 
 	NPI_dates = pd.read_csv('../data/NPI_dates.csv')
 
@@ -40,9 +43,10 @@ if __name__ == '__main__':
 		sample = data['chain']
 		log_posterior = data['lnprobability']
 
-		official_date = pd.to_datetime(NPI_dates.loc[NPI_dates['Country'] == country.replace('_', ' '), 'Last'].values[0])
-
-		official_date_days = (official_date - start_date).days
+		first_date = pd.to_datetime(NPI_dates.loc[NPI_dates['Country'] == country.replace('_', ' '), 'First'].values[0])
+		last_date = pd.to_datetime(NPI_dates.loc[NPI_dates['Country'] == country.replace('_', ' '), 'Last'].values[0])
+		first_date_days = (first_date - start_date).days
+		last_date_days = (last_date - start_date).days
 
 		τ_sample = sample[:, :, -1]
 		nburn = τ_sample.shape[1]//2
@@ -53,20 +57,20 @@ if __name__ == '__main__':
 		plt.xlabel('Iteration')
 		plt.title('Trace plot: {}'.format(country));
 		plt.axvline(nburn/thin, color='k');
-		plt.show()
+		if not quiet: plt.show()
 
 		plt.plot(τ_sample[:,nburn::thin].T)
 		plt.ylabel('τ')
 		plt.xlabel('Iteration')
[…]

### 2020-05-24 src/Fig_tau_posterior.py
save fig to file

@@ -113,7 +113,7 @@ if __name__ == '__main__':
 
 		ax.set_title(country.replace('_', ' '))
 		sns.despine()
-		plt.show()
-		# fig_filename = os.path.join(output_folder, 'figures', '{}_τ_posterior.pdf'.format(country))
-		# print("Saving to", fig_filename)
-		# fig.savefig(fig_filename, dpi=100, **savefig_bbox(*txt))
+		if not quiet: plt.show()
+		fig_filename = os.path.join(output_folder, 'figures', '{}_τ_posterior.pdf'.format(country))
+		print("Saving to", fig_filename)
+		fig.savefig(fig_filename, dpi=100, **savefig_bbox(*txt))

### 2020-05-24 src/Fig_tau_posterior.py
back to line for NPI

@@ -95,8 +95,8 @@ if __name__ == '__main__':
 		# for hpi_ in hpis:
 		# ax.fill_between([hpi_[0], hpi_[1]], 0, ymax, alpha=0.4, color='k')
 		# ax.axvline(first_date_days, color=red)
-		# ax.axvline(last_date_days, color=red)		
-		ax.fill_between([first_date_days, last_date_days], 0, ymax, alpha=0.4, color=red)
+		ax.axvline(last_date_days, color=red)		
+		# ax.fill_between([first_date_days, last_date_days], 0, ymax, alpha=0.4, color=red)
 		ax.set(xlabel=r'Effective start of NPI, $\tau$', 
 		       ylabel=r'Posterior density, $(P(\tau \mid \vec{X})$', 
 		       xlim=(τ_sample.min()-1, τ_sample.max()+1),
@@ -107,9 +107,9 @@ if __name__ == '__main__':
 		ax.set_xticks(days);
 		txt = ax.set_xticklabels([(start_date + d).strftime('%b %d') for d in deltas], rotation=45)
 
-		# ax.annotate(r'$\hat{\tau}$', (τ_med+0.1, ymax*0.875))
+		ax.annotate(r'$\hat{\tau}$', (τ_med+0.1, ymax*0.875))
 		# ax.annotate(r'$\tau^0$', (first_date_days+0.1, ymax*0.875))
-		# ax.annotate(r'$\tau^*$', (last_date_days+0.1, ymax*0.875))
+		ax.annotate(r'$\tau^*$', (last_date_days+0.1, ymax*0.875))
 
 		ax.set_title(country.replace('_', ' '))
 		sns.despine()

### 2020-05-24 ms/ms.tex
update bib
formatting, organization, phrasing

@@
 \usepackage[unicode=true,colorlinks=true,urlcolor=Blue,citecolor=black,linkcolor=black]{hyperref}
-\def\equationautorefname~#1\null{Eq.~(#1)\null}
+\def\equationautorefname~#1\null{Eq.~#1\null}
 \PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
@@
  \renewcommand\linenumberfont{\normalfont\small\sffamily}
-\linenumbers
+%\linenumbers
 % \modulolinenumbers[2]
@@
 
-During February and March 2020, several countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules to control the \covid pandemic caused by the \sars virus.
+During February and March 2020, several countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules, to control the \covid pandemic caused by the \sars virus.
 Overall, these interventions seem to have successfully reduced the spread of the pandemic.
@@
 For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective date on Mar~16 ($\pm$0.47 days 95\% CI).
-In contrast, Spain announced a lockdown on Mar~14, but we infer an effective start date on Mar~8 or~9 ($\pm$ 1.08 days 95\% CI).
+In contrast, Spain announced a lockdown on Mar~14, but we infer an effective start date on Mar~8 ($\pm$ 1.08 days 95\% CI).
 We discuss potential causes and consequences of our results.
@@
 \pagebreak
-% Introduction
-\section*{Introduction}
-
-The \covid pandemic has resulted in implementation of extreme non-pharmaceutical interventions (NPIs) in many affected countries.
-These interventions, from social distancing to lockdowns, are applied in a rapid and widespread fashion.
-NPIs are designed and assessed using epidemiological models, which follow the dynamics of the viral infection to forecast the effect of different mitigation and suppression strategies on the levels of infection, hospitalization, and fatality.
-These epidemiological models usually assume that the effect of NPIs on disease transmission begins at the officially declared date~\citep{Li2020,Gatto2020,Flaxman2020}.
-
-Adoption of public health recommendations is often critical for effective response to infectious diseases, and has been studied in the context of HIV~\citep{Kaufman2014} and vaccination~\citep{Dunn2015,Wiyeh2018}, for example.
-However, behavioural and social change does not occur immediately, but rather requires time to diffuse in the population through media, social networks, and social interactions. 
-Moreover, compliance to NPIs may differ between different interventions and between people with different backgrounds.
-For example, in a survey of 2,108 adults in the UK during Mar 2020, \citet{Atchison2020} found that those over 70 years old were more likely to adopt social distancing than young adults (18-34 years old), and that those with lower income were less likely to be able to work from home and to self-isolate.
-Similarly, compliance to NPIs may be impacted by personal experiences.
-\citet{Smith2020} have surveyed 6,149 UK adults in late April and found that people who believe they have already had \covid are more likely to think they are immune, and less likely to comply with social distancing measures.
-Compliance may also depend on risk perception as perceived by the the number of domestic cases or even by reported cases in other regions and countries.
-Interestingly, the perceived risk of \covid infection has likely caused a reduction in the number of influenza-like illness cases in the US starting from mid-February~\citep{Zipfel2020}.
-
-Here, we hypothesize that there is a significant difference between the official start of NPIs and their adoption by the public and therefore their effect on transmission dynamics.
-We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to infer the effective start date of NPIs from publicly available \covid case data in several geographical regions.
-We compare these estimates to the official dates, and find that they include both late and early effects of NPIs on \covid transmission dynamics.
-We conclude by demonstrating how differences between the official and effective start of NPIs can confuse assessments of the effectiveness of the NPIs in a simple epidemic control framework.
-
-%\pagebreak
-% Models and Methods
-\section*{Models and Methods}
-
-%%% Data %%%%
-\paragraph*{Data.} 
-We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions.
-These incidence data summarise the number of individuals $X_t$ tested positive for \sars RNA (using RT-qPCR) at each day $t$.
-Data for Wuhan, China retrieved from \citet{Pei2020}, data for 11 European countries retrieved from \citet{Flaxman2020}. 
-Regions in which there were multiple sequences of days with zero confirmed cases (e.g.
-France), we cropped the data to begin with the last sequence so that our analysis focuses on the first sustained outbreak rather than isolated imported cases.
-For dates of official NPI dates see \autoref{table:NPI_dates}.
 
 % Table - NPI dates %%%
-\begin{table}[h]
+\begin{table}[b!]
 \centering
@@
 
-%%% Fig - NPI dates %%%
+% Introduction
+\section*{Introduction}
+
+The \covid pandemic has resulted in implementation of extreme non-pharmaceutical interventions (NPIs) in many affected countries.
+These interventions, from social distancing to lockdowns, are applied in a rapid and widespread fashion.
+NPIs are designed and assessed using epidemiological models, which follow the dynamics of infection to forecast the effect of different mitigation and suppression strategies on the levels of infection, hospitalization, and fatality.
+These epidemiological models usually assume that the effect of NPIs on infection dynamics begins at the officially declared date~\citep{Flaxman2020,Gatto2020,Li2020}.
+
+Adoption of public-health recommendations is often critical for effective response to infectious diseases, and has been studied in the context of HIV~\citep{Kaufman2014} and vaccination~\citep{Dunn2015,Wiyeh2018}, for example.
+However, behavioural and social change does not occur immediately, but rather requires time to diffuse in the population through media, social networks, and social interactions. 
+Moreover, compliance to NPIs may differ between different interventions and between people with different backgrounds.
+For example, in a survey of 2,108 adults in the UK during Mar~2020, \citet{Atchison2020} found that those over 70 years old were more likely to adopt social distancing than young adults (18-34 years old), and that those with lower income were less likely to be able to work from home and to self-isolate.
+Similarly, compliance to NPIs may be impacted by personal experiences.
+\citet{Smith2020} have surveyed 6,149 UK adults in late Apr~2020 and found that people who believe they have already had \covid are more likely to think they are immune, and less likely to comply with social distancing guidelines.
+Compliance may also depend on risk perception as perceived by the the number of domestic cases or even by reported cases in other regions and countries.
+Interestingly, the perceived risk of \covid infection has likely caused a reduction in the number of influenza-like illness cases in the US starting from mid-Feb~2020~\citep{Zipfel2020}.
+
+Here, we hypothesise that there is a significant difference between the official start of NPIs and their effective adoption by the public and therefore their effect on infection dynamics.
+We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and a \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to infer the effective start date of NPIs from publicly available \covid case data in 12 geographical regions.
+We compare these estimates to the official dates, and find that they include both late and early effects of NPIs on infection dynamics.
+We conclude by demonstrating how differences between the official and effective start of NPIs can confound assessments of the effectiveness of the NPIs in a simple epidemic control framework.
+
+% Results
+\section*{Results}
+
+%%% Fig - tau summary %%%
+\begin{figure}[b!]
+    \centering
+	\includegraphics[width=0.6\textwidth]{../figures/Fig-tau-summary.pdf}
+    \caption{
+    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
+    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs is shown for different regions.
+The effective dates in Italy and Wuhan are significantly delayed compared to the official dates, whereas in Denmark, France, Sweden, Spain, and Germany, the effective date is earlier than the official date.
+	Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}.
+$\tau^*$ is the last NPI date (\autoref{table:NPI_dates}).
+Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e.
+interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
+    }
+    \label{fig:tau-summary}
+\end{figure}
+
+Several studies have described the effects of non-pharmaceutical interventions in different geographical regions~\citep{Flaxman2020,Gatto2020,Li2020}. 
+Some of these studies have assumed that the parameters of the epidemiological model change at a specific date (\autoref{eq:NPI_model}), and set the change date $\tau$ to the official NPI date $\tau^*$ (\autoref{table:NPI_dates}).
+They then fit the model once for time $t<\tau^*$ and once for time $t \ge \tau^*$.
+For example, \citet{Li2020} estimate the infection dynamics in China before and after $\tau^*$, which is set at Jan 23, 2020.
+Thereby, they effectively estimate the transmission and reporting rates before and after $\tau^*$ separately.
+
+Here, we estimate the joint posterior distribution of the \emph{effective} start date of the NPIs $\tau$ and the transmission and reporting rates before and after $\tau$ from the entire data, rather than splitting the data at~$\tau$.
+We then estimate the marginal posterior probability of $\tau$ by marginalising the joint posterior, and estimate $\hat{\tau}$ as the posterior median.
+
+We compared the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e.
+transmission and reporting rates are constant).
+The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}).
+When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free~$\tau$ was preferred in 8 out of 12 of the regions (although only narrowly for 5 of the 8).
+The exceptions are Austria, Belgium, Norway, and United Kingdom.
+
+We compared the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and found that in most regions the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
+The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well as Switzerland (see below).
+Norway also has a relatively wide credible interval, maybe because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
+In the following, we describe our findings in more detail.
+
+%%% Late
+\paragraph*{Late effective start of NPIs.}
+In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$.
+
+In Italy, the first case was officially confirmed on Feb~21.
+School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
+That is, the first and last official NPI dates are Mar~8 and Mar~11.
+However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.47 days 95\% CI ; \autoref{fig:late}).
+
+Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~2 ($\pm$2.85 days 95\% CI; \autoref{fig:late}).
+
+%%% Fig - Late %%%
 \begin{figure}[h]
     \centering
-	\includegraphics[width=0.5\textwidth]{../figures/Fig-NPI_dates.pdf}
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/posterior/Italy_τ_posterior.pdf}
+    \end{subfigure}
+  	~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/posterior/Wuhan_τ_posterior.pdf}
+    \end{subfigure}
     \caption{
-    \textbf{Official start of non-pharmaceutical interventions.}
-	See \autoref{table:NPI_dates} for more details. Wuhan, China is not shown.
+	\textbf{Late effect of non-pharmaceutical interventions in Italy and in Wuhan, China.}
+    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
+Red line shows the official last NPI date $\tau^*$.
+Black line shows the estimate $\hat{\tau}$.
+Shaded area show a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
     }
-    \label{fig:NPI_dates}
+    \label{fig:late}
 \end{figure}
+
+%%% Early
+\paragraph*{Early effective start of NPIs.}
+In contrast, in other regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
+
+In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score: 2--1) with a crowd of 50,965 in Seville.
+A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
+Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8-9 ($\pm$1.08 days 95 \%CI), rather than Mar~14 (\autoref{fig:early}).
+
+Similarly, in France we also estimate the effective start of NPIs $\hat{\tau}$ on Mar~8 or~9 ($\pm$6.27 days 95\% CI, \autoref{fig:early}).
+Although the credible interval is wider compared to Spain, spanning from Mar~2 to Mar~15, the official lockdown start at Mar~17 is later still, and even the earliest NPI, banning of public events, only started on Mar~13~\citep{Flaxman2020}.
+
+Interestingly, the effect of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~8-9, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
+The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
+This may suggest correlations between effective starts of NPIs due to global or international events.
+
+%%% Fig - Early %%%
+\begin{figure}[h]
+    \centering
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/posterior/Spain_τ_posterior.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+		\includegraphics[width=\textwidth]{../figures/posterior/France_τ_posterior.pdf}
+    \end{subfigure}
+    \caption{
+    \textbf{Early effect of non-pharmaceutical interventions in France and Spain.}
+    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
+Red line shows the official last NPI date $\tau^*$.
+Black line shows the estimated $\hat{\tau}$.
+Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
+	}
+	\label{fig:early}
+\end{figure}
+
+%%% Exact
+\paragraph*{Like a Swiss watch.}
+We find one case in which the official and effective dates match: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
+Indeed, the posterior median $\hat{\tau}$ is Mar~20 ($\pm$8.46 days 95\% CI), and the posterior distribution shows two density peaks: a smaller one between Mar~10 and Mar~14, and a bigger one between Mar~17 and Mar~22~(\autoref{fig:Switzerland}).
+It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
+
+%%% Assessment
+\paragraph*{Consequences of late and early effect of NPIs on real-time assessment.}
+
+The success of non-pharmaceutical interventions is assessed by health officials using various metrics, such as the decline in the growth rate of daily cases.
+These assessments are made a specific number of days after the intervention began, to accommodate for the expected serial interval~\citep{Banholzer2020} (i.e.
+time between successive cases in a chain of transmission), which is estimated at about 4-7 days~\citep{Gatto2020}.
+% Table 1
+
+However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
+This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~16 ($\hat{\tau}$).
+If health officials assumed the dynamics to immediately change at $\tau^*$, they will have expected the number of cases be within the red lines (posterior predictions assuming $\tau=\tau^*$).
+This would have lead to a significant underestimation, which might have been interpreted by as ineffectiveness of the NPI, leading to further escalations.
+However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data (stars).
+
+%%% Fig - Prediction %%%
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=0.5\textwidth]{../figures/Fig-prediction.pdf}
+    \caption{
+    \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
+    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
+    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
+    Shaded box illustrates a serial interval of seven days.
+    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
+    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
+    Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
+    } 
+    \label{fig:prediction}
+\end{figure}
+
+\pagebreak
+% Discussion
+\section*{Discussion}
+
+We have inferred the effective start date of NPIs in several geographical regions using an SEIR epidemiological model and an MCMC parameter estimation framework.
+We find examples of both late and early effect of NPIs (\autoref{fig:tau-summary}).
+
+For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred five days or more after the official date (\autoref{fig:late}).
+This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
+Late effect of NPIs may also be due to the time required by both the government and the citizens to organise for a lockdown, and for the new guidelines to diffuse in the population.
+
+In contrast, in most investigated countries (e.g., Spain and France), we inferred reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}).
+An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
+Adoption of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may have been influenced by increased risk perception due to domestic or international COVID\=/19-related reports.
+Indeed, the evidence supports a change in infection dynamics (i.e.
+a model with fixed or free $\tau$) even for Sweden~(\autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
+
+Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of \covid\citep{Gatto2020}.
+However, late and early effects such as we have inferred may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).
+
+We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (i.e.
+free or fixed model in \autoref{table:WAIC}).
+It could be interesting to check if the evidence favours a model with \emph{two} change-points, rather than one. 
+Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
+However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as changing weather, news of new treatments, and international outbreaks.
+
+As several countries begin to relieve lockdowns and ease restrictions, we expect similar delays and advances to occur: in some countries the population will behave as if restrictions were eased even before the official date, and in some countries the population will continue to self-restrict even after restrictions are officially removed.
+
+\paragraph*{Conclusions.}
+We have inferred the effective start date of NPIs and found that they often differ from the official dates.
+Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to infectious disease.
+Therefore, we emphasize the need to further study variability in compliance and behavior over both time and space.
+This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Arthur2020,Fenichela2011,Walters2013}.
+
+%\pagebreak
+% Models and Methods
+{\small
+\section*{Models and Methods}
+
+%%% Data %%%%
+\paragraph*{Data.} 
+We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions during Jan--Mar 2020.
+These incidence data summarise the number of individuals $X_t$ tested positive for \sars RNA (using RT-qPCR) at each day $t$.
+Data for Wuhan, China retrieved from \citet{Pei2020}, data for 11 European countries retrieved from \citet{Flaxman2020}. 
+Where there were multiple sequences of days with zero confirmed cases (e.g.
+France), we cropped the data to begin with the last sequence so that our analysis focuses on the first sustained outbreak rather than isolated imported cases.
+For official NPI dates see \autoref{table:NPI_dates}.
 
@@
 \paragraph*{SEIR model.} \label{sec:model}
-We model SARS\=/CoV\=/2 infection dynamics by following the number of susceptible $S$, exposed $E$, reported infected $I_r$, and unreported infected $I_u$ individuals in a population of size $N$.
+We model SARS\=/CoV\=/2 infection dynamics by following the number of susceptible $S$, exposed $E$, reported infected $I_r$, unreported infected $I_u$, and recovered $R$ individuals in a population of size $N$.
 This model distinguishes between reported and unreported infected individuals: the reported infected are those that have enough symptoms to eventually be tested and thus appear in daily case reports, to which we fit the model.
+This model is inspired by \citet{Li2020} and \citet{Pei2020}, who used a similar model with multiple regions and constant transmission and reporting rates to study \covid dynamics in China and in the continental US.
 
-Susceptible ($S$) individuals become exposed due to contact with reported or unreported infected individuals ($I_r$ or $I_u$) at a rate $\beta_t$ or $\mu \beta_t$.
-The parameter $0 < \mu < 1$ represents the decreased transmission rate from unreported infected individuals, who are often subclinical or even asymptomatic~\citep{Thompson2020,Ferretti2020}.
+Susceptible ($S$) individuals become exposed due to contact with reported or unreported infected individuals ($I_r$ or $I_u$) at a rate $\beta_t$ or $\mu \beta_t$, respectively.
+The parameter $0 < \mu < 1$ represents the decreased transmission rate from unreported infected individuals, who are often subclinical or even asymptomatic~\citep{Ferretti2020,Thompson2020}.
 The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioural changes of both susceptible and infected individuals.
 Exposed individuals, after an average incubation period of $Z$ days, become reported infected with probability $\alpha_t$ or unreported infected with probability $(1-\alpha_t)$.
-The reporting rate $0 < \alpha_t < 1$ may also change over time due to changes in human behavior.
-Infected individuals remain infectious for an average period of $D$ days, after which they either recover, or becomes ill enough to be quarantined.
-They therefore no longer infect other individuals, and the model does not track their frequency.
-The model is described by the following equations:
+The reporting rate $0 < \alpha_t < 1$ may also change over time due to changes in human behaviour.
+Infected individuals remain infectious for an average period of $D$ days, after which they either recover, or become ill enough to be quarantined.
+In either case, they no longer infect other individuals, and therefore effectively become recorved ($R$).
+The model is described by the following set of equations:
 
@@
 The initial numbers of exposed $E(0)$ and unreported infected $I_u(0)$ are free model parameters (i.e.
-estimated during model fitting), whereas the initial number of reported infected is assumed to be zero $I_r(0)=0$, and the number of susceptible is $S(0)=N-E(0)-I_u(0)$.
-This model is inspired by \citet{Li2020} and \citet{Pei2020}, who used a similar model with multiple regions and constant transmission $\beta$ and reporting rate $\alpha$ to infer \covid dynamics in China and the continental US, respectively.
+inferred from the data), whereas the initial number of reported infected and recovered is assumed to be zero, $I_r(0)=R(0)=0$, and the number of susceptible is $S(0)=N-E(0)-I_u(0)$.
 
@@
 \paragraph*{Likelihood function.}
-% TODO update
-For a given vector $\theta$ of model parameters the \emph{expected} cumulative number of reported infected individuals ($I_r$) until day $t$ is, following~\autoref{eq:model},
+
+For a given vector $\theta$ of model parameters the \emph{expected} cumulative number of reported infected individuals ($I_r$) until day $t$, following~\autoref{eq:model}, is
 \begin{equation} \label{eq:Yt}
@@
 $(Y_t(\theta) - \tilde{X}_{t-1})$.
-We therefore assume that $X_t$ conditioned on $\tilde{X}_{t-1}$ is Poisson distributed,
+We therefore assume that $X_t$ conditioned on $\tilde{X}_{t-1}$ is Poisson distributed, such that
 \begin{equation} \label{eq:Xt} \begin{aligned}
@@
 \Big(X_t \mid \tilde{X}_{t-1}, \theta \Big) & \sim 
-\mathit{Poi}\Big( \big(Y_t(\theta) - \tilde{X}_{t-1}\big) \cdot p_t \Big), \quad t>1.
+\mathit{Poi}\Big( \big(Y_t(\theta) - \tilde{X}_{t-1}\big) \cdot p_t \Big), \quad t=2,\ldots,T.
 \end{aligned}\end{equation}
 
-Hence, the \emph{likelihood function} $\mathbb{L}(\theta \mid \vec{X})$ for the parameter vector $\theta$ given the confirmed case data $\vec{X} = (X_1, \ldots, X_T)$ is defined by the probability to observe  $\vec{X}$ given $\theta$,
+Hence, the \emph{likelihood function} $\mathcal{L}(\theta \mid \vec{X})$ for a parameter vector $\theta$ given the confirmed case data $\vec{X} = (X_1, \ldots, X_T)$ is defined by the probability to observe  $\vec{X}$ given $\theta$,
 \begin{equation} \label{eq:likelihood}
-\mathbb{L}(\theta \mid \vec{X}) = 
+\mathcal{L}(\theta \mid \vec{X}) = 
 P(\vec{X} \mid \theta) = 
@@
 \paragraph*{NPI model.}
-To model non-pharmaceutical interventions (NPIs), we set the beginning of the NPIs to day $\tau$ and define
+To model non-pharmaceutical interventions (NPIs), we set the start of the NPIs to day $\tau$ and define
 \begin{equation} \label{eq:NPI_model}
 \beta_t = \begin{cases} 
-  \beta, & t < \tau \\ % TODO make sure < and not \le, and the next line \ge and not >
+  \beta, & t < \tau \\ 
   \beta \lambda, & t \ge \tau
@@
 \alpha_t = \begin{cases} 
-  \alpha_1, & t < \tau \\ % TODO make sure < and not \le, and the next line \ge and not >
+  \alpha_1, & t < \tau \\ 
   \alpha_2, & t \ge \tau
@@
 p_t = \begin{cases} 
-  1/9, & t < \tau \\ % TODO make sure < and not \le, and the next line \ge and not >
+  1/9, & t < \tau \\ 
   1/6, & t \ge \tau
@@
 We start our model $\Delta t$ days~\citep{Gatto2020} before the outbreak (defined as consecutive days with increasing confirmed cases) in each country.
-The model in \autoref{eq:model} is parameterized by the vector $\theta$, where
+The model in \autoref{eq:model} is parameterised by the vector $\theta$, where
 \begin{equation} \label{eq:theta}
@@
 The likelihood function is defined in \autoref{eq:likelihood}.
-The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using an \textit{affine-invariant ensemble sampler for Markov chain Monte Carlo} (MCMC)~\citep{Goodman2010} implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
+The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using the affine-invariant ensemble sampler for Markov chain Monte Carlo (MCMC)~\citep{Goodman2010} implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
 
@@
 \begin{equation} \label{eq:priors}
-\begin{aligned} % TODO verify before submitting
+\begin{aligned} 
 Z & \sim \mathit{Uniform}(2, 5) \\
@@
 where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
-% TODO something formal for the next line?
-We have also tested an uninformative uniform prior $U(1,T-2)$.
+We have also tested an uninformative uniform prior $\mathit{Uniform}(1,T-2)$.
 The uninformative prior could result in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom.
@@
 $\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
-We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood, higher DIC (see below), and unreasonable parameter estimates, and therefore chose $U(1,5)$ as the prior.
+We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior.
 
@@
 A minority (<5\%) of MCMC chains that fail to fully converge can lead to overestimation of the variance (the second term in \autoref{eq:WAIC}).
-Therefore, we exclude from the WAIC computation chains with mean log-likelihood that is three standard deviations or more from the overall mean. %TODO
+Therefore, we exclude from the computation of WAIC chains with mean log-likelihood that is three standard deviations or more from the overall mean. %TODO
 
 We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}), and plot the simulated dynamics~(\autoref{fig:ppc}).
-Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the compactness of the predictions) are good ways to visually compare models.
+Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the tightness of the predictions) are good ways to visually compare models.
 
@@
 
-%\pagebreak
-% Results
-\section*{Results}
-
-%%% Fig - tau summary %%%
-\begin{figure}[b!]
-    \centering
-	\includegraphics[width=0.6\textwidth]{../figures/Fig-tau-summary.pdf}
-    \caption{
-    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
-    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPI is shown for different regions.
-The effective NPI dates in Italy and Wuhan are significantly delayed compared to the official dates, whereas in Denmark, France, Spain, and Germany, the effective date is earlier than the official date.
-	$\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}.
-$\tau^*$ is the last NPI date, see \autoref{table:NPI_dates}.
-Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e.
-area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
-    }
-    \label{fig:tau-summary}
-\end{figure}
-
-Several studies have described the effects of non-pharmaceutical interventions in different geographical regions~\citep{Flaxman2020,Gatto2020,Li2020}. 
-These studies have assumed that the parameters of the epidemiological model change at a specific date, as in \autoref{eq:NPI_model}, and set the change date $\tau$ to the official NPI date $\tau^*$ (\autoref{table:NPI_dates}).
-They then fit the model once for time $t<\tau^*$ and once for time $t \ge \tau^*$.
-For example, \citet{Li2020} estimate the dynamics in China before and after $\tau^*$ at Jan 23. Thereby, they effectively estimate $(\beta, \alpha_1)$ and $(\lambda, \alpha_2)$ separately.
-Here we estimate the posterior distribution $P(\tau \mid \vec{X})$ of the \emph{effective} start date of the NPIs by jointly estimating $\tau, \beta, \lambda, \alpha_1, \alpha_2$ on the entire data per region (e.g.
-Italy, Austria), rather than splitting the data at $\tau^*$.
-We then estimate the posterior probability $P(\tau \mid \vec{X})$ by marginalizing the joint posterior, and estimate $\hat{\tau}$ as the posterior median.
-
-We compared the posterior predictive plots of a model with a free $\tau$ with those of a model with $\tau$ fixed at $\tau^*$ and without $\tau$.
-The model with free $\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}).
-When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free parameter for the start of the NPI was preferred over the other models in 8 out of 12 of the regions (although only narrowly for 5 of the 8); the exceptions are Austria, Belgium, Norway, and United Kingdom.
-
-We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
-The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well another country (see below).
-Norway also has a relatively wide credible interval, which could be expected as it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
-In the following, we describe our findings on late and early effective start of NPI in detail.
-
-%%% Late
-\paragraph*{Late effective start of NPIs.}
-In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$ (\autoref{fig:tau-summary}).
-
-In Italy, the first case was officially confirmed on Feb~21.
-School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
-That is, the first and last official dates are Mar~8 and Mar~11.
-However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.47 days 95\% CI ; \autoref{fig:late}).
-
-Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later at Feb~2 ($\pm$2.85 days 95\% CI \autoref{fig:late}).
-
-%%% Fig - Late %%%
-\begin{figure}[h]
-    \centering
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Wuhan_τ_posterior.pdf}
-    \end{subfigure}
-  	~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Italy_τ_posterior.pdf}
-    \end{subfigure}
-    \caption{
-	\textbf{Late effect of non-pharmaceutical interventions in Italy and Wuhan, China.}
-    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
-Red line shows the official last NPI date $\tau^*$.
-Black line shows the estimated $\hat{\tau}$.
-Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
-    }
-    \label{fig:late}
-\end{figure}
-
-%%% Early
-\paragraph*{Early effective start of NPIs.}
-In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
-In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
-A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8 or~9 ($\pm$1.08 days 95\%CI), rather than Mar~14 (\autoref{fig:early}).
-
-Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ on Mar~8 or Mar~9 ($\pm$6.27 days 95\% CI, \autoref{fig:early}).
-Although the credible interval is wider compared to Spain, spanning from Mar~2 to Mar~15, the official lockdown start at Mar~17 is later still, and even the earliest NPI, banning of public events, only started on Mar~13~\citep{Flaxman2020}.
-
-Interestingly, the effective start of NPIs $\hat{\tau}$ in both France and Spain is estimated at Mar~8, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
-The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
-This may suggest that correlation exist between effective start of NPIs due to global or international events.
-
-%%% Fig - Early %%%
-\begin{figure}[h]
-    \centering
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/France_τ_posterior.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Spain_τ_posterior.pdf}
-    \end{subfigure}
-    \caption{
-    \textbf{Early effect of non-pharmaceutical interventions in France and Spain.}
-    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
-Red line shows the official last NPI date $\tau^*$.
-Black line shows the estimated $\hat{\tau}$.
-Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
-	}
-	\label{fig:early}
-\end{figure}
-
-%%% Exact
-\paragraph*{Like a Swiss watch.}
-We find one case in which the official and effective dates match: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
-Indeed, the posterior median $\hat{\tau}$ is Mar~20 ($\pm$8.46 days 95\% CI), and the posterior distribution shows two density peaks: a smaller one between Mar~10 and Mar~14, and a bigger one between Mar~17 and Mar~22~(\autoref{fig:Switzerland}).
-It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
+} % small
 
@@
 \begin{landscape}
-\begin{table}[h]
+\begin{table}[]
 \centering
@@
 
-%%% Assessment
-\paragraph*{Consequences of late and early effect of NPIs on real-time assessment.}
-
-The success of non-pharmaceutical interventions is assessed by health officials using various metrics, such as the decline in the growth rate of daily cases.
-These assessments are made a specific number of days after the intervention began, to accommodate for the expected serial interval~\citep{Banholzer2020} (i.e.
-time between successive cases in a chain of transmission), which is estimated at about 4-7 days~\citep{Gatto2020}.
-% Table 1
-
-However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
-This is illustrated in \autoref{fig:prediction} using data and parameters from Italy.
-Here, a lockdown is officially ordered on Mar~10 ($\tau^*$), but its late effect on the transmission dynamics starts on Mar~16 ($\hat{\tau}$).
-If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases be within the red lines (posterior predictions assuming $\tau=\tau^*$).
-This leads to a significant underestimation, which might be interpreted by officials as ineffectiveness of NPIs, leading to further escalations.
-However, the number of cases will actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data.
-
-%%% Fig - Prediction %%%
-\begin{figure}[h]
-    \centering
-	\includegraphics[width=0.5\textwidth]{../figures/Fig-prediction.pdf}
-    \caption{
-    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
-    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
-    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
-    Shaded box illustrates a serial interval of seven days.
-    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model under-estimates the number of cases seven days after the start of the NPI.
-    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
-    Here, model parameters are estimates for Italy (\autoref{table:estimated-params}).
-    } 
-    \label{fig:prediction}
-\end{figure}
-
-\pagebreak
-% Discussion
-\section*{Discussion}
-
-We have estimated the effective start date of NPIs in several geographical regions using an SEIR epidemiological model and an MCMC parameter estimation framework.
-We find examples of both late and early effect of NPIs (\autoref{fig:tau-summary}).
-
-For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred more than five days after the official date (\autoref{fig:late}).
-This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
-Late effect of NPIs might also be due to the time required by both the government and the citizens to organise for a lockdown, and for the new guidelines to diffuse in the population.
-
-In contrast, in most investigated countries (e.g., Spain and France), we infer reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}).
-An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
-These behavioural adaptations may have a stronger influence than previously thought.
-In addition, diffusion of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may have been influenced by increased risk perception due to domestic or international COVID\=/19-related reports.
-Indeed, the evidence supports a change in transmission dynamics (i.e.
-a model with free $\tau$) even for Sweden~(\autoref{fig:ppc}), in which a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
-
-Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of \covid~\citep{Gatto2020}.
-However, the late and early effects we have estimated may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).
-
-We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (\autoref{table:WAIC}).
-It could be interesting to check if the evidence favours a model with \emph{two} change-points, rather than one. 
-Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
-However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as weather, or domestic and international events that affect risk perception.
-
-As several countries (e.g.
-Austria) begin to relieve lockdowns and ease restrictions, we expect similar delays and advances to occur: in some countries people will begin to behave as if restrictions were eased even before the official date, and in some countries people will continue to self-restrict even after restrictions are officially removed.
-
-\paragraph*{Conclusions.}
-We have estimated the effective start date of NPIs and found that they often differ from the official dates.
-Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to infectious disease.
-Therefore, we emphasize the need to further study variability in compliance and behavior over both time and space.
-This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Fenichela2011,Walters2013,Arthur2020}.
-
 %\pagebreak
@@
 \bibliographystyle{agsm}
-\bibliography{/Users/yoavram/Documents/library}
-%\bibliography{ms}
+%\bibliography{/Users/yoavram/Documents/library}
+\bibliography{ms}
 
@@
 \beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
+
+%%%% Fig - NPI dates %%%
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=0.5\textwidth]{../figures/Fig-NPI_dates.pdf}
+    \caption{
+    \textbf{Official start of non-pharmaceutical interventions.}
+	See \autoref{table:NPI_dates} for more details. Wuhan, China is not shown.
+    } 
+    \label{fig:NPI_dates}
+\end{figure}
 
@@
 \textbf{WAIC values for the different models.}
-	WAIC (widely applicable information criterion)~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
+	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
-Bold emphasise cases in which the \emph{Free} model had the lowest WAIC.
+Bold values emphasize cases in which the \emph{Free} model has the lowest WAIC.
 $*$ and $**$ mark if the difference was smaller or greater than 2, which is a popular significance level for model comparison~\citep{Kass1995}.
@@
 Black line shows the estimated $\hat{\tau}$.
-Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
+Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
      }
@@
     \caption{
-    \textbf{Figure S3. Posterior prediction check plots}
+    \textbf{Figure S4. Posterior prediction plots.}
     Markers represent data ($\vec{X}$).
 Black line represent a smoothing of the data points using a Savitzky-Golay filter.
-Color lines represent posterior predictions from a model with fixed $\tau$, in red, and free $\tau$, in blue.
+Colored lines represent posterior predictions from a model with fixed $\tau$ in red, and free $\tau$ in blue.
 These predictions are made by drawing 1,000 samples from the parameter posterior distribution and then generating a daily case count using the SEIR model in \autoref{eq:model}.

### 2020-05-24 ms/ms.tex
for at, revise, describe WAIC for uniform prior

@@
 
-We compared the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e.
+We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e.
 transmission and reporting rates are constant).
 The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}).
-When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free~$\tau$ was preferred in 8 out of 12 of the regions (although only narrowly for 5 of the 8).
+When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is preferred in 8 out of 12 of the regions (although only narrowly for 5 of the 8).
 The exceptions are Austria, Belgium, Norway, and United Kingdom.
 
-We compared the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and found that in most regions the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
+We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
 The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well as Switzerland (see below).
@@
 \paragraph*{Late effective start of NPIs.}
-In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$.
+In both Wuhan, China, and in Italy we estimate that the effective start of NPIs $\hat{\tau}$ is significantly later than the official date $\tau^*$.
 
@@
 Black line shows the estimate $\hat{\tau}$.
-Shaded area show a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
+Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
     }
@@
     \centering
-	\includegraphics[width=0.5\textwidth]{../figures/Fig-prediction.pdf}
+	\includegraphics[width=0.4\textwidth]{../figures/Fig-prediction.pdf}
     \caption{
@@
 
-For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred five days or more after the official date (\autoref{fig:late}).
+For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five days or more after the official date (\autoref{fig:late}).
 This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
-Late effect of NPIs may also be due to the time required by both the government and the citizens to organise for a lockdown, and for the new guidelines to diffuse in the population.
+Late effect of NPIs may also be due to the time required by both the government and the citizens to organise for a lockdown, and for the new guidelines to be adopted by the population.
 
@@
 free or fixed model in \autoref{table:WAIC}).
-It could be interesting to check if the evidence favours a model with \emph{two} change-points, rather than one. 
+It could be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
 Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
@@
 We have inferred the effective start date of NPIs and found that they often differ from the official dates.
-Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to infectious disease.
+Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to an epidemic.
 Therefore, we emphasize the need to further study variability in compliance and behavior over both time and space.
@@
 where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
-We have also tested an uninformative uniform prior $\mathit{Uniform}(1,T-2)$.
-The uninformative prior could result in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom.
+We also tested an uninformative uniform prior, $\mathit{Uniform}(1,T-2)$.
+WAIC (\autoref{eq:WAIC}) of a model with this uniform prior was either higher, or lower by less than 2, compared to WAIC of a model with the truncated normal prior.
+The uninformative prior resulted in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom.
 This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
-We therefore decided to use the more informative truncated normal prior.
+We therefore decided to use the more informative truncated normal prior for $\tau$.
 Other priors follow \citet{Li2020}, with the following exceptions.
 $\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
-We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior.
+We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior for $\Delta t$.
 

### 2020-06-24 ms/ms.tex
started updating for Apr11 data

@@
     \centering
-	\includegraphics[width=0.6\textwidth]{../figures/Fig-tau-summary.pdf}
+	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Apr11.pdf}
     \caption{
@@
     	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs is shown for different regions.
-The effective dates in Italy and Wuhan are significantly delayed compared to the official dates, whereas in Denmark, France, Sweden, Spain, and Germany, the effective date is earlier than the official date.
+The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers).
+In contrast, the effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although this is significant only for Germany (i.e., zero is not included in 75\% CI).
 	Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}.
@@
 Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e.
-interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
+interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.) \autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, rather then Apr~11, 2020.
     }
@@
 When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is preferred in 8 out of 12 of the regions (although only narrowly for 5 of the 8).
-The exceptions are Austria, Belgium, Norway, and United Kingdom.
+The exceptions are Austria, Belgium, Norway, and United Kingdom. % TODO UPDATE
 
-We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
-The exceptions are, as with the comparison to the simpler models, Austria, Belgium, and United Kingdom, as well as Switzerland (see below).
-Norway also has a relatively wide credible interval, maybe because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
+We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions (10 of 12) the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
+The exceptions are%, as with the comparison to the simpler models, % TODO UPDATE
+Switzerland (see below), and Norway. The latter also has a relatively wide credible interval, maybe because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
 In the following, we describe our findings in more detail.
@@
 \paragraph*{Late effective start of NPIs.}
-In both Wuhan, China, and in Italy we estimate that the effective start of NPIs $\hat{\tau}$ is significantly later than the official date $\tau^*$.
+In half of the examined regions, we estimate that the effective start of NPIs $\hat{\tau}$ is significantly later than the official date $\tau^*$.
 
@@
 That is, the first and last official NPI dates are Mar~8 and Mar~11.
-However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.47 days 95\% CI ; \autoref{fig:late}).
+However, we estimate the effective date $\hat{\tau}$ at Mar~17 ($\pm$2.99 days 95\% CI ; \autoref{fig:late}).
 
-Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~2 ($\pm$2.85 days 95\% CI; \autoref{fig:late}).
+In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~2 ($\pm$2.85 days 95\% CI; \autoref{fig:late}).
+% TODO CHECK
+
+In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score: 2--1) with a crowd of 50,965 in Seville.
+A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
+Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~24 ($\pm$2.96 days 95 \%CI), rather than Mar~14 (\autoref{fig:late}).
+
+Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ on Mar~24 ($\pm$2.29 days 95\% CI, \autoref{fig:late}).
+This is about a week later then the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
+
+Interestingly, the effect of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~24, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
+The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
+This may suggest correlations between effective starts of NPIs due to global or international events.
 
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Italy_τ_posterior.pdf}
+        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Italy_τ_posterior.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Wuhan_τ_posterior.pdf}
+        \includegraphics[width=\textwidth]{../figures/posterior/Mar28/Wuhan_τ_posterior.pdf}
+    \end{subfigure}
+    \\
+    \begin{subfigure}{0.45\textwidth} % TODO UPDATE
+        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Spain_τ_posterior.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/France_τ_posterior.pdf}
     \end{subfigure}
     \caption{
-	\textbf{Late effect of non-pharmaceutical interventions in Italy and in Wuhan, China.}
+	\textbf{Late effect of non-pharmaceutical interventions.}
     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
@@
 \paragraph*{Early effective start of NPIs.}
-In contrast, in other regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
+In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
 
-In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score: 2--1) with a crowd of 50,965 in Seville.
-A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8-9 ($\pm$1.08 days 95 \%CI), rather than Mar~14 (\autoref{fig:early}).
+Germany...
 
-Similarly, in France we also estimate the effective start of NPIs $\hat{\tau}$ on Mar~8 or~9 ($\pm$6.27 days 95\% CI, \autoref{fig:early}).
-Although the credible interval is wider compared to Spain, spanning from Mar~2 to Mar~15, the official lockdown start at Mar~17 is later still, and even the earliest NPI, banning of public events, only started on Mar~13~\citep{Flaxman2020}.
+Sweden...
 
-Interestingly, the effect of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~8-9, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
-The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
-This may suggest correlations between effective starts of NPIs due to global or international events.
+Denmark...
 
@@
     \centering
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Spain_τ_posterior.pdf}
+    \begin{subfigure}{0.45\textwidth} % TODO UPDATE xlim
+        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Germany_τ_posterior.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-		\includegraphics[width=\textwidth]{../figures/posterior/France_τ_posterior.pdf}
+		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/Sweden_τ_posterior.pdf}
+    \end{subfigure}
+
+	\begin{subfigure}{0.45\textwidth} % TODO UPDATE xlim
+        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Switzerland_τ_posterior.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/Norway_τ_posterior.pdf}
     \end{subfigure}
     \caption{
-    \textbf{Early effect of non-pharmaceutical interventions in France and Spain.}
+    \textbf{Early and exact effect of non-pharmaceutical interventions.}
     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
@@
 \paragraph*{Like a Swiss watch.}
-We find one case in which the official and effective dates match: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
-Indeed, the posterior median $\hat{\tau}$ is Mar~20 ($\pm$8.46 days 95\% CI), and the posterior distribution shows two density peaks: a smaller one between Mar~10 and Mar~14, and a bigger one between Mar~17 and Mar~22~(\autoref{fig:Switzerland}).
+We find one case in which the official and effective dates match (and the credible interval is narrow): Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
+Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\pm$2.51 days 95\% CI, see \autoref{fig:early}).
 It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
+
+The estimated effective date in Norway is Mar~22 ($\pm$10.79 days 95\% CI), which doesn't significantly differ from the official date of Mar~24.
+However, the posterior distribution is very wide (\autoref{fig:early}): it covers the range between Mar~9, three days before the first NPI, and Mar~29, five days  after the last NPI (see 
+\autoref{table:NPI_dates} and \autoref{fig:NPI_dates} for NPI dates).
+The uncertainty in the estimate of the effective start of NPI may be due to the long duration between the first and last NPI; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is quite narrow (\autoref{fig:early}).
 
@@
 However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
-This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~16 ($\hat{\tau}$).
+This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~17 ($\hat{\tau}$).
 If health officials assumed the dynamics to immediately change at $\tau^*$, they will have expected the number of cases be within the red lines (posterior predictions assuming $\tau=\tau^*$).
@@
 
-%%% Fig - Prediction %%%
+%%% Fig - Prediction %%% TODO UPDATE?
 \begin{figure}[h]
@@
 
+In most investigated countries we find a late effect of NPI on the transmission dynamics.
 For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five days or more after the official date (\autoref{fig:late}).
-This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
+This difference might be explained, in some case, by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
 Late effect of NPIs may also be due to the time required by both the government and the citizens to organise for a lockdown, and for the new guidelines to be adopted by the population.
 
-In contrast, in most investigated countries (e.g., Spain and France), we inferred reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}).
+In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}), although this is only significant for Germany (\autoref{fig:tau-summary}).
 An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
@@
 a model with fixed or free $\tau$) even for Sweden~(\autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
+%TODO check
 
@@
 free or fixed model in \autoref{table:WAIC}).
+%TODO check
 It could be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
@@
 
-As several countries begin to relieve lockdowns and ease restrictions, we expect similar delays and advances to occur: in some countries the population will behave as if restrictions were eased even before the official date, and in some countries the population will continue to self-restrict even after restrictions are officially removed.
+As countries relieve lockdowns and ease restrictions and the number of cases increases, we expect similar delays and advances to occur: in some countries the population will behave as if restrictions were eased even before the official date, and in some countries the population will continue to self-restrict even after restrictions are officially removed.
 
@@
 \paragraph*{Data.} 
-We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions during Jan--Mar 2020.
+We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions during Jan--Apr 2020.
 These incidence data summarise the number of individuals $X_t$ tested positive for \sars RNA (using RT-qPCR) at each day $t$.
@@
 We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior for $\Delta t$.
+Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11.
 
@@
 
-%% Table - estimated parameters %%%
+%% Table - estimated parameters %%% TODO UPDATE
 \begin{landscape}
@@
 
-% Table - WAIC %%%
+% Table - WAIC %%% TODO UPDATE
 \begin{table}[h]
@@
 
-%%% Fig Fr-vs-Es %%%
+%%% Fig - tau summary Mar 28 %%%
+\begin{figure}[b!]
+    \centering
+	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Mar28.pdf}
+    \caption{
+    \textbf{Official vs. effective start of non-pharmaceutical interventions estimated up to Mar~28.}
+    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs estimated from case data up to Mar~28, 2020, shown for different regions.
+Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}.
+$\tau^*$ is the last NPI date (\autoref{table:NPI_dates}).
+Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e.
+interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
+    }
+    \label{fig:tau-summary-mar28}
+\end{figure}
+
+%%% Fig Fr-vs-Es %%% TODO UPDATE
 \begin{figure}[h]
@@
 
-%%% Fig Switzerland %%%
- \begin{figure}[h]
-     \centering
-         \includegraphics[width=0.75\textwidth]{../figures/posterior/Switzerland_τ_posterior.pdf}
-     \caption{
- 	\textbf{Effective date of non-pharmaceutical interventions in Switzerland matches the official date}
-     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
+%%%% Fig Switzerland %%%
+% \begin{figure}[h]
+%     \centering
+%         \includegraphics[width=0.75\textwidth]{../figures/posterior/Apr11/Switzerland_τ_posterior.pdf}
+%     \caption{
+% 	\textbf{Effective date of non-pharmaceutical interventions in Switzerland matches the official date}
+%     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
 Red line shows the official last NPI date $\tau^*$.
@@
 Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
-     }
-     \label{fig:Switzerland}
- \end{figure}
+%     }
+%     \label{fig:Switzerland}
+% \end{figure}
 
-%%% Fig PPC %%%
+%%% Fig PPC %%% TODO UPDATE
 \begin{figure}[h]	

### 2020-06-24 src/Fig_tau_summary.py
determine red and blue by ci75

@@ -59,6 +59,7 @@ if __name__ == '__main__':
 	df['τ official days'] = [date_to_int(x) for x in df['τ official']]
 	df['τ official - mean days'] = df['τ official days'] - df['τ mean days']
 	df['τ official - median days'] = df['τ official days'] - df['τ median days']
+	df['τ median - official days'] = df['τ median days'] - df['τ official days']
 	df['τ official - MAP days'] = df['τ official days'] - df['τ MAP days']
 	df['τ mean'] = [date_to_date(x) for x in df['τ mean']]
 	df['τ median'] = [date_to_date(x) for x in df['τ median']]
@@ -78,33 +79,36 @@ if __name__ == '__main__':
 		plt.plot(np.arange(30, 90), np.arange(30, 90), ls='--', color='k')
 		plt.show()
 
-	col = 'τ official - median days'
+	col = 'τ median - official days'
 	ci75_col = 'τ CI median (75%)'
 	ci95_col = 'τ CI median (95%)'
 
 	fig, ax = plt.subplots(figsize=(6, 6))
-	df_ = df.sort_values(col)
+	df_ = df.sort_values(col, ascending=False)
 
-	val = -df_[col]
+	val = df_[col]
 	ci75 = df_[ci75_col]
 	ci95 = df_[ci95_col]
 	country = df_['country']
 	ax.hlines(country, val-ci95, val+ci95)
 	ax.hlines(country, val-ci75, val+ci75, lw=3)
 
-	idx = df_[col] >= 1
-	val = -df_.loc[idx, col]
+	# idx = df_[col] >= 1
+	idx = df_[col] + ci75 < 0
+	val = df_.loc[idx, col]
 	country = df_.loc[idx, 'country']
 	ax.plot(val, country, 'o', markersize=10, color=blue)
 
 
-	idx = df_[col] <= -1
-	val = -df_.loc[idx, col]
+	# # idx = df_[col] <= -1
+	idx = 0 < df_[col] - ci75
+	val = df_.loc[idx, col]
 	country = df_.loc[idx, 'country']
 	ax.plot(val, country, 'o', markersize=10, color=red)
 
-	idx = (-1 < df_[col]) & (df_[col] < 1)
-	val = -df_.loc[idx, col]
+	# # idx = (-1 < df_[col]) & (df_[col] < 1)
+	idx = (0 <= df_[col] + ci75) & (df_[col] - ci75 <= 0)
+	val = df_.loc[idx, col]
[…]

### 2020-06-29 ms/ms.tex
update ISF
a bit more update for Apr11 date

@@
     }
-    \label{fig:tau-summary}
+    \label{fig:tau-summary} %TODO maybe to put some footnote about the realibility of the result for Sweden and Denmark?
 \end{figure}
@@
 The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}).
-When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is preferred in 8 out of 12 of the regions (although only narrowly for 5 of the 8).
-The exceptions are Austria, Belgium, Norway, and United Kingdom. % TODO UPDATE
+For all of the 11 European countries the out-of-sample root mean squared error (RMSE) is lower for the model with a free~$\tau$ (\autoref{table:RMSE}).
+We don't check it for Wuhan, China, because we have no out-of-sample data for it.
+% TODO reference?
+; add table ; find more data for Wuhan?
+When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is strongly preferred in all of 12 countries except Sweden and Denmark.
+Both Sweden and Denmark, don't have one visual peak in daily cases for the dates the models are evaluated, so probably it is not enough data for the free~$\tau$ model.
 
@@
 
-In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~2 ($\pm$2.85 days 95\% CI; \autoref{fig:late}).
+In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~3 ($\pm$2.85 days 95\% CI; \autoref{fig:late}).
 % TODO CHECK
@@
 In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
-
-Germany...
-
-Sweden...
-
-Denmark...
+Germany... %TODO
+Sweden and Denmark show early effective start of NPIs.
+However, for both of the contries, the dataset that we use doesn't contain dates with the peak of the pandemic.
+WAIC model comparison method doesn't prefer the model with a free~$\tau$ on the model with a fixed~$\tau$.
+Therefore we can't trust in the outcome of the free~$\tau$ model for those countries.
 
@@
 
-We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (i.e.
+We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 10 out of 12 regions (i.e.
 free or fixed model in \autoref{table:WAIC}).
-%TODO check
 It could be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
@@
 We thank Lilach Hadany and Oren Kolodny for discussions and comments.
-This work was supported in part by the Israel Science Foundation 552/19 and 1399/17.
+This work was supported in part by the Israel Science Foundation 3811/19 and 1399/17.
 }

### 2020-07-07 src/Table-WAIC.py
update job ids

@@ -59,7 +59,7 @@ def WAIC(logliks):
 if __name__ == '__main__':    
     output_folder = r'/Users/yoavram/Library/Mobile Documents/com~apple~CloudDocs/EffectiveNPI-Data/output'
     
-    job_ids = ['2020-05-14-n1-normal-1M', '2020-05-14-n1-notau-1M', '2020-05-15-n1-fixed-tau-1M']
+    job_ids = ['2020-05-26-Apr11', '2020-06-27-Apr11-notau', '2020-06-25-Apr11-fixedtau']
     print('Job IDs:')
     print(job_ids)    
     countries = 'Austria Belgium Denmark France Germany Italy Norway Spain Sweden Switzerland United_Kingdom Wuhan'.split(' ')
@@ -79,9 +79,9 @@ if __name__ == '__main__':
             ))
 
     df = pd.DataFrame(results)
-    df.loc[df['job_id'] == '2020-05-14-n1-normal-1M', 'job_id'] = 'Free'
-    df.loc[df['job_id'] == '2020-05-14-n1-notau-1M', 'job_id'] = 'No'
-    df.loc[df['job_id'] == '2020-05-15-n1-fixed-tau-1M', 'job_id'] = 'Fixed'
+    df.loc[df['job_id'] == '2020-05-26-Apr11', 'job_id'] = 'Free'
+    df.loc[df['job_id'] == '2020-06-27-Apr11-notau', 'job_id'] = 'No'
+    df.loc[df['job_id'] == '2020-06-25-Apr11-fixedtau', 'job_id'] = 'Fixed'
     df = df.rename(columns={'country':'Country', 'job_id':'Model'})
     df['Country'] = [x.replace('_', ' ') for x in df['Country']]
     df.loc[df['Country']=='Wuhan', 'Country'] = 'Wuhan China'

### 2020-07-07 ms/ms.tex
update and revise -  YR and IK together

--- 2020-06-24 ms/ms.tex
@@
 %%% Fig - tau summary %%%
+% generated with /python Fig_tau_summary.py 2020-05-26-Apr11/
 \begin{figure}[b!]
@@
 In contrast, the effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although this is significant only for Germany (i.e., zero is not included in 75\% CI).
+The confidence intervals for Sweden, Denmark, and Norway are especially wide, see text and \autoref{fig:early} for possible explanation.
 	Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}.
@@
 Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e.
-interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.) \autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, rather then Apr~11, 2020.
+interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.) \autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, 2020, rather then Apr~11.
     }
@@
 transmission and reporting rates are constant).
-The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}).
-When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is preferred in 8 out of 12 of the regions (although only narrowly for 5 of the 8).
-The exceptions are Austria, Belgium, Norway, and United Kingdom. % TODO UPDATE
+All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~23, and these predictions were then compared to the real case data. 
+The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
+In all 11 of the European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
+% TODO RMSE reference? (Ilia)
+% TODO add table for RMSE (Ilia)
+% TODO consider dropping WAIC for RMSE (both)
+When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark.
+Notably, the data for Sweden and Denmark does not have a peak during the evaluated dates, possibly leading to wide 95\% confidence intervals on $\hat{tau}$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).
 
 We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions (10 of 12) the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
-The exceptions are%, as with the comparison to the simpler models, % TODO UPDATE
-Switzerland (see below), and Norway. The latter also has a relatively wide credible interval, maybe because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
+The exceptions are Switzerland (see below), and Norway.
+The latter also has a wide 95\% credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
 In the following, we describe our findings in more detail.
@@
 
-In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~2 ($\pm$2.85 days 95\% CI; \autoref{fig:late}).
-% TODO CHECK
+In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~3 ($\pm$2.85 days 95\% CI; \autoref{fig:late}).
 
@@
     \\
-    \begin{subfigure}{0.45\textwidth} % TODO UPDATE
+    \begin{subfigure}{0.45\textwidth}
         \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Spain_τ_posterior.pdf}
@@
 
-Germany...
+Germany... %TODO (Yoav) \autoref{fig:early}
 
-Sweden...
-
-Denmark...
+Sweden and Denmark show early effective start of NPIs.
+However, for both countries, the casa data does not include a "peak", that is, the number of daily cases continues to grow during the evaluated dates.
+We suspect this increases the uncertainty in our inference.
+Accordingly, WAIC of our model with a free $\tau$ parameter was higher (less fit) than the WAIC of a model with $\tau$ fixed at $\tau^*$ the official date.
 
@@
     \centering
-    \begin{subfigure}{0.45\textwidth} % TODO UPDATE xlim
+    \begin{subfigure}{0.45\textwidth}
         \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Germany_τ_posterior.pdf}
@@
 
-	\begin{subfigure}{0.45\textwidth} % TODO UPDATE xlim
+	\begin{subfigure}{0.45\textwidth}
         \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Switzerland_τ_posterior.pdf}
@@
 time between successive cases in a chain of transmission), which is estimated at about 4-7 days~\citep{Gatto2020}.
-% Table 1
+% Gatto2020 Table 1
 
@@
 
-%%% Fig - Prediction %%% TODO UPDATE?
+%%% Fig - Prediction
 \begin{figure}[h]
@@
 An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
-Adoption of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may have been influenced by increased risk perception due to domestic or international COVID\=/19-related reports.
+Adoption of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may be influenced by increased risk perception due to domestic or international COVID\=/19-related reports~\citep{Arthur2020}.
 Indeed, the evidence supports a change in infection dynamics (i.e.
-a model with fixed or free $\tau$) even for Sweden~(\autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
-%TODO check
+a model with fixed or free $\tau$) even for Sweden~(\autoref{table:RMSE}, \autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
 
@@
 
-We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (i.e.
-free or fixed model in \autoref{table:WAIC}).
-%TODO check
-It could be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
+We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in most regions (i.e.
+free or fixed model in \autoref{table:RMSE} and \autoref{table:WAIC}).
+It may be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
 Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
@@
 
+% UPDATE
 As countries relieve lockdowns and ease restrictions and the number of cases increases, we expect similar delays and advances to occur: in some countries the population will behave as if restrictions were eased even before the official date, and in some countries the population will continue to self-restrict even after restrictions are officially removed.
@@
 Infected individuals remain infectious for an average period of $D$ days, after which they either recover, or become ill enough to be quarantined.
-In either case, they no longer infect other individuals, and therefore effectively become recorved ($R$).
+In either case, they no longer infect other individuals, and therefore effectively become recovered ($R$).
 The model is described by the following set of equations:
@@
 A minority (<5\%) of MCMC chains that fail to fully converge can lead to overestimation of the variance (the second term in \autoref{eq:WAIC}).
-Therefore, we exclude from the computation of WAIC chains with mean log-likelihood that is three standard deviations or more from the overall mean. %TODO
+Therefore, we exclude from the computation of WAIC chains with mean log-likelihood that is three standard deviations or more from the overall mean.
 
-We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}), and plot the simulated dynamics~(\autoref{fig:ppc}).
+We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$ fitted to data up to Apr~11, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}) from Apr~12 to Apr~24, and plot the predicted dynamics~(\autoref{fig:ppc}).
 Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the tightness of the predictions) are good ways to visually compare models.
+We also compute the expected posterior root mean squared error of these predictions~(\autoref{table:RMSE}).
 
@@
 Samples from the posterior distributions will be deposited on FigShare.
-
 } % small
 
-%% Table - estimated parameters %%% TODO UPDATE
+%% Table - estimated parameters
+% generated with /python Table_estimated_params.py 2020-05-26-Apr11/
 \begin{landscape}
@@
 
-% Table - WAIC %%% TODO UPDATE
+% Table - WAIC %%% TODO regenerate table
+% generated with /python Table-WAIC.py/
 \begin{table}[h]
@@
     every last row/.style={after row=\hline},
-    ]{../figures/Table-WAIC_tmp.csv}
+    ]{../figures/Table-WAIC.csv}
 \caption{
@@
 	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
-WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular significance level for model comparison~\citep{Kass1995}.
 Bold values emphasize cases in which the \emph{Free} model has the lowest WAIC.
-$*$ and $**$ mark if the difference was smaller or greater than 2, which is a popular significance level for model comparison~\citep{Kass1995}.
 }
@@
 
+% Table - RMSE %%% TODO regenerate table
+% generated with /python Table-RMSE.py/
+%\begin{table}[h]
+%\centering
+%\pgfplotstabletypeset[
+%    col sep=comma,
+%    string type,
+%    every head row/.style={before row=\hline,after row=\hline},
+%    every last row/.style={after row=\hline},
+%    ]{../figures/Table-RMSE.csv}
+%\caption{
+%\textbf{Posterior RMSE of out-of-sample predictions with the different models.}
+%	Expected posterior predictive RMSE (root mean squared error) for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
+%	In all cases, the model with free parameter $\tau$ has the lowest RMSE.
+%	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions from Apr~12 to Apr~24 by sampling model parameters from the posterior distribution.
+%	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model.
+%}
+%\label{table:RMSE}
+%\end{table}
+
 %%% Fig - tau summary Mar 28 %%%
+% generated with
+% > python Fig_tau_summary.py 2020-06-23-Mar28
+% > mv ../figures/Fig-tau-summary.pdf ../figures/Fig-summary-Mar28.pdf
 \begin{figure}[b!]
@@
 
-%%% Fig Fr-vs-Es %%% TODO UPDATE
+%%% Fig Fr-vs-Es
+% generated with Fig-fr-vs-es.ipynb
 \begin{figure}[h]
@@
 
-%%%% Fig Switzerland %%%
-% \begin{figure}[h]
-%     \centering
-%         \includegraphics[width=0.75\textwidth]{../figures/posterior/Apr11/Switzerland_τ_posterior.pdf}
-%     \caption{
-% 	\textbf{Effective date of non-pharmaceutical interventions in Switzerland matches the official date}
-%     Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
-Red line shows the official last NPI date $\tau^*$.
-Black line shows the estimated $\hat{\tau}$.
-Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
-%     }
-%     \label{fig:Switzerland}
-% \end{figure}
-
 %%% Fig PPC %%% TODO UPDATE
+% generated with ppc.sh
 \begin{figure}[h]	
@@
     \end{subfigure}
-    %%%  
-	\subcaption{Wuhan, China}    
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Wuhan_ppc.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Wuhan_ppc.pdf}
-    \end{subfigure}
+%    %%%  
+%	\subcaption{Wuhan, China}    
+%    \begin{subfigure}{0.45\textwidth}
+%        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Wuhan_ppc.pdf}
+%    \end{subfigure}
+%    ~
+%    \begin{subfigure}{0.45\textwidth}
+%        \includegraphics[width=\textwidth]{../figures/ppc/free/Wuhan_ppc.pdf}
+%    \end{subfigure}
     %%%
@@
     \textbf{Figure S4. Posterior prediction plots.}
-    Markers represent data ($\vec{X}$).
-Black line represent a smoothing of the data points using a Savitzky-Golay filter.
-Colored lines represent posterior predictions from a model with fixed $\tau$ in red, and free $\tau$ in blue.
-These predictions are made by drawing 1,000 samples from the parameter posterior distribution and then generating a daily case count using the SEIR model in \autoref{eq:model}.
+    Markers represent data ($\vec{X}$): circles for data up to Apr~11, 2020, and stars after Apr~12; Apr~11 is represented using a vertical dashed line.
+Black line represents a smoothing of the data points using a Savitzky-Golay filter.
+Coloured lines represent posterior predictions from a model with fixed $\tau$ in red, and free $\tau$ in blue.
+The models were fitted with data up to Apr~11.
+The predictions are generated by drawing 1,000 parameters sets from the posterior distribution, and then generating a daily case count using the SEIR model in \autoref{eq:model}.
 Note the differences in the y-axis scale.

### 2020-07-08 src/Table-RMSE.py
print filename

@@ -20,4 +20,6 @@ def bold_all(df, columns):
 
 df_out.set_index('Country', inplace=True)
 bold_all(df_out, df_out.columns)
-df_out.to_csv('../figures/Table-RMSE_tmp.csv',index='Country')
+output_fname = '../figures/Table-RMSE.csv'
+df_out.to_csv(output_fname,index='Country')
+print(output_fname)

### 2020-07-08 src/ppc.sh
changed Table-WAIC

@@ -1,14 +1,16 @@
-for country in Austria Belgium Denmark France Germany Italy Norway Spain Sweden Switzerland United_Kingdom
-# for country in Spain
+# for country in Austria Belgium Denmark France Germany Italy Norway Spain Sweden Switzerland United_Kingdom
+for country in Wuhan
 do
 	# python Fig_ppc.py 2020-06-23-28Mar $country red
-	# python Fig_tau_posterior.py 2020-06-23-28Mar $country  -q
+	python Fig_tau_posterior.py 2020-05-26-Apr11 $country  -q
+	python Fig_tau_posterior.py 2020-06-27-Apr11-notau $country  -q
+	python Fig_tau_posterior.py 2020-06-25-Apr11-fixedtau $country  -q
 
 	# python Fig_ppc.py 2020-06-25-11Mar-fixedtau $country blue
-	python Fig_ppc.py 2020-06-27-11Mar-notau $country green
+	# python Fig_ppc.py 2020-06-27-Apr11-notau $country red
 
 	# python Fig_tau_posterior.py 2020-06-25-11Mar-fixedtau $country  -q
-	python Fig_tau_posterior.py 2020-06-27-11Mar-notau $country  -q
+	# python Fig_tau_posterior.py 2020-06-27-11Mar-notau $country  -q
 
 	# python ppc.py 2020-05-14-n1-notau-1M $country green >> no_tau_loss.txt
 	# python ppc.py 2020-05-15-n1-fixed-tau-1M $country red >> fixed_tau_loss.txt

### 2020-07-08 src/Table-WAIC.py
cleaner output filename

@@ -99,7 +99,7 @@ if __name__ == '__main__':
             df.loc[idx, i] = ['\\textbf{'+'{:.2f}'.format(x)+'}' for x in df.loc[idx, i]] 
 
     bold_all(df, df.columns)
-    df=df[df.columns.reindex(['No','Fixed','Free'])[0]]
+    df = df[df.columns.reindex(['No','Fixed','Free'])[0]]
 
     # for country, row in df.iterrows():
     #     if row['Free'] + 2 < min(row['Fixed'], row['No']):
@@ -109,6 +109,7 @@ if __name__ == '__main__':
     #     else:
     #         df.loc[country, 'Free'] ='{:.2f}'.format(row['Free'])
     
-    df.to_csv('../figures/Table-WAIC_tmp.csv', index='Country', float_format="%.2f")
-    print("Saved to ../figures/Table-WAIC_tmp.csv")
+    output_fname = "../figures/Table-WAIC.csv"
+    df.to_csv(output_fname, index='Country', float_format="%.2f")
+    print(output_fname)
     

### 2020-07-08 ms/ms.tex
update WAIC table formatting; added RMSE table

@@
 In all 11 of the European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
-% TODO RMSE reference? (Ilia)
-% TODO add table for RMSE (Ilia)
-% TODO consider dropping WAIC for RMSE (both)
 When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark.
@@
 \paragraph*{Early effective start of NPIs.}
-In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
+In some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
+The only significant early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\pm$1.05 days 95 \%CI; \autoref{fig:early}).
+This date occurs between the the first and last official NPI dates, which are Mar~12 and Mar~22 (\autoref{table:NPI_dates}).
+Therefore, when we refer to this case as "early", we mean that the effective date (Mar~19) occurs \emph{before} the official lockdown date (Mar~22), not that it occurs before all the NPIs.
+Interestingly, Germany has the longest duration between first and last NPIs except in Norway (10 and 12 days, respectively; see \autoref{table:NPI_dates}).Nevertheless, the credible intervals for the effective date in Germany is very narrow ($\pm$1.05 days 95 \%CI; \autoref{fig:early}), whereas they are very wide in Norway (XXX).
 
-Germany... %TODO (Yoav) \autoref{fig:early}
-
-Sweden and Denmark show early effective start of NPIs.
-However, for both countries, the casa data does not include a "peak", that is, the number of daily cases continues to grow during the evaluated dates.
+We also estimate early effective start of NPIs in Sweden and Denmark.
+However, in both countries, the casa data does not include a "peak", that is, the number of daily cases continues to grow during the evaluated dates.
 We suspect this increases the uncertainty in our inference.
@@
 
-% UPDATE
-As countries relieve lockdowns and ease restrictions and the number of cases increases, we expect similar delays and advances to occur: in some countries the population will behave as if restrictions were eased even before the official date, and in some countries the population will continue to self-restrict even after restrictions are officially removed.
+As different countries experiment with various interventions strategies, we expect similar delays and advances to occur: in some cases the population will be late to comply with are guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally declared.
 
@@
 % Table - WAIC %%% TODO regenerate table
-% generated with /python Table-WAIC.py/
+% generated with /python Table-WAIC.py 2020-05-26-Apr11/
 \begin{table}[h]
@@
 
-% Table - RMSE %%% TODO regenerate table
-% generated with /python Table-RMSE.py/
-%\begin{table}[h]
-%\centering
-%\pgfplotstabletypeset[
-%    col sep=comma,
-%    string type,
-%    every head row/.style={before row=\hline,after row=\hline},
-%    every last row/.style={after row=\hline},
-%    ]{../figures/Table-RMSE.csv}
-%\caption{
-%\textbf{Posterior RMSE of out-of-sample predictions with the different models.}
-%	Expected posterior predictive RMSE (root mean squared error) for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
-%	In all cases, the model with free parameter $\tau$ has the lowest RMSE.
-%	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions from Apr~12 to Apr~24 by sampling model parameters from the posterior distribution.
-%	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model.
-%}
-%\label{table:RMSE}
-%\end{table}
+% Table - RMSE 
+% generated with /python Table-RMSE.py 2020-05-26-Apr11/
+\begin{table}[h]
+\centering
+\pgfplotstabletypeset[
+    col sep=comma,
+    string type,
+    every head row/.style={before row=\hline,after row=\hline},
+    every last row/.style={after row=\hline},
+    ]{../figures/Table-RMSE.csv}
+\caption{
+\textbf{Posterior RMSE of out-of-sample predictions with the different models.}
+	Expected posterior predictive RMSE (root mean squared error) for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
+	In all cases, the model with free parameter $\tau$ has the lowest RMSE.
+	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions from Apr~12 to Apr~24 by sampling model parameters from the posterior distribution.
+	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model.
+}
+\label{table:RMSE}
+\end{table}
 

### 2020-07-08 src/Table-WAIC.py
fixed rounding digits in WAIC and RMSE

@@ -96,6 +96,7 @@ if __name__ == '__main__':
         minidxs = df.idxmin(axis=1)
         for i in columns:
             idx = i==minidxs
+            df.loc[~idx, i] = ['{:.2f}'.format(x) for x in df.loc[~idx, i]] 
             df.loc[idx, i] = ['\\textbf{'+'{:.2f}'.format(x)+'}' for x in df.loc[idx, i]] 
 
     bold_all(df, df.columns)

### 2020-07-08 src/Fig_NPI_dates.py
improve fname

@@ -48,5 +48,6 @@ if __name__ == '__main__':
 	ax.grid(True)
 	txt = ax.get_yticklabels()
 	sns.despine()
-	fig.savefig('../figures/Fig-NPI_dates.pdf', dpi=100, **savefig_bbox(*txt))
-	print("Saved NPI dates to ../figures/Fig-NPI_dates.pdf")
+	output_fname = "../figures/Fig-NPI_dates.pdf"
+	fig.savefig(output_fname, dpi=100, **savefig_bbox(*txt))
+	print(output_fname)

### 2020-07-08 ms/ms.tex
update PPC plots; add table titles for RMSE and WAIC

@@
 %%%% Fig - NPI dates %%%
+% generated with /python Fig_NPI_dates.py/
 \begin{figure}[h]
@@
 
-% Table - WAIC %%% TODO regenerate table
+% Table - WAIC 
 % generated with /python Table-WAIC.py 2020-05-26-Apr11/
@@
 \centering
+\caption*{\textbf{WAIC}}
 \pgfplotstabletypeset[
@@
 \centering
+\caption*{\textbf{RMSE}}
 \pgfplotstabletypeset[
@@
 	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model. 
+	Bold values emphasize cases in which the \emph{Free} model has the lowest RMSE.
 }
@@
 
-%%% Fig PPC %%% TODO UPDATE
+%%% Fig PPC
 % generated with ppc.sh
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Austria_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Austria_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Austria_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Austria_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Belgium_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Belgium_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Belgium_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Belgium_ppc_long.pdf}
     \end{subfigure}
@@
      \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Denmark_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Denmark_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Denmark_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Denmark_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/France_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/France_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/France_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/France_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Germany_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Germany_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Germany_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Germany_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Italy_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Italy_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Italy_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Italy_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Norway_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Norway_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Norway_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Norway_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Spain_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Spain_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Spain_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Spain_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Sweden_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Sweden_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Sweden_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Sweden_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Switzerland_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Switzerland_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Switzerland_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Switzerland_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/United_Kingdom_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/United_Kingdom_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/United_Kingdom_ppc.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/United_Kingdom_ppc_long.pdf}
     \end{subfigure}
-%    %%%  
-%	\subcaption{Wuhan, China}    
-%    \begin{subfigure}{0.45\textwidth}
-%        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Wuhan_ppc.pdf}
-%    \end{subfigure}
-%    ~
-%    \begin{subfigure}{0.45\textwidth}
-%        \includegraphics[width=\textwidth]{../figures/ppc/free/Wuhan_ppc.pdf}
-%    \end{subfigure}
-    %%%
     \caption{
-    \textbf{Figure S4. Posterior prediction plots.}
-    Markers represent data ($\vec{X}$): circles for data up to Apr~11, 2020, and stars after Apr~12; Apr~11 is represented using a vertical dashed line.
-Black line represents a smoothing of the data points using a Savitzky-Golay filter.
-Coloured lines represent posterior predictions from a model with fixed $\tau$ in red, and free $\tau$ in blue.
-The models were fitted with data up to Apr~11.
+    \textbf{Figure S4. Posterior prediction plots for 11 European countries.}
+    Circles and stars represent daily case data up to and after Apr~11, 2020, respectively, and the vertical dashed line represents Apr~11.
+    Black lines represents a smoothing of the data points using a Savitzky-Golay filter. 
+    Colored lines represent posterior predictions from a model with fixed $\tau$ -- blue -- and free $\tau$ -- green.
+Models were fitted with data up to Apr~11.
 The predictions are generated by drawing 1,000 parameters sets from the posterior distribution, and then generating a daily case count using the SEIR model in \autoref{eq:model}.
 Note the differences in the y-axis scale.
+    Posterior predictions with the free $\tau$ model are good for all countries except Denmark and Sweden, but poor for the fixed $\tau$ model.
+    The predictions of the model without $\tau$ (not shown) are even worse.
     }	

### 2020-07-08 ms/ms.tex
revised exact and early results

@@
 
-During February and March 2020, several countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules, to control the \covid pandemic caused by the \sars virus.
+During Feb-Apr 2020, many countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules, to control the \covid pandemic caused by the \sars virus.
 Overall, these interventions seem to have successfully reduced the spread of the pandemic.
-We hypothesise that the official and effective start date of such interventions can significantly differ, for example due to slow adoption by the population, or due to unpreparedness of the authorities and the public.
+We hypothesise that the official and effective start date of such interventions can significantly differ, for example due to slow adoption by the population, or because the authorities and the public are unprepared.
 We fit an SEIR model to case data from~12 countries to infer the effective start dates of interventions and contrast them with the official dates.
-We find both late and early effects of interventions.
-For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective date on Mar~16 ($\pm$0.47 days 95\% CI).
-In contrast, Spain announced a lockdown on Mar~14, but we infer an effective start date on Mar~8 ($\pm$ 1.08 days 95\% CI).
+We find mostly late, but also early effects of interventions.
+For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective date on Mar~17 ($\pm$2.99 days 95\% CI).
+In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\pm$ 1.05 days 95\% CI).
 We discuss potential causes and consequences of our results.
@@
 The only significant early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\pm$1.05 days 95 \%CI; \autoref{fig:early}).
-This date occurs between the the first and last official NPI dates, which are Mar~12 and Mar~22 (\autoref{table:NPI_dates}).
+This estimate falls between the first and last official NPI dates, Mar~12 and Mar~22 (\autoref{table:NPI_dates}).
 Therefore, when we refer to this case as "early", we mean that the effective date (Mar~19) occurs \emph{before} the official lockdown date (Mar~22), not that it occurs before all the NPIs.
-Interestingly, Germany has the longest duration between first and last NPIs except in Norway (10 and 12 days, respectively; see \autoref{table:NPI_dates}).Nevertheless, the credible intervals for the effective date in Germany is very narrow ($\pm$1.05 days 95 \%CI; \autoref{fig:early}), whereas they are very wide in Norway (XXX).
+Interestingly, Germany has the second longest duration between first and last NPIs after Norway (10 and 12 days respectively; see \autoref{table:NPI_dates}).
+Nevertheless, the credible interval for the effective date in Germany is narrow ($\pm$1.05 days 95 \%CI), whereas it is quite wide in Norway ($\pm$10.79 days; \autoref{fig:early}).
 
 We also estimate early effective start of NPIs in Sweden and Denmark.
-However, in both countries, the casa data does not include a "peak", that is, the number of daily cases continues to grow during the evaluated dates.
-We suspect this increases the uncertainty in our inference.
-Accordingly, WAIC of our model with a free $\tau$ parameter was higher (less fit) than the WAIC of a model with $\tau$ fixed at $\tau^*$ the official date.
+However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and the fit of the model with free $\tau$ wasn't significantly better than the fit of the model with $\tau$ fixed at the official date (\autoref{table:RMSE}, \autoref{table:WAIC}).
+In Sweden, the data does not "peak", i.e., the number of daily cases continues to grow up to Apr~11 (\autoref{fig:ppc}). 
+In Denmark, the opposite occurs: there are seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
+We suspect that these missing and additional "peaks" increase the uncertainty in our inference.
 
@@
 %%% Exact
-\paragraph*{Like a Swiss watch.}
-We find one case in which the official and effective dates match (and the credible interval is narrow): Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
+\paragraph*{Early effective start of NPIs.}
+We find one case in which the official and effective dates match "\emph{like a Swiss watch}", and the credible interval is narrow: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
 Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\pm$2.51 days 95\% CI, see \autoref{fig:early}).
@@
 \autoref{table:NPI_dates} and \autoref{fig:NPI_dates} for NPI dates).
-The uncertainty in the estimate of the effective start of NPI may be due to the long duration between the first and last NPI; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is quite narrow (\autoref{fig:early}).
+The uncertainty in the estimate of the effective start of NPIs may be due to the long duration between the first and last NPI; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is quite narrow (\autoref{fig:early}).
 

### 2020-07-09 ms/ms.tex
YMB comments

@@
 
+%TODO 
+% clearly mention that you estimate an aggregated effect for all measures and a single date for when the effect of the measures begins, even though in real life the measures were spread out in time and each one probably has a different effect.
+
 % Introduction
@@
 Here, we estimate the joint posterior distribution of the \emph{effective} start date of the NPIs $\tau$ and the transmission and reporting rates before and after $\tau$ from the entire data, rather than splitting the data at~$\tau$.
-We then estimate the marginal posterior probability of $\tau$ by marginalising the joint posterior, and estimate $\hat{\tau}$ as the posterior median.
+We then estimate $\hat{\tau}$ as the median of the marginal posterior distribution of $\tau$.
 
@@
 School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
-That is, the first and last official NPI dates are Mar~8 and Mar~11.
+That is, the first and last official NPI dates are Mar~5 and Mar~11.
 However, we estimate the effective date $\hat{\tau}$ at Mar~17 ($\pm$2.99 days 95\% CI ; \autoref{fig:late}).
@@
 The success of non-pharmaceutical interventions is assessed by health officials using various metrics, such as the decline in the growth rate of daily cases.
-These assessments are made a specific number of days after the intervention began, to accommodate for the expected serial interval~\citep{Banholzer2020} (i.e.
-time between successive cases in a chain of transmission), which is estimated at about 4-7 days~\citep{Gatto2020}.
-% Gatto2020 Table 1
+These assessments are made a specific number of days after the intervention began, to accommodate for the expected generation interval (i.e.
+average time between infection events in a chain of transmission), which is estimated at about 4-7 days~\citep{Bar-On2020}.
 
@@
 \paragraph*{SEIR model.} \label{sec:model}
-We model SARS\=/CoV\=/2 infection dynamics by following the number of susceptible $S$, exposed $E$, reported infected $I_r$, unreported infected $I_u$, and recovered $R$ individuals in a population of size $N$.
+We model \sars infection dynamics by following the number of susceptible $S$, exposed $E$, reported infected $I_r$, unreported infected $I_u$, and recovered $R$ individuals in a population of size $N$.
 This model distinguishes between reported and unreported infected individuals: the reported infected are those that have enough symptoms to eventually be tested and thus appear in daily case reports, to which we fit the model.
@@
 The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioural changes of both susceptible and infected individuals.
-Exposed individuals, after an average incubation period of $Z$ days, become reported infected with probability $\alpha_t$ or unreported infected with probability $(1-\alpha_t)$.
+Exposed individuals, after an average latent period of $Z$ days, become reported infected with probability $\alpha_t$ or unreported infected with probability $(1-\alpha_t)$.
 The reporting rate $0 < \alpha_t < 1$ may also change over time due to changes in human behaviour.
@@
 \section*{Acknowledgements}
-We thank Lilach Hadany and Oren Kolodny for discussions and comments.
+We thank Yinon M. Bar-On, Lilach Hadany and Oren Kolodny for discussions and comments.
 This work was supported in part by the Israel Science Foundation 3811/19 and 1399/17.
@@
 \bibliographystyle{agsm}
-%\bibliography{/Users/yoavram/Documents/library}
-\bibliography{ms}
+\bibliography{/Users/yoavram/Documents/library}
+%\bibliography{ms}
 

### 2020-07-09 ms/ms.tex
OU comments

@@
 
-%TODO 
-% clearly mention that you estimate an aggregated effect for all measures and a single date for when the effect of the measures begins, even though in real life the measures were spread out in time and each one probably has a different effect.
-
 % Introduction
@@
 Here, we hypothesise that there is a significant difference between the official start of NPIs and their effective adoption by the public and therefore their effect on infection dynamics.
-We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and a \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to infer the effective start date of NPIs from publicly available \covid case data in 12 geographical regions.
+We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) model and a \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to infer the effective start date of NPIs from publicly available \covid case data in 12 geographical regions.
 We compare these estimates to the official dates, and find that they include both late and early effects of NPIs on infection dynamics.
@@
 The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers).
-In contrast, the effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although this is significant only for Germany (i.e., zero is not included in 75\% CI).
-The confidence intervals for Sweden, Denmark, and Norway are especially wide, see text and \autoref{fig:early} for possible explanation.
+In contrast, the estimated effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although uncertainty is low only for Germany (i.e., zero is not included in 95\% CI).
+The credible intervals for Sweden, Denmark, and Norway are especially wide, see text and \autoref{fig:early} for possible explanation.
 	Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}.
 $\tau^*$ is the last NPI date (\autoref{table:NPI_dates}).
-Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e.
-interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.) \autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, 2020, rather then Apr~11.
+Thin and bold lines show 95\% and 75\% credible intervals, respectively.
+\autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, 2020, rather then Apr~11.
     }
@@
 We then estimate $\hat{\tau}$ as the median of the marginal posterior distribution of $\tau$.
+This is done under the simplifying assumption that all interventions start at a specific date, despite the reality that the durations between the first and last NPIs were between 4 and 12 days (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
 
@@
 transmission and reporting rates are constant).
-All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~23, and these predictions were then compared to the real case data. 
+All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~24, and these predictions were then compared to the real case data. 
 The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
@@
 
-We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions (10 of 12) the effective start of NPIs  significantly differs from the official date (\autoref{fig:tau-summary}), that is, the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$ (\autoref{fig:tau-summary}).
+We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPIs differs from the official date: in 10 of 12 countries the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$; (7 of 12 countries when considering a 95\% interval; \autoref{fig:tau-summary}).
 The exceptions are Switzerland (see below), and Norway.
@@
 Interestingly, the effect of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~24, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
-The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
+The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
 This may suggest correlations between effective starts of NPIs due to global or international events.
@@
 Black line shows the estimate $\hat{\tau}$.
-Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
+Shaded area shows a 95\% credible interval.
     }
@@
 In some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
-The only significant early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\pm$1.05 days 95 \%CI; \autoref{fig:early}).
+The only conclusive early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\pm$1.05 days 95 \%CI; \autoref{fig:early}).
 This estimate falls between the first and last official NPI dates, Mar~12 and Mar~22 (\autoref{table:NPI_dates}).
@@
 We also estimate early effective start of NPIs in Sweden and Denmark.
-However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and the fit of the model with free $\tau$ wasn't significantly better than the fit of the model with $\tau$ fixed at the official date (\autoref{table:RMSE}, \autoref{table:WAIC}).
-In Sweden, the data does not "peak", i.e., the number of daily cases continues to grow up to Apr~11 (\autoref{fig:ppc}). 
+However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and the fit of the model with free $\tau$ wasn't better than the fit of the model with $\tau$ fixed at the official date (\autoref{table:WAIC}).
+In Sweden, the data do not "peak", i.e., the number of daily cases continues to grow up to Apr~11 (\autoref{fig:ppc}). 
 In Denmark, the opposite occurs: there are seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
@@
 Black line shows the estimated $\hat{\tau}$.
-Shaded area shows a 95\% credible interval (interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$).
+Shaded area shows a 95\% credible interval.
 	}
@@
 %%% Exact
-\paragraph*{Early effective start of NPIs.}
+\paragraph*{Exact effective start of NPIs.}
 We find one case in which the official and effective dates match "\emph{like a Swiss watch}", and the credible interval is narrow: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
@@
 
-The estimated effective date in Norway is Mar~22 ($\pm$10.79 days 95\% CI), which doesn't significantly differ from the official date of Mar~24.
-However, the posterior distribution is very wide (\autoref{fig:early}): it covers the range between Mar~9, three days before the first NPI, and Mar~29, five days  after the last NPI (see 
+The estimated effective date in Norway is Mar~22, which is close to the official date of Mar~24.
+However, the posterior distribution is very wide ($\pm$10.79 days 95\% CI; \autoref{fig:early}): it covers the range between Mar~9, three days before the first NPI, and Mar~29, five days  after the last NPI (see
 \autoref{table:NPI_dates} and \autoref{fig:NPI_dates} for NPI dates).
-The uncertainty in the estimate of the effective start of NPIs may be due to the long duration between the first and last NPI; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is quite narrow (\autoref{fig:early}).
+The uncertainty in the estimate of the effective start of NPIs might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is quite narrow (\autoref{fig:early}).
 
@@
 
-However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
+However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a generation interval of 4-7 days and neglect the late or early population response to the NPI.
 This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~17 ($\hat{\tau}$).
@@
     Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
-    Shaded box illustrates a serial interval of seven days.
+    Shaded box illustrates a generation interval of seven days.
     \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
@@
 
-We have inferred the effective start date of NPIs in several geographical regions using an SEIR epidemiological model and an MCMC parameter estimation framework.
-We find examples of both late and early effect of NPIs (\autoref{fig:tau-summary}).
+We have inferred the effective start date of NPIs in several geographical regions using an SEIR model under an MCMC parameter estimation framework.
+We find examples of both late and early effective start of NPIs relative to the official date (\autoref{fig:tau-summary}).
 
-In most investigated countries we find a late effect of NPI on the transmission dynamics.
+In most investigated countries we find a late effective start of NPIs on the transmission dynamics.
 For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five days or more after the official date (\autoref{fig:late}).
-This difference might be explained, in some case, by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
-Late effect of NPIs may also be due to the time required by both the government and the citizens to organise for a lockdown, and for the new guidelines to be adopted by the population.
+This difference might be explained, in some cases, by low compliance: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, resulting in people leaving these provinces before the lockdown started~\citep{Gatto2020}.
+Late effect of NPIs may also be due to the time required by both the government and the citizens to prepare for a lockdown, and for new guidelines to be adopted by the population.
 
-In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}), although this is only significant for Germany (\autoref{fig:tau-summary}).
+In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}), although this is only conclusive in Germany (\autoref{fig:tau-summary}).
 An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
@@
 Indeed, the evidence supports a change in infection dynamics (i.e.
-a model with fixed or free $\tau$) even for Sweden~(\autoref{table:RMSE}, \autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
+a model with fixed or free $\tau$) even for Sweden~(\autoref{table:WAIC}, \autoref{table:RMSE}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
+%TODO mobility decreased
 
-Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of \covid\citep{Gatto2020}.
+Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic generation interval of \covid~\citep{Bar-On2020}.
 However, late and early effects such as we have inferred may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).
 
-We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in most regions (i.e.
+As expected, we have found that the evidence supports a model in which transmission rate changes at a specific time point $\tau$ over a model with a constant transmission rate (i.e.
 free or fixed model in \autoref{table:RMSE} and \autoref{table:WAIC}).
-It may be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
-Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
-However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as changing weather, news of new treatments, and international outbreaks.
+It may be interesting to check if the evidence favors a model with \emph{two} or more change-points, rather than one. 
+Multiple such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
+However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementations of NPIs~\citep{Flaxman2020}.
 
-As different countries experiment with various interventions strategies, we expect similar delays and advances to occur: in some cases the population will be late to comply with are guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally declared.
+As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with the guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally announced.
 
-\paragraph*{Conclusions.}
-We have inferred the effective start date of NPIs and found that they often differ from the official dates.
 Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to an epidemic.
@@
 We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions during Jan--Apr 2020.
-These incidence data summarise the number of individuals $X_t$ tested positive for \sars RNA (using RT-qPCR) at each day $t$.
+These incidence data summarise the number of individuals $X_t$ tested positive for \sars (using RT-qPCR) at each day $t$.
 Data for Wuhan, China retrieved from \citet{Pei2020}, data for 11 European countries retrieved from \citet{Flaxman2020}. 
@@
 
-We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$ fitted to data up to Apr~11, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}) from Apr~12 to Apr~24, and plot the predicted dynamics~(\autoref{fig:ppc}).
+We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$ fitted to data up to Apr~11, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}) up to Apr~24, and plot the predicted dynamics~(\autoref{fig:ppc}).
 Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the tightness of the predictions) are good ways to visually compare models.
@@
 All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
-Samples from the posterior distributions will be deposited on FigShare.
 } % small
@@
 	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
-WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular significance level for model comparison~\citep{Kass1995}.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.
 Bold values emphasize cases in which the \emph{Free} model has the lowest WAIC.
@@
 	In all cases, the model with free parameter $\tau$ has the lowest RMSE.
-	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions from Apr~12 to Apr~24 by sampling model parameters from the posterior distribution.
+	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions up to Apr~24 by sampling model parameters from the posterior distribution.
 	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model. 
@@
 $\tau^*$ is the last NPI date (\autoref{table:NPI_dates}).
-Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e.
-interval in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
+Thin and bold lines show 95\% and 75\% credible intervals, respectively.
     }
@@
 Note the differences in the y-axis scale.
-    Posterior predictions with the free $\tau$ model are good for all countries except Denmark and Sweden, but poor for the fixed $\tau$ model.
+    Posterior predictions with the free $\tau$ model predict the out-of-sample data well for all countries except Denmark and Sweden, but poorly for the fixed $\tau$ model.
     The predictions of the model without $\tau$ (not shown) are even worse.

### 2020-07-09 ms/ms.tex
OU comments
revised for submission

@@
 
-Here, we estimate the joint posterior distribution of the \emph{effective} start date of the NPIs $\tau$ and the transmission and reporting rates before and after $\tau$ from the entire data, rather than splitting the data at~$\tau$.
+Here, we estimate the joint posterior distribution of the effective start date of NPIs, $\tau$, and the transmission and reporting rates before and after $\tau$ from the entire data, rather than splitting the data at~$\tau$.
 We then estimate $\hat{\tau}$ as the median of the marginal posterior distribution of $\tau$.
@@
 In all 11 of the European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
-When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark.
-Notably, the data for Sweden and Denmark does not have a peak during the evaluated dates, possibly leading to wide 95\% confidence intervals on $\hat{tau}$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).
+When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark (\autoref{table:WAIC}).
+Notably, the data for Sweden and Denmark do not have a single "peak" during the evaluated dates, possibly leading to wide 95\% credible intervals on $\hat{\tau}$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).
 
-We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPIs differs from the official date: in 10 of 12 countries the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$; (7 of 12 countries when considering a 95\% interval; \autoref{fig:tau-summary}).
-The exceptions are Switzerland (see below), and Norway.
+We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPIs differs from the official date: in 10 of 12 countries the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$ (7 of 12 countries when considering a 95\% interval; \autoref{fig:tau-summary}).
+The exceptions are Switzerland and Norway (see below).
 The latter also has a wide 95\% credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
+
 In the following, we describe our findings in more detail.
@@
 \paragraph*{Late effective start of NPIs.}
-In half of the examined regions, we estimate that the effective start of NPIs $\hat{\tau}$ is significantly later than the official date $\tau^*$.
+In half of the examined regions, we estimate that the effective start of NPIs $\hat{\tau}$ is later than the official date $\tau^*$.
 
@@
 That is, the first and last official NPI dates are Mar~5 and Mar~11.
-However, we estimate the effective date $\hat{\tau}$ at Mar~17 ($\pm$2.99 days 95\% CI ; \autoref{fig:late}).
+However, we estimate the effective date $\hat{\tau}$ six days after the lockdown, at Mar~17 ($\pm$2.99 days 95\% CI; \autoref{fig:late}).
 
@@
 
-In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score: 2--1) with a crowd of 50,965 in Seville.
+In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score 2--1) with a crowd of 50,965 in Seville.
 A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~24 ($\pm$2.96 days 95 \%CI), rather than Mar~14 (\autoref{fig:late}).
+Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ ten days later, at Mar~24 ($\pm$2.96 days 95 \%CI), rather than Mar~14 (\autoref{fig:late}).
 
-Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ on Mar~24 ($\pm$2.29 days 95\% CI, \autoref{fig:late}).
-This is about a week later then the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
+Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ at Mar~24 ($\pm$2.29 days 95\% CI, \autoref{fig:late}).
+This is a week later than the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
 
-Interestingly, the effect of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~24, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
+Interestingly, the effective start of of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~24, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
 The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
@@
 
-We also estimate early effective start of NPIs in Sweden and Denmark.
-However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and the fit of the model with free $\tau$ wasn't better than the fit of the model with $\tau$ fixed at the official date (\autoref{table:WAIC}).
+We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
+However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free $\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
+
 In Sweden, the data do not "peak", i.e., the number of daily cases continues to grow up to Apr~11 (\autoref{fig:ppc}). 
@@
 We suspect that these missing and additional "peaks" increase the uncertainty in our inference.
+
+The estimated effective date in Norway is Mar~22, which is two days earlier than the official date of Mar~24.
+However, the posterior distribution is very wide ($\pm$10.79 days 95\% CI; \autoref{fig:early}): it covers the range between Mar~9, three days before the first NPI, and Mar~29, five days  after the last NPI (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
+The uncertainty in the estimate of the effective start of NPIs might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is quite narrow (\autoref{fig:early}).
 
@@
 \paragraph*{Exact effective start of NPIs.}
-We find one case in which the official and effective dates match "\emph{like a Swiss watch}", and the credible interval is narrow: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
+We find one case in which the official and effective dates match "\emph{like a Swiss watch}" and the credible interval is narrow.
+Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
 Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\pm$2.51 days 95\% CI, see \autoref{fig:early}).
 It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
-
-The estimated effective date in Norway is Mar~22, which is close to the official date of Mar~24.
-However, the posterior distribution is very wide ($\pm$10.79 days 95\% CI; \autoref{fig:early}): it covers the range between Mar~9, three days before the first NPI, and Mar~29, five days  after the last NPI (see
-\autoref{table:NPI_dates} and \autoref{fig:NPI_dates} for NPI dates).
-The uncertainty in the estimate of the effective start of NPIs might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is quite narrow (\autoref{fig:early}).
 
@@
 This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~17 ($\hat{\tau}$).
-If health officials assumed the dynamics to immediately change at $\tau^*$, they will have expected the number of cases be within the red lines (posterior predictions assuming $\tau=\tau^*$).
-This would have lead to a significant underestimation, which might have been interpreted by as ineffectiveness of the NPI, leading to further escalations.
+If health officials assume that the dynamics change exactly at $\tau^*$, they would expect the number of cases to be within the red lines (posterior predictions assuming $\tau=\tau^*$).
+This would lead to a significant underestimation, which might be interpreted as ineffectiveness of the NPI, leading to further escalations.
 However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data (stars).
@@
     \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
-    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
+    Real number of daily cases in Italy in black (markers: data; line: time moving average). 
     Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
@@
 
-In most investigated countries we find a late effective start of NPIs on the transmission dynamics.
-For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five days or more after the official date (\autoref{fig:late}).
+In most investigated countries we find a late effective start of NPIs.
+For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five or more days after the official date (\autoref{fig:late}).
 This difference might be explained, in some cases, by low compliance: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, resulting in people leaving these provinces before the lockdown started~\citep{Gatto2020}.
@@
 
-In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}), although this is only conclusive in Germany (\autoref{fig:tau-summary}).
+In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented, although this is only conclusive in Germany (\autoref{fig:early}).
 An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
@@
 a model with fixed or free $\tau$) even for Sweden~(\autoref{table:WAIC}, \autoref{table:RMSE}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
-%TODO mobility decreased
 
-Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic generation interval of \covid~\citep{Bar-On2020}.
+As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with the guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally announced.
+
+Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic generation interval of \covid\citep{Bar-On2020}.
 However, late and early effects such as we have inferred may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).
 
-As expected, we have found that the evidence supports a model in which transmission rate changes at a specific time point $\tau$ over a model with a constant transmission rate (i.e.
-free or fixed model in \autoref{table:RMSE} and \autoref{table:WAIC}).
+As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (\autoref{table:WAIC} and \autoref{table:RMSE}).
 It may be interesting to check if the evidence favors a model with \emph{two} or more change-points, rather than one. 
-Multiple such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
+Multiple change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
 However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementations of NPIs~\citep{Flaxman2020}.
-
-As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with the guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally announced.
 
@@
 We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions during Jan--Apr 2020.
-These incidence data summarise the number of individuals $X_t$ tested positive for \sars (using RT-qPCR) at each day $t$.
-Data for Wuhan, China retrieved from \citet{Pei2020}, data for 11 European countries retrieved from \citet{Flaxman2020}. 
+These incidence data summarise the number of individuals $X_t$ tested positive for \sars (using RT-qPCR tests) on each day $t$.
+Data for Wuhan, China retrieved from \citet{Pei2020}. Data for 11 European countries retrieved from \citet{Flaxman2020}. 
 Where there were multiple sequences of days with zero confirmed cases (e.g.
@@
 We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates, and therefore chose $\mathit{Uniform}(1,5)$ as the prior for $\Delta t$.
+We also tried to estimate the value of $p_t$ before and after $\tau$, instead of keeping it fixed at $1/9$ and $1/6$.
+The model with fixed values was supported by the evidence (lower WAIC, see below) in 9 of 12 countries.
+Moreover, the estimates for Wuhan, China were $1/9$ and $1/6$, as in \citet{Li2020}.
+\\
 Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11.
@@
 Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the tightness of the predictions) are good ways to visually compare models.
-We also compute the expected posterior root mean squared error of these predictions~(\autoref{table:RMSE}).
+We also compute the expected posterior RMSE (root mean squared error) of these predictions~(\autoref{table:RMSE}).
 
@@
 \section*{Acknowledgements}
-We thank Yinon M. Bar-On, Lilach Hadany and Oren Kolodny for discussions and comments.
-This work was supported in part by the Israel Science Foundation 3811/19 and 1399/17.
+We thank Yinon M. Bar-On, Lilach Hadany, and Oren Kolodny for discussions and comments.
+This work was supported in part by the Israel Science Foundation (3811/19, 552/19, and 1399/17).
 }

### 2020-07-09 ms/ms.tex
added arrows to ppc

@@
     \textbf{Figure S4. Posterior prediction plots for 11 European countries.}
-    Circles and stars represent daily case data up to and after Apr~11, 2020, respectively, and the vertical dashed line represents Apr~11.
-    Black lines represents a smoothing of the data points using a Savitzky-Golay filter. 
-    Colored lines represent posterior predictions from a model with fixed $\tau$ -- blue -- and free $\tau$ -- green.
+    The vertical dashed line represents Apr~11, 2020.
+    Circles and stars represent daily case data up to and after Apr~11, respectively
+    Black and white arrows denote the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs, respectively.
+    Black lines represent a smoothing of the data points using a Savitzky-Golay filter. 
+    Coloured lines represent posterior predictions from a model with fixed $\tau$ (blue) and free $\tau$ (green).
 Models were fitted with data up to Apr~11.
-The predictions are generated by drawing 1,000 parameters sets from the posterior distribution, and then generating a daily case count using the SEIR model in \autoref{eq:model}.
+The predictions are generated by drawing 1,000 parameters sets from the posterior distribution, and then generating a daily case count using the SEIR model up to Apr~24.
 Note the differences in the y-axis scale.

### 2020-07-12 src/Re.py
compute Re pre and post NPI

@@ -1,0 +1,85 @@
+#!/usr/bin/env python
+# coding: utf-8
+import sys
+import os
+from datetime import datetime, timedelta
+import urllib
+
+import matplotlib as mpl
+# mpl.use("Agg")
+import matplotlib.pyplot as plt
+import numpy as np
+import scipy.stats
+from scipy.integrate import odeint
+import scipy.signal
+import pandas as pd
+import seaborn as sns
+sns.set_context('paper', font_scale=1.3)
+red, blue, green = sns.color_palette('Set1', 3)
+colors = {'red':red, 'blue':blue, 'green':green}
+
+from click_spinner import spinner
+
+from inference import get_last_NPI_date
+from inference import get_first_NPI_date
+from inference import params_bounds
+from inference import get_model_class
+from inference import find_start_day
+from model.normal_prior_model import NormalPriorModel
+from model.fixed_tau_model import FixedTauModel
+from sklearn.metrics import mean_squared_error
+
+
+def load_chain(job_id=None, fname=None, burn_fraction=0.6):
+	with spinner():
+		if fname is None:
+			fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
+		inference_data = np.load(fname)
+		chain = inference_data['chain']
+		var_names = inference_data['var_names']
+		nsteps, ndim, N, Td1, Td2, model_type = inference_data['params']
+		X = inference_data['incidences']
+		start_date = inference_data['start_date']
+		logliks = inference_data['logliks']
+		# print("Loaded {} with parameters:".format(fname))
+		# print(var_names)
+		nchains, nsteps, ndim = chain.shape
+		nburn = int(nsteps * burn_fraction)
+		chain = chain[:, nburn:, :]
+		chain = chain.reshape((-1, ndim))
+		logliks = logliks.reshape((nchains, nsteps))
+		logliks = logliks[:, nburn:].ravel()
+		return chain, logliks, Td1, Td2, model_type, X, start_date, N
+
+def posterior_Re(chain, nreps):
+	# params: ['Z', 'D', 'μ', 'β', 'α1', 'λ', 'α2', ...]
+	θ = chain[np.random.choice(chain.shape[0], nreps)]
[…]

### 2020-07-12 src/Re.sh

@@ -1,0 +1,8 @@
+for country in Austria Belgium Denmark France Germany Italy Norway Spain Sweden Switzerland United_Kingdom Wuhan
+do
+	python Re.py 2020-05-26-Apr11 $country
+	python Re.py 2020-06-25-Apr11-fixedtau $country
+done
+
+
+

### 2020-07-12 ms/ms.tex
added Re figure

@@
 %%% Assessment
-\paragraph*{Consequences of late and early effect of NPIs on real-time assessment.}
+\paragraph*{Consequences of late and early effective start on assessment of NPIs impact.}
 
-The success of non-pharmaceutical interventions is assessed by health officials using various metrics, such as the decline in the growth rate of daily cases.
-These assessments are made a specific number of days after the intervention began, to accommodate for the expected generation interval (i.e.
-average time between infection events in a chain of transmission), which is estimated at about 4-7 days~\citep{Bar-On2020}.
+The reproductive number is XXX
+We can infer the reproductive numbers before ($R_1$) and after ($R_2$) the implementation of NPIs from the estimated model parameters \citep{Li2020},
+\begin{equation} \label{eq:Re}
+\begin{aligned}
+R_1 &= \alpha_1 \beta D + (1-\alpha_1) \mu \beta D, \\
+R_2 &= \alpha_2 \lambda \beta D + (1-\alpha_2) \mu \lambda \beta D.
+\end{aligned}
+\end{equation}
 
-However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a generation interval of 4-7 days and neglect the late or early population response to the NPI.
+The impact of NPIs can then be measured by the relative reduction in the reproductive number \citep{Flaxman2020}, $\frac{R_1 - R_2}{R1}$. 
+We compared the relative reduction in the reproductive number estimated with either the fixed $\tau$ model or the free $\tau$ model.
+That is, the y-axis shows the estimate assuming that NPIs started at the official date $\tau^*$, whereas the x-axis shows the estimate when inferring the effective start of the NPIs from the data.
+\autoref{fig:Re} demonstrates that estimates from the fixed $\tau$ model are consistently lower than estimated from a free $\tau$ model, except for Sweden and Denmark, in which the inference uncertainty in is high anyway.
+This suggests that if health officials and researchers assume that the impact of NPIs starts at the official date, then they are likely to under-estimate their impact. 
+This might be interpreted as ineffectiveness of the NPI, leading to further escalations.
+
+Health officials might also assess the impact of NPIs by comparing their expectations (e.g. model predictions) to data at a specific number of days after the intervention began. 
+However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate such assessments.
 This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~17 ($\hat{\tau}$).
@@
 
+%%% Fig - Re
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=\textwidth]{../figures/Fig_Re.pdf}
+    \caption{
+    \textbf{Impact of NPIs is under-estimated when assuming they start at the official date.}
+    Shown are estimates of the relative reductions in reproductive numbers, which are measure of the impact NPIs have on the disease transmission.
+     The y-axis shows the estimate when assuming the start of NPIs is fixed at the official date (fixed $\tau$); the x-axis show the estimate when inferring the start of NPIs from the data (free $\tau$).
+    The relative reductions in $R_e$ are consistently lower for the fixed $\tau$ model.
+    The reproduction numbers are $R_{1}=\alpha_1 \beta D + (1-\alpha_1) \mu \beta D$ prior to start of NPI and $R_2=\alpha_2 \lambda \beta D + (1-\alpha_2) \mu \lambda \beta D$ after start of NPI.
+The relative reduction is $(R{1}-R{2}) / R{1}$.
+    } 
+    \label{fig:Re}
+\end{figure}
+
 %%% Fig - Prediction
+
 \begin{figure}[h]
@@
 
-As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with the guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally announced.
-
-Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic generation interval of \covid\citep{Bar-On2020}.
-However, late and early effects such as we have inferred may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).
-
 As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (\autoref{table:WAIC} and \autoref{table:RMSE}).
@@
 However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementations of NPIs~\citep{Flaxman2020}.
+
+As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with the guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally announced.
+Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic generation interval of \covid\citep{Bar-On2020}.
+However, late and early effects such as we have inferred may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).
 

### 2020-07-12 src/Fig_Re.py
generate Fig_Re

@@ -1,0 +1,74 @@
+#!/usr/bin/env python
+# coding: utf-8
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+sns.set_context('paper', font_scale=1.3)
+red, blue, green = sns.color_palette('Set1', 3)
+import pycountry # https://pypi.org/project/pycountry/
+
+import os
+from glob import glob
+from datetime import datetime, timedelta
+
+from rakott.mpl import savefig_bbox, fig_panel_labels
+
+colors = sns.color_palette('Paired')
+
+
+if __name__ == '__main__':
+    job_id_free = '2020-05-26-Apr11'
+    job_id_fixed = '2020-06-25-Apr11-fixedtau'
+
+    dfs = []
+    countries = ['Austria', 'Belgium', 'Denmark', 'France', 'Germany', 'Italy', 'Norway', 'Spain', 'Sweden', 'Switzerland', 'United Kingdom']    
+    for country in countries:   
+        fname = r'../output/{}/figures/Re_{}.csv'.format(job_id_free, country.replace(' ', '_')) 
+        df = pd.read_csv(fname)        
+        df['country'] = country
+        df['model'] = 'free'
+        dfs.append(df)
+        fname = r'../output/{}/figures/Re_{}.csv'.format(job_id_fixed, country.replace(' ', '_')) 
+        df = pd.read_csv(fname)
+        df['country'] = country
+        df['model'] = 'fixed'
+        dfs.append(df)
+
+    df = pd.concat(dfs)
+    grp = df.groupby(['country', 'model'])
+    agg = grp.agg(
+        low=pd.NamedAgg(column="rel_reduc_Re", aggfunc=lambda x: np.percentile(x, 25)),
+        median=pd.NamedAgg(column="rel_reduc_Re", aggfunc=np.median),
+        high=pd.NamedAgg(column="rel_reduc_Re", aggfunc=lambda x: np.percentile(x, 75))
+    ).reset_index()
+    df_free = agg[agg['model'] == 'free']
+    df_fixed = agg[agg['model'] == 'fixed']
+    countries = agg['country'].unique()
+
+    fig, ax = plt.subplots(1, 1, figsize=(8, 8*2/3), sharex=True, sharey=True)
+    ax.errorbar(
+        df_free['median'], 
+        df_fixed['median'], 
[…]

### 2020-07-12 ms/ms.tex
revised and highlighted changes

--- 2020-07-09 ms/ms.tex
@@
 In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\pm$ 1.05 days 95\% CI).
-We discuss potential causes and consequences of our results.
-
+\hl{
+We demonstrate that differences between the official and effective start of NPIs can distort conclusions about their impact, and discuss potential causes and consequences of our results.
+}
 \end{abstract}
@@
 We compare these estimates to the official dates, and find that they include both late and early effects of NPIs on infection dynamics.
-We conclude by demonstrating how differences between the official and effective start of NPIs can confound assessments of the effectiveness of the NPIs in a simple epidemic control framework.
+\hl{We conclude by demonstrating how differences between the official and effective start of NPIs can confound assessments of their impact.}
 
@@
 %%% Assessment
-\paragraph*{Consequences of late and early effect of NPIs on real-time assessment.}
+\paragraph*{\hl{Consequences of late and early effective start on assessment of NPIs impact.}}
 
-The success of non-pharmaceutical interventions is assessed by health officials using various metrics, such as the decline in the growth rate of daily cases.
-These assessments are made a specific number of days after the intervention began, to accommodate for the expected generation interval (i.e.
-average time between infection events in a chain of transmission), which is estimated at about 4-7 days~\citep{Bar-On2020}.
+The \emph{effective reproduction number} $R$ is the number of secondary
+cases generated by an infectious case after an epidemic is already underway \citep{Bar-On2020}.
+We infer the effective reproduction numbers before and after the implementation of NPIs from the estimated model parameters (\autoref{eq:Re}).
+The impact of NPIs can then be measured by the relative reduction in the reproduction number \citep{Flaxman2020}.
 
-However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a generation interval of 4-7 days and neglect the late or early population response to the NPI.
+We compared the estimated NPIs impact with either the fixed~$\tau$ model or the free~$\tau$ model.
+That is, the estimated impact assuming that NPIs started at the official date $\tau^*$ versus the estimate when inferring the effective start of the NPIs from the data.
+\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model are consistently lower than estimates from a free~$\tau$ model, except for Sweden and Denmark, in which the inference uncertainty in is high.
+This suggests that if health officials and researchers assume that NPIs start at their official date, then their impact is likely to be under-estimated.
+This might be interpreted as ineffectiveness of the NPIs, leading to further escalations.
+
+Health officials might also assess the impact of NPIs by comparing model expectations to actual number of daily cases at a specific number of days after the intervention began. 
+However, a significant difference between the official and effective start of interventions can invalidate such assessments.
 This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~17 ($\hat{\tau}$).
@@
 
-%%% Fig - Prediction
+%%% Fig - Re
+% generated with
+% > source ./Re.sh
+% > python Fig_Re.py
 \begin{figure}[h]
     \centering
-	\includegraphics[width=0.4\textwidth]{../figures/Fig-prediction.pdf}
+	\includegraphics[width=\textwidth]{../figures/Fig_Re.pdf}
     \caption{
-    \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
+    \textbf{\hl{Impact of NPIs is under-estimated when assuming they start at the official date.}}
+    Shown are estimates of the relative reductions in $R$ (the effective reproduction number), which measures the impact of NPIs on disease transmission.
+     The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed $\tau$); the x-axis shows estimates when inferring the effective start of NPIs from the data (free $\tau$).
+The dashed line shows a 1-to-1 correspondence.
+     Markers and bars denote the posterior median and 50\% credible intervals.
+    The relative reductions in $R$ are consistently lower for the fixed $\tau$ model (below the line), except for Sweden and Denmark in which uncertainty is high.
+    } 
+    \label{fig:Re}
+\end{figure}
+
+%%% Fig - Prediction
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.5\textwidth]{../figures/Fig-prediction.pdf}
+    \caption{
+    \textbf{\hl{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}}
     Real number of daily cases in Italy in black (markers: data; line: time moving average).
@@
 However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementations of NPIs~\citep{Flaxman2020}.
+
+\hl{As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with the guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally announced.}
+Attempts to asses the impact of NPIs~\citep{Banholzer2020,Flaxman2020} \hl{generally assume they start at their official date, and that a significant change in the dynamics can be observed roughly seven-days after the start of NPIs} (due to the characteristic generation interval of \covid\citep{Bar-On2020}).
+\hl{However, late and early effective start of NPIs, such as we have inferred, may bias these assessments and lead to wrong conclusions about the impact of NPIs} (\autoref{fig:Re}, \autoref{fig:prediction}).
 
@@
 The values for $p_t$ follow \citet{Li2020}, who estimated the average time between infection and reporting in Wuhan, China, at 9 days before the start of NPIs and 6 days after start of NPIs.
+
+Following \citet{Li2020}, the effective reproduction numbers before and after the start of NPIs are
+\begin{equation} \label{eq:Re}
+\begin{aligned}
+R_1 &= \alpha_1 \beta D + (1-\alpha_1) \mu \beta D, \\
+R_2 &= \alpha_2 \lambda \beta D + (1-\alpha_2) \mu \lambda \beta D.
+\end{aligned}
+\end{equation}
+The relative reduction in the effective reproduction number due to NPIs is then $\frac{R_1 - R_2}{R1}$.
 

### 2020-07-13 ms/ms.tex
revised and removed highlights

@@
 In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\pm$ 1.05 days 95\% CI).
-\hl{
 We demonstrate that differences between the official and effective start of NPIs can distort conclusions about their impact, and discuss potential causes and consequences of our results.
-}
 \end{abstract}
@@
 We compare these estimates to the official dates, and find that they include both late and early effects of NPIs on infection dynamics.
-\hl{We conclude by demonstrating how differences between the official and effective start of NPIs can confound assessments of their impact.}
+We conclude by demonstrating how differences between the official and effective start of NPIs can confound assessments of their impacts.
 
@@
 %%% Assessment
-\paragraph*{\hl{Consequences of late and early effective start on assessment of NPIs impact.}}
+\paragraph*{Assessment of impact of NPIs.}
 
@@
 cases generated by an infectious case after an epidemic is already underway \citep{Bar-On2020}.
-We infer the effective reproduction numbers before and after the implementation of NPIs from the estimated model parameters (\autoref{eq:Re}).
-The impact of NPIs can then be measured by the relative reduction in the reproduction number \citep{Flaxman2020}.
+We infer the effective reproduction numbers before and after the implementation of NPIs from model parameters (\autoref{eq:Re}).
+We then estimate the impact of NPIs as the relative reduction in the reproduction number \citep{Flaxman2020}. 
+We compare the impacts estimated using the fixed~$\tau$ model and the free~$\tau$ model.
+That is, we compare the impact estimate assuming that NPIs started at the official date $\tau^*$, versus the estimate when inferring the effective start of NPIs from the data.
+\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model are consistently lower than estimates from the free~$\tau$ model, except in Sweden and Denmark, in which estimation uncertainty is high.
 
-We compared the estimated NPIs impact with either the fixed~$\tau$ model or the free~$\tau$ model.
-That is, the estimated impact assuming that NPIs started at the official date $\tau^*$ versus the estimate when inferring the effective start of the NPIs from the data.
-\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model are consistently lower than estimates from a free~$\tau$ model, except for Sweden and Denmark, in which the inference uncertainty in is high.
-This suggests that if health officials and researchers assume that NPIs start at their official date, then their impact is likely to be under-estimated.
-This might be interpreted as ineffectiveness of the NPIs, leading to further escalations.
+These results suggest that the impact of past NPIs by health officials and researchers is likely to be under-estimated if they assume NPIs effectively start at their official dates.
+The estimated impact can then be interpreted as ineffectiveness of the NPIs, leading to further escalations.
 
@@
 If health officials assume that the dynamics change exactly at $\tau^*$, they would expect the number of cases to be within the red lines (posterior predictions assuming $\tau=\tau^*$).
-This would lead to a significant underestimation, which might be interpreted as ineffectiveness of the NPI, leading to further escalations.
-However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data (stars).
+This would lead to a significant under-estimation, which might be interpreted as ineffectiveness of the NPI, leading to further escalations.
+However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which are a good fit to the real data (stars).
 
@@
     \caption{
-    \textbf{\hl{Impact of NPIs is under-estimated when assuming they start at the official date.}}
+    \textbf{Impact of NPIs is under-estimated when assuming they start at the official date.}
     Shown are estimates of the relative reductions in $R$ (the effective reproduction number), which measures the impact of NPIs on disease transmission.
      The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed $\tau$); the x-axis shows estimates when inferring the effective start of NPIs from the data (free $\tau$).
-The dashed line shows a 1-to-1 correspondence.
+The dashed line shows a one-to-one correspondence.
      Markers and bars denote the posterior median and 50\% credible intervals.
-    The relative reductions in $R$ are consistently lower for the fixed $\tau$ model (below the line), except for Sweden and Denmark in which uncertainty is high.
+    The relative reductions in $R$ are consistently lower for the fixed $\tau$ model (below the dashed line), except in Sweden and Denmark in which uncertainty is high.
     } 
@@
     \caption{
-    \textbf{\hl{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}}
+    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
     Real number of daily cases in Italy in black (markers: data; line: time moving average).
@@
     Shaded box illustrates a generation interval of seven days.
-    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
-    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
+    \textbf{(A)} Using the official date $\tau^*$ for the effective start of the NPI, the model under-estimates the number of cases seven days after the start of the NPI.
+    \textbf{(B)} Using the estimated date $\hat{\tau}$ for the effective start of the NPI, the model precisely estimates the number of cases seven days after the start of the NPI.
     Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
@@
 
-\hl{As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with the guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally announced.}
-Attempts to asses the impact of NPIs~\citep{Banholzer2020,Flaxman2020} \hl{generally assume they start at their official date, and that a significant change in the dynamics can be observed roughly seven-days after the start of NPIs} (due to the characteristic generation interval of \covid\citep{Bar-On2020}).
-\hl{However, late and early effective start of NPIs, such as we have inferred, may bias these assessments and lead to wrong conclusions about the impact of NPIs} (\autoref{fig:Re}, \autoref{fig:prediction}).
+As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with new guidelines, whereas in other cases the population will adopt either restrictions or relaxations even before they are formally announced.
+Attempts to asses the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date, and that a significant change in the dynamics can be observed roughly seven days after the start of NPIs (due to the characteristic generation interval of \covid\citep{Bar-On2020}).
+However, late and early effective start of NPIs, such as we have inferred, can bias these assessments and lead to wrong conclusions about the impact of NPIs (\autoref{fig:Re}, \autoref{fig:prediction}).
 
@@
 \end{equation}
-The relative reduction in the effective reproduction number due to NPIs is then $\frac{R_1 - R_2}{R1}$.
+The relative reduction in the effective reproduction number due to NPIs is $\frac{R_1 - R_2}{R1}$.
 

### 2020-08-05 ms/ms.tex
sig statement; orcid

@@
 \affil[3]{Porter School of the Environment and Earth Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
-\affil[a]{Corresponding author: yoav@yoavram.com}
+\affil[a]{Corresponding author: yoav@yoavram.com, ORCID 0000-0002-9653-4458}
 
@@
 \maketitle
+
+Keywords: SEIR, COVID-19, public health, pandemic, infectious disease
+
+Classification: Biological Sciences / Population Biology; Social Sciences / Environmental Sciences
+
+%\pagebreak
 
@@
 We demonstrate that differences between the official and effective start of NPIs can distort conclusions about their impact, and discuss potential causes and consequences of our results.
+\end{abstract}
+
+\renewcommand{\abstractname}{Significance statement}
+\begin{abstract}
+We have inferred the effective start date of non-pharmaceutical interventions, like lockdowns, during the initial \covid outbreak by fitting an epidemiological model to daily case data in 12 countries.
+Our results show that the effective start of interventions in many cases occurs 3-10 days later then the official day.
+Moreover, we demonstrate that such delays can lead to under-estimation of the intervention impact, and thus cause decision makers to implement stricter interventions.
+We suggest that a late effective start of interventions may be the result of low compliance to authority or conformity to social norms.
+Our study therefore highlights the importance of further research into the role of human behavior and social organization on the current and future epidemics.
 \end{abstract}

### 2020-08-26 ms/ms.tex
prep for ploscompbiol

--- 2020-07-13 ms/ms.tex
@@
 \maketitle
+
+Keywords: SEIR, COVID-19, public health, pandemic, infectious disease, NPI
+
+%Classification: Biological Sciences / Population Biology; Social Sciences / Environmental Sciences
+
+%\pagebreak
 
@@
 \end{abstract}
+
+%\renewcommand{\abstractname}{Significance statement}
+%\begin{abstract}
+%We have inferred the effective start date of non-pharmaceutical interventions, like lockdowns, during the initial \covid outbreak by fitting an epidemiological model to daily case data in 12 countries.
+Our results show that the effective start of interventions in many cases occurs 3-10 days later then the official day.
+Moreover, we demonstrate that such delays can lead to under-estimation of the intervention impact, and thus cause decision makers to implement stricter interventions.
+We suggest that a late effective start of interventions may be the result of low compliance to authority or conformity to social norms.
+Our study therefore highlights the importance of further research into the role of human behavior and social organization on the current and future epidemics.
+%\end{abstract}
 
@@
 Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to an epidemic.
-Therefore, we emphasize the need to further study variability in compliance and behavior over both time and space.
+Therefore, we emphasize the need to further study heterogeneity in compliance and behavior over both time and space.
 This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Arthur2020,Fenichela2011,Walters2013}.
@@
 We thank Yinon M. Bar-On, Lilach Hadany, and Oren Kolodny for discussions and comments.
-This work was supported in part by the Israel Science Foundation (3811/19, 552/19, and 1399/17).
+This work was supported in part by the Israel Science Foundation (3811/19 and 552/19, YR).
 }

### 2021-01-14 ms/ms.tex
declarations section for BMC and todos

@@
 \pagebreak
+
+%TODO HDI
 
@@
 \end{equation}
-The relative reduction in the effective reproduction number due to NPIs is $\frac{R_1 - R_2}{R1}$.
+The relative reduction in the effective reproduction number due to NPIs is $\frac{R_1 - R_2}{R_1}$.
 
@@
 \Delta t & \sim \mathit{Uniform}(1, 5) \\
-\tau &\sim \mathit{TruncatedNormal}\Big(\frac{\tau^*+\tau^0}{2}, \frac{\tau^*-\tau^0}{2}, 1, T-2\Big),
+\tau &\sim \mathit{TruncatedNormal}\Big(\frac{\tau^*+\tau^0}{2}, \frac{\tau^*-\tau^0}{2}, 5, T-2\Big),
 \end{aligned}
 \end{equation}
-where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 1 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
+where the prior for $\tau$ is a truncated normal distribution shaped so that the date of the first and last NPI, $\tau^0$ and $\tau^*$ (\autoref{table:NPI_dates}), are at minus and plus one standard deviation, and taking values only between 5 and  $T-2$, where $T$ is the number of days in the data $\vec{X}$.
 We also tested an uninformative uniform prior, $\mathit{Uniform}(1,T-2)$.
@@
 Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11.
+
+%TODO Convergence and merge to previous paragraph
+The integrated autocorrelation time (REF) was averaged across all parameters and chains, and found to be between VALUE FOR NORWAY and VALUE FOR GERMANY (Fig.
+SX).
+This means that in our 50 chains, with 7M samples per chain and 2M burnin samples, contain between XXX and XXX effectively uncorrelated samples (PLEASE CALCULATE THESE XXX and XXX numbers).
+We examined the trace plots for TAU in all regions (Fig.
+SX).
+All chains seem to converge to the same stationary distribution, in most cases before 2M steps (thus our choice of burnin time).
+In the case of Spain, single chain converged at around X steps (VERIFY!!).
+This single chain was considered part of the burnin and removed in the analysis.
+We also ran additional chains, which we initialized with a different seed and ran for 2M steps.
+The estimated posterior distributions of TAU are very similar to our main analysis (Fig.
+SX), further increasing our confidence in the convergence of our inference.
 
@@
 We also compute the expected posterior RMSE (root mean squared error) of these predictions~(\autoref{table:RMSE}).
+} % small
 
-%%% Source code %%%
-\paragraph*{Source code.} 
-We use Python 3 with the NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee packages.
-All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
+% Declarations
+{\small
+\section*{Declarations}
+\textbf{Ethics approval and consent to participate.} Not applicable.
+
+\textbf{Consent for publication.} Not applicable.
+
+\textbf{Availability of data and materials.} We use Python 3 with the NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee packages.
+Source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
+We used freely available data, see source code repository and above \emph{Data} section for details.
+
+\textbf{Competing interests.} The authors declare that they have no competing interests.
+
+\textbf{Funding.} This work was supported in part by the Israel Science Foundation (3811/19 and 552/19, YR). The funding body had no role in designing, performing, or analyzing the study.
+
+\textbf{Authors' contributions.} UO and YR designed the research. IK and YR performed the research and wrote the manuscript. All authors read and approved the final manuscript.
+
+\textbf{Acknowledgements.} 
+We thank Yinon M. Bar-On, Lilach Hadany, and Oren Kolodny for discussions and comments.
 } % small
@@
 %\pagebreak
-% Acknowledgements
-{\small
-\section*{Acknowledgements}
-We thank Yinon M. Bar-On, Lilach Hadany, and Oren Kolodny for discussions and comments.
-This work was supported in part by the Israel Science Foundation (3811/19 and 552/19, YR).
-}
 

### 2021-01-14 ms/ms.tex
started revising for BMC

@@
 \author[2,3]{Uri Obolski}
-\author[1,a]{Yoav Ram}
+\author[1,4,a]{Yoav Ram}
 
@@
 \affil[3]{Porter School of the Environment and Earth Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
+\affil[4]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
 \affil[a]{Corresponding author: yoav@yoavram.com, ORCID 0000-0002-9653-4458}
@@
 During Feb-Apr 2020, many countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules, to control the \covid pandemic caused by the \sars virus.
-Overall, these interventions seem to have successfully reduced the spread of the pandemic.
+Overall, these interventions seem to have reduced the spread of the pandemic.
 We hypothesise that the official and effective start date of such interventions can significantly differ, for example due to slow adoption by the population, or because the authorities and the public are unprepared.
-We fit an SEIR model to case data from~12 countries to infer the effective start dates of interventions and contrast them with the official dates.
+We fit an SEIR model to case data from~12 regions to infer the effective start dates of interventions and contrast them with the official dates.
 We find mostly late, but also early effects of interventions.
@@
 
-In the following, we describe our findings in more detail.
-
 %%% Late
@@
 \paragraph*{Exact effective start of NPIs.}
-We find one case in which the official and effective dates match "\emph{like a Swiss watch}" and the credible interval is narrow.
+We find one case in which the official and effective dates match and the credible interval is narrow.
 Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
@@
 
-In most investigated countries we find a late effective start of NPIs.
+In most investigated regions we find a late effective start of NPIs.
 For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five or more days after the official date (\autoref{fig:late}).
@@
 The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using the affine-invariant ensemble sampler for Markov chain Monte Carlo (MCMC)~\citep{Goodman2010} implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
+Reported credible intervals, or CI, are calculated as the highest density intervals, or HDI~\citep{Kruschke2015}.
 
@@
 The uninformative prior resulted in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom. 
-This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
+%This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
 We therefore decided to use the more informative truncated normal prior for $\tau$.
@@
 \\
-Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11.
 
@@
 SX), further increasing our confidence in the convergence of our inference.
+
+Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11.
 

### 2021-01-17 src/plot_utils.py
added delete_chain_less_than parameter for removing bad chain like in Spain

@@ -24,7 +24,8 @@ from model.uniform_prior_model import UniformPriorModel
 from inference import get_first_NPI_date, get_last_NPI_date, params_bounds, params_bounds, get_model_class
 
 country_name = None
-def load_data(file_name, _country_name, burn_fraction=0.6, lim_steps=None):
+# Currently lim_steps and delete_chain_less_than are not implemented to work together
+def load_data(file_name, _country_name, burn_fraction=0.6, lim_steps=None, delete_chain_less_than=None): 
     # it's the only global point. we initialize all the params here once and don't update it later (only when load_data again for different file_name)
     # TODO PLEASE DONT USE global anywhere in your code
     global first_NPI, last_NPI, incidences, start_date, var_names, nsteps, ndim, N, Td1, Td2, ndays, sample, lnprobability, logliks, model_type, model, country_name
@@ -35,9 +36,16 @@ def load_data(file_name, _country_name, burn_fraction=0.6, lim_steps=None):
     var_names = list(data['var_names'])
     nsteps, ndim, N, Td1, Td2, model_type = data['params']
     chain = data['chain']
-    nwalkers = chain.shape[0]
-    ndays = len(incidences)
+    nwalkers = chain.shape[0] #nwalkers before deleting bad chain
+    ndays = len(incidences)        
     nburn = int(nsteps * burn_fraction)
+
+    if delete_chain_less_than:
+        if len((chain[:,1_000_000, var_names.index('τ')]<delete_chain_less_than).nonzero())>1:
+            raise AssertionError('too many bad chains')
+        bad_chain_ind = (chain[:,1_000_000, var_names.index('τ')]<delete_chain_less_than).nonzero()[0][0]
+        chain = np.delete(chain, bad_chain_ind, axis=0)
+
     sample = chain[:, nburn:, :].reshape(-1, ndim)
     # try:
     #     sample[:,var_names.index('τ')] = sample[:,var_names.index('τ')].astype(int) #in inference we allways convert it to int
@@ -49,11 +57,18 @@ def load_data(file_name, _country_name, burn_fraction=0.6, lim_steps=None):
     #     None
 
     lnprobability = data['lnprobability'][:, nburn:]
-    logliks = data['logliks'].reshape(nwalkers,nsteps)[:,nburn:].reshape(-1)
+    logliks = data['logliks'].reshape(nwalkers,nsteps)[:,nburn:]
[…]

### 2021-01-17 ms/get_bibtex.py
better error msg

@@ -26,7 +26,11 @@ def main(keys_filename, bibtex_filename, output_filename, verbose):
 
 	out_bib = BibDatabase()
 	for key in citation_keys:
-		e = main_bib.entries_dict[key]
+		try:
+			e = main_bib.entries_dict[key]
+		except KeyError as err:
+			print("** Failed to find '{}', check for typos **".format(key))
+			raise err
 		out_bib.entries.append(e)
 	
 	if verbose:

### 2021-01-17 ms/ms.tex
revised following reviews, not finished

@@
  \renewcommand\linenumberfont{\normalfont\small\sffamily}
-%\linenumbers
-% \modulolinenumbers[2]
+\linenumbers
+\modulolinenumbers[2]
 
@@
 Overall, these interventions seem to have reduced the spread of the pandemic.
-We hypothesise that the official and effective start date of such interventions can significantly differ, for example due to slow adoption by the population, or because the authorities and the public are unprepared.
+We hypothesise that the official and effective start date of such interventions can be noticeably different, for example due to slow adoption by the population, or because the authorities and the public are unprepared.
 We fit an SEIR model to case data from~12 regions to infer the effective start dates of interventions and contrast them with the official dates.
@@
 In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\pm$ 1.05 days 95\% CI).
-We demonstrate that differences between the official and effective start of NPIs can distort conclusions about their impact, and discuss potential causes and consequences of our results.
+We demonstrate that differences between the official and effective start of NPIs can lead to under-estimating their impact, and discuss potential causes and consequences of our results.
 \end{abstract}
@@
 The date of the first intervention is for a ban of public events, or encouragement of social distancing, or for school closures.
-In all countries except Sweden, the date of the last intervention is for a lockdown.
+In all countries except Sweden, the date of the last intervention ($\tau^*$) is for a lockdown.
 In Sweden, where a lockdown was not ordered during the studied dates, the last date is for school closures.
@@
 
+% TODO rename label to tau-tau* (remove hat)
+% TODO update figure
+% TODO update caption tau values
 %%% Fig - tau summary %%%
@@
     \textbf{Official vs. effective start of non-pharmaceutical interventions.}
-    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs is shown for different regions.
+    	The difference between $\tau$ the effective and $\tau^*$ the official start of NPIs is shown for different regions.
 The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers).
@@
 The credible intervals for Sweden, Denmark, and Norway are especially wide, see text and \autoref{fig:early} for possible explanation.
-	Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}.
-$\tau^*$ is the last NPI date (\autoref{table:NPI_dates}).
-Thin and bold lines show 95\% and 75\% credible intervals, respectively.
+	Here, the markers show $\hat{\tau}$, the marginal posterior median, see \autoref{table:estimated-params}.
+$\tau^*$ is the last NPI date (a lockdown except in Sweden, see \autoref{table:NPI_dates}).
+Thin and bold lines show 95\% and 75\% credible intervals (HDI), respectively.
 \autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, 2020, rather then Apr~11.
@@
 Several studies have described the effects of non-pharmaceutical interventions in different geographical regions~\citep{Flaxman2020,Gatto2020,Li2020}. 
-Some of these studies have assumed that the parameters of the epidemiological model change at a specific date (\autoref{eq:NPI_model}), and set the change date $\tau$ to the official NPI date $\tau^*$ (\autoref{table:NPI_dates}).
+Some of these studies have assumed that the parameters of the epidemiological model change at a specific date (\autoref{eq:NPI_model}), and set the change date $\tau$ to the official NPI date $\tau^*$, usually the lockdown start date (\autoref{table:NPI_dates}).
 They then fit the model once for time $t<\tau^*$ and once for time $t \ge \tau^*$.
@@
 
-We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPIs differs from the official date: in 10 of 12 countries the 75\% credible interval on $\hat{\tau}$ does not include $\tau^*$ (7 of 12 countries when considering a 95\% interval; \autoref{fig:tau-summary}).
+We compare the official $\tau^*$ and effective $\tau$ start of NPIs and find that in most regions the effective start of NPIs differs from the official date: in 10 of 12 countries the 75\% credible interval on $\tau$ does not include $\tau^*$ (7 of 12 countries when considering a 95\% interval; \autoref{fig:tau-summary}).
 The exceptions are Switzerland and Norway (see below).
 The latter also has a wide 95\% credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
+Reported credible intervals, or CI, are calculated as the highest density intervals, or HDI~\citep{Kruschke2015}.
 
@@
 \paragraph*{Late effective start of NPIs.}
-In half of the examined regions, we estimate that the effective start of NPIs $\hat{\tau}$ is later than the official date $\tau^*$.
+In half of the examined regions, we estimate that the effective start of NPIs $\tau$ is later than the official date $\tau^*$.
 
@@
 This is a week later than the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
-
-Interestingly, the effective start of of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~24, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
-The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
-This may suggest correlations between effective starts of NPIs due to global or international events.
 
@@
 \paragraph*{Early effective start of NPIs.}
-In some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
+In some regions we estimate an effective start of NPIs $\tau$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
 The only conclusive early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\pm$1.05 days 95 \%CI; \autoref{fig:early}).
@@
 Interestingly, Germany has the second longest duration between first and last NPIs after Norway (10 and 12 days respectively; see \autoref{table:NPI_dates}).
-Nevertheless, the credible interval for the effective date in Germany is narrow ($\pm$1.05 days 95 \%CI), whereas it is quite wide in Norway ($\pm$10.79 days; \autoref{fig:early}).
+However, the credible interval for the effective date in Germany is narrow ($\pm$1.05 days 95 \%CI), whereas it is quite wide in Norway ($\pm$10.79 days; \autoref{fig:early}).
+That $\tau$ is significantly earlier than $\tau^*$ can be suggest that early NPIs had been more effective in reducing the transmission rates in Germany compared to other countries.
+Another possible interpretation is that the German population anticipated the lockdown and began to act in accordance before it started.
 
@@
 However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free $\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
-
-In Sweden, the data do not "peak", i.e., the number of daily cases continues to grow up to Apr~11 (\autoref{fig:ppc}). 
+Indeed, Denmark and Sweden had the highest estimates of $\lambda$, that is, the  smallest effect of NPIs on transmission rates was the estimated, which  probably hinders our ability to estimate $\tau$ in these countries.
+Moreover, in Sweden, the data do not "peak", i.e., the number of daily cases continues to grow up to Apr~11 (\autoref{fig:ppc}). 
 In Denmark, the opposite occurs: there are seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
@@
 
-The \emph{effective reproduction number} $R$ is the number of secondary
-cases generated by an infectious case after an epidemic is already underway \citep{Bar-On2020}.
-We infer the effective reproduction numbers before and after the implementation of NPIs from model parameters (\autoref{eq:Re}).
+The \emph{effective reproduction number} $R$ is the average number of secondary
+cases caused by an infected individual after an epidemic is already underway \citep{Bar-On2020a}.
+We infer the model-based effective reproduction numbers before and after the implementation of NPIs from model parameters (\autoref{eq:Re}).
 We then estimate the impact of NPIs as the relative reduction in the reproduction number \citep{Flaxman2020}. 
@@
 
-These results suggest that the impact of past NPIs by health officials and researchers is likely to be under-estimated if they assume NPIs effectively start at their official dates.
+These results suggest that the impact of past NPIs is likely to be under-estimated by health officials and researchers if they assume NPIs effectively start at their official dates.
 The estimated impact can then be interpreted as ineffectiveness of the NPIs, leading to further escalations.
@@
 If health officials assume that the dynamics change exactly at $\tau^*$, they would expect the number of cases to be within the red lines (posterior predictions assuming $\tau=\tau^*$).
-This would lead to a significant under-estimation, which might be interpreted as ineffectiveness of the NPI, leading to further escalations.
+This would lead to a significant under-estimation, which might be interpreted as ineffectiveness of the NPI, leading to more aggressive NPIs being applied.
 However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which are a good fit to the real data (stars).
@@
 %%% Fig - Prediction
-
 \begin{figure}[h]
@@
     \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
-    Real number of daily cases in Italy in black (markers: data; line: time moving average).
+    Real number of daily cases in Italy in black (markers: data; line: smoothed using a Savitzky-Golay filter with window length 3).
     Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution).
-    Shaded box illustrates a generation interval of seven days.
-    \textbf{(A)} Using the official date $\tau^*$ for the effective start of the NPI, the model under-estimates the number of cases seven days after the start of the NPI.
-    \textbf{(B)} Using the estimated date $\hat{\tau}$ for the effective start of the NPI, the model precisely estimates the number of cases seven days after the start of the NPI.
+    Shaded box illustrates a serial interval of seven days~\citep{Ali2020} after the start of NPIs.
+    \textbf{(A)} Using the official date $\tau^*$ for the effective start of the NPI, Mar~11, the model under-estimates the number of cases seven days after the start of the NPI.
+    \textbf{(B)} Using the estimated date $\hat{\tau}$ for the effective start of the NPI, Mar~17, the model precisely estimates the number of cases seven days after the start of the NPI.
     Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
@@
 For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five or more days after the official date (\autoref{fig:late}).
-This difference might be explained, in some cases, by low compliance: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, resulting in people leaving these provinces before the lockdown started~\citep{Gatto2020}.
+This difference might be explained, in some cases, by low compliance or non-adherence to guidelines: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, resulting in people leaving these provinces before the lockdown started~\citep{Gatto2020}.
 Late effect of NPIs may also be due to the time required by both the government and the citizens to prepare for a lockdown, and for new guidelines to be adopted by the population.
@@
 
+Interestingly, the effective start of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~24, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
+The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
+This may suggest correlations between effective starts of NPIs due to global or international events.
+
 As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with new guidelines, whereas in other cases the population will adopt either restrictions or relaxations even before they are formally announced.
-Attempts to asses the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date, and that a significant change in the dynamics can be observed roughly seven days after the start of NPIs (due to the characteristic generation interval of \covid\citep{Bar-On2020}).
+Attempts to asses the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date, and that a significant change in the dynamics can be observed roughly seven days after the start of NPIs (due to the characteristic serial interval of \covid\citep{Ali2020}).
 However, late and early effective start of NPIs, such as we have inferred, can bias these assessments and lead to wrong conclusions about the impact of NPIs (\autoref{fig:Re}, \autoref{fig:prediction}).
@@
 We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions during Jan--Apr 2020.
-These incidence data summarise the number of individuals $X_t$ tested positive for \sars (using RT-qPCR tests) on each day $t$.
-Data for Wuhan, China retrieved from \citet{Pei2020}. Data for 11 European countries retrieved from \citet{Flaxman2020}. 
+These incidence data summarise the number of individuals $X_t$ tested positive for \sars (using RT-qPCR tests) on each day $t$
+Data for Wuhan, China, Jan~10 to Feb~8, retrieved from \citet{Pei2020}. Data for 11 European countries, Feb~20 to Apr~24, retrieved from \citet{Flaxman2020}.
 Where there were multiple sequences of days with zero confirmed cases (e.g.
@@
 In either case, they no longer infect other individuals, and therefore effectively become recovered ($R$).
-The model is described by the following set of equations:
-
+The model is described by the following set of equations,
 \begin{equation} \label{eq:model}
 \begin{aligned}
-\frac{dS}{dt} & = -\beta_t S \frac{I_r}{N} - \mu \beta_t S \frac{I_u}{N} \\
-\frac{dE}{dt} & = \beta_t S \frac{I_r}{N} + \mu \beta_t S \frac{I_u}{N}  - \frac{E}{Z} \\
+\frac{dS}{dt} & = -\beta_t S \Big(\frac{I_r}{N} + \mu \frac{I_u}{N}\Big) \\
+\frac{dE}{dt} & = \beta_t S \Big(\frac{I_r}{N} + \mu \frac{I_u}{N}\Big)  - \frac{E}{Z} \\
 \frac{dI_r}{dt} & = \alpha_t \frac{E}{Z} - \frac{I_r}{D} \\
-\frac{dI_u}{dt} & = (1-\alpha_t) \frac{E}{Z} - \frac{I_r}{D} \\
-\frac{dR}{dt} & = \frac{I_r}{D} + \frac{I_r}{D} .
+\frac{dI_u}{dt} & = \big(1-\alpha_t\big) \frac{E}{Z} - \frac{I_u}{D} \\
+\frac{dR}{dt} & = \frac{I_r}{D} + \frac{I_u}{D} \;,
 \end{aligned}
 \end{equation}
+where $N$ is the population size.
 The initial numbers of exposed $E(0)$ and unreported infected $I_u(0)$ are free model parameters (i.e.
@@
 $(Y_t(\theta) - \tilde{X}_{t-1})$.
-We therefore assume that $X_t$ conditioned on $\tilde{X}_{t-1}$ is Poisson distributed, such that
+We assume that $X_t$ conditioned on $\tilde{X}_{t-1}$ is Poisson distributed, such that
 \begin{equation} \label{eq:Xt} \begin{aligned}
@@
 To estimate the model parameters from the daily case data $\vec{X}$, we apply a Bayesian inference approach.
+Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11 (for Wuhan, China, model fitting was applied for data up to Feb~8.)
 We start our model $\Delta t$ days~\citep{Gatto2020} before the outbreak (defined as consecutive days with increasing confirmed cases) in each country.
-The model in \autoref{eq:model} is parameterised by the vector $\theta$, where
+The model in Eqs.~\ref{eq:model} and~\ref{eq:NPI_model} is parameterised by the vector $\theta$, where
 \begin{equation} \label{eq:theta}
-\theta=\Big(Z, D, \mu, \{\beta_t\}, \{\alpha_t\}, \{p_t\}, E(0), I_u(0), \tau, \Delta t \Big).
+\theta=\Big(Z, D, \mu, \beta, \lambda, \alpha_1, \alpha_2, E(0), I_u(0), \Delta t, \tau \Big) \;.
 \end{equation}
@@
 The likelihood function is defined in \autoref{eq:likelihood}.
-The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using the affine-invariant ensemble sampler for Markov chain Monte Carlo (MCMC)~\citep{Goodman2010} implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
-Reported credible intervals, or CI, are calculated as the highest density intervals, or HDI~\citep{Kruschke2015}.
-
 We defined the following prior distributions on the model parameters $P(\theta)$: 
@@
 Moreover, the estimates for Wuhan, China were $1/9$ and $1/6$, as in \citet{Li2020}.
-\\
 
-%TODO Convergence and merge to previous paragraph
-The integrated autocorrelation time (REF) was averaged across all parameters and chains, and found to be between VALUE FOR NORWAY and VALUE FOR GERMANY (Fig.
-SX).
-This means that in our 50 chains, with 7M samples per chain and 2M burnin samples, contain between XXX and XXX effectively uncorrelated samples (PLEASE CALCULATE THESE XXX and XXX numbers).
-We examined the trace plots for TAU in all regions (Fig.
-SX).
-All chains seem to converge to the same stationary distribution, in most cases before 2M steps (thus our choice of burnin time).
-In the case of Spain, single chain converged at around X steps (VERIFY!!).
-This single chain was considered part of the burnin and removed in the analysis.
-We also ran additional chains, which we initialized with a different seed and ran for 2M steps.
-The estimated posterior distributions of TAU are very similar to our main analysis (Fig.
-SX), further increasing our confidence in the convergence of our inference.
-
-Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11.
+The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using the affine-invariant ensemble sampler for Markov chain Monte Carlo (MCMC)~\citep{Goodman2010} implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
+We used the default configuration using the stretch move with stretch scale parameter $a=2$.
+For the main analysis we use 50 chains (or walkers) per region, with 7M samples per chain (no thinning was applied).
+% TODO add refs and values
+The \emph{integrated autocorrelation time}~\citep{Foreman-Mackey2013,Goodman2010} was averaged across parameters and chains for each region (FIGURE). 
+It was found to be between XXX for Norway and XXX for Germany. 
+We examined the trace plots for $\tau$ in all regions (FIGURE).
+All chains seem to converge to the same stationary distribution, in most cases before 2M steps. Thus, we discarded the first 2M samples as burnin samples.
+The only exception is Spain, in which a single chain converged at around X steps.
+We considered this chain as part of the burnin and removed in the analysis.
+We also ran additional chains with 2M samples and a different initialization (i.e.
+seed).
+The estimated posterior distributions of $\tau$ are very similar to our main analysis, further increasing our confidence in the convergence of our inference.
+We also examined the joint posterior distributions of $\tau$ and $\lambda$ to check for correlations (FIGURE). %TODO what do we find?
 
@@
 We compare models by reporting their relative WAIC; lower is better~(\autoref{table:WAIC}).
-A minority (<5\%) of MCMC chains that fail to fully converge can lead to overestimation of the variance (the second term in \autoref{eq:WAIC}).
-Therefore, we exclude from the computation of WAIC chains with mean log-likelihood that is three standard deviations or more from the overall mean.
 
@@
 \textbf{Acknowledgements.} 
-We thank Yinon M. Bar-On, Lilach Hadany, and Oren Kolodny for discussions and comments.
+We thank Yinon M. Bar-On, Lilach Hadany, and Oren Kolodny for discussions and comments. % Tim CD Lucas
 } % small
@@
     \textbf{Official vs. effective start of non-pharmaceutical interventions estimated up to Mar~28.}
-    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPIs estimated from case data up to Mar~28, 2020, shown for different regions.
-Here, $\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}.
-$\tau^*$ is the last NPI date (\autoref{table:NPI_dates}).
-Thin and bold lines show 95\% and 75\% credible intervals, respectively.
+    	The difference between $\tau$ the effective and $\tau^*$ the official start of NPIs estimated from case data up to Mar~28, 2020, shown for different regions.
+Here, $\hat{\tau}$ is the marginal posterior median, see \autoref{table:estimated-params}.
+$\tau^*$ is the last NPI date (a lockdown everywhere by Sweden, see \autoref{table:NPI_dates}).
+Thin and bold lines show 95\% and 75\% credible intervals (HDI), respectively.
     }
@@
     \caption{
-    \textbf{\covid confirmed cases in France and Spain.}
+    \textbf{\covid daily confirmed cases in France and Spain.}
     Number of cases proportional to population size (as of 2018). 
     Vertical line shows Mar~8, the effective start of NPIs $\hat{\tau}$ in both countries.
+    Data from~\citet{Flaxman2020}.
     } 
@@
     Black and white arrows denote the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs, respectively.
-    Black lines represent a smoothing of the data points using a Savitzky-Golay filter. 
+    Black lines represent a smoothing of the data points using a Savitzky-Golay filter with window length 3. 
     Coloured lines represent posterior predictions from a model with fixed $\tau$ (blue) and free $\tau$ (green).

### 2021-01-17 src/plot_utils.py
fixed WAIC

@@ -61,14 +61,13 @@ def load_data(file_name, _country_name, burn_fraction=0.6, lim_steps=None, delet
     if delete_chain_less_than:
             lnprobability = np.delete(lnprobability, bad_chain_ind, axis=0)
             logliks = np.delete(logliks, bad_chain_ind, axis=0)
-    logliks = logliks.reshape(-1)
 
     if lim_steps:
         #TODO add delete_chain_less logic
         sample = chain[:, int(lim_steps * burn_fraction):lim_steps, :].reshape(-1, ndim)
         lnprobability = data['lnprobability'][:, int(lim_steps * burn_fraction):lim_steps]
-        logliks = data['logliks'].reshape(nwalkers,nsteps)[:,int(lim_steps * burn_fraction):lim_steps].reshape(-1)
-        
+        logliks = data['logliks'].reshape(nwalkers,nsteps)[:,int(lim_steps * burn_fraction):lim_steps]
+
     last_NPI = (get_last_NPI_date(country_name) - pd.to_datetime(start_date)).days
     first_NPI = (get_first_NPI_date(country_name) - pd.to_datetime(start_date)).days
 
@@ -162,9 +161,7 @@ def inliers(logliks, PLOT=False):
     return idx
 
 def calc_WAIC():
-    nwalkers = logliks.size//nsteps
-    l_logliks = logliks.reshape(nwalkers,nsteps)
-    l_logliks = l_logliks[inliers(l_logliks)]
+    l_logliks = logliks[inliers(logliks)]
     S = l_logliks.size
     llpd = -np.log(S) + logsumexp(l_logliks)
     p1 = 2*(-np.log(S) + logsumexp(l_logliks) - l_logliks.mean())

### 2021-01-17 src/plot_utils.py
change nburn logic a bit

@@ -24,8 +24,7 @@ from model.uniform_prior_model import UniformPriorModel
 from inference import get_first_NPI_date, get_last_NPI_date, params_bounds, params_bounds, get_model_class
 
 country_name = None
-# Currently lim_steps and delete_chain_less_than are not implemented to work together
-def load_data(file_name, _country_name, burn_fraction=0.6, lim_steps=None, delete_chain_less_than=None): 
+def load_data(file_name, _country_name, nburn=2_000_000, lim_steps=None, delete_chain_less_than=None): 
     # it's the only global point. we initialize all the params here once and don't update it later (only when load_data again for different file_name)
     # TODO PLEASE DONT USE global anywhere in your code
     global first_NPI, last_NPI, incidences, start_date, var_names, nsteps, ndim, N, Td1, Td2, ndays, sample, lnprobability, logliks, model_type, model, country_name
@@ -38,7 +37,6 @@ def load_data(file_name, _country_name, burn_fraction=0.6, lim_steps=None, delet
     chain = data['chain']
     nwalkers = chain.shape[0] #nwalkers before deleting bad chain
     ndays = len(incidences)        
-    nburn = int(nsteps * burn_fraction)
 
     if delete_chain_less_than:
         if len((chain[:,1_000_000, var_names.index('τ')]<delete_chain_less_than).nonzero())>1:
@@ -46,7 +44,7 @@ def load_data(file_name, _country_name, burn_fraction=0.6, lim_steps=None, delet
         bad_chain_ind = (chain[:,1_000_000, var_names.index('τ')]<delete_chain_less_than).nonzero()[0][0]
         chain = np.delete(chain, bad_chain_ind, axis=0)
 
-    sample = chain[:, nburn:, :].reshape(-1, ndim)
+    sample = chain[:, nburn:lim_steps, :].reshape(-1, ndim)
     # try:
     #     sample[:,var_names.index('τ')] = sample[:,var_names.index('τ')].astype(int) #in inference we allways convert it to int
     # except ValueError: #if the model doesn't have such parameter
@@ -56,17 +54,11 @@ def load_data(file_name, _country_name, burn_fraction=0.6, lim_steps=None, delet
     # except ValueError:#if the model doesn't have such parameter
     #     None
 
-    lnprobability = data['lnprobability'][:, nburn:]
[…]

### 2021-01-20 ms/ms.tex
autocorr figure; moved ppc fig; fixed captioning issue in ppc fig

@@
 
-Keywords: SEIR, COVID-19, public health, pandemic, infectious disease, NPI
-
-%Classification: Biological Sciences / Population Biology; Social Sciences / Environmental Sciences
+Keywords: SEIR, COVID-19, public health, epidemic, infectious disease, NPI
 
@@
 \begin{abstract}
-
 During Feb-Apr 2020, many countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules, to control the \covid pandemic caused by the \sars virus.
 Overall, these interventions seem to have reduced the spread of the pandemic.
-We hypothesise that the official and effective start date of such interventions can be noticeably different, for example due to slow adoption by the population, or because the authorities and the public are unprepared.
+We hypothesize that the official and effective start date of such interventions can be noticeably different, for example due to slow adoption by the population, or because the authorities and the public are unprepared.
 We fit an SEIR model to case data from~12 regions to infer the effective start dates of interventions and contrast them with the official dates.
 We find mostly late, but also early effects of interventions.
-For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective date on Mar~17 ($\pm$2.99 days 95\% CI).
-In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\pm$ 1.05 days 95\% CI).
+For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective date on Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\% CI).
+In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95\% CI).
 We demonstrate that differences between the official and effective start of NPIs can lead to under-estimating their impact, and discuss potential causes and consequences of our results.
@@
 
-%\renewcommand{\abstractname}{Significance statement}
-%\begin{abstract}
-%We have inferred the effective start date of non-pharmaceutical interventions, like lockdowns, during the initial \covid outbreak by fitting an epidemiological model to daily case data in 12 countries.
-Our results show that the effective start of interventions in many cases occurs 3-10 days later then the official day.
-Moreover, we demonstrate that such delays can lead to under-estimation of the intervention impact, and thus cause decision makers to implement stricter interventions.
-We suggest that a late effective start of interventions may be the result of low compliance to authority or conformity to social norms.
-Our study therefore highlights the importance of further research into the role of human behavior and social organization on the current and future epidemics.
-%\end{abstract}
-
 \pagebreak
-
-%TODO HDI
 
@@
 
-% TODO rename label to tau-tau* (remove hat)
-% TODO update figure
-% TODO update caption tau values
 %%% Fig - tau summary %%%
-% generated with /python Fig_tau_summary.py 2020-05-26-Apr11/
+% generated with /python Fig_tau_summary.py FOLDER NAME
 \begin{figure}[b!]
@@
 The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers).
-In contrast, the estimated effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although uncertainty is low only for Germany (i.e., zero is not included in 95\% CI).
+In contrast, the estimated effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although uncertainty is low only for Germany (i.e., zero is not in 95\% CI).
 The credible intervals for Sweden, Denmark, and Norway are especially wide, see text and \autoref{fig:early} for possible explanation.
-	Here, the markers show $\hat{\tau}$, the marginal posterior median, see \autoref{table:estimated-params}.
-$\tau^*$ is the last NPI date (a lockdown except in Sweden, see \autoref{table:NPI_dates}).
-Thin and bold lines show 95\% and 75\% credible intervals (HDI), respectively.
-\autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, 2020, rather then Apr~11.
+	Here, the markers show $\hat{\tau}$, the marginal posterior median (\autoref{table:estimated-params}).
+$\tau^*$ is the last NPI date (a lockdown, except in Sweden; \autoref{table:NPI_dates}).
+Thin and bold lines show 95\% and 75\% credible intervals, respectively.
+\autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28 rather than Apr~11.
     }
@@
 Here, we estimate the joint posterior distribution of the effective start date of NPIs, $\tau$, and the transmission and reporting rates before and after $\tau$ from the entire data, rather than splitting the data at~$\tau$.
+This is done under the simplifying assumption that all interventions start at a specific date, despite the reality that the durations between the first and last NPIs were between 4 and 12 days (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
 We then estimate $\hat{\tau}$ as the median of the marginal posterior distribution of $\tau$.
-This is done under the simplifying assumption that all interventions start at a specific date, despite the reality that the durations between the first and last NPIs were between 4 and 12 days (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
+Credible intervals (CI) are calculated as the highest density intervals~\citep{Kruschke2015}, and their upper and lower boundaries are reported as $\substack{+\text{upper} \\ -\text{lower}}$ in days relative to $\hat{\tau}$.
 
@@
 All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~24, and these predictions were then compared to the real case data. 
+% TODO update when ppc fig is ready
 The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
@@
 We compare the official $\tau^*$ and effective $\tau$ start of NPIs and find that in most regions the effective start of NPIs differs from the official date: in 10 of 12 countries the 75\% credible interval on $\tau$ does not include $\tau^*$ (7 of 12 countries when considering a 95\% interval; \autoref{fig:tau-summary}).
-The exceptions are Switzerland and Norway (see below).
-The latter also has a wide 95\% credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
-Reported credible intervals, or CI, are calculated as the highest density intervals, or HDI~\citep{Kruschke2015}.
-
-%%% Late
-\paragraph*{Late effective start of NPIs.}
-In half of the examined regions, we estimate that the effective start of NPIs $\tau$ is later than the official date $\tau^*$.
-
-In Italy, the first case was officially confirmed on Feb~21.
-School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
-That is, the first and last official NPI dates are Mar~5 and Mar~11.
-However, we estimate the effective date $\hat{\tau}$ six days after the lockdown, at Mar~17 ($\pm$2.99 days 95\% CI; \autoref{fig:late}).
-
-In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later, at Feb~3 ($\pm$2.85 days 95\% CI; \autoref{fig:late}).
-
-In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score 2--1) with a crowd of 50,965 in Seville.
-A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ ten days later, at Mar~24 ($\pm$2.96 days 95 \%CI), rather than Mar~14 (\autoref{fig:late}).
-
-Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ at Mar~24 ($\pm$2.29 days 95\% CI, \autoref{fig:late}).
-This is a week later than the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
+The exceptions are Norway and Switzerland (see below).
+The former also has a wide 95\% credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
 
@@
 Red line shows the official last NPI date $\tau^*$.
-Black line shows the estimate $\hat{\tau}$.
+Black line shows the estimated effective start date $\hat{\tau}$.
 Shaded area shows a 95\% credible interval.
@@
 
-%%% Early
-\paragraph*{Early effective start of NPIs.}
-In some regions we estimate an effective start of NPIs $\tau$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
-The only conclusive early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\pm$1.05 days 95 \%CI; \autoref{fig:early}).
-This estimate falls between the first and last official NPI dates, Mar~12 and Mar~22 (\autoref{table:NPI_dates}).
-Therefore, when we refer to this case as "early", we mean that the effective date (Mar~19) occurs \emph{before} the official lockdown date (Mar~22), not that it occurs before all the NPIs.
-Interestingly, Germany has the second longest duration between first and last NPIs after Norway (10 and 12 days respectively; see \autoref{table:NPI_dates}).
-However, the credible interval for the effective date in Germany is narrow ($\pm$1.05 days 95 \%CI), whereas it is quite wide in Norway ($\pm$10.79 days; \autoref{fig:early}).
-That $\tau$ is significantly earlier than $\tau^*$ can be suggest that early NPIs had been more effective in reducing the transmission rates in Germany compared to other countries.
-Another possible interpretation is that the German population anticipated the lockdown and began to act in accordance before it started.
+%%% Late
+\paragraph*{Late effective start of NPIs.}
+In half of the examined regions, we estimate that the effective start of NPIs $\tau$ is later than the official date $\tau^*$.
 
-We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
-However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free $\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
-Indeed, Denmark and Sweden had the highest estimates of $\lambda$, that is, the  smallest effect of NPIs on transmission rates was the estimated, which  probably hinders our ability to estimate $\tau$ in these countries.
-Moreover, in Sweden, the data do not "peak", i.e., the number of daily cases continues to grow up to Apr~11 (\autoref{fig:ppc}). 
-In Denmark, the opposite occurs: there are seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
-We suspect that these missing and additional "peaks" increase the uncertainty in our inference.
+In Italy, the first case was officially confirmed on Feb~21.
+School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
+That is, the first and last official NPI dates are Mar~5 and Mar~11.
+However, we estimate the effective date $\hat{\tau}$ six days after the lockdown, at Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\% CI; \autoref{fig:late}).
 
-The estimated effective date in Norway is Mar~22, which is two days earlier than the official date of Mar~24.
-However, the posterior distribution is very wide ($\pm$10.79 days 95\% CI; \autoref{fig:early}): it covers the range between Mar~9, three days before the first NPI, and Mar~29, five days  after the last NPI (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
-The uncertainty in the estimate of the effective start of NPIs might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is quite narrow (\autoref{fig:early}).
+In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\% CI; \autoref{fig:late}).
+
+In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score 2--1) with a crowd of 50,965 in Seville.
+A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
+Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95 \%CI, \autoref{fig:late}).
+
+Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ at Mar~24 ($\substack{+2.65 \\ -1.44}$~days 95\% CI, \autoref{fig:late}).
+This is a week later than the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
 
@@
 Red line shows the official last NPI date $\tau^*$.
-Black line shows the estimated $\hat{\tau}$.
+Black line shows the estimated effective start date $\hat{\tau}$.
 Shaded area shows a 95\% credible interval.
@@
 
+%%% Early
+\paragraph*{Early effective start of NPIs.}
+In some regions we estimate an effective start of NPIs $\tau$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
+The only conclusive early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95 \%CI; \autoref{fig:early}).
+This estimate falls between the first and last official NPI dates, Mar~12 and Mar~22 (\autoref{table:NPI_dates}).
+Therefore, when we refer to this case as "early", we mean that the effective date (Mar~19) occurs \emph{before} the official lockdown date (Mar~22), not that it occurs before all NPIs.
+Interestingly, Germany has the second longest duration between first and last NPIs after Norway (10 and 12 days respectively; \autoref{table:NPI_dates}).
+However, the credible interval for the effective date in Germany is narrow (1.91 days), whereas it is very wide in Norway (17.61 days).
+That $\tau$ is significantly earlier than $\tau^*$ can suggest that early NPIs in Germany more effectively reduced transmission rates compared to other countries.
+Another possible interpretation is that the German population anticipated the lockdown and began to act accordingly before it started.
+
+We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
+However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free $\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
+%TODO check WAIC
+Indeed, Denmark and Sweden had the highest estimates of $\lambda$, that is, the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries.
+Moreover, in Sweden the number of daily cases continues to grow up to Apr~11, rather then "peak" (\autoref{fig:ppc}). 
+In Denmark, the opposite occurs: there are seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
+We suspect that these missing and additional "peaks" increase the uncertainty in our inference.
+
+The estimated effective date in Norway is Mar~22, two days earlier than the official date of Mar~24.
+However, the posterior distribution is very wide ($\substack{+5.59 \\ -12.03}$~days 95\% CI): it covers the range between Mar~10, two days before the first NPI, and Mar~27, three days after the last NPI (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
+The uncertainty in the estimate of the effective start of NPIs might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is very narrow (\autoref{fig:early}).
+
 %%% Exact
@@
 Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
-Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\pm$2.51 days 95\% CI, see \autoref{fig:early}).
+Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\substack{+3.2 \\ -1.78}$~days 95\% CI, see \autoref{fig:early}).
 It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
@@
 cases caused by an infected individual after an epidemic is already underway \citep{Bar-On2020a}.
-We infer the model-based effective reproduction numbers before and after the implementation of NPIs from model parameters (\autoref{eq:Re}).
+We infer model-based effective reproduction numbers before and after the implementation of NPIs from model parameters (\autoref{eq:Re}).
 We then estimate the impact of NPIs as the relative reduction in the reproduction number \citep{Flaxman2020}. 
@@
 \autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model are consistently lower than estimates from the free~$\tau$ model, except in Sweden and Denmark, in which estimation uncertainty is high.
-
+%TODO update when fig Re is ready
 These results suggest that the impact of past NPIs is likely to be under-estimated by health officials and researchers if they assume NPIs effectively start at their official dates.
-The estimated impact can then be interpreted as ineffectiveness of the NPIs, leading to further escalations.
+The estimated impact can then be interpreted as ineffectiveness of the NPIs, leading to more aggressive NPIs being applied.
 
@@
 If health officials assume that the dynamics change exactly at $\tau^*$, they would expect the number of cases to be within the red lines (posterior predictions assuming $\tau=\tau^*$).
-This would lead to a significant under-estimation, which might be interpreted as ineffectiveness of the NPI, leading to more aggressive NPIs being applied.
+This would lead to a significant under-estimation, which might be interpreted as ineffectiveness of NPIs, and therefore more strict guidelines.
 However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which are a good fit to the real data (stars).
@@
 % > python Fig_Re.py
+% TODO update
 \begin{figure}[h]
@@
 The dashed line shows a one-to-one correspondence.
-     Markers and bars denote the posterior median and 50\% credible intervals.
+     Markers and bars denote the posterior median and 50\% credible intervals (HDI).
     The relative reductions in $R$ are consistently lower for the fixed $\tau$ model (below the dashed line), except in Sweden and Denmark in which uncertainty is high.
@@
 %%% Fig - Prediction
+% TODO update
 \begin{figure}[h]
@@
 
-In most investigated regions we find a late effective start of NPIs.
+In most investigated regions we find late effective start of NPIs.
 For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five or more days after the official date (\autoref{fig:late}).
@@
 In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented, although this is only conclusive in Germany (\autoref{fig:early}).
-An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population.
-Adoption of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may be influenced by increased risk perception due to domestic or international COVID\=/19-related reports~\citep{Arthur2020}.
+An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population, possibly due to earlier NPIs or NPIs being applied in other regions.
+Adoption of these behaviours may occur via media and social networks, rather than official guidelines, and may be influenced by increased risk perception due to domestic or international \covid reports~\citep{Arthur2020}.
 Indeed, the evidence supports a change in infection dynamics (i.e.
@@
 
-As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (\autoref{table:WAIC} and \autoref{table:RMSE}).
-It may be interesting to check if the evidence favors a model with \emph{two} or more change-points, rather than one. 
+As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (Tables~\ref{table:WAIC} and~\ref{table:RMSE}).
+It may be interesting to check if the evidence supports a model with \emph{two} or more change-points, rather than one. 
 Multiple change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
@@
 
-Interestingly, the effective start of NPIs $\hat{\tau}$ in both France and Spain is estimated to have started on Mar~24, although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
+Interestingly, the effective start of NPIs in France and Spain is estimated to have started on Mar~24 and~25, respectively,
+although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
 The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
@@
 Attempts to asses the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date, and that a significant change in the dynamics can be observed roughly seven days after the start of NPIs (due to the characteristic serial interval of \covid\citep{Ali2020}).
-However, late and early effective start of NPIs, such as we have inferred, can bias these assessments and lead to wrong conclusions about the impact of NPIs (\autoref{fig:Re}, \autoref{fig:prediction}).
+However, late and early effective start of NPIs, such as we have inferred, can bias these assessments and lead to wrong conclusions about the impact of NPIs (Figures~\ref{fig:Re} and~\ref{fig:prediction}).
 
@@
 We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions during Jan--Apr 2020.
-These incidence data summarise the number of individuals $X_t$ tested positive for \sars (using RT-qPCR tests) on each day $t$
-Data for Wuhan, China, Jan~10 to Feb~8, retrieved from \citet{Pei2020}. Data for 11 European countries, Feb~20 to Apr~24, retrieved from \citet{Flaxman2020}.
+These incidence data summarise the number of individuals $X_t$ tested positive for \sars (using RT-qPCR tests) on each day $t$.
+Data for Wuhan, China, from Jan~10 to Feb~8, retrieved from \citet{Pei2020}. Data for 11 European countries, from Feb~20 to Apr~24, retrieved from \citet{Flaxman2020}.
 Where there were multiple sequences of days with zero confirmed cases (e.g.
@@
 % TODO add refs and values
-The \emph{integrated autocorrelation time}~\citep{Foreman-Mackey2013,Goodman2010} was averaged across parameters and chains for each region (FIGURE). 
-It was found to be between XXX for Norway and XXX for Germany. 
+The \emph{integrated autocorrelation time}~\citep{Foreman-Mackey2013,Goodman2010} was averaged across parameters and chains for each region. 
+It was found to be between XXX for Norway and XXX for Germany (\autoref{fig:autocorr}). % TODO this gives effective number of samples...
 We examined the trace plots for $\tau$ in all regions (FIGURE).
@@
 The only exception is Spain, in which a single chain converged at around X steps.
-We considered this chain as part of the burnin and removed in the analysis.
+We considered this chain as part of the burnin and removed it from the analysis.
 We also ran additional chains with 2M samples and a different initialization (i.e.
 seed).
-The estimated posterior distributions of $\tau$ are very similar to our main analysis, further increasing our confidence in the convergence of our inference.
+The estimated posterior distributions of $\tau$ were very similar to our main analysis, further increasing our confidence in the convergence of our inference.
 We also examined the joint posterior distributions of $\tau$ and $\lambda$ to check for correlations (FIGURE). %TODO what do we find?
 
-%%% Model selection %%%
+%%% Model comparison %%%
 \paragraph*{Model comparison.}
@@
 
-\textbf{Availability of data and materials.} We use Python 3 with the NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee packages.
+\textbf{Availability of data and materials.} We use Python 3 with NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee.
 Source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
@@
 %% Table - estimated parameters
-% generated with /python Table_estimated_params.py 2020-05-26-Apr11/
+% generated with /python Table_estimated_params.py FOLDER NAME
 \begin{landscape}
@@
 \centering
+\footnotesize{
 \input{../figures/Table-estimated-params.tex}
+}
 \caption{
@@
 All estimates are posterior medians.
-75\% and 95\% credible intervals given for $\tau$, in days.
-$\tau^*$ is the official last NPI date, see \autoref{table:NPI_dates}.
+75\% and 95\% credible intervals (HDI) are given for $\tau$ in days relative to $\hat{\tau}$.
+$\tau^*$ is the official last NPI date (\autoref{table:NPI_dates}).
 }
@@
 % Table - WAIC 
-% generated with /python Table-WAIC.py 2020-05-26-Apr11/
+% generated with /python Table-WAIC.py FOLDER NAME
+% TODO update
 \begin{table}[h]
@@
 % Table - RMSE 
-% generated with /python Table-RMSE.py 2020-05-26-Apr11/
+% generated with /python Table-RMSE.py FOLDER NAME
+% TODO update
 \begin{table}[h]
@@
     	The difference between $\tau$ the effective and $\tau^*$ the official start of NPIs estimated from case data up to Mar~28, 2020, shown for different regions.
-Here, $\hat{\tau}$ is the marginal posterior median, see \autoref{table:estimated-params}.
+Here, $\hat{\tau}$ is the marginal posterior median.
 $\tau^*$ is the last NPI date (a lockdown everywhere by Sweden, see \autoref{table:NPI_dates}).
-Thin and bold lines show 95\% and 75\% credible intervals (HDI), respectively.
+Thin and bold lines show 95\% and 75\% credible intervals (HDI~\citep{Kruschke2015}), respectively.
+Inference performed similarly to main inference, but with data up to Mar~28 and only 1M samples per chain with 600K burnin.
     }
@@
 \end{figure}
+
+%%% Fig PPC
+% TODO update fixed, Wuhan
+% generated with ppc.sh
+\begin{figure*}[h]	
+    \centering
+	\caption*{Austria}
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Austria_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Austria_ppc_long.pdf}
+    \end{subfigure}
+    %%%
+	\caption*{Belgium}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Belgium_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Belgium_ppc_long.pdf}
+    \end{subfigure}
+    %%%
+	\caption*{Denmark}    
+     \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Denmark_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Denmark_ppc_long.pdf}
+    \end{subfigure}
+	%%%
+	\caption*{France}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/France_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/France_ppc_long.pdf}
+    \end{subfigure}
+\end{figure*}
+    %%%
+\begin{figure*}
+    \ContinuedFloat 
+	\caption*{Germany}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Germany_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Germany_ppc_long.pdf}
+    \end{subfigure}
+    %%% 
+	\caption*{Italy}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Italy_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Italy_ppc_long.pdf}
+    \end{subfigure}
+    %%% 
+	\caption*{Norway}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Norway_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Norway_ppc_long.pdf}
+    \end{subfigure}
+    %%%  
+	\caption*{Spain}     
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Spain_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Spain_ppc_long.pdf}
+    \end{subfigure}
+\end{figure*}
+    %%%
+\begin{figure*}
+    \ContinuedFloat 
+	\caption*{Sweden}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Sweden_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Sweden_ppc_long.pdf}
+    \end{subfigure}
+    %%%  
+	\caption*{Switzerland}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Switzerland_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/Switzerland_ppc_long.pdf}
+    \end{subfigure}
+    %%%  
+	\caption*{United Kingdom}    
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/fixed/United_Kingdom_ppc_long.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/United_Kingdom_ppc_long.pdf}
+    \end{subfigure}
+    \caption{    
+    \textbf{Posterior prediction plots for 11 European countries.}
+    The vertical dashed line represents Apr~11, 2020.
+    Circles and stars represent daily case data up to and after Apr~11, respectively
+    Black and white arrows denote the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs, respectively.
+    Black lines represent a smoothing of the data points using a Savitzky-Golay filter with window length 3. 
+    Coloured lines represent posterior predictions from a model with fixed $\tau$ (blue) and free $\tau$ (green).
+Models were fitted with data up to Apr~11.
+The predictions are generated by drawing 1,000 parameters sets from the posterior distribution, and then generating a daily case count using the SEIR model (\autoref{eq:model}) up to Apr~24.
+Note the differences in the y-axis scale.
+    Posterior predictions with the free $\tau$ model predict the out-of-sample data well for all countries except Denmark and Sweden, but poorly with the fixed $\tau$ model.
+    The predictions of the model without $\tau$ (not shown) are even worse.
+    }
+    \label{fig:ppc}	    
+\end{figure*}
 
@@
 
-%%% Fig PPC
-% generated with ppc.sh
+%%% Fig Autocorrelation
+% generated with Fig_autocorr.py
+% TODO update after changes requested email 20 Jan
 \begin{figure}[h]
     \centering
-	\subcaption{Austria}
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Austria_ppc_long.pdf}
-    \end{subfigure}
-    \label{fig:ppc}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Austria_ppc_long.pdf}
-    \end{subfigure}
-    %%%
-	\subcaption{Belgium}    
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Belgium_ppc_long.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Belgium_ppc_long.pdf}
-    \end{subfigure}
-    %%%
-	\subcaption{Denmark}    
-     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Denmark_ppc_long.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Denmark_ppc_long.pdf}
-    \end{subfigure}
-	%%%
-	\subcaption{France}    
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/France_ppc_long.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/France_ppc_long.pdf}
-    \end{subfigure}
-\end{figure}
-    %%%
-\begin{figure}
-    \ContinuedFloat 
-	\subcaption{Germany}    
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Germany_ppc_long.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Germany_ppc_long.pdf}
-    \end{subfigure}
-    %%% 
-	\subcaption{Italy}    
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Italy_ppc_long.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Italy_ppc_long.pdf}
-    \end{subfigure}
-    %%% 
-	\subcaption{Norway}    
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Norway_ppc_long.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Norway_ppc_long.pdf}
-    \end{subfigure}
-    %%%  
-	\subcaption{Spain}     
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Spain_ppc_long.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Spain_ppc_long.pdf}
-    \end{subfigure}
-\end{figure}
-    %%%
-\begin{figure}
-    \ContinuedFloat 
-	\subcaption{Sweden}    
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Sweden_ppc_long.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Sweden_ppc_long.pdf}
-    \end{subfigure}
-    %%%  
-	\subcaption{Switzerland}    
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/Switzerland_ppc_long.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Switzerland_ppc_long.pdf}
-    \end{subfigure}
-    %%%  
-	\subcaption{United Kingdom}    
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/fixed/United_Kingdom_ppc_long.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/United_Kingdom_ppc_long.pdf}
-    \end{subfigure}
+	\includegraphics[width=\textwidth]{../figures/Fig-autocorr.pdf}
     \caption{
-    \textbf{Figure S4. Posterior prediction plots for 11 European countries.}
-    The vertical dashed line represents Apr~11, 2020.
-    Circles and stars represent daily case data up to and after Apr~11, respectively
-    Black and white arrows denote the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs, respectively.
-    Black lines represent a smoothing of the data points using a Savitzky-Golay filter with window length 3. 
-    Coloured lines represent posterior predictions from a model with fixed $\tau$ (blue) and free $\tau$ (green).
-Models were fitted with data up to Apr~11.
-The predictions are generated by drawing 1,000 parameters sets from the posterior distribution, and then generating a daily case count using the SEIR model up to Apr~24.
-Note the differences in the y-axis scale.
-    Posterior predictions with the free $\tau$ model predict the out-of-sample data well for all countries except Denmark and Sweden, but poorly for the fixed $\tau$ model.
-    The predictions of the model without $\tau$ (not shown) are even worse.
+    \textbf{Integrated autocorrelation time.}
+    The integrated autocorrelation time (IAT)~\citep{Foreman-Mackey2013,Goodman2010}, averaged across model parameters and MCMC chains, as a function of the number of samples in the chains.
+    IAT is less than 200K in all cases, while chain length is 5M after burn-in period of 2M.
+With 50 chains per region, this gives a at least 1,250 uncorrelated samples for estimating the posterior distribution.
     } 
+    \label{fig:autocorr}
 \end{figure}

### 2021-01-20 ms/ms.tex
add trace plots

@@
 It was found to be between XXX for Norway and XXX for Germany (\autoref{fig:autocorr}). % TODO this gives effective number of samples...
-We examined the trace plots for $\tau$ in all regions (FIGURE).
+We examined the trace plots for $\tau$ in all regions (\autoref{fig:trace}).
 All chains seem to converge to the same stationary distribution, in most cases before 2M steps. Thus, we discarded the first 2M samples as burnin samples.
@@
 
+%%% Fig Trace plots
+%%% Fig PPC
+% TODO update fixed, Wuhan
+% generated with ppc.sh
+\begin{figure}[p]	
+    \centering
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Austria_trace3.png}
+    \end{subfigure}
+    ~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Belgium_trace3.png}
+    \end{subfigure}
+    ~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Denmark_trace3.png}
+    \end{subfigure}
+	~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/France_trace3.png}
+    \end{subfigure}
+    ~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Germany_trace3.png}
+    \end{subfigure}
+    ~%%% 
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Italy_trace3.png}
+    \end{subfigure}
+    ~%%% 
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Norway_trace3.png}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Spain_trace3.png}
+    \end{subfigure}
+    ~%%%
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Sweden_trace3.png}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Switzerland_trace3.png}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/United_Kingdom_trace3.png}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Wuhan_trace3.png}
+    \end{subfigure}
+    \caption{    % TODO update according to requests from Ilia Jan 20 email
+    \textbf{Trace plots for $\tau$.}
+    The value of $\tau$ at each iteration (i.e.
+sample or step) of 50 chains per region.
+Each chain is plotted with a different color.
+In France and Austria some chains converged relatively late but before the burn-in period (2M iterations).
+In Spain, a single chain converged very late (after roughly 6M iterations).
+This chain was removed from further analysis.
+The y-axis range is set to the support of the prior on $\tau$.
+    }
+    \label{fig:trace}	    
+\end{figure}
+
+% TODO tau-lambda density plots
+
 \end{document}

### 2021-01-20 ms/ms.tex
finalized autocorr

@@
 
-The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using the affine-invariant ensemble sampler for Markov chain Monte Carlo (MCMC)~\citep{Goodman2010} implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
-We used the default configuration using the stretch move with stretch scale parameter $a=2$.
+The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using the affine-invariant ensemble sampler for Markov chain Monte Carlo (MCMC)~\citep{Goodman2010}, implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
+We use the default configuration using the stretch move with stretch scale parameter $a=2$.
 For the main analysis we use 50 chains (or walkers) per region, with 7M samples per chain (no thinning was applied).
-% TODO add refs and values
-The \emph{integrated autocorrelation time}~\citep{Foreman-Mackey2013,Goodman2010} was averaged across parameters and chains for each region. 
-It was found to be between XXX for Norway and XXX for Germany (\autoref{fig:autocorr}). % TODO this gives effective number of samples...
+The \emph{integrated autocorrelation time} (IAT)~\citep{Foreman-Mackey2013,Goodman2010} was averaged across parameters and chains for each region. 
+The IAT was between 40K for Norway and 187K for Germany~(\autoref{fig:autocorr}).
 We examined the trace plots for $\tau$ in all regions (\autoref{fig:trace}).
-All chains seem to converge to the same stationary distribution, in most cases before 2M steps. Thus, we discarded the first 2M samples as burnin samples.
-The only exception is Spain, in which a single chain converged at around X steps.
-We considered this chain as part of the burnin and removed it from the analysis.
-We also ran additional chains with 2M samples and a different initialization (i.e.
+All chains seem to converge to the same stationary distribution, in most cases before 2M steps. Thus, we discarded the first 2M samples as burn-in samples.
+The only exception is Spain, in which a single chain converged at around 6M steps.
+We considered this chain as part of the burn-in and removed it from the analysis.
+Therefore, 50 chains with 5M samples and IAT between 40K and 187K give an effective number of uncorrelated samples between 1,335 and 6,250.
+We ran additional chains with 2M samples and a different initialization (i.e.
 seed).
@@
 Thin and bold lines show 95\% and 75\% credible intervals (HDI~\citep{Kruschke2015}), respectively.
-Inference performed similarly to main inference, but with data up to Mar~28 and only 1M samples per chain with 600K burnin.
+Inference performed similarly to main inference, but with data up to Mar~28 and only 1M samples per chain with 600K burn-in.
     }
@@
 %%% Fig PPC
-% TODO update fixed, Wuhan
+% TODO update fixed
 % generated with ppc.sh
@@
     \caption{
-    \textbf{Integrated autocorrelation time.}
-    The integrated autocorrelation time (IAT)~\citep{Foreman-Mackey2013,Goodman2010}, averaged across model parameters and MCMC chains, as a function of the number of samples in the chains.
-    IAT is less than 200K in all cases, while chain length is 5M after burn-in period of 2M.
-With 50 chains per region, this gives a at least 1,250 uncorrelated samples for estimating the posterior distribution.
+    \textbf{Integrated autocorrelation time (IAT)}~\citep{Foreman-Mackey2013,Goodman2010}, averaged across model parameters and MCMC chains, as a function of the number of samples in the chains.
+    IAT is less than 187K in all cases, while chain length is 5M after burn-in period of 2M.
+With 50 chains per region, this gives a at least 1,335 uncorrelated samples for estimating the posterior distribution.
     } 

### 2021-01-20 ms/ms.tex
update trace plots and todos

@@
 % generated with Fig_autocorr.py
-% TODO update after changes requested email 20 Jan
 \begin{figure}[h]
@@
 %%% Fig Trace plots
-%%% Fig PPC
-% TODO update fixed
-% generated with ppc.sh
 \begin{figure}[p]	
@@
     \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Austria_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/Austria_trace.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Belgium_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/Belgium_trace.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Denmark_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/Denmark_trace.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/France_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/France_trace.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Germany_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/Germany_trace.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Italy_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/Italy_trace.pdf}
     \end{subfigure}
@@
 	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Norway_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/Norway_trace.pdf}
     \end{subfigure}
@@
 	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Spain_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/Spain_trace.pdf}
     \end{subfigure}
@@
 	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Sweden_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/Sweden_trace.pdf}
     \end{subfigure}
@@
 	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Switzerland_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/Switzerland_trace.pdf}
     \end{subfigure}
@@
 	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/United_Kingdom_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/United_Kingdom_trace.pdf}
     \end{subfigure}
@@
 	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Wuhan_trace3.png}
+        \includegraphics[width=\textwidth]{../figures/traces/Wuhan_trace.pdf}
     \end{subfigure}
-    \caption{    % TODO update according to requests from Ilia Jan 20 email
+    \caption{
     \textbf{Trace plots for $\tau$.}
-    The value of $\tau$ at each iteration (i.e.
-sample or step) of 50 chains per region.
-Each chain is plotted with a different color.
+    The value of $\tau$ at for sequential samples of 50 chains per region.
+To facilitate visualization, chains were thinned 1:10,000 for the trace plots, but not for posterior estimation.
+Line transparency was set at $\alpha=0.1$ and chains cycle through different colors.
 In France and Austria some chains converged relatively late but before the burn-in period (2M iterations).
-In Spain, a single chain converged very late (after roughly 6M iterations).
-This chain was removed from further analysis.
-The y-axis range is set to the support of the prior on $\tau$.
+In Spain, a single chain converged very late (after roughly 6M iterations) and was removed from further analysis.
+The y-axis range is the support of the prior on $\tau$.
     }

### 2021-01-20 src/Fig_autocorr.py
more changes on plots

@@ -18,20 +18,22 @@ class FixedScalarFormatter(matplotlib.ticker.ScalarFormatter):
 def plot_auto(country_name): 
     file_name = '7M/inference/{}.autocorr'.format(country_name)
     auto = np.loadtxt(file_name)
-    iters = np.array([0]+[i*200000 for i in range(1,len(auto)+1)])
-    taus = np.array([0]+list(auto))
+    iters = np.array([0]+[i*200000 for i in range(1,len(auto)+1)])/1_000_000
+    taus = np.array([0]+list(auto))/1000
+    print(max(taus))
 #     plt.plot(iters, iters/100.0,'--k')
     plt.plot(iters, taus, label=country_name)
     plt.xlabel("number of iterations")
     plt.ylabel(r"auto-correlation mean");
-    plt.legend(loc='center left', bbox_to_anchor=(1, 0.5))
+    lgd = plt.legend(loc='center left', bbox_to_anchor=(1, 0.5))
     plt.xlim(0,iters[-1])
+    return lgd
 
 for c in ['Austria','Belgium','Denmark','France','Germany','Italy','Norway',
           'Spain','Sweden','Switzerland','United_Kingdom','Wuhan']:
-    plot_auto(c)
+    lgd = plot_auto(c)
 plt.ylim(0)
 
-plt.gca().xaxis.set_major_formatter(FixedScalarFormatter(6)) #doesn't work :\\\\
-plt.gca().yaxis.set_major_formatter(FixedScalarFormatter(3)) 
-plt.gcf().savefig('7M/figures/auto-correlation.pdf')
+# plt.gca().xaxis.set_major_formatter(FixedScalarFormatter(6)) #doesn't work :\\\\
+# plt.gca().yaxis.set_major_formatter(FixedScalarFormatter(3)) 
+plt.gcf().savefig('7M/figures/Fig-autocorr.pdf',bbox_extra_artists=(lgd,), bbox_inches='tight')

### 2021-01-20 src/Fig_trace_tau.py

@@ -3,52 +3,39 @@ import matplotlib.pyplot as plt
 import matplotlib
 from matplotlib.ticker import ScalarFormatter
 import seaborn as sns
-sns.set_context('paper', font_scale=1.3)
 import sys
+from datetime import timedelta, datetime
+import pandas as pd
 
-def plot_τ_trace(path, country_name):
+sns.set_context('paper', font_scale=1.6)
+
+
+def τ_to_string(τ,start_date):
+    return (pd.to_datetime(start_date) + timedelta(days=τ)).strftime('%b %d')
+
+def plot_τ_trace(path, country_name, thin=10_000, alpha=0.1):
     file_name_npz = '{}/inference/{}.npz'.format(path, country_name)
     data = np.load(file_name_npz)
     chain = data['chain']
     var_names = list(data['var_names'])
+    start_date = data['start_date']
+    maxx = len(data['incidences'])-2
+    minn = 10
     
     fig, ax = plt.subplots()
-    plt.plot(chain[:,::100,var_names.index('τ')].T, color='k', alpha=0.15);
+    plt.plot(chain[:,::thin,var_names.index('τ')].T, alpha=alpha);
     labels = ['−10000', '0', '1', '2', '3', '4', '5', '6', '7', '8']
     ax.set_xticklabels(labels);
-    ax.text(x = 0.93, y = -0.15, s = '1e6', transform=ax.transAxes);
-    plt.xlim(-2000, 70000);
-    plt.xlabel('number of iterations')
+    plt.xlim(-2_00_000/thin, 7_000_000/thin);
+    plt.ylim(minn,maxx)
+    plt.xlabel('Samples (1M)')
     plt.ylabel('τ')
     plt.title(country_name);
+    days = np.linspace(minn,maxx,5)
+    labels = [τ_to_string(d,start_date) for d in days]
+    plt.yticks(days,labels);
     plt.tight_layout()
-    fig.savefig('{}/figures/{}_trace1.png'.format(path, country_name),dpi=100)
-    
-    fig, ax = plt.subplots()
-    plt.plot(chain[:,::100,var_names.index('τ')].T, color='k', alpha=0.05);
-    labels = ['−10000', '0', '1', '2', '3', '4', '5', '6', '7', '8']
-    ax.set_xticklabels(labels);
-    ax.text(x = 0.93, y = -0.15, s = '1e6', transform=ax.transAxes);
-    plt.xlim(-2000, 70000);
[…]

### 2021-01-20 src/plot_utils.py
fixing write_csv for fixed tau model

@@ -106,12 +106,22 @@ def write_csv_data(file_name):
             τ_median =  model.τ
             τ_MAP =  model.τ
             τ_MAP_i = model.τ
+            τ_hpd_75_from,τ_hpd_75_to =  model.τ, model.τ
+            τ_hpd_75_from_date = format(τ_to_string(model.τ))
+            τ_hpd_75_to_date = format(τ_to_string(model.τ))
+            τ_hpd_95_from_date = format(τ_to_string(model.τ))
+            τ_hpd_95_to_date = format(τ_to_string(model.τ))
         except AttributeError: #model without τ
             τ_posterior =np.array([0])
             τ_mean = 0
             τ_median =  0
             τ_MAP =  0
             τ_MAP_i = 0
+            τ_hpd_75_from,τ_hpd_75_to =  0,0
+            τ_hpd_75_from_date = 0
+            τ_hpd_75_to_date = 0
+            τ_hpd_95_from_date = 0
+            τ_hpd_95_to_date = 0
 
     τ_official_from1Jar = (pd.to_datetime(start_date) - pd.Timestamp('2020-01-01')).days + last_NPI
     τ_mean_from1Jar = (pd.to_datetime(start_date) - pd.Timestamp('2020-01-01')).days + τ_posterior.mean()

### 2021-01-20 src/plot_utils.py
fixing write_csv more

@@ -106,7 +106,7 @@ def write_csv_data(file_name):
             τ_median =  model.τ
             τ_MAP =  model.τ
             τ_MAP_i = model.τ
-            τ_hpd_75_from,τ_hpd_75_to =  model.τ, model.τ
+            τ_hpd_75_from, τ_hpd_75_to, τ_hpd_95_from,τ_hpd_95_to =  model.τ, model.τ, model.τ, model.τ           
             τ_hpd_75_from_date = format(τ_to_string(model.τ))
             τ_hpd_75_to_date = format(τ_to_string(model.τ))
             τ_hpd_95_from_date = format(τ_to_string(model.τ))
@@ -117,7 +117,7 @@ def write_csv_data(file_name):
             τ_median =  0
             τ_MAP =  0
             τ_MAP_i = 0
-            τ_hpd_75_from,τ_hpd_75_to =  0,0
+            τ_hpd_75_from, τ_hpd_75_to, τ_hpd_95_from, τ_hpd_95_to =  0,0
             τ_hpd_75_from_date = 0
             τ_hpd_75_to_date = 0
             τ_hpd_95_from_date = 0

### 2021-01-20 src/plot_utils.py
added real num of steps if emcee did less steps that planed

@@ -27,7 +27,7 @@ country_name = None
 def load_data(file_name, _country_name, nburn=2_000_000, lim_steps=None, delete_chain_less_than=None): 
     # it's the only global point. we initialize all the params here once and don't update it later (only when load_data again for different file_name)
     # TODO PLEASE DONT USE global anywhere in your code
-    global first_NPI, last_NPI, incidences, start_date, var_names, nsteps, ndim, N, Td1, Td2, ndays, sample, lnprobability, logliks, model_type, model, country_name
+    global first_NPI, last_NPI, incidences, start_date, var_names, realnsteps, nsteps, ndim, N, Td1, Td2, ndays, sample, lnprobability, logliks, model_type, model, country_name
     country_name = _country_name
     data = np.load(file_name)
     incidences = data['incidences']
@@ -36,6 +36,7 @@ def load_data(file_name, _country_name, nburn=2_000_000, lim_steps=None, delete_
     nsteps, ndim, N, Td1, Td2, model_type = data['params']
     chain = data['chain']
     nwalkers = chain.shape[0] #nwalkers before deleting bad chain
+    realnsteps = data['chain'].shape[1]
     ndays = len(incidences)        
 
     if delete_chain_less_than:
@@ -55,7 +56,7 @@ def load_data(file_name, _country_name, nburn=2_000_000, lim_steps=None, delete_
     #     None
 
     lnprobability = data['lnprobability'][:, nburn:lim_steps]
-    logliks = data['logliks'].reshape(nwalkers,nsteps)[:,nburn:lim_steps]
+    logliks = data['logliks'].reshape(nwalkers,realnsteps)[:,nburn:lim_steps]
     if delete_chain_less_than:
             lnprobability = np.delete(lnprobability, bad_chain_ind, axis=0)
             logliks = np.delete(logliks, bad_chain_ind, axis=0)
@@ -74,7 +75,7 @@ def write_csv_header(file_name):
 
     with open(file_name, mode='w') as file: # use pd to write csv files?
         writer = csv.writer(file, delimiter=',', quotechar='"', quoting=csv.QUOTE_MINIMAL)
[…]

### 2021-01-20 src/Fig_autocorr.py
improved Fig-autocorr

@@ -16,24 +16,25 @@ class FixedScalarFormatter(matplotlib.ticker.ScalarFormatter):
         self.orderOfMagnitude = self.power
         
 def plot_auto(country_name): 
-    file_name = '7M/inference/{}.autocorr'.format(country_name)
+    file_name = '7MFixed/inference/{}.autocorr'.format(country_name)
     auto = np.loadtxt(file_name)
     iters = np.array([0]+[i*200000 for i in range(1,len(auto)+1)])/1_000_000
     taus = np.array([0]+list(auto))/1000
-    print(max(taus))
+    print(max(taus),country_name)
 #     plt.plot(iters, iters/100.0,'--k')
     plt.plot(iters, taus, label=country_name)
-    plt.xlabel("number of iterations")
-    plt.ylabel(r"auto-correlation mean");
+    plt.xlabel("Samples (1M)")
+    plt.ylabel("Average IAT (1K)");
     lgd = plt.legend(loc='center left', bbox_to_anchor=(1, 0.5))
     plt.xlim(0,iters[-1])
     return lgd
 
 for c in ['Austria','Belgium','Denmark','France','Germany','Italy','Norway',
-          'Spain','Sweden','Switzerland','United_Kingdom','Wuhan']:
+          'Spain','Sweden','Switzerland','United_Kingdom']:
     lgd = plot_auto(c)
 plt.ylim(0)
 
 # plt.gca().xaxis.set_major_formatter(FixedScalarFormatter(6)) #doesn't work :\\\\
 # plt.gca().yaxis.set_major_formatter(FixedScalarFormatter(3)) 
-plt.gcf().savefig('7M/figures/Fig-autocorr.pdf',bbox_extra_artists=(lgd,), bbox_inches='tight')
+# plt.tight_layout()
+plt.gcf().savefig('7MFixed/figures/Fig-autocorr.pdf',bbox_extra_artists=(lgd,), bbox_inches='tight')

### 2021-01-21 src/Fig_autocorr.py
more changes on plots

@@ -14,23 +14,23 @@ class FixedScalarFormatter(matplotlib.ticker.ScalarFormatter):
 
     def _set_order_of_magnitude(self):
         self.orderOfMagnitude = self.power
-        
+
 def plot_auto(country_name): 
     file_name = '7MFixed/inference/{}.autocorr'.format(country_name)
     auto = np.loadtxt(file_name)
     iters = np.array([0]+[i*200000 for i in range(1,len(auto)+1)])/1_000_000
     taus = np.array([0]+list(auto))/1000
-    print(max(taus),country_name)
+    print(max(taus),country_name, len(auto))
 #     plt.plot(iters, iters/100.0,'--k')
     plt.plot(iters, taus, label=country_name)
     plt.xlabel("Samples (1M)")
     plt.ylabel("Average IAT (1K)");
     lgd = plt.legend(loc='center left', bbox_to_anchor=(1, 0.5))
-    plt.xlim(0,iters[-1])
+    plt.xlim(0,7)
     return lgd
 
 for c in ['Austria','Belgium','Denmark','France','Germany','Italy','Norway',
-          'Spain','Sweden','Switzerland','United_Kingdom']:
+          'Spain','Sweden','Switzerland','United_Kingdom','Wuhan']:
     lgd = plot_auto(c)
 plt.ylim(0)
 

### 2021-01-21 src/Fig_ppc.py

@@ -128,7 +128,8 @@ if __name__ == '__main__':
     ndays = len(X)
     
     chain_fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
-    delete_chain_less_than = 15 if country=='Spain' else None
+    if job_id=='7M' and country=='Spain':  #TODO make input parameter
+        delete_chain_less_than = 15 
     chain, _, Td1, Td2, model_type, _, start_date, N = load_chain(fname=chain_fname,delete_chain_less_than=delete_chain_less_than)
     X_mean = scipy.signal.savgol_filter(X, 3, 1)
     

### 2021-01-21 src/Re.py

@@ -28,28 +28,7 @@ from inference import find_start_day
 from model.normal_prior_model import NormalPriorModel
 from model.fixed_tau_model import FixedTauModel
 from sklearn.metrics import mean_squared_error
-
-
-def load_chain(job_id=None, fname=None, burn_fraction=0.6):
-	with spinner():
-		if fname is None:
-			fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
-		inference_data = np.load(fname)
-		chain = inference_data['chain']
-		var_names = inference_data['var_names']
-		nsteps, ndim, N, Td1, Td2, model_type = inference_data['params']
-		X = inference_data['incidences']
-		start_date = inference_data['start_date']
-		logliks = inference_data['logliks']
-		# print("Loaded {} with parameters:".format(fname))
-		# print(var_names)
-		nchains, nsteps, ndim = chain.shape
-		nburn = int(nsteps * burn_fraction)
-		chain = chain[:, nburn:, :]
-		chain = chain.reshape((-1, ndim))
-		logliks = logliks.reshape((nchains, nsteps))
-		logliks = logliks[:, nburn:].ravel()
-		return chain, logliks, Td1, Td2, model_type, X, start_date, N
+from Fig_ppc import load_chain
 
 def posterior_Re(chain, nreps):
 	# params: ['Z', 'D', 'μ', 'β', 'α1', 'λ', 'α2', ...]
@@ -61,7 +40,6 @@ def posterior_Re(chain, nreps):
 def Re(α, β, D, μ):
     # Li et al 2020, SI pg 4
     return α*β*D + (1-α)*μ*β*D 
-
 
 if __name__ == '__main__':
 	nreps = 1000
@@ -77,6 +55,9 @@ if __name__ == '__main__':
 		print(Re_file, "already exists")
 	else:
 		chain_fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
+		delete_chain_less_than = None
+		if job_id=='7M' and country=='Spain':  #TODO make input parameter
+			delete_chain_less_than = 15 
 		chain, _, _, _, model_type, _, _, _ = load_chain(fname=chain_fname)
 		Re_pre, Re_post = posterior_Re(chain, nreps)
 		rel_reduc_Re = 1 - Re_post / Re_pre

### 2021-01-21 src/Fig_Re.py
added hdi to fig re

@@ -8,7 +8,7 @@ import seaborn as sns
 sns.set_context('paper', font_scale=1.3)
 red, blue, green = sns.color_palette('Set1', 3)
 import pycountry # https://pypi.org/project/pycountry/
-
+from arviz import hdi
 import os
 from glob import glob
 from datetime import datetime, timedelta
@@ -19,8 +19,8 @@ colors = sns.color_palette('Paired')
 
 
 if __name__ == '__main__':
-    job_id_free = '2020-05-26-Apr11'
-    job_id_fixed = '2020-06-25-Apr11-fixedtau'
+    job_id_free = '7M'
+    job_id_fixed = '7MFixed'
 
     dfs = []
     countries = ['Austria', 'Belgium', 'Denmark', 'France', 'Germany', 'Italy', 'Norway', 'Spain', 'Sweden', 'Switzerland', 'United Kingdom']    
@@ -39,9 +39,9 @@ if __name__ == '__main__':
     df = pd.concat(dfs)
     grp = df.groupby(['country', 'model'])
     agg = grp.agg(
-        low=pd.NamedAgg(column="rel_reduc_Re", aggfunc=lambda x: np.percentile(x, 25)),
+        low=pd.NamedAgg(column="rel_reduc_Re", aggfunc=lambda x: hdi(x, 0.75)[0]),
         median=pd.NamedAgg(column="rel_reduc_Re", aggfunc=np.median),
-        high=pd.NamedAgg(column="rel_reduc_Re", aggfunc=lambda x: np.percentile(x, 75))
+        high=pd.NamedAgg(column="rel_reduc_Re", aggfunc=lambda x: hdi(x, 0.75)[1])
     ).reset_index()
     df_free = agg[agg['model'] == 'free']
     df_fixed = agg[agg['model'] == 'fixed']

### 2021-01-21 src/Fig_Re.py
small fixes

@@ -39,9 +39,9 @@ if __name__ == '__main__':
     df = pd.concat(dfs)
     grp = df.groupby(['country', 'model'])
     agg = grp.agg(
-        low=pd.NamedAgg(column="rel_reduc_Re", aggfunc=lambda x: hdi(x, 0.75)[0]),
+        low=pd.NamedAgg(column="rel_reduc_Re", aggfunc=lambda x: hdi(np.array(x), 0.50)[0]),
         median=pd.NamedAgg(column="rel_reduc_Re", aggfunc=np.median),
-        high=pd.NamedAgg(column="rel_reduc_Re", aggfunc=lambda x: hdi(x, 0.75)[1])
+        high=pd.NamedAgg(column="rel_reduc_Re", aggfunc=lambda x: hdi(np.array(x), 0.50)[1])
     ).reset_index()
     df_free = agg[agg['model'] == 'free']
     df_fixed = agg[agg['model'] == 'fixed']

### 2021-01-21 src/Table-RMSE.py

@@ -1,7 +1,7 @@
 import os
 import pandas as pd
 
-no_fixed_free = ['2020-06-27-Apr11-notau', '2020-06-25-Apr11-fixedtau', '2020-05-26-Apr11']
+no_fixed_free = ['7MNoTau', '7MFixed', '7M']
 main_folder = r'../output'
 nodf, fixeddf, freedf = [pd.read_csv(os.path.join(main_folder, inner_folder, 'figures', 'ppc_rmse.csv'),sep='\t',header=None) 
        for inner_folder in no_fixed_free]

### 2021-01-21 src/Table-WAIC.py

@@ -22,14 +22,13 @@ sns.set_context('paper', font_scale=1.3)
 red, blue, green = sns.color_palette('Set1', 3)
 
 
-def load_chain(job_id, country, burn_fraction=0.6):
+def load_chain(job_id, country, nburn=2_000_000):
     fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
     inference_data = np.load(fname)
     nsteps, ndim, N, Td1, Td2, model_type = inference_data['params']
     logliks = inference_data['logliks']
     nchains = logliks.size // nsteps
     logliks = logliks.reshape(nchains, nsteps)
-    nburn = int(nsteps*burn_fraction)
     logliks = logliks[:, nburn:]
     return logliks
 
@@ -59,7 +58,7 @@ def WAIC(logliks):
 if __name__ == '__main__':    
     output_folder = r'../output'
     
-    job_ids = ['2020-05-26-Apr11', '2020-06-27-Apr11-notau', '2020-06-25-Apr11-fixedtau']
+    job_ids = ['7M', '7MNoTau', '7MFixed']
     print('Job IDs:')
     print(job_ids)    
     countries = 'Austria Belgium Denmark France Germany Italy Norway Spain Sweden Switzerland United_Kingdom Wuhan'.split(' ')
@@ -79,9 +78,9 @@ if __name__ == '__main__':
             ))
 
     df = pd.DataFrame(results)
-    df.loc[df['job_id'] == '2020-05-26-Apr11', 'job_id'] = 'Free'
-    df.loc[df['job_id'] == '2020-06-27-Apr11-notau', 'job_id'] = 'No'
-    df.loc[df['job_id'] == '2020-06-25-Apr11-fixedtau', 'job_id'] = 'Fixed'
+    df.loc[df['job_id'] == '7M', 'job_id'] = 'Free'
+    df.loc[df['job_id'] == '7MNoTau', 'job_id'] = 'No'
+    df.loc[df['job_id'] == '7MFixed', 'job_id'] = 'Fixed'
     df = df.rename(columns={'country':'Country', 'job_id':'Model'})
     df['Country'] = [x.replace('_', ' ') for x in df['Country']]
     df.loc[df['Country']=='Wuhan', 'Country'] = 'Wuhan China'

### 2021-01-21 ms/ms.tex
update Re, added joint plots; edited text accordingly

@@
 
-Keywords: SEIR, COVID-19, public health, epidemic, infectious disease, NPI
-
 %\pagebreak
@@
 We demonstrate that differences between the official and effective start of NPIs can lead to under-estimating their impact, and discuss potential causes and consequences of our results.
+\\\\
+{\small
+Keywords: SEIR, COVID-19, public health, epidemic, infectious disease, NPI
+}
 \end{abstract}
@@
 All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~24, and these predictions were then compared to the real case data. 
-% TODO update when ppc fig is ready
 The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
+%TODO update when WAIC and RMSE are ready
 In all 11 of the European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
 When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark (\autoref{table:WAIC}).
+Indeed, when examining the joint posterior distribution of $\tau$ and $\lambda$ (the latter is the reduction in transmission due to NPIs, see \autoref{eq:model}), high values of $\lambda$ are estimated in Denmark and Sweden (0.7 and 0.74, respectively; see~\autoref{table:estimated-params}), probably making the inference of $\tau$ more difficult due to unidentifiability, resulting in a wide posterior for $\tau$.
+Some correlation between the parameters is also evident in Norway, where $\lambda$ estimates are low (0.31).
 Notably, the data for Sweden and Denmark do not have a single "peak" during the evaluated dates, possibly leading to wide 95\% credible intervals on $\hat{\tau}$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).
@@
 
-In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\% CI; \autoref{fig:late}).
+In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\% CI), although there is low but noticeable posterior probability for Jan~24-25 (\autoref{fig:late}).
+Higher estimates of $\tau$ (Jan~4) correspond to a wide posterior for $\lambda$ (\autoref{fig:joint}).
 
@@
 %TODO check WAIC
-Indeed, Denmark and Sweden had the highest estimates of $\lambda$, that is, the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries.
-Moreover, in Sweden the number of daily cases continues to grow up to Apr~11, rather then "peak" (\autoref{fig:ppc}). 
+Indeed, Denmark and Sweden had the highest estimates of $\lambda$, that is, the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (see also joint posterior distributions in \autoref{fig:joint}).
+Moreover, in Sweden the number of daily cases continues to grow up to Apr~11, rather than "peak" (\autoref{fig:ppc}). 
 In Denmark, the opposite occurs: there are seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
@@
 The uncertainty in the estimate of the effective start of NPIs might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is very narrow (\autoref{fig:early}).
+There may also be an unidentifiability issue, as there is some negative correlation between the estimates of $\tau$ and $\lambda$ (\autoref{fig:joint}).
 
@@
 That is, we compare the impact estimate assuming that NPIs started at the official date $\tau^*$, versus the estimate when inferring the effective start of NPIs from the data.
-\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model are consistently lower than estimates from the free~$\tau$ model, except in Sweden and Denmark, in which estimation uncertainty is high.
-%TODO update when fig Re is ready
+\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model (y-axis) are consistently lower than estimates from the free~$\tau$ model (x-axis), except in Sweden and Denmark, in which estimation uncertainty is high.
 These results suggest that the impact of past NPIs is likely to be under-estimated by health officials and researchers if they assume NPIs effectively start at their official dates.
@@
     \textbf{Impact of NPIs is under-estimated when assuming they start at the official date.}
-    Shown are estimates of the relative reductions in $R$ (the effective reproduction number), which measures the impact of NPIs on disease transmission.
+    Shown are estimates of the relative reductions in $R$ (the effective reproduction number), which measures the impact of NPIs on disease transmission, in 11 European countries.
      The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed $\tau$); the x-axis shows estimates when inferring the effective start of NPIs from the data (free $\tau$).
@@
 As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with new guidelines, whereas in other cases the population will adopt either restrictions or relaxations even before they are formally announced.
-Attempts to asses the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date, and that a significant change in the dynamics can be observed roughly seven days after the start of NPIs (due to the characteristic serial interval of \covid\citep{Ali2020}).
+Attempts to assess the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date, and that a significant change in the dynamics can be observed roughly seven days after the start of NPIs (due to the characteristic serial interval of \covid\citep{Ali2020}).
 However, late and early effective start of NPIs, such as we have inferred, can bias these assessments and lead to wrong conclusions about the impact of NPIs (Figures~\ref{fig:Re} and~\ref{fig:prediction}).
@@
 The uninformative prior resulted in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom. 
-%This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
 We therefore decided to use the more informative truncated normal prior for $\tau$.
@@
 The \emph{integrated autocorrelation time} (IAT)~\citep{Foreman-Mackey2013,Goodman2010} was averaged across parameters and chains for each region. 
-The IAT was between 40K for Norway and 187K for Germany~(\autoref{fig:autocorr}).
+The IAT was between 32K for Wuhan, China, and 187K for Germany~(\autoref{fig:autocorr}).
 We examined the trace plots for $\tau$ in all regions (\autoref{fig:trace}).
@@
 We considered this chain as part of the burn-in and removed it from the analysis.
-Therefore, 50 chains with 5M samples and IAT between 40K and 187K give an effective number of uncorrelated samples between 1,335 and 6,250.
+Therefore, 50 chains with 5M samples and IAT between 40K and 187K give an effective samples size between 1,335 and 7,812.
 We ran additional chains with 2M samples and a different initialization (i.e.
@@
 The estimated posterior distributions of $\tau$ were very similar to our main analysis, further increasing our confidence in the convergence of our inference.
-We also examined the joint posterior distributions of $\tau$ and $\lambda$ to check for correlations (FIGURE). %TODO what do we find?
+We also examined the joint posterior distributions of $\tau$ and $\lambda$ (\autoref{fig:joint}).
 
@@
 \textbf{Acknowledgements.} 
-We thank Yinon M. Bar-On, Lilach Hadany, and Oren Kolodny for discussions and comments. % Tim CD Lucas
+We thank Yinon M. Bar-On, Lilach Hadany, Oren Kolodny, and Zohar Yakhini. % Tim CD Lucas
 } % small
@@
 %%% Fig PPC
-% TODO update fixed
 % generated with ppc.sh
@@
 
+%%% Fig Joint density plots
+\begin{figure}[p]	
+    \centering
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/Austria_joint.pdf}
+    \end{subfigure}
+    ~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/Belgium_joint.pdf}
+    \end{subfigure}
+    ~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/Denmark_joint.pdf}
+    \end{subfigure}
+	~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/France_joint.pdf}
+    \end{subfigure}
+    ~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/Germany_joint.pdf}
+    \end{subfigure}
+    ~%%% 
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/Italy_joint.pdf}
+    \end{subfigure}
+    ~%%% 
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/Norway_joint.pdf}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/Spain_joint.pdf}
+    \end{subfigure}
+    ~%%%
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/Sweden_joint.pdf}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/Switzerland_joint.pdf}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/United_Kingdom_joint.pdf}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/joint/Wuhan_joint.pdf}
+    \end{subfigure}
+    \caption{
+    \textbf{Joint posterior density plots for $\tau$ and $\lambda$.} The high values of $\lambda$ estimated in Denmark and Sweden reduce the effect of the NPIs thereby making the inference of $\tau$ more difficult, resulting in wide posterior distributions.
+A correlation between the parameters is also evident in Norway.
+In Belgium and in Wuhan a later $\tau$ gives a wide estimate for $\lambda$.
+In comparison, Austria, France, Germany, Italy, Spain, Switzerland, and UK have a narrow joint distribution.
+    }
+    \label{fig:joint}
+\end{figure}
+
 %%% Fig Fr-vs-Es
@@
 
-% TODO tau-lambda density plots
-
 \end{document}

### 2021-01-21 ms/ms.tex
UO comments

@@
 \affil[1]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya 4610101, Israel}
-\affil[2]{School of Public Health, Tel Aviv University, Tel Aviv 6997801, Israel}{
-\affil[3]{Porter School of the Environment and Earth Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
+\affil[2]{School of Public Health, Faculty of Medicine, Tel Aviv University, Tel Aviv 6997801, Israel}{
+\affil[3]{Porter School of the Environment and Earth Sciences, Faculty of Exact Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
 \affil[4]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv 6997801, Israel}
@@
 When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark (\autoref{table:WAIC}).
-Indeed, when examining the joint posterior distribution of $\tau$ and $\lambda$ (the latter is the reduction in transmission due to NPIs, see \autoref{eq:model}), high values of $\lambda$ are estimated in Denmark and Sweden (0.7 and 0.74, respectively; see~\autoref{table:estimated-params}), probably making the inference of $\tau$ more difficult due to unidentifiability, resulting in a wide posterior for $\tau$.
-Some correlation between the parameters is also evident in Norway, where $\lambda$ estimates are low (0.31).
-Notably, the data for Sweden and Denmark do not have a single "peak" during the evaluated dates, possibly leading to wide 95\% credible intervals on $\hat{\tau}$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).
+Indeed, we estimate low effect of NPIs on transmission in Denmark and Sweden (i.e. $\lambda=0.7$ and $0.74$, respectively; see~\autoref{table:estimated-params}).
+This may interfere with the inference of $\tau$ due to unidentifiability. 
+Notably, the data for Sweden and Denmark do not have a single "peak" during the evaluated dates, possibly leading to wide credible intervals on $\tau$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).
 
@@
 The exceptions are Norway and Switzerland (see below).
-The former also has a wide 95\% credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
+The former also has a wide credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
 
@@
 
-In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\% CI), although there is low but noticeable posterior probability for Jan~24-25 (\autoref{fig:late}).
-Higher estimates of $\tau$ (Jan~4) correspond to a wide posterior for $\lambda$ (\autoref{fig:joint}).
+In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\% CI).
+Yet, there is low but noticeable posterior probability for Jan~25 (\autoref{fig:late}), for which the effect of NPIs on transmission is considerably lower (\autoref{fig:joint}).
 
@@
 A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95 \%CI, \autoref{fig:late}).
-
-Similarly, in France we estimate the effective start of NPIs $\hat{\tau}$ at Mar~24 ($\substack{+2.65 \\ -1.44}$~days 95\% CI, \autoref{fig:late}).
+Nevertheless, we estimate the effective start of NPI eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95 \%CI, \autoref{fig:late}).
+
+Similarly, in France we estimate the effective start of NPIs at Mar~24 ($\substack{+2.65 \\ -1.44}$~days 95\% CI, \autoref{fig:late}).
 This is a week later than the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
@@
 Therefore, when we refer to this case as "early", we mean that the effective date (Mar~19) occurs \emph{before} the official lockdown date (Mar~22), not that it occurs before all NPIs.
-Interestingly, Germany has the second longest duration between first and last NPIs after Norway (10 and 12 days respectively; \autoref{table:NPI_dates}).
-However, the credible interval for the effective date in Germany is narrow (1.91 days), whereas it is very wide in Norway (17.61 days).
-That $\tau$ is significantly earlier than $\tau^*$ can suggest that early NPIs in Germany more effectively reduced transmission rates compared to other countries.
+Interestingly, Germany had the second longest duration between first and last NPIs after Norway (10 and 12 days respectively; \autoref{table:NPI_dates}).
+However, the credible interval for the effective start date in Germany is narrow (1.91 days), whereas it is very wide in Norway (17.61 days).
+The significantly earlier estimate of $\hat{\tau}$ relative to $\tau^*$ can suggest that early NPIs in Germany more effectively reduced transmission rates compared to other countries.
 Another possible interpretation is that the German population anticipated the lockdown and began to act accordingly before it started.
+% TODO rephrase
 
@@
 %TODO check WAIC
-Indeed, Denmark and Sweden had the highest estimates of $\lambda$, that is, the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (see also joint posterior distributions in \autoref{fig:joint}).
-Moreover, in Sweden the number of daily cases continues to grow up to Apr~11, rather than "peak" (\autoref{fig:ppc}). 
-In Denmark, the opposite occurs: there are seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
+Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries ().%TODO ref to table
+Moreover, in Sweden the number of daily cases continued to grow up to Apr~11, rather than "peak" (\autoref{fig:ppc}). 
+In Denmark, the opposite occurred: there were seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
 We suspect that these missing and additional "peaks" increase the uncertainty in our inference.
@@
 However, the posterior distribution is very wide ($\substack{+5.59 \\ -12.03}$~days 95\% CI): it covers the range between Mar~10, two days before the first NPI, and Mar~27, three days after the last NPI (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
-The uncertainty in the estimate of the effective start of NPIs might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, but the corresponding posterior distribution is very narrow (\autoref{fig:early}).
-There may also be an unidentifiability issue, as there is some negative correlation between the estimates of $\tau$ and $\lambda$ (\autoref{fig:joint}).
+The high uncertainty might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, and the corresponding posterior distribution is very narrow (\autoref{fig:early}).
+There may also be an unidentifiability issues between $\tau$ and the effect of NPIs on transmission (\autoref{fig:joint}).
 
@@
 
+Interestingly, the effective start of NPIs in France and Spain is estimated to have started on Mar~24 and~25, respectively,
+although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
+The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
+This may suggest correlations between effective starts of NPIs due to global or international events.
+
 As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (Tables~\ref{table:WAIC} and~\ref{table:RMSE}).
@@
 However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementations of NPIs~\citep{Flaxman2020}.
-
-Interestingly, the effective start of NPIs in France and Spain is estimated to have started on Mar~24 and~25, respectively,
-although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
-The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
-This may suggest correlations between effective starts of NPIs due to global or international events.
 
@@
 The \emph{integrated autocorrelation time} (IAT)~\citep{Foreman-Mackey2013,Goodman2010} was averaged across parameters and chains for each region. 
-The IAT was between 32K for Wuhan, China, and 187K for Germany~(\autoref{fig:autocorr}).
+The average IAT was between 32K for Wuhan, China, and 187K for Germany~(\autoref{fig:autocorr}).
 We examined the trace plots for $\tau$ in all regions (\autoref{fig:trace}).
-All chains seem to converge to the same stationary distribution, in most cases before 2M steps. Thus, we discarded the first 2M samples as burn-in samples.
-The only exception is Spain, in which a single chain converged at around 6M steps.
+All chains seem to converge to the stationary distribution, in most cases before 2M samples. Thus, we discarded the first 2M samples as burn-in samples.
+The only exception is Spain, in which a single chain converged at around 6M samples.
 We considered this chain as part of the burn-in and removed it from the analysis.
-Therefore, 50 chains with 5M samples and IAT between 40K and 187K give an effective samples size between 1,335 and 7,812.
+Therefore, 50 chains with 5M samples and IAT between 32K and 187K give an effective sample size between 1,336 and 7,812.
 We ran additional chains with 2M samples and a different initialization (i.e.
@@
 The estimated posterior distributions of $\tau$ were very similar to our main analysis, further increasing our confidence in the convergence of our inference.
-We also examined the joint posterior distributions of $\tau$ and $\lambda$ (\autoref{fig:joint}).
 

### 2021-01-21 ms/ms.tex
kruschke citation fixed

@@
 We then estimate $\hat{\tau}$ as the median of the marginal posterior distribution of $\tau$.
-Credible intervals (CI) are calculated as the highest density intervals~\citep{Kruschke2015}, and their upper and lower boundaries are reported as $\substack{+\text{upper} \\ -\text{lower}}$ in days relative to $\hat{\tau}$.
+Credible intervals (CI) are calculated as the highest density intervals~\citep{Kruschke2014}, and their upper and lower boundaries are reported as $\substack{+\text{upper} \\ -\text{lower}}$ in days relative to $\hat{\tau}$.
 
@@
 $\tau^*$ is the last NPI date (a lockdown everywhere by Sweden, see \autoref{table:NPI_dates}).
-Thin and bold lines show 95\% and 75\% credible intervals (HDI~\citep{Kruschke2015}), respectively.
+Thin and bold lines show 95\% and 75\% credible intervals (HDI~\citep{Kruschke2014}), respectively.
 Inference performed similarly to main inference, but with data up to Mar~28 and only 1M samples per chain with 600K burn-in.

### 2021-01-21 ms/ms.tex
resolve some small todos

@@
 The significantly earlier estimate of $\hat{\tau}$ relative to $\tau^*$ can suggest that early NPIs in Germany more effectively reduced transmission rates compared to other countries.
-Another possible interpretation is that the German population anticipated the lockdown and began to act accordingly before it started.
-% TODO rephrase
+Another possible interpretation is that the German population anticipated the lockdown and reacted before it started.
 
@@
 %TODO check WAIC
-Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries ().%TODO ref to table
+Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (\autoref{table:estimated-params}).
 Moreover, in Sweden the number of daily cases continued to grow up to Apr~11, rather than "peak" (\autoref{fig:ppc}). 
@@
 % > python Fig_Re.py
-% TODO update
 \begin{figure}[h]

### 2021-01-21 src/Table-RMSE.py
Table wait fixed nsteps

@@ -2,7 +2,7 @@ import os
 import pandas as pd
 
 no_fixed_free = ['7MNoTau', '7MFixed', '7M']
-main_folder = r'../output'
+main_folder = r'../../output-tmp'
 nodf, fixeddf, freedf = [pd.read_csv(os.path.join(main_folder, inner_folder, 'figures', 'ppc_rmse.csv'),sep='\t',header=None) 
        for inner_folder in no_fixed_free]
 

### 2021-01-21 src/Table-WAIC.py

@@ -26,6 +26,7 @@ def load_chain(job_id, country, nburn=2_000_000):
     fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
     inference_data = np.load(fname)
     nsteps, ndim, N, Td1, Td2, model_type = inference_data['params']
+    nchains, nsteps, ndim = inference_data['data'].shape
     logliks = inference_data['logliks']
     nchains = logliks.size // nsteps
     logliks = logliks.reshape(nchains, nsteps)
@@ -56,7 +57,7 @@ def WAIC(logliks):
     return -2*(llpd + -p1), -2*(llpd + -p2)
 
 if __name__ == '__main__':    
-    output_folder = r'../output'
+    output_folder = r'../../output-tmp'
     
     job_ids = ['7M', '7MNoTau', '7MFixed']
     print('Job IDs:')

### 2021-01-21 src/Table-WAIC.py
fixed

@@ -26,7 +26,7 @@ def load_chain(job_id, country, nburn=2_000_000):
     fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
     inference_data = np.load(fname)
     nsteps, ndim, N, Td1, Td2, model_type = inference_data['params']
-    nchains, nsteps, ndim = inference_data['data'].shape
+    nchains, nsteps, ndim = inference_data['chain'].shape
     logliks = inference_data['logliks']
     nchains = logliks.size // nsteps
     logliks = logliks.reshape(nchains, nsteps)

### 2021-01-21 src/Fig_ppc.py
small change

@@ -128,6 +128,7 @@ if __name__ == '__main__':
     ndays = len(X)
     
     chain_fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
+    delete_chain_less_than = 'None'
     if job_id=='7M' and country=='Spain':  #TODO make input parameter
         delete_chain_less_than = 15 
     chain, _, Td1, Td2, model_type, _, start_date, N = load_chain(fname=chain_fname,delete_chain_less_than=delete_chain_less_than)

### 2021-01-21 src/Fig_ppc.py
small change

@@ -128,7 +128,7 @@ if __name__ == '__main__':
     ndays = len(X)
     
     chain_fname = os.path.join(output_folder, job_id, 'inference', '{}.npz'.format(country))
-    delete_chain_less_than = 'None'
+    delete_chain_less_than = None
     if job_id=='7M' and country=='Spain':  #TODO make input parameter
         delete_chain_less_than = 15 
     chain, _, Td1, Td2, model_type, _, start_date, N = load_chain(fname=chain_fname,delete_chain_less_than=delete_chain_less_than)

### 2021-01-21 ms/ms.tex
update Table S1 S2; update Spain postrior to add red line

@@
 
-We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e.
+We compare the posterior predictive plots of a model with  free±$\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e.
 transmission and reporting rates are constant).
@@
 The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
-%TODO update when WAIC and RMSE are ready
-In all 11 of the European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
-When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark (\autoref{table:WAIC}).
+In all 11 european countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
+When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only marginal preference), Sweden, and Denmark (\autoref{table:WAIC}).
 Indeed, we estimate low effect of NPIs on transmission in Denmark and Sweden (i.e. $\lambda=0.7$ and $0.74$, respectively; see~\autoref{table:estimated-params}).
@@
 We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
-However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free $\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
-%TODO check WAIC
+However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free±$\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
 Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (\autoref{table:estimated-params}).
@@
     Shown are estimates of the relative reductions in $R$ (the effective reproduction number), which measures the impact of NPIs on disease transmission, in 11 European countries.
-     The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed $\tau$); the x-axis shows estimates when inferring the effective start of NPIs from the data (free $\tau$).
+     The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed~$\tau$); the x-axis shows estimates when inferring the effective start of NPIs from the data (free $\tau$).
 The dashed line shows a one-to-one correspondence.
      Markers and bars denote the posterior median and 50\% credible intervals (HDI).
-    The relative reductions in $R$ are consistently lower for the fixed $\tau$ model (below the dashed line), except in Sweden and Denmark in which uncertainty is high.
+    The relative reductions in $R$ are consistently lower for the fixed~$\tau$ model (below the dashed line), except in Sweden and Denmark in which uncertainty is high.
     } 
@@
 %%% Fig - Prediction
-% TODO update
 \begin{figure}[h]
@@
 We use the default configuration using the stretch move with stretch scale parameter $a=2$.
-For the main analysis we use 50 chains (or walkers) per region, with 7M samples per chain (no thinning was applied).
+For the main analysis we use 50 chains (or walkers) per region, with 7M samples per chain (no thinning was applied; 6.8M for Wuhan).
 The \emph{integrated autocorrelation time} (IAT)~\citep{Foreman-Mackey2013,Goodman2010} was averaged across parameters and chains for each region. 
-The average IAT was between 32K for Wuhan, China, and 187K for Germany~(\autoref{fig:autocorr}).
+The average IAT was between 31.9K for Wuhan and 187K for Germany~(\autoref{fig:autocorr}).
 We examined the trace plots for $\tau$ in all regions (\autoref{fig:trace}).
@@
 We considered this chain as part of the burn-in and removed it from the analysis.
-Therefore, 50 chains with 5M samples and IAT between 32K and 187K give an effective sample size between 1,336 and 7,812.
+Therefore, 50 chains with 5M samples and IAT between 32K and 187K give an effective sample size between 1,336 and 7,523.
 We ran additional chains with 2M samples and a different initialization (i.e.
@@
 The estimated posterior distributions of $\tau$ were very similar to our main analysis, further increasing our confidence in the convergence of our inference.
+For the models with fixed~$\tau$ and without~$\tau$, in some countries, we use less than 7M samples per chain because the IAT converged sooner.
+In these cases the effective sample size was at between 2,454 (UK) and 10,954 (Spain) in the fixed~$\tau$ model and between 6,230 (Norway) and 94,339 (Italy) in the model without~$\tau$.
 
@@
 
+% Table - RMSE 
+% generated with /python Table-RMSE.py FOLDER NAME
+\begin{table}[h]
+\centering
+\caption*{\textbf{RMSE}}
+\pgfplotstabletypeset[
+    col sep=comma,
+    string type,
+    every head row/.style={before row=\hline,after row=\hline},
+    every last row/.style={after row=\hline},
+    ]{../figures/Table-RMSE.csv}
+\caption{
+\textbf{Posterior RMSE of out-of-sample predictions with the different models.}
+	Expected posterior predictive RMSE (root mean squared error) for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
+	In all cases, the model with free parameter $\tau$ has the lowest RMSE.
+	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions up to Apr~24 by sampling model parameters from the posterior distribution.
+	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model. 
+	Bold values emphasize cases in which the \emph{Free} model has the lowest RMSE.
+}
+\label{table:RMSE}
+\end{table}
+
 % Table - WAIC 
 % generated with /python Table-WAIC.py FOLDER NAME
-% TODO update
 \begin{table}[h]
@@
 
-% Table - RMSE 
-% generated with /python Table-RMSE.py FOLDER NAME
-% TODO update
-\begin{table}[h]
-\centering
-\caption*{\textbf{RMSE}}
-\pgfplotstabletypeset[
-    col sep=comma,
-    string type,
-    every head row/.style={before row=\hline,after row=\hline},
-    every last row/.style={after row=\hline},
-    ]{../figures/Table-RMSE.csv}
-\caption{
-\textbf{Posterior RMSE of out-of-sample predictions with the different models.}
-	Expected posterior predictive RMSE (root mean squared error) for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
-	In all cases, the model with free parameter $\tau$ has the lowest RMSE.
-	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions up to Apr~24 by sampling model parameters from the posterior distribution.
-	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model. 
-	Bold values emphasize cases in which the \emph{Free} model has the lowest RMSE.
-}
-\label{table:RMSE}
-\end{table}
-
 %%% Fig - tau summary Mar 28 %%%
@@
     Circles and stars represent daily case data up to and after Apr~11, respectively
-    Black and white arrows denote the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs, respectively.
+    Black and white arrows denote the official ($\tau^*$) and effective ($\hat{\tau}$) start of NPIs, respectively.
     Black lines represent a smoothing of the data points using a Savitzky-Golay filter with window length 3. 
-    Coloured lines represent posterior predictions from a model with fixed $\tau$ (blue) and free $\tau$ (green).
+    Coloured lines represent posterior predictions from a model with fixed~$\tau$ (blue) and free~$\tau$ (green).
 Models were fitted with data up to Apr~11.
@@
 Note the differences in the y-axis scale.
-    Posterior predictions with the free $\tau$ model predict the out-of-sample data well for all countries except Denmark and Sweden, but poorly with the fixed $\tau$ model.
-    The predictions of the model without $\tau$ (not shown) are even worse.
+    Posterior predictions with the free~$\tau$ model predict the out-of-sample data well for all countries except Denmark and Sweden, but poorly with the fixed~$\tau$ model.
+    The predictions of the model without~$\tau$ (not shown) are even worse.
     }

### 2021-01-21 src/Fig_tau_posterior.py
fixed Fig5

@@ -41,7 +41,7 @@ def calc_hpd(trace, mass_frac) :
 
 if __name__ == '__main__':
 	job_id = sys.argv[1]
-	output_folder = r'../output-tmp/{}/'.format(job_id)
+	output_folder = r'../../output-tmp/{}/'.format(job_id)
 	country = sys.argv[2]
 	quiet = len(sys.argv) > 3 and sys.argv[3] == '-q'	
 	nburn = 2_000_000

### 2021-01-21 src/Re.py

@@ -44,7 +44,7 @@ def Re(α, β, D, μ):
 if __name__ == '__main__':
 	nreps = 1000
 	
-	output_folder = r'../output'
+	output_folder = r'../../output-tmp'
 	job_id = sys.argv[1]	
 	country = sys.argv[2]
 	if len(sys.argv) > 3:

### 2021-01-22 ms/ms.tex
typos fix

@@
 
-We compare the posterior predictive plots of a model with  free±$\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e.
+We compare the posterior predictive plots of a model with  free~$\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e.
 transmission and reporting rates are constant).
@@
 A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPI eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95 \%CI, \autoref{fig:late}).
+Nevertheless, we estimate the effective start of NPI eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95\% CI, \autoref{fig:late}).
 
@@
 We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
-However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free±$\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
+However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free~$\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
 Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (\autoref{table:estimated-params}).
@@
     The vertical dashed line represents Apr~11, 2020.
-    Circles and stars represent daily case data up to and after Apr~11, respectively
+    Circles and stars represent daily case data up to and after Apr~11, respectively.
     Black and white arrows denote the official ($\tau^*$) and effective ($\hat{\tau}$) start of NPIs, respectively.

### 2021-01-25 ms/ms.tex
underestimate impact vs delay fig

@@
 That is, the first and last official NPI dates are Mar~5 and Mar~11.
-However, we estimate the effective date $\hat{\tau}$ six days after the lockdown, at Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\% CI; \autoref{fig:late}).
+However, we estimate the effective date ($\hat{\tau}$) six~days after the lockdown, at Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\% CI; \autoref{fig:late}).
 
@@
 We infer model-based effective reproduction numbers before and after the implementation of NPIs from model parameters (\autoref{eq:Re}).
-We then estimate the impact of NPIs as the relative reduction in the reproduction number \citep{Flaxman2020}. 
+We then estimate the impact of NPIs as the relative reduction in the reproduction number \citep{Flaxman2020}, $\Delta R = (R_1-R_2)/R_1$. 
 We compare the impacts estimated using the fixed~$\tau$ model and the free~$\tau$ model.
@@
 \autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model (y-axis) are consistently lower than estimates from the free~$\tau$ model (x-axis), except in Sweden and Denmark, in which estimation uncertainty is high.
+That is, assuming NPIs start at their official date leads to error in the estimation of their impact.
+Moreover, these errors increase with the duration of the delay between the official and effective dates (\autoref{fig:Re2}).
 These results suggest that the impact of past NPIs is likely to be under-estimated by health officials and researchers if they assume NPIs effectively start at their official dates.
 The estimated impact can then be interpreted as ineffectiveness of the NPIs, leading to more aggressive NPIs being applied.
-
-Health officials might also assess the impact of NPIs by comparing model expectations to actual number of daily cases at a specific number of days after the intervention began. 
-However, a significant difference between the official and effective start of interventions can invalidate such assessments.
-This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~17 ($\hat{\tau}$).
-If health officials assume that the dynamics change exactly at $\tau^*$, they would expect the number of cases to be within the red lines (posterior predictions assuming $\tau=\tau^*$).
-This would lead to a significant under-estimation, which might be interpreted as ineffectiveness of NPIs, and therefore more strict guidelines.
-However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which are a good fit to the real data (stars).
 
@@
     \textbf{Impact of NPIs is under-estimated when assuming they start at the official date.}
-    Shown are estimates of the relative reductions in $R$ (the effective reproduction number), which measures the impact of NPIs on disease transmission, in 11 European countries.
+    Shown are estimates of the relative reduction in the effective reproduction number ($R$, see \autoref{eq:Re}), which measures the impact of NPIs on disease transmission, in 11 European countries.
      The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed~$\tau$); the x-axis shows estimates when inferring the effective start of NPIs from the data (free $\tau$).
@@
     \label{fig:Re}
-\end{figure}
-
-%%% Fig - Prediction
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.5\textwidth]{../figures/Fig-prediction.pdf}
-    \caption{
-    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
-    Real number of daily cases in Italy in black (markers: data; line: smoothed using a Savitzky-Golay filter with window length 3).
-    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution).
-    Shaded box illustrates a serial interval of seven days~\citep{Ali2020} after the start of NPIs.
-    \textbf{(A)} Using the official date $\tau^*$ for the effective start of the NPI, Mar~11, the model under-estimates the number of cases seven days after the start of the NPI.
-    \textbf{(B)} Using the estimated date $\hat{\tau}$ for the effective start of the NPI, Mar~17, the model precisely estimates the number of cases seven days after the start of the NPI.
-    Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
-    }
-    \label{fig:prediction}
 \end{figure}
@@
 Attempts to assess the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date, and that a significant change in the dynamics can be observed roughly seven days after the start of NPIs (due to the characteristic serial interval of \covid\citep{Ali2020}).
-However, late and early effective start of NPIs, such as we have inferred, can bias these assessments and lead to wrong conclusions about the impact of NPIs (Figures~\ref{fig:Re} and~\ref{fig:prediction}).
+However, late and early effective start of NPIs, such as we have inferred, can bias these assessments and lead to wrong conclusions about the impact of NPIs (\autoref{fig:Re}).
 
@@
 
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=\textwidth]{../figures/Fig_Re2.pdf}
+    \caption{
+    \textbf{Under-estimation of impact of NPIs increases with the delay in their effective start.} The x-axis shows the days between the estimated effective start of NPIs, $\hat\tau$, and the official date, $\tau^*$.
+The y-axis shows the error in estimation of the impact of NPIs when assuming they start at their official date (i.e.
+the difference between the median estimate from a model with fixed~$\tau$ minus the estimate from a model with free~$\tau$, the y- and x-axis of \autoref{fig:Re}, respectively.
+    Impact of NPIs defined as $\frac{R_1 - R_2}{R_1}$, where $R_1$ and $R_2$ are the effective reproduction numbers before and after the NPI, respectively (\autoref{eq:Re}).
+	The dashed black line shows a linear regression, slope=-0.0313 with 95\% CI [-0.051, -0.012], $R^2=0.554$.
+    } 
+    \label{fig:Re2}
+\end{figure}
+
 %%% Fig Fr-vs-Es

### 2021-01-25 ms/ms.tex
small typos

@@
     \textbf{Under-estimation of impact of NPIs increases with the delay in their effective start.} The x-axis shows the days between the estimated effective start of NPIs, $\hat\tau$, and the official date, $\tau^*$.
-The y-axis shows the error in estimation of the impact of NPIs when assuming they start at their official date (i.e.
-the difference between the median estimate from a model with fixed~$\tau$ minus the estimate from a model with free~$\tau$, the y- and x-axis of \autoref{fig:Re}, respectively.
+The y-axis shows the error in estimation of the impact of NPIs when assuming they start at their official date, i.e.
+the difference between the median estimate from a model with fixed~$\tau$ minus the estimate from a model with free~$\tau$, the y-axis and x-axis of \autoref{fig:Re}, respectively.
     Impact of NPIs defined as $\frac{R_1 - R_2}{R_1}$, where $R_1$ and $R_2$ are the effective reproduction numbers before and after the NPI, respectively (\autoref{eq:Re}).

### 2021-01-26 ms/ms.tex
editing before submission to bmc public health: abstract, phrasing, preprints update

@@
 \begin{abstract}
+\noindent
+\textbf{Background.}
 During Feb-Apr 2020, many countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules, to control the \covid pandemic caused by the \sars virus.
 Overall, these interventions seem to have reduced the spread of the pandemic.
-We hypothesize that the official and effective start date of such interventions can be noticeably different, for example due to slow adoption by the population, or because the authorities and the public are unprepared.
-We fit an SEIR model to case data from~12 regions to infer the effective start dates of interventions and contrast them with the official dates.
-We find mostly late, but also early effects of interventions.
-For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective date on Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\% CI).
-In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95\% CI).
-We demonstrate that differences between the official and effective start of NPIs can lead to under-estimating their impact, and discuss potential causes and consequences of our results.
+We hypothesized that the official and effective start date of such interventions can be noticeably different, for example due to slow adoption by the population, or because the authorities and the public are unprepared, and that these difference can lead to errors in the estimation of the impact of NPIs.
+\\
+\textbf{Methods.}
+SEIR models were fitted to case data from~12 regions to infer the effective start dates of interventions and contrast them with the official dates. \\
+\textbf{Results.}
+We infer mostly late, but also early effects of interventions.
+For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective start date on Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\%~CI).
+In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95\%~CI).
+Moreover, we find that the impact of NPIs can be under-estimated if assuming they start at their official date.\\
+\textbf{Conclusions.}
+Differences between the official and effective start of NPIs are likely, and that neglecting such differences can lead to under-estimation of the impact of NPIs.
 \\\\
@@
 \centering
+\caption{
+\textbf{Official start of non-pharmaceutical interventions.}}
 \pgfplotstabletypeset[
@@
     ]{../data/NPI_dates.csv}
-\caption{
-\textbf{Official start of non-pharmaceutical interventions.}
+\caption*{
 The date of the first intervention is for a ban of public events, or encouragement of social distancing, or for school closures.
@@
 
-Here, we hypothesise that there is a significant difference between the official start of NPIs and their effective adoption by the public and therefore their effect on infection dynamics.
+Here, we hypothesize that there is a significant difference between the official start of NPIs and their effective adoption by the public and therefore their effect on infection dynamics.
 We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) model and a \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to infer the effective start date of NPIs from publicly available \covid case data in 12 geographical regions.
@@
 The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers).
-In contrast, the estimated effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although uncertainty is low only for Germany (i.e., zero is not in 95\% CI).
+In contrast, the estimated effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although uncertainty is low only for Germany (i.e., zero is not in 95\%~CI).
 The credible intervals for Sweden, Denmark, and Norway are especially wide, see text and \autoref{fig:early} for possible explanation.
@@
 transmission and reporting rates are constant).
-All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~24, and these predictions were then compared to the real case data. 
+All three models were fitted to case data up to Apr~11, 2020, used to predict out-of-sample case data up to Apr~24, 2020, and these predictions were then compared to the real case data. 
 The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
@@
 
-We compare the official $\tau^*$ and effective $\tau$ start of NPIs and find that in most regions the effective start of NPIs differs from the official date: in 10 of 12 countries the 75\% credible interval on $\tau$ does not include $\tau^*$ (7 of 12 countries when considering a 95\% interval; \autoref{fig:tau-summary}).
+We compare the official ($\tau^*$) and effective ($\tau$) start of NPIs and find that in most regions the effective start date differs from the official date: in 10 of 12 countries the 75\%~credible interval on $\tau$ does not include $\tau^*$ (7 of 12 countries when considering a 95\%~credible interval; \autoref{fig:tau-summary}).
 The exceptions are Norway and Switzerland (see below).
@@
 Black line shows the estimated effective start date $\hat{\tau}$.
-Shaded area shows a 95\% credible interval.
+Shaded area shows a 95\%~credible interval.
     }
@@
 That is, the first and last official NPI dates are Mar~5 and Mar~11.
-However, we estimate the effective date ($\hat{\tau}$) six~days after the lockdown, at Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\% CI; \autoref{fig:late}).
-
-In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\% CI).
+However, we estimate the effective date ($\hat{\tau}$) six~days after the lockdown, at Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\%~CI; \autoref{fig:late}).
+
+In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\%~CI).
 Yet, there is low but noticeable posterior probability for Jan~25 (\autoref{fig:late}), for which the effect of NPIs on transmission is considerably lower (\autoref{fig:joint}).
 
-In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score 2--1) with a crowd of 50,965 in Seville.
+In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the International Women's Day~\citep{RaphaelMinder}, and a football match between Real Betis and Real Madrid (final score 2--1) with a crowd of 50,965 in Seville~\citep{ESPNRealBetis}.
 A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPI eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95\% CI, \autoref{fig:late}).
-
-Similarly, in France we estimate the effective start of NPIs at Mar~24 ($\substack{+2.65 \\ -1.44}$~days 95\% CI, \autoref{fig:late}).
+Nevertheless, we estimate the effective start of NPIs eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95\%~CI, \autoref{fig:late}).
+
+Similarly, in France we estimate the effective start of NPIs at Mar~24 ($\substack{+2.65 \\ -1.44}$~days 95\%~CI, \autoref{fig:late}).
 This is a week later than the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
@@
 \paragraph*{Early effective start of NPIs.}
-In some regions we estimate an effective start of NPIs $\tau$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
-The only conclusive early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95 \%CI; \autoref{fig:early}).
+In some regions we estimate an effective start of NPIs that is \emph{earlier} then the official date (\autoref{fig:tau-summary}).
+The only conclusive early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95~\%CI, \autoref{fig:early}).
 This estimate falls between the first and last official NPI dates, Mar~12 and Mar~22 (\autoref{table:NPI_dates}).
@@
 We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
-However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free~$\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
+However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free~$\tau$ over the model with $\tau$ fixed at the official date (\autoref{table:WAIC}).
 Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (\autoref{table:estimated-params}).
-Moreover, in Sweden the number of daily cases continued to grow up to Apr~11, rather than "peak" (\autoref{fig:ppc}). 
+Moreover, in Sweden the number of daily cases continued to grow through Apr~11, rather than "peak" (\autoref{fig:ppc}). 
 In Denmark, the opposite occurred: there were seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
@@
 The estimated effective date in Norway is Mar~22, two days earlier than the official date of Mar~24.
-However, the posterior distribution is very wide ($\substack{+5.59 \\ -12.03}$~days 95\% CI): it covers the range between Mar~10, two days before the first NPI, and Mar~27, three days after the last NPI (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
+However, the posterior distribution is very wide ($\substack{+5.59 \\ -12.03}$~days 95\%~CI): it covers the range between Mar~10, two days before the first NPI, and Mar~27, three days after the last NPI (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
 The high uncertainty might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, and the corresponding posterior distribution is very narrow (\autoref{fig:early}).
-There may also be an unidentifiability issues between $\tau$ and the effect of NPIs on transmission (\autoref{fig:joint}).
+There may also be an unidentifiability issue between $\tau$ and the effect of NPIs on transmission (\autoref{fig:joint}).
 
@@
 Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
-Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\substack{+3.2 \\ -1.78}$~days 95\% CI, see \autoref{fig:early}).
-It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
+Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\substack{+3.2 \\ -1.78}$~days 95\%~CI, see \autoref{fig:early}). 
+Notably, Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
 
@@
 \paragraph*{Assessment of impact of NPIs.}
-
 The \emph{effective reproduction number} $R$ is the average number of secondary
@@
 We infer model-based effective reproduction numbers before and after the implementation of NPIs from model parameters (\autoref{eq:Re}).
-We then estimate the impact of NPIs as the relative reduction in the reproduction number \citep{Flaxman2020}, $\Delta R = (R_1-R_2)/R_1$. 
+We then estimate the impact of NPIs as the relative reduction in the reproduction number \citep{Flaxman2020}, $\Delta R = (R_1-R_2)/R_1$, where $R_1$ and $R_2$ are the reproduction numbers before and after the start of NPIs.
 We compare the impacts estimated using the fixed~$\tau$ model and the free~$\tau$ model.
-That is, we compare the impact estimate assuming that NPIs started at the official date $\tau^*$, versus the estimate when inferring the effective start of NPIs from the data.
-\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model (y-axis) are consistently lower than estimates from the free~$\tau$ model (x-axis), except in Sweden and Denmark, in which estimation uncertainty is high.
+That is, we compare the impact estimate $\Delta R$ assuming that NPIs started at the official date $\tau^*$, versus the estimate when inferring the effective start of NPIs from the data.
+
+\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model (y-axis) are consistently lower than estimates from the free~$\tau$ model (x-axis), except in Sweden (SE) and Denmark (DK), in which estimation uncertainty is high.
 That is, assuming NPIs start at their official date leads to error in the estimation of their impact.
 Moreover, these errors increase with the duration of the delay between the official and effective dates (\autoref{fig:Re2}).
-These results suggest that the impact of past NPIs is likely to be under-estimated by health officials and researchers if they assume NPIs effectively start at their official dates.
+These results suggest that the impact of past NPIs is likely to be under-estimated by health officials and researchers if they assume NPIs start at their official dates.
 The estimated impact can then be interpreted as ineffectiveness of the NPIs, leading to more aggressive NPIs being applied.
@@
     \caption{
-    \textbf{Impact of NPIs is under-estimated when assuming they start at the official date.}
-    Shown are estimates of the relative reduction in the effective reproduction number ($R$, see \autoref{eq:Re}), which measures the impact of NPIs on disease transmission, in 11 European countries.
-     The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed~$\tau$); the x-axis shows estimates when inferring the effective start of NPIs from the data (free $\tau$).
+    \textbf{Impact of NPIs is under-estimated when assuming they start at their official date.}
+    Shown are estimates of the relative reduction in the effective reproduction number ($R$, see \autoref{eq:Re}), which measures the impact of NPIs on the epidemic.
+     The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed~$\tau$).
+The x-axis shows estimates when inferring the effective start of NPIs from the data (free~$\tau$).
 The dashed line shows a one-to-one correspondence.
@@
     The relative reductions in $R$ are consistently lower for the fixed~$\tau$ model (below the dashed line), except in Sweden and Denmark in which uncertainty is high.
+    \autoref{fig:Re2} shows how the delay in effective start affect the estimation error. 
+    AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.
     } 
@@
 
-\pagebreak
+%\pagebreak
 % Discussion
@@
 
-We have inferred the effective start date of NPIs in several geographical regions using an SEIR model under an MCMC parameter estimation framework.
+We have inferred the effective start date of NPIs in 12~regions using SEIR models under an MCMC parameter estimation framework.
 We find examples of both late and early effective start of NPIs relative to the official date (\autoref{fig:tau-summary}).
@@
 For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five or more days after the official date (\autoref{fig:late}).
-This difference might be explained, in some cases, by low compliance or non-adherence to guidelines: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, resulting in people leaving these provinces before the lockdown started~\citep{Gatto2020}.
+These differences might be explained, in some cases, by low compliance or non-adherence to guidelines: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, which led to people leaving these provinces before the lockdown started~\citep{Gatto2020}.
 Late effect of NPIs may also be due to the time required by both the government and the citizens to prepare for a lockdown, and for new guidelines to be adopted by the population.
@@
 Indeed, the evidence supports a change in infection dynamics (i.e.
-a model with fixed or free $\tau$) even for Sweden~(\autoref{table:WAIC}, \autoref{table:RMSE}, \autoref{fig:ppc}), where a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
+a model with fixed or free $\tau$) even for Sweden~(\autoref{table:RMSE}, \autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented within the investigated time\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
 
@@
 The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
-This may suggest correlations between effective starts of NPIs due to global or international events.
-
-As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (Tables~\ref{table:WAIC} and~\ref{table:RMSE}).
+This may suggest correlations between effective starts of NPIs in different countries due to international or global events.
+
+As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (Tables~\ref{table:RMSE} and~\ref{table:WAIC}).
 It may be interesting to check if the evidence supports a model with \emph{two} or more change-points, rather than one. 
 Multiple change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
-However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementations of NPIs~\citep{Flaxman2020}.
+However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementation of NPIs~\citep{Flaxman2020}.
 
 As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with new guidelines, whereas in other cases the population will adopt either restrictions or relaxations even before they are formally announced.
-Attempts to assess the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date, and that a significant change in the dynamics can be observed roughly seven days after the start of NPIs (due to the characteristic serial interval of \covid\citep{Ali2020}).
-However, late and early effective start of NPIs, such as we have inferred, can bias these assessments and lead to wrong conclusions about the impact of NPIs (\autoref{fig:Re}).
+Attempts to assess the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date.
+However, late effective start of NPIs, such as we have inferred, can lead to under-estimation of the impact of NPIs (Figures~\ref{fig:Re} and~\ref{fig:Re2}).
+Such under-estimation may lead decision-makers to enforce stricter guidelines, rather than enforce earlier implementation of the guidelines~\citep{Pei2020a}.
 
@@
 The posterior distribution of the model parameters $P(\theta \mid \vec{X})$ is estimated using the affine-invariant ensemble sampler for Markov chain Monte Carlo (MCMC)~\citep{Goodman2010}, implemented in the \texttt{emcee} Python package~\citep{Foreman-Mackey2013}.
-We use the default configuration using the stretch move with stretch scale parameter $a=2$.
+We use the default configuration with the stretch move and a stretch scale parameter $a=2$.
 For the main analysis we use 50 chains (or walkers) per region, with 7M samples per chain (no thinning was applied; 6.8M for Wuhan).
@@
 \centering
+\caption{
+\textbf{Parameter estimates for different regions.}}
 \footnotesize{
@@
 }
-\caption{
-\textbf{Parameter estimates for different regions.}
+\caption*{
 See \autoref{eq:model} for model parameters.
@@
 \centering
-\caption*{\textbf{RMSE}}
+\caption{
+\textbf{Posterior RMSE of out-of-sample predictions with the different models.}}
+%\caption*{\textbf{RMSE}}
 \pgfplotstabletypeset[
@@
     ]{../figures/Table-RMSE.csv}
-\caption{
-\textbf{Posterior RMSE of out-of-sample predictions with the different models.}
+\caption*{
 	Expected posterior predictive RMSE (root mean squared error) for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
@@
 \centering
-\caption*{\textbf{WAIC}}
+\caption{
+\textbf{WAIC values for the different models.}}
 \pgfplotstabletypeset[
@@
     ]{../figures/Table-WAIC.csv} 
-\caption{
-\textbf{WAIC values for the different models.}
+\caption*{
 	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
@@
     Impact of NPIs defined as $\frac{R_1 - R_2}{R_1}$, where $R_1$ and $R_2$ are the effective reproduction numbers before and after the NPI, respectively (\autoref{eq:Re}).
-	The dashed black line shows a linear regression, slope=-0.0313 with 95\% CI [-0.051, -0.012], $R^2=0.554$.
+	The dashed black line shows a linear regression, slope=-0.0313 with (-0.051, -0.012)~95\%~CI, $R^2=0.554$.
+	AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.
     } 

### 2021-01-27 ms/ms.tex
revise format for bmc

@@
 
-% Introduction
-\section*{Introduction}
+% Background
+\section*{Background}
 
@@
 
-% Results
-\section*{Results}
-
-%%% Fig - tau summary %%%
-% generated with /python Fig_tau_summary.py FOLDER NAME
-\begin{figure}[b!]
-    \centering
-	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Apr11.pdf}	
-    \caption{
-    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
-    	The difference between $\tau$ the effective and $\tau^*$ the official start of NPIs is shown for different regions.
-The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers).
-In contrast, the estimated effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although uncertainty is low only for Germany (i.e., zero is not in 95\%~CI).
-The credible intervals for Sweden, Denmark, and Norway are especially wide, see text and \autoref{fig:early} for possible explanation.
-	Here, the markers show $\hat{\tau}$, the marginal posterior median (\autoref{table:estimated-params}).
-$\tau^*$ is the last NPI date (a lockdown, except in Sweden; \autoref{table:NPI_dates}).
-Thin and bold lines show 95\% and 75\% credible intervals, respectively.
-\autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28 rather than Apr~11.
-    }
-    \label{fig:tau-summary}
-\end{figure}
-
-Several studies have described the effects of non-pharmaceutical interventions in different geographical regions~\citep{Flaxman2020,Gatto2020,Li2020}. 
-Some of these studies have assumed that the parameters of the epidemiological model change at a specific date (\autoref{eq:NPI_model}), and set the change date $\tau$ to the official NPI date $\tau^*$, usually the lockdown start date (\autoref{table:NPI_dates}).
-They then fit the model once for time $t<\tau^*$ and once for time $t \ge \tau^*$.
-For example, \citet{Li2020} estimate the infection dynamics in China before and after $\tau^*$, which is set at Jan 23, 2020.
-Thereby, they effectively estimate the transmission and reporting rates before and after $\tau^*$ separately.
-
-Here, we estimate the joint posterior distribution of the effective start date of NPIs, $\tau$, and the transmission and reporting rates before and after $\tau$ from the entire data, rather than splitting the data at~$\tau$.
-This is done under the simplifying assumption that all interventions start at a specific date, despite the reality that the durations between the first and last NPIs were between 4 and 12 days (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
-We then estimate $\hat{\tau}$ as the median of the marginal posterior distribution of $\tau$.
-Credible intervals (CI) are calculated as the highest density intervals~\citep{Kruschke2014}, and their upper and lower boundaries are reported as $\substack{+\text{upper} \\ -\text{lower}}$ in days relative to $\hat{\tau}$.
-
-We compare the posterior predictive plots of a model with  free~$\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e.
-transmission and reporting rates are constant).
-All three models were fitted to case data up to Apr~11, 2020, used to predict out-of-sample case data up to Apr~24, 2020, and these predictions were then compared to the real case data. 
-The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
-In all 11 european countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
-When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only marginal preference), Sweden, and Denmark (\autoref{table:WAIC}).
-Indeed, we estimate low effect of NPIs on transmission in Denmark and Sweden (i.e. $\lambda=0.7$ and $0.74$, respectively; see~\autoref{table:estimated-params}).
-This may interfere with the inference of $\tau$ due to unidentifiability. 
-Notably, the data for Sweden and Denmark do not have a single "peak" during the evaluated dates, possibly leading to wide credible intervals on $\tau$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).
-
-We compare the official ($\tau^*$) and effective ($\tau$) start of NPIs and find that in most regions the effective start date differs from the official date: in 10 of 12 countries the 75\%~credible interval on $\tau$ does not include $\tau^*$ (7 of 12 countries when considering a 95\%~credible interval; \autoref{fig:tau-summary}).
-The exceptions are Norway and Switzerland (see below).
-The former also has a wide credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
-
-%%% Fig - Late %%%
-\begin{figure}[h]
-    \centering
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Italy_τ_posterior.pdf}
-    \end{subfigure}
-  	~
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Mar28/Wuhan_τ_posterior.pdf}
-    \end{subfigure}
-    \\
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Spain_τ_posterior.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/France_τ_posterior.pdf}
-    \end{subfigure}
-    \caption{
-	\textbf{Late effect of non-pharmaceutical interventions.}
-    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
-Red line shows the official last NPI date $\tau^*$.
-Black line shows the estimated effective start date $\hat{\tau}$.
-Shaded area shows a 95\%~credible interval.
-    }
-    \label{fig:late}
-\end{figure}
-
-%%% Late
-\paragraph*{Late effective start of NPIs.}
-In half of the examined regions, we estimate that the effective start of NPIs $\tau$ is later than the official date $\tau^*$.
-
-In Italy, the first case was officially confirmed on Feb~21.
-School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
-That is, the first and last official NPI dates are Mar~5 and Mar~11.
-However, we estimate the effective date ($\hat{\tau}$) six~days after the lockdown, at Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\%~CI; \autoref{fig:late}).
-
-In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\%~CI).
-Yet, there is low but noticeable posterior probability for Jan~25 (\autoref{fig:late}), for which the effect of NPIs on transmission is considerably lower (\autoref{fig:joint}).
-
-In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the International Women's Day~\citep{RaphaelMinder}, and a football match between Real Betis and Real Madrid (final score 2--1) with a crowd of 50,965 in Seville~\citep{ESPNRealBetis}.
-A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
-Nevertheless, we estimate the effective start of NPIs eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95\%~CI, \autoref{fig:late}).
-
-Similarly, in France we estimate the effective start of NPIs at Mar~24 ($\substack{+2.65 \\ -1.44}$~days 95\%~CI, \autoref{fig:late}).
-This is a week later than the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
-
-%%% Fig - Early %%% 
-\begin{figure}[h]
-    \centering
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Germany_τ_posterior.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/Sweden_τ_posterior.pdf}
-    \end{subfigure}
-
-	\begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Switzerland_τ_posterior.pdf}
-    \end{subfigure}
-    ~
-    \begin{subfigure}{0.45\textwidth}
-		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/Norway_τ_posterior.pdf}
-    \end{subfigure}
-    \caption{
-    \textbf{Early and exact effect of non-pharmaceutical interventions.}
-    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
-Red line shows the official last NPI date $\tau^*$.
-Black line shows the estimated effective start date $\hat{\tau}$.
-Shaded area shows a 95\% credible interval.
-	}
-	\label{fig:early}
-\end{figure}
-
-%%% Early
-\paragraph*{Early effective start of NPIs.}
-In some regions we estimate an effective start of NPIs that is \emph{earlier} then the official date (\autoref{fig:tau-summary}).
-The only conclusive early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95~\%CI, \autoref{fig:early}).
-This estimate falls between the first and last official NPI dates, Mar~12 and Mar~22 (\autoref{table:NPI_dates}).
-Therefore, when we refer to this case as "early", we mean that the effective date (Mar~19) occurs \emph{before} the official lockdown date (Mar~22), not that it occurs before all NPIs.
-Interestingly, Germany had the second longest duration between first and last NPIs after Norway (10 and 12 days respectively; \autoref{table:NPI_dates}).
-However, the credible interval for the effective start date in Germany is narrow (1.91 days), whereas it is very wide in Norway (17.61 days).
-The significantly earlier estimate of $\hat{\tau}$ relative to $\tau^*$ can suggest that early NPIs in Germany more effectively reduced transmission rates compared to other countries.
-Another possible interpretation is that the German population anticipated the lockdown and reacted before it started.
-
-We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
-However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free~$\tau$ over the model with $\tau$ fixed at the official date (\autoref{table:WAIC}).
-Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (\autoref{table:estimated-params}).
-Moreover, in Sweden the number of daily cases continued to grow through Apr~11, rather than "peak" (\autoref{fig:ppc}). 
-In Denmark, the opposite occurred: there were seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
-We suspect that these missing and additional "peaks" increase the uncertainty in our inference.
-
-The estimated effective date in Norway is Mar~22, two days earlier than the official date of Mar~24.
-However, the posterior distribution is very wide ($\substack{+5.59 \\ -12.03}$~days 95\%~CI): it covers the range between Mar~10, two days before the first NPI, and Mar~27, three days after the last NPI (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
-The high uncertainty might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, and the corresponding posterior distribution is very narrow (\autoref{fig:early}).
-There may also be an unidentifiability issue between $\tau$ and the effect of NPIs on transmission (\autoref{fig:joint}).
-
-%%% Exact
-\paragraph*{Exact effective start of NPIs.}
-We find one case in which the official and effective dates match and the credible interval is narrow.
-Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
-Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\substack{+3.2 \\ -1.78}$~days 95\%~CI, see \autoref{fig:early}). 
-Notably, Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
-
-%%% Assessment
-\paragraph*{Assessment of impact of NPIs.}
-The \emph{effective reproduction number} $R$ is the average number of secondary
-cases caused by an infected individual after an epidemic is already underway \citep{Bar-On2020a}.
-We infer model-based effective reproduction numbers before and after the implementation of NPIs from model parameters (\autoref{eq:Re}).
-We then estimate the impact of NPIs as the relative reduction in the reproduction number \citep{Flaxman2020}, $\Delta R = (R_1-R_2)/R_1$, where $R_1$ and $R_2$ are the reproduction numbers before and after the start of NPIs.
-We compare the impacts estimated using the fixed~$\tau$ model and the free~$\tau$ model.
-That is, we compare the impact estimate $\Delta R$ assuming that NPIs started at the official date $\tau^*$, versus the estimate when inferring the effective start of NPIs from the data.
-
-\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model (y-axis) are consistently lower than estimates from the free~$\tau$ model (x-axis), except in Sweden (SE) and Denmark (DK), in which estimation uncertainty is high.
-That is, assuming NPIs start at their official date leads to error in the estimation of their impact.
-Moreover, these errors increase with the duration of the delay between the official and effective dates (\autoref{fig:Re2}).
-These results suggest that the impact of past NPIs is likely to be under-estimated by health officials and researchers if they assume NPIs start at their official dates.
-The estimated impact can then be interpreted as ineffectiveness of the NPIs, leading to more aggressive NPIs being applied.
-
-%%% Fig - Re
-% generated with
-% > source ./Re.sh
-% > python Fig_Re.py
-\begin{figure}[h]
-    \centering
-	\includegraphics[width=\textwidth]{../figures/Fig_Re.pdf}
-    \caption{
-    \textbf{Impact of NPIs is under-estimated when assuming they start at their official date.}
-    Shown are estimates of the relative reduction in the effective reproduction number ($R$, see \autoref{eq:Re}), which measures the impact of NPIs on the epidemic.
-     The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed~$\tau$).
-The x-axis shows estimates when inferring the effective start of NPIs from the data (free~$\tau$).
-The dashed line shows a one-to-one correspondence.
-     Markers and bars denote the posterior median and 50\% credible intervals (HDI).
-    The relative reductions in $R$ are consistently lower for the fixed~$\tau$ model (below the dashed line), except in Sweden and Denmark in which uncertainty is high.
-    \autoref{fig:Re2} shows how the delay in effective start affect the estimation error. 
-    AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.
-    } 
-    \label{fig:Re}
-\end{figure}
-
-%\pagebreak
-% Discussion
-\section*{Discussion}
-
-We have inferred the effective start date of NPIs in 12~regions using SEIR models under an MCMC parameter estimation framework.
-We find examples of both late and early effective start of NPIs relative to the official date (\autoref{fig:tau-summary}).
-
-In most investigated regions we find late effective start of NPIs.
-For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five or more days after the official date (\autoref{fig:late}).
-These differences might be explained, in some cases, by low compliance or non-adherence to guidelines: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, which led to people leaving these provinces before the lockdown started~\citep{Gatto2020}.
-Late effect of NPIs may also be due to the time required by both the government and the citizens to prepare for a lockdown, and for new guidelines to be adopted by the population.
-
-In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented, although this is only conclusive in Germany (\autoref{fig:early}).
-An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population, possibly due to earlier NPIs or NPIs being applied in other regions.
-Adoption of these behaviours may occur via media and social networks, rather than official guidelines, and may be influenced by increased risk perception due to domestic or international \covid reports~\citep{Arthur2020}.
-Indeed, the evidence supports a change in infection dynamics (i.e.
-a model with fixed or free $\tau$) even for Sweden~(\autoref{table:RMSE}, \autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented within the investigated time\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
-
-Interestingly, the effective start of NPIs in France and Spain is estimated to have started on Mar~24 and~25, respectively,
-although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
-The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
-This may suggest correlations between effective starts of NPIs in different countries due to international or global events.
-
-As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (Tables~\ref{table:RMSE} and~\ref{table:WAIC}).
-It may be interesting to check if the evidence supports a model with \emph{two} or more change-points, rather than one. 
-Multiple change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
-However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementation of NPIs~\citep{Flaxman2020}.
-
-As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with new guidelines, whereas in other cases the population will adopt either restrictions or relaxations even before they are formally announced.
-Attempts to assess the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date.
-However, late effective start of NPIs, such as we have inferred, can lead to under-estimation of the impact of NPIs (Figures~\ref{fig:Re} and~\ref{fig:Re2}).
-Such under-estimation may lead decision-makers to enforce stricter guidelines, rather than enforce earlier implementation of the guidelines~\citep{Pei2020a}.
-
-Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to an epidemic.
-Therefore, we emphasize the need to further study heterogeneity in compliance and behavior over both time and space.
-This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Arthur2020,Fenichela2011,Walters2013}.
-
-%\pagebreak
-% Models and Methods
-{\small
-\section*{Models and Methods}
+% Methods
+\section*{Methods}
 
@@
 We also compute the expected posterior RMSE (root mean squared error) of these predictions~(\autoref{table:RMSE}).
-} % small
-
-% Declarations
-{\small
-\section*{Declarations}
-\textbf{Ethics approval and consent to participate.} Not applicable.
-
-\textbf{Consent for publication.} Not applicable.
-
-\textbf{Availability of data and materials.} We use Python 3 with NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee.
-Source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
-We used freely available data, see source code repository and above \emph{Data} section for details.
-
-\textbf{Competing interests.} The authors declare that they have no competing interests.
-
-\textbf{Funding.} This work was supported in part by the Israel Science Foundation (3811/19 and 552/19, YR). The funding body had no role in designing, performing, or analyzing the study.
-
-\textbf{Authors' contributions.} UO and YR designed the research. IK and YR performed the research and wrote the manuscript. All authors read and approved the final manuscript.
-
-\textbf{Acknowledgements.} 
-We thank Yinon M. Bar-On, Lilach Hadany, Oren Kolodny, and Zohar Yakhini. % Tim CD Lucas
-} % small
+
+% Results
+\section*{Results}
+
+%%% Fig - tau summary %%%
+% generated with /python Fig_tau_summary.py FOLDER NAME
+\begin{figure}[b!]
+    \centering
+	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Apr11.pdf}	
+    \caption{
+    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
+    	The difference between $\tau$ the effective and $\tau^*$ the official start of NPIs is shown for different regions.
+The effective date is delayed in UK, Austria, Italy, France, Belgium, Spain, and Wuhan, China, compared to the official date (red markers).
+In contrast, the estimated effective dates in Sweden, Denmark, and Germany are earlier than the official dates (blue markers), although uncertainty is low only for Germany (i.e., zero is not in 95\%~CI).
+The credible intervals for Sweden, Denmark, and Norway are especially wide, see text and \autoref{fig:early} for possible explanation.
+	Here, the markers show $\hat{\tau}$, the marginal posterior median (\autoref{table:estimated-params}).
+$\tau^*$ is the last NPI date (a lockdown, except in Sweden; \autoref{table:NPI_dates}).
+Thin and bold lines show 95\% and 75\% credible intervals, respectively.
+\autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28 rather than Apr~11.
+    }
+    \label{fig:tau-summary}
+\end{figure}
+
+Several studies have described the effects of non-pharmaceutical interventions in different geographical regions~\citep{Flaxman2020,Gatto2020,Li2020}. 
+Some of these studies have assumed that the parameters of the epidemiological model change at a specific date (\autoref{eq:NPI_model}), and set the change date $\tau$ to the official NPI date $\tau^*$, usually the lockdown start date (\autoref{table:NPI_dates}).
+They then fit the model once for time $t<\tau^*$ and once for time $t \ge \tau^*$.
+For example, \citet{Li2020} estimate the infection dynamics in China before and after $\tau^*$, which is set at Jan 23, 2020.
+Thereby, they effectively estimate the transmission and reporting rates before and after $\tau^*$ separately.
+
+Here, we estimate the joint posterior distribution of the effective start date of NPIs, $\tau$, and the transmission and reporting rates before and after $\tau$ from the entire data, rather than splitting the data at~$\tau$.
+This is done under the simplifying assumption that all interventions start at a specific date, despite the reality that the durations between the first and last NPIs were between 4 and 12 days (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
+We then estimate $\hat{\tau}$ as the median of the marginal posterior distribution of $\tau$.
+Credible intervals (CI) are calculated as the highest density intervals~\citep{Kruschke2014}, and their upper and lower boundaries are reported as $\substack{+\text{upper} \\ -\text{lower}}$ in days relative to $\hat{\tau}$.
+
+We compare the posterior predictive plots of a model with  free~$\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e.
+transmission and reporting rates are constant).
+All three models were fitted to case data up to Apr~11, 2020, used to predict out-of-sample case data up to Apr~24, 2020, and these predictions were then compared to the real case data. 
+The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
+In all 11 european countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
+When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only marginal preference), Sweden, and Denmark (\autoref{table:WAIC}).
+Indeed, we estimate low effect of NPIs on transmission in Denmark and Sweden (i.e. $\lambda=0.7$ and $0.74$, respectively; see~\autoref{table:estimated-params}).
+This may interfere with the inference of $\tau$ due to unidentifiability. 
+Notably, the data for Sweden and Denmark do not have a single "peak" during the evaluated dates, possibly leading to wide credible intervals on $\tau$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).
+
+We compare the official ($\tau^*$) and effective ($\tau$) start of NPIs and find that in most regions the effective start date differs from the official date: in 10 of 12 countries the 75\%~credible interval on $\tau$ does not include $\tau^*$ (7 of 12 countries when considering a 95\%~credible interval; \autoref{fig:tau-summary}).
+The exceptions are Norway and Switzerland (see below).
+The former also has a wide credible interval, perhaps because it has the longest duration between the first and last NPIs~(\autoref{table:NPI_dates}).
+
+%%% Fig - Late %%%
+\begin{figure}[h]
+    \centering
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Italy_τ_posterior.pdf}
+    \end{subfigure}
+  	~
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/posterior/Mar28/Wuhan_τ_posterior.pdf}
+    \end{subfigure}
+    \\
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Spain_τ_posterior.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/France_τ_posterior.pdf}
+    \end{subfigure}
+    \caption{
+	\textbf{Late effect of non-pharmaceutical interventions.}
+    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
+Red line shows the official last NPI date $\tau^*$.
+Black line shows the estimated effective start date $\hat{\tau}$.
+Shaded area shows a 95\%~credible interval.
+    }
+    \label{fig:late}
+\end{figure}
+
+%%% Late
+\paragraph*{Late effective start of NPIs.}
+In half of the examined regions, we estimate that the effective start of NPIs $\tau$ is later than the official date $\tau^*$.
+
+In Italy, the first case was officially confirmed on Feb~21.
+School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
+That is, the first and last official NPI dates are Mar~5 and Mar~11.
+However, we estimate the effective date ($\hat{\tau}$) six~days after the lockdown, at Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\%~CI; \autoref{fig:late}).
+
+In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\%~CI).
+Yet, there is low but noticeable posterior probability for Jan~25 (\autoref{fig:late}), for which the effect of NPIs on transmission is considerably lower (\autoref{fig:joint}).
+
+In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the International Women's Day~\citep{RaphaelMinder}, and a football match between Real Betis and Real Madrid (final score 2--1) with a crowd of 50,965 in Seville~\citep{ESPNRealBetis}.
+A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
+Nevertheless, we estimate the effective start of NPIs eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95\%~CI, \autoref{fig:late}).
+
+Similarly, in France we estimate the effective start of NPIs at Mar~24 ($\substack{+2.65 \\ -1.44}$~days 95\%~CI, \autoref{fig:late}).
+This is a week later than the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
+
+%%% Fig - Early %%% 
+\begin{figure}[h]
+    \centering
+    \begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Germany_τ_posterior.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/Sweden_τ_posterior.pdf}
+    \end{subfigure}
+
+	\begin{subfigure}{0.45\textwidth}
+        \includegraphics[width=\textwidth]{../figures/posterior/Apr11/Switzerland_τ_posterior.pdf}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.45\textwidth}
+		\includegraphics[width=\textwidth]{../figures/posterior/Apr11/Norway_τ_posterior.pdf}
+    \end{subfigure}
+    \caption{
+    \textbf{Early and exact effect of non-pharmaceutical interventions.}
+    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples.
+Red line shows the official last NPI date $\tau^*$.
+Black line shows the estimated effective start date $\hat{\tau}$.
+Shaded area shows a 95\% credible interval.
+	}
+	\label{fig:early}
+\end{figure}
+
+%%% Early
+\paragraph*{Early effective start of NPIs.}
+In some regions we estimate an effective start of NPIs that is \emph{earlier} then the official date (\autoref{fig:tau-summary}).
+The only conclusive early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95~\%CI, \autoref{fig:early}).
+This estimate falls between the first and last official NPI dates, Mar~12 and Mar~22 (\autoref{table:NPI_dates}).
+Therefore, when we refer to this case as "early", we mean that the effective date (Mar~19) occurs \emph{before} the official lockdown date (Mar~22), not that it occurs before all NPIs.
+Interestingly, Germany had the second longest duration between first and last NPIs after Norway (10 and 12 days respectively; \autoref{table:NPI_dates}).
+However, the credible interval for the effective start date in Germany is narrow (1.91 days), whereas it is very wide in Norway (17.61 days).
+The significantly earlier estimate of $\hat{\tau}$ relative to $\tau^*$ can suggest that early NPIs in Germany more effectively reduced transmission rates compared to other countries.
+Another possible interpretation is that the German population anticipated the lockdown and reacted before it started.
+
+We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
+However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free~$\tau$ over the model with $\tau$ fixed at the official date (\autoref{table:WAIC}).
+Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (\autoref{table:estimated-params}).
+Moreover, in Sweden the number of daily cases continued to grow through Apr~11, rather than "peak" (\autoref{fig:ppc}). 
+In Denmark, the opposite occurred: there were seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
+We suspect that these missing and additional "peaks" increase the uncertainty in our inference.
+
+The estimated effective date in Norway is Mar~22, two days earlier than the official date of Mar~24.
+However, the posterior distribution is very wide ($\substack{+5.59 \\ -12.03}$~days 95\%~CI): it covers the range between Mar~10, two days before the first NPI, and Mar~27, three days after the last NPI (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
+The high uncertainty might be due to the long duration between the first and last NPIs; however, Germany had the second longest duration between first and last NPIs, and the corresponding posterior distribution is very narrow (\autoref{fig:early}).
+There may also be an unidentifiability issue between $\tau$ and the effect of NPIs on transmission (\autoref{fig:joint}).
+
+%%% Exact
+\paragraph*{Exact effective start of NPIs.}
+We find one case in which the official and effective dates match and the credible interval is narrow.
+Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
+Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\substack{+3.2 \\ -1.78}$~days 95\%~CI, see \autoref{fig:early}). 
+Notably, Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
+
+%%% Assessment
+\paragraph*{Assessment of impact of NPIs.}
+The \emph{effective reproduction number} $R$ is the average number of secondary
+cases caused by an infected individual after an epidemic is already underway \citep{Bar-On2020a}.
+We infer model-based effective reproduction numbers before and after the implementation of NPIs from model parameters (\autoref{eq:Re}).
+We then estimate the impact of NPIs as the relative reduction in the reproduction number \citep{Flaxman2020}, $\Delta R = (R_1-R_2)/R_1$, where $R_1$ and $R_2$ are the reproduction numbers before and after the start of NPIs.
+We compare the impacts estimated using the fixed~$\tau$ model and the free~$\tau$ model.
+That is, we compare the impact estimate $\Delta R$ assuming that NPIs started at the official date $\tau^*$, versus the estimate when inferring the effective start of NPIs from the data.
+
+\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model (y-axis) are consistently lower than estimates from the free~$\tau$ model (x-axis), except in Sweden (SE) and Denmark (DK), in which estimation uncertainty is high.
+That is, assuming NPIs start at their official date leads to error in the estimation of their impact.
+Moreover, these errors increase with the duration of the delay between the official and effective dates (\autoref{fig:Re2}).
+These results suggest that the impact of past NPIs is likely to be under-estimated by health officials and researchers if they assume NPIs start at their official dates.
+The estimated impact can then be interpreted as ineffectiveness of the NPIs, leading to more aggressive NPIs being applied.
+
+%%% Fig - Re
+% generated with
+% > source ./Re.sh
+% > python Fig_Re.py
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=\textwidth]{../figures/Fig_Re.pdf}
+    \caption{
+    \textbf{Impact of NPIs is under-estimated when assuming they start at their official date.}
+    Shown are estimates of the relative reduction in the effective reproduction number ($R$, see \autoref{eq:Re}), which measures the impact of NPIs on the epidemic.
+     The y-axis shows estimates when assuming the start of NPIs is fixed at the official date (fixed~$\tau$).
+The x-axis shows estimates when inferring the effective start of NPIs from the data (free~$\tau$).
+The dashed line shows a one-to-one correspondence.
+     Markers and bars denote the posterior median and 50\% credible intervals (HDI).
+    The relative reductions in $R$ are consistently lower for the fixed~$\tau$ model (below the dashed line), except in Sweden and Denmark in which uncertainty is high.
+    \autoref{fig:Re2} shows how the delay in effective start affect the estimation error. 
+    AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.
+    } 
+    \label{fig:Re}
+\end{figure}
+
+% Discussion and Conclusions
+\section*{Discussion and Conclusions}
+
+We have inferred the effective start date of NPIs in 12~regions using SEIR models under an MCMC parameter estimation framework.
+We find examples of both late and early effective start of NPIs relative to the official date (\autoref{fig:tau-summary}).
+
+In most investigated regions we find late effective start of NPIs.
+For example, in Italy and in Wuhan, China, the effective start of the lockdowns seems to have occurred five or more days after the official date (\autoref{fig:late}).
+These differences might be explained, in some cases, by low compliance or non-adherence to guidelines: In Italy, for example, the government plan to implement a lockdown in the Northern provinces leaked to the public, which led to people leaving these provinces before the lockdown started~\citep{Gatto2020}.
+Late effect of NPIs may also be due to the time required by both the government and the citizens to prepare for a lockdown, and for new guidelines to be adopted by the population.
+
+In contrast, in some regions we inferred reduced transmission rates even before official lockdowns were implemented, although this is only conclusive in Germany (\autoref{fig:early}).
+An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population, possibly due to earlier NPIs or NPIs being applied in other regions.
+Adoption of these behaviours may occur via media and social networks, rather than official guidelines, and may be influenced by increased risk perception due to domestic or international \covid reports~\citep{Arthur2020}.
+Indeed, the evidence supports a change in infection dynamics (i.e.
+a model with fixed or free $\tau$) even for Sweden~(\autoref{table:RMSE}, \autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented within the investigated time\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
+
+Interestingly, the effective start of NPIs in France and Spain is estimated to have started on Mar~24 and~25, respectively,
+although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
+The number of daily cases was similar in both countries until Mar~8, but diverged by Mar~13, reaching much higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
+This may suggest correlations between effective starts of NPIs in different countries due to international or global events.
+
+As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (Tables~\ref{table:RMSE} and~\ref{table:WAIC}).
+It may be interesting to check if the evidence supports a model with \emph{two} or more change-points, rather than one. 
+Multiple change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
+However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementation of NPIs~\citep{Flaxman2020}.
+
+As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with new guidelines, whereas in other cases the population will adopt either restrictions or relaxations even before they are formally announced.
+Attempts to assess the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date.
+However, late effective start of NPIs, such as we have inferred, can lead to under-estimation of the impact of NPIs (Figures~\ref{fig:Re} and~\ref{fig:Re2}).
+Such under-estimation may lead decision-makers to enforce stricter guidelines, rather than enforce earlier implementation of the guidelines~\citep{Pei2020a}.
+
+Our results highlight the complex interaction between personal, regional, and global determinants of behavioral response to an epidemic.
+Therefore, we emphasize the need to further study heterogeneity in compliance and behavior over both time and space.
+This can be accomplished both by surveying differences in compliance within and between populations~\citep{Atchison2020}, and by incorporating specific behavioral models into epidemiological models~\citep{Arthur2020,Fenichela2011,Walters2013}.
 
@@
 
-%\pagebreak
+% Abbreviations
+{\small
+\section*{Abbreviations}
+NPI: non-pharmaceutical interventions, CI: credible interval, MCMC: Markov chain Monte Carlo, SEIR: susceptible-exposed-infected-recovered, WAIC: widely applicable information criterion, RMSE: root mean squared error, IAT: integrated autocorrelation time, \covid: coronavirus disease 2019, \sars: severe acute respiratory syndrome coronavirus 2.
+}
+
+% Declarations
+{\small
+\section*{Declarations}
+\textbf{Ethics approval and consent to participate.} Not applicable.
+
+\textbf{Consent for publication.} Not applicable.
+
+\textbf{Availability of data and materials.} We use Python 3 with NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee.
+Source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
+Data for Wuhan, China, retrieved from \citet{Pei2020}.
+Data for 11 European countries retrieved from \citet{Flaxman2020}.
+
+\textbf{Competing interests.} The authors declare that they have no competing interests.
+
+\textbf{Funding.} This work was supported in part by the Israel Science Foundation (3811/19 and 552/19, YR). The funding body had no role in designing, performing, or analyzing the study.
+
+\textbf{Authors' contributions.} UO and YR designed the research. IK and YR performed the research and wrote the manuscript. All authors read and approved the final manuscript.
+
+\textbf{Acknowledgements.} 
+We thank Yinon M. Bar-On, Lilach Hadany, Oren Kolodny, and Zohar Yakhini. % Tim CD Lucas
+} % small
 
@@
 
-%%% Fig - tau summary Mar 28 %%%
-% generated with
-% > python Fig_tau_summary.py 2020-06-23-Mar28
-% > mv ../figures/Fig-tau-summary.pdf ../figures/Fig-summary-Mar28.pdf
-\begin{figure}[b!]
-    \centering
-	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Mar28.pdf}
-    \caption{
-    \textbf{Official vs. effective start of non-pharmaceutical interventions estimated up to Mar~28.}
-    	The difference between $\tau$ the effective and $\tau^*$ the official start of NPIs estimated from case data up to Mar~28, 2020, shown for different regions.
-Here, $\hat{\tau}$ is the marginal posterior median.
-$\tau^*$ is the last NPI date (a lockdown everywhere by Sweden, see \autoref{table:NPI_dates}).
-Thin and bold lines show 95\% and 75\% credible intervals (HDI~\citep{Kruschke2014}), respectively.
-Inference performed similarly to main inference, but with data up to Mar~28 and only 1M samples per chain with 600K burn-in.
-    }
-    \label{fig:tau-summary-mar28}
+%%% Fig Autocorrelation
+% generated with Fig_autocorr.py
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=\textwidth]{../figures/Fig-autocorr.pdf}
+    \caption{
+    \textbf{Integrated autocorrelation time (IAT)}~\citep{Foreman-Mackey2013,Goodman2010}, averaged across model parameters and MCMC chains, as a function of the number of samples in the chains.
+    IAT is less than 187K in all cases, while chain length is 5M after burn-in period of 2M.
+With 50 chains per region, this gives a at least 1,335 uncorrelated samples for estimating the posterior distribution.
+    } 
+    \label{fig:autocorr}
+\end{figure}
+
+%%% Fig Trace plots
+\begin{figure}[p]	
+    \centering
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Austria_trace.pdf}
+    \end{subfigure}
+    ~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Belgium_trace.pdf}
+    \end{subfigure}
+    ~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Denmark_trace.pdf}
+    \end{subfigure}
+	~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/France_trace.pdf}
+    \end{subfigure}
+    ~%%%
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Germany_trace.pdf}
+    \end{subfigure}
+    ~%%% 
+    \begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Italy_trace.pdf}
+    \end{subfigure}
+    ~%%% 
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Norway_trace.pdf}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Spain_trace.pdf}
+    \end{subfigure}
+    ~%%%
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Sweden_trace.pdf}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Switzerland_trace.pdf}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/United_Kingdom_trace.pdf}
+    \end{subfigure}
+    ~%%%  
+	\begin{subfigure}{0.32\textwidth}
+        \includegraphics[width=\textwidth]{../figures/traces/Wuhan_trace.pdf}
+    \end{subfigure}
+    \caption{
+    \textbf{Trace plots for $\tau$.}
+    The value of $\tau$ at for sequential samples of 50 chains per region.
+To facilitate visualization, chains were thinned 1:10,000 for the trace plots, but not for posterior estimation.
+Line transparency was set at $\alpha=0.1$ and chains cycle through different colors.
+In France and Austria some chains converged relatively late but before the burn-in period (2M iterations).
+In Spain, a single chain converged very late (after roughly 6M iterations) and was removed from further analysis.
+The y-axis range is the support of the prior on $\tau$.
+    }
+    \label{fig:trace}	    
 \end{figure}
@@
 
+%%% Fig - tau summary Mar 28 %%%
+% generated with
+% > python Fig_tau_summary.py 2020-06-23-Mar28
+% > mv ../figures/Fig-tau-summary.pdf ../figures/Fig-summary-Mar28.pdf
+\begin{figure}[b!]
+    \centering
+	\includegraphics[width=0.6\textwidth]{../figures/Fig-summary-Mar28.pdf}
+    \caption{
+    \textbf{Official vs. effective start of non-pharmaceutical interventions estimated up to Mar~28.}
+    	The difference between $\tau$ the effective and $\tau^*$ the official start of NPIs estimated from case data up to Mar~28, 2020, shown for different regions.
+Here, $\hat{\tau}$ is the marginal posterior median.
+$\tau^*$ is the last NPI date (a lockdown everywhere by Sweden, see \autoref{table:NPI_dates}).
+Thin and bold lines show 95\% and 75\% credible intervals (HDI~\citep{Kruschke2014}), respectively.
+Inference performed similarly to main inference, but with data up to Mar~28 and only 1M samples per chain with 600K burn-in.
+    }
+    \label{fig:tau-summary-mar28}
+\end{figure}
+
 %%% Fig Joint density plots
@@
 
-%%% Fig Autocorrelation
-% generated with Fig_autocorr.py
-\begin{figure}[h]
-    \centering
-	\includegraphics[width=\textwidth]{../figures/Fig-autocorr.pdf}
-    \caption{
-    \textbf{Integrated autocorrelation time (IAT)}~\citep{Foreman-Mackey2013,Goodman2010}, averaged across model parameters and MCMC chains, as a function of the number of samples in the chains.
-    IAT is less than 187K in all cases, while chain length is 5M after burn-in period of 2M.
-With 50 chains per region, this gives a at least 1,335 uncorrelated samples for estimating the posterior distribution.
-    } 
-    \label{fig:autocorr}
-\end{figure}
-
-%%% Fig Trace plots
-\begin{figure}[p]	
-    \centering
-    \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Austria_trace.pdf}
-    \end{subfigure}
-    ~%%%
-    \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Belgium_trace.pdf}
-    \end{subfigure}
-    ~%%%
-    \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Denmark_trace.pdf}
-    \end{subfigure}
-	~%%%
-    \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/France_trace.pdf}
-    \end{subfigure}
-    ~%%%
-    \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Germany_trace.pdf}
-    \end{subfigure}
-    ~%%% 
-    \begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Italy_trace.pdf}
-    \end{subfigure}
-    ~%%% 
-	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Norway_trace.pdf}
-    \end{subfigure}
-    ~%%%  
-	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Spain_trace.pdf}
-    \end{subfigure}
-    ~%%%
-	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Sweden_trace.pdf}
-    \end{subfigure}
-    ~%%%  
-	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Switzerland_trace.pdf}
-    \end{subfigure}
-    ~%%%  
-	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/United_Kingdom_trace.pdf}
-    \end{subfigure}
-    ~%%%  
-	\begin{subfigure}{0.32\textwidth}
-        \includegraphics[width=\textwidth]{../figures/traces/Wuhan_trace.pdf}
-    \end{subfigure}
-    \caption{
-    \textbf{Trace plots for $\tau$.}
-    The value of $\tau$ at for sequential samples of 50 chains per region.
-To facilitate visualization, chains were thinned 1:10,000 for the trace plots, but not for posterior estimation.
-Line transparency was set at $\alpha=0.1$ and chains cycle through different colors.
-In France and Austria some chains converged relatively late but before the burn-in period (2M iterations).
-In Spain, a single chain converged very late (after roughly 6M iterations) and was removed from further analysis.
-The y-axis range is the support of the prior on $\tau$.
-    }
-    \label{fig:trace}	    
-\end{figure}
-
 \end{document}

### 2021-02-06 ms/ms.tex
autocorrelation change for fixed tau model

@@
 For the models with fixed~$\tau$ and without~$\tau$, in some countries, we use less than 7M samples per chain because the IAT converged sooner.
-In these cases the effective sample size was at between 2,454 (UK) and 10,954 (Spain) in the fixed~$\tau$ model and between 6,230 (Norway) and 94,339 (Italy) in the model without~$\tau$.
+In these cases the effective sample size was at between 1,451 (Germany) and 7,443 (Wuhan) in the fixed~$\tau$ model and between 6,230 (Norway) and 94,339 (Italy) in the model without~$\tau$.
 

### 2021-02-08 ms/ms.tex
corrected fixed model conclusions

@@
 We also tried to estimate the value of $p_t$ before and after $\tau$, instead of keeping it fixed at $1/9$ and $1/6$.
-The model with fixed values was supported by the evidence (lower WAIC, see below) in 9 of 12 countries.
+The model with fixed values was supported by the evidence (lower WAIC, see~\autoref{eq:WAIC}) in 9 of 12 countries.
 Moreover, the estimates for Wuhan, China were $1/9$ and $1/6$, as in \citet{Li2020}.
@@
 The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
-In all 11 european countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
-When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only marginal preference), Sweden, and Denmark (\autoref{table:WAIC}).
+In 7 of 11 European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
+The RMSE is lower for the fixed~$\tau$ model in Denmark, Norway, and Sweden, where the difference is small (<9\%) and Switzerland, where the estimated and official dates match (\autoref{fig:tau-summary}).
+A similar trend appears when we compare the models using WAIC~(\autoref{eq:WAIC}: the model with free~$\tau$ is strongly preferred in 9 out of 12 regions, the exceptions being Denmark, Norway, and Switzerland (\autoref{table:WAIC}).
 Indeed, we estimate low effect of NPIs on transmission in Denmark and Sweden (i.e. $\lambda=0.7$ and $0.74$, respectively; see~\autoref{table:estimated-params}).
@@
 We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
-However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free~$\tau$ over the model with $\tau$ fixed at the official date (\autoref{table:WAIC}).
+However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and the evidence did not support the model with free~$\tau$ over the model with $\tau$ fixed at the official date (Tables~\ref{table:RMSE} and~\ref{table:WAIC}).
 Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (\autoref{table:estimated-params}).
@@
 
-\pagebreak
+%\pagebreak
 \section*{Supplementary Material}
@@
 	Expected posterior predictive RMSE (root mean squared error) for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
-	In all cases, the model with free parameter $\tau$ has the lowest RMSE.
+	In 7 of 11 cases, the model with free parameter $\tau$ has the lowest RMSE.
 	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions up to Apr~24 by sampling model parameters from the posterior distribution.
 	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model. 
-	Bold values emphasize cases in which the \emph{Free} model has the lowest RMSE.
+	Bold values highlight the model with the lowest RMSE.
 }
@@
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.
-Bold values emphasize cases in which the \emph{Free} model has the lowest WAIC.
+Bold values highlight the model with the lowest WAIC.
 } 

### 2021-02-08 ms/ms.tex
corrected fixed model conclusions, Re figures

@@
 
-\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model (y-axis) are consistently lower than estimates from the free~$\tau$ model (x-axis), except in Sweden (SE) and Denmark (DK), in which estimation uncertainty is high.
-That is, assuming NPIs start at their official date leads to error in the estimation of their impact.
-Moreover, these errors increase with the duration of the delay between the official and effective dates (\autoref{fig:Re2}).
-These results suggest that the impact of past NPIs is likely to be under-estimated by health officials and researchers if they assume NPIs start at their official dates.
+\autoref{fig:Re} demonstrates that estimates from the fixed~$\tau$ model (y-axis) may be significantly lower than estimates from the free~$\tau$ model (x-axis).
+This is the case in six of 12 regions, although the effect size is small in two of these regions (France, FR, and Italy, IT).
+Moreover, we find that NPI impact estimation errors increase with the duration of the delay between the official and effective dates (\autoref{fig:Re2}).
+These results suggest that the impact of past NPIs can be under-estimated by health officials and researchers if they assume NPIs start at their official dates.
 The estimated impact can then be interpreted as ineffectiveness of the NPIs, leading to more aggressive NPIs being applied.
@@
      Markers and bars denote the posterior median and 50\% credible intervals (HDI).
-    The relative reductions in $R$ are consistently lower for the fixed~$\tau$ model (below the dashed line), except in Sweden and Denmark in which uncertainty is high.
-    \autoref{fig:Re2} shows how the delay in effective start affect the estimation error. 
+    The relative reductions in $R$ are lower for the fixed~$\tau$ model (below the dashed line) In six of 12 regions.
+    \autoref{fig:Re2} shows how the delay in effective start affects the estimation error. 
     AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.

### 2021-02-09 ms/ms.tex
revise linreg values in fig S7
two linreg in fig s7

@@
     Impact of NPIs defined as $\frac{R_1 - R_2}{R_1}$, where $R_1$ and $R_2$ are the effective reproduction numbers before and after the NPI, respectively (\autoref{eq:Re}).
-	The dashed black line shows a linear regression, slope=-0.0313 with (-0.051, -0.012)~95\%~CI, $R^2=0.554$.
+	The dashed gray and black lines show a linear regression with and without SE, DK and CN, respectively
+	(with: slope=$-0.019\pm0.026$~95\%~CI, intercept=$-0.105 \pm0.175$~95\%~CI, $R^2=0.211$;
+     without: slope=$-0.016 \pm0.006$~95\%~CI, intercept=$-0.001 \pm0.036$~95\%~CI, $R^2=0.845$.)
 	AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.

### 2021-07-14 ms/ms.tex
updated for IJID

@@
 Overall, these interventions seem to have reduced the spread of the pandemic.
-We hypothesized that the official and effective start date of such interventions can be noticeably different, for example due to slow adoption by the population, or because the authorities and the public are unprepared, and that these difference can lead to errors in the estimation of the impact of NPIs.
+We hypothesized that the official and effective start date of such interventions can be noticeably different, for example due to slow adoption by the population, or because the authorities and the public are unprepared, and that these differences can lead to errors in the estimation of the impact of NPIs.
 \\
 \textbf{Methods.}
-SEIR models were fitted to case data from~12 regions to infer the effective start dates of interventions and contrast them with the official dates. \\
+SEIR models were fitted to case data from~12 regions to infer the effective start dates of interventions and contrast them with the official dates.
+The impact of NPIs was estimated from the inferred model parameters.
+\\
 \textbf{Results.}
@@
 \textbf{Conclusions.}
-Differences between the official and effective start of NPIs are likely, and that neglecting such differences can lead to under-estimation of the impact of NPIs.
+Differences between the official and effective start of NPIs are likely.
+Neglecting such differences can lead to under-estimation of the impact of NPIs, which could cause decision-makers to escalate interventions and guidelines.
 \\\\
 {\small
-Keywords: SEIR, COVID-19, public health, epidemic, infectious disease, NPI
+\textbf{Keywords}: SEIR, COVID-19, public health, epidemic, infectious disease, NPI
 }
@@
 
-As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with new guidelines, whereas in other cases the population will adopt either restrictions or relaxations even before they are formally announced.
+As different countries experiment with various intervention strategies, we expect similar shifts to continue to occur: in some cases the population will be late to comply with new guidelines, whereas in other cases the population will adopt either restrictions or relaxations even before they are formally announced.
 Attempts to assess the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date.

### 2021-09-02 ms/ms.tex
first attempt to respond to reviewers

@@
 \textbf{Results.}
-We infer mostly late, but also early effects of interventions.
+We infer mostly late effective start of interventions.
 For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective start date on Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\%~CI).
-In contrast, Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95\%~CI).
+In one case we infer an early effective start date: Germany announced a lockdown on Mar~22, but we infer an effective start date on Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95\%~CI).
 Moreover, we find that the impact of NPIs can be under-estimated if assuming they start at their official date.\\
@@
 We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) model and a \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to infer the effective start date of NPIs from publicly available \covid case data in 12 geographical regions.
-We compare these estimates to the official dates, and find that they include both late and early effects of NPIs on infection dynamics.
+We compare these estimates to the official dates, and find that they include mostly late effects of NPIs on infection dynamics.
 We conclude by demonstrating how differences between the official and effective start of NPIs can confound assessments of their impacts.
@@
 These incidence data summarise the number of individuals $X_t$ tested positive for \sars (using RT-qPCR tests) on each day $t$.
-Data for Wuhan, China, from Jan~10 to Feb~8, retrieved from \citet{Pei2020}. Data for 11 European countries, from Feb~20 to Apr~24, retrieved from \citet{Flaxman2020}.
+Data for Wuhan, China, from Jan~10 to Feb~8, 2020, retrieved from \citet{Pei2020}. Data for 11 European countries, from Feb~20 to Apr~24, 2020, retrieved from \citet{Flaxman2020}.
 Where there were multiple sequences of days with zero confirmed cases (e.g.
@@
 where $0 < \lambda < 1$.
+That is, we assume that the transmission rate, $\beta$, the reporting rate, $\alpha$, and the confirmation rate, $p_t$, change after the implementation of NPIs, for example due to increased awareness and reduced contact rate in the population, and implementation of measures such as testing by the authorities.
 The values for $p_t$ follow \citet{Li2020}, who estimated the average time between infection and reporting in Wuhan, China, at 9 days before the start of NPIs and 6 days after start of NPIs.
@@
 To estimate the model parameters from the daily case data $\vec{X}$, we apply a Bayesian inference approach.
-Model fitting was calibrated for case data up to Mar~28, and then applied to data up to Apr~11 (for Wuhan, China, model fitting was applied for data up to Feb~8.)
+Model fitting was calibrated for case data up to Mar~28, 2020, and then applied to data up to Apr~11, 2020 (for Wuhan, China, model fitting was applied for data up to Feb~8, 2020.)
 We start our model $\Delta t$ days~\citep{Gatto2020} before the outbreak (defined as consecutive days with increasing confirmed cases) in each country.
@@
 Moreover, the estimates for Wuhan, China were $1/9$ and $1/6$, as in \citet{Li2020}.
+Hence, we fixed the values of $p_t$ to those used in \citet{Li2020} (see \autoref{eq:NPI_model}) in the analysis reported below.
 
@@
 Thin and bold lines show 95\% and 75\% credible intervals, respectively.
-\autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28 rather than Apr~11.
+\autoref{fig:tau-summary-mar28} shows a similar summary when estimating $\hat{\tau}$ using case data up to Mar~28, 2020, rather than Apr~11, 2020.
     }
@@
 
-In Italy, the first case was officially confirmed on Feb~21.
+In Italy, the first case was officially confirmed on Feb~21, 2020.
 School closures were implemented on Mar~5~\citep{Flaxman2020}, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
@@
 
-In Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\%~CI).
+In Wuhan, China, a lockdown was ordered on Jan 23, 2020~\citep{Li2020}, but we estimate the effective start of NPIs to be ten days later, at Feb~2 ($\substack{+2.29 \\ -2.97}$~days 95\%~CI).
 Yet, there is low but noticeable posterior probability for Jan~25 (\autoref{fig:late}), for which the effect of NPIs on transmission is considerably lower (\autoref{fig:joint}).
 
-In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the International Women's Day~\citep{RaphaelMinder}, and a football match between Real Betis and Real Madrid (final score 2--1) with a crowd of 50,965 in Seville~\citep{ESPNRealBetis}.
+In Spain, social distancing was encouraged starting on Mar~8, 2020~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the International Women's Day~\citep{RaphaelMinder}, and a football match between Real Betis and Real Madrid (final score 2--1) with a crowd of 50,965 in Seville~\citep{ESPNRealBetis}.
 A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
@@
 
-Similarly, in France we estimate the effective start of NPIs at Mar~24 ($\substack{+2.65 \\ -1.44}$~days 95\%~CI, \autoref{fig:late}).
+Similarly, in France we estimate the effective start of NPIs at Mar~24, 2020 ($\substack{+2.65 \\ -1.44}$~days 95\%~CI, \autoref{fig:late}).
 This is a week later than the official lockdown, which started at Mar~17, and more than 10 days after the earliest NPI, banning of public events, which started on Mar~13~\citep{Flaxman2020}.
@@
 In some regions we estimate an effective start of NPIs that is \emph{earlier} then the official date (\autoref{fig:tau-summary}).
-The only conclusive early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\substack{+0.92 \\ -0.99}$~days 95~\%CI, \autoref{fig:early}).
+The only conclusive early case is Germany, in which we estimate the effective start of NPIs at Mar~19, 2020 ($\substack{+0.92 \\ -0.99}$~days 95~\%CI, \autoref{fig:early}).
 This estimate falls between the first and last official NPI dates, Mar~12 and Mar~22 (\autoref{table:NPI_dates}).
@@
 
-The estimated effective date in Norway is Mar~22, two days earlier than the official date of Mar~24.
+The estimated effective date in Norway is Mar~22, 2020, two days earlier than the official date of Mar~24.
 However, the posterior distribution is very wide ($\substack{+5.59 \\ -12.03}$~days 95\%~CI): it covers the range between Mar~10, two days before the first NPI, and Mar~27, three days after the last NPI (\autoref{table:NPI_dates}, \autoref{fig:NPI_dates}).
@@
 We find one case in which the official and effective dates match and the credible interval is narrow.
-Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
+Switzerland ordered a national lockdown on Mar~20, 2020, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
 Indeed, the posterior median $\hat{\tau}$ is Mar~19 ($\substack{+3.2 \\ -1.78}$~days 95\%~CI, see \autoref{fig:early}). 
@@
 We have inferred the effective start date of NPIs in 12~regions using SEIR models under an MCMC parameter estimation framework.
-We find examples of both late and early effective start of NPIs relative to the official date (\autoref{fig:tau-summary}).
+We find examples of mostly late but also early effective start of NPIs relative to the official date (\autoref{fig:tau-summary}).
 
@@
 Indeed, the evidence supports a change in infection dynamics (i.e.
-a model with fixed or free $\tau$) even for Sweden~(\autoref{table:RMSE}, \autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented within the investigated time\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
-
-Interestingly, the effective start of NPIs in France and Spain is estimated to have started on Mar~24 and~25, respectively,
+a model with fixed or free $\tau$) even for Sweden~(\autoref{table:RMSE}, \autoref{table:WAIC}, \autoref{fig:ppc}), where a lockdown was not implemented within the investigated time\footnote{Sweden banned public events on Mar~12, 2020, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.
+
+Interestingly, the effective start of NPIs in France and Spain is estimated to have started on Mar~24 and~25, 2020, respectively,
 although the official NPI dates differ significantly: the first NPI in France is only one day before the last NPI in Spain.
@@
 As expected, we have found that the evidence supports a model in which the transmission rate changes at a specific time point over a model with a constant transmission rate (Tables~\ref{table:RMSE} and~\ref{table:WAIC}).
-It may be interesting to check if the evidence supports a model with \emph{two} or more change-points, rather than one. 
+It may be interesting to check if the evidence supports a model with \emph{two} or more change-points, rather than one, or even a continuous change.
 Multiple change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
-However, interpretation of such models will be harder, as multiple change-points are also likely to result in parameter unidentifiability, for example due to simultaneous implementation of NPIs~\citep{Flaxman2020}.
-
-As different countries experiment with various intervention strategies, we expect similar shifts to continue to occur: in some cases the population will be late to comply with new guidelines, whereas in other cases the population will adopt either restrictions or relaxations even before they are formally announced.
+However, both inference and interpretation of such models will be harder, as multiple change-points or continuous change are also likely to result in parameter unidentifiability, for example due to simultaneous implementation of NPIs~\citep{Flaxman2020}.
+
+As different countries experiment with various intervention strategies, we expect similar shifts to continue to occur: in some cases the population will be late to comply with new guidelines, whereas in other cases the population will adopt restrictions, and possibly relaxations, even before they are formally announced.
 Attempts to assess the impact of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume they start at their official date.
@@
 \textbf{Acknowledgements.} 
-We thank Yinon M. Bar-On, Lilach Hadany, Oren Kolodny, and Zohar Yakhini. % Tim CD Lucas
+We thank Yinon M. Bar-On, Lilach Hadany, Oren Kolodny, Zohar Yakhini, Tim CD Lucas, and two anonymous reviewers. 
 } % small
@@
 Models were fitted with data up to Apr~11.
-The predictions are generated by drawing 1,000 parameters sets from the posterior distribution, and then generating a daily case count using the SEIR model (\autoref{eq:model}) up to Apr~24.
+The predictions are generated by drawing 1,000 parameters sets from the posterior distribution, and then generating a daily case count using the SEIR model (\autoref{eq:model}) up to Apr~24, 2020.
 Note the differences in the y-axis scale.
@@
     Number of cases proportional to population size (as of 2018). 
-    Vertical line shows Mar~8, the effective start of NPIs $\hat{\tau}$ in both countries.
+    Vertical line shows Mar~8, 2020, the effective start of NPIs $\hat{\tau}$ in both countries.
     Data from~\citet{Flaxman2020}.

### 2021-09-05 ms/ms.tex
UO changes
IK comments

@@
 \textbf{Results.}
-We infer mostly late effective start of interventions.
+We infer mostly late effective start dates of interventions.
 For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective start date on Mar~17 ($\substack{+3.05 \\ -2.01}$~days 95\%~CI).
@@
      Markers and bars denote the posterior median and 50\% credible intervals (HDI).
-    The relative reductions in $R$ are lower for the fixed~$\tau$ model (below the dashed line) In six of 12 regions.
-    \autoref{fig:Re2} shows how the delay in effective start affects the estimation error. 
+    The relative reductions in $R$ are lower for the fixed~$\tau$ model (below the dashed line) in at least seven of 12 regions.
+    Note that the estimated effective start date is uncertain for Sweden, Denmark, and Norway (\autoref{fig:tau-summary}).
+    The effect of the delay in effective start affects on the estimation error is shown in \autoref{fig:Re2}.
     AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.
@@
      without: slope=$-0.016 \pm0.006$~95\%~CI, intercept=$-0.001 \pm0.036$~95\%~CI, $R^2=0.845$.)
+     Note that the estimated effective start date is uncertain for Sweden, Denmark, and Norway (\autoref{fig:tau-summary}).
 	AT: Austria, BE: Belgium, CH: Switzerland, CN: Wuhan, China, DE: Germany, DK: Denmark, ES: Spain, FR: France, GB: United Kingdom, IT: Italy, NO: Norway, SE: Sweden.

### 2021-09-05 ms/ms.tex
added free to name of ppc free files for submission

@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Austria_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_Austria_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Belgium_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_Belgium_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Denmark_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_Denmark_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/France_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_France_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Germany_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_Germany_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Italy_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_Italy_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Norway_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_Norway_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Spain_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_Spain_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Sweden_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_Sweden_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/Switzerland_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_Switzerland_ppc_long.pdf}
     \end{subfigure}
@@
     \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\textwidth]{../figures/ppc/free/United_Kingdom_ppc_long.pdf}
+        \includegraphics[width=\textwidth]{../figures/ppc/free/free_United_Kingdom_ppc_long.pdf}
     \end{subfigure}
